### 2020-09-26 doc_arXiv/doc_arXiv.Rmd
added preprint

---
title: Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values
authors:
  - name: Byron C. Jaeger
    thanks: Source code available at https://github.com/bcjaeger/INTERMACS-missing-data
    department: Department of Biostatistics
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL 35211
    email: bcjaeger@uab.edu
abstract: |
  \textbf{Background} Risk predictions play an important role in clinical decision making. When developing risk prediction models, practitioners often impute missing values to the mean. The purpose of this article is to evaluate the impact of applying different strategies to impute missing values on the prognostic accuracy of prediction models fitted to the imputed data. A secondary objective was to compare the accuracy of different imputation methods. To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).  \newline\textbf{Methods and Results} We applied all pairwise combinations of <fill in> different strategies to impute missing values and three different strategies to fit a risk prediction model for mortality and transplant after receiving mechanical circulatory support. Model performance was compared using Concordance (i.e. C-index), Brier Score, and Net reclassification index. Results indicated that multiple  imputation consistently resulted in  superior prognostic models compared  to  other missing data strategies, particularly imputation to the mean. <fill in results> \newline\textbf{Conclusion} Selecting an optimal strategy to handle missing values can have a substantial impact on improving model accuracy.  In the current analysis, multiple imputation emerged as an optimal strategy to handle missing values in the INTERMACS data. The evaluation and selection of the optimal missing data strategy in this work has the potential to improve risk predictions for other longitudinal registries.
keywords:
  - Missing Data, 
  - INTERMACS, 
  - imputation, 
  - heart failure, 
  - mortality, 
  - risk prediction
bibliography: references.bib
output: rticles::arxiv_article
header-includes:
   - \usepackage{placeins}
   - \newcommand{\ie}{\textit{i.e., }}
   - \newcommand{\eg}{\textit{e.g., }}
   - \newcommand{\cstat}{\widehat{\textrm{C}}(t)}
   - \newcommand{\bstat}{\widehat{\textrm{BS}}(t)}
   - \newcommand{\ibstat}{\mathcal{\widehat{BS}}(\tau)}
---

# Introduction
\label{sec:introduction}

Heart disease is a leading cause of death in the United States. Heart failure, a primary component of heart disease, affects over 6 million Americans, and for ~10\% of these patients medical management is no longer effective \cite{benjamin2017heart,national2017health}. Mechanical circulatory support (MCS) is a surgical intervention in which a mechanical device is implanted in parallel to the heart to improve circulation \cite{patel2014contemporary}. Typically, MCS is used while a patient waits for a heart transplant (bridge-to-transplant) or in some cases as an alternative to transplant (destination therapy) \cite{slaughter2009advanced}. Over 250,000 patients could benefit from MCS \cite{miller2011left}. However, less than 4,000 new patients receive a long-term MCS device each year, with widely heterogeneous outcomes \cite{stewart2011keeping}. The 2-year survival on MCS ranges from 61\% for destination therapy to 78\% for bridge-to-transplant \cite{patel2014contemporary}. Therefore, there is great need for reliable predictions of patient-specific probability to experience adverse events after receiving MCS. This information can be used to improve patient selection for MCS,inform the design of next generation pumps, and refine patient management strategies.

### 2020-09-26 doc_arXiv/doc_arXiv.tex
added preprint

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}

\title{Improving Outcome Predictions for Patients Receiving Mechanical
Circulatory Support by Optimizing Imputation of Missing Values}

\author{
    Byron C. Jaeger
    \thanks{Source code available at
\url{https://github.com/bcjaeger/INTERMACS-missing-data}}
   \\
    Department of Biostatistics \\
    University of Alabama at Birmingham \\
  Birmingham, AL 35211 \\
  \texttt{\href{mailto:bcjaeger@uab.edu}{\nolinkurl{bcjaeger@uab.edu}}} \\
  }

\begin{document}
\maketitle

\def\tightlist{}

\begin{abstract}
\textbf{Background} Risk predictions play an important role in clinical
decision making. When developing risk prediction models, practitioners
often impute missing values to the mean. The purpose of this article is
to evaluate the impact of applying different strategies to impute
missing values on the prognostic accuracy of prediction models fitted to
the imputed data. A secondary objective was to compare the accuracy of
different imputation methods. To complete these objectives, we used data
from the Interagency Registry for Mechanically Assisted Circulatory
Support (INTERMACS). \newline\textbf{Methods and Results} We applied all
pairwise combinations of different strategies to impute missing values
and three different strategies to fit a risk prediction model for
mortality and transplant after receiving mechanical circulatory support.
Model performance was compared using Concordance (i.e.~C-index), Brier
Score, and Net reclassification index. Results indicated that multiple
imputation consistently resulted in superior prognostic models compared
to other missing data strategies, particularly imputation to the mean.
\newline\textbf{Conclusion} Selecting an optimal strategy to handle
missing values can have a substantial impact on improving model
accuracy. In the current analysis, multiple imputation emerged as an
optimal strategy to handle missing values in the INTERMACS data. The
evaluation and selection of the optimal missing data strategy in this
work has the potential to improve risk predictions for other
longitudinal registries.
\end{abstract}

\keywords{
    Missing Data,
   \and
    INTERMACS,
   \and
    imputation,
   \and
    heart failure,
   \and
    mortality,
   \and
    risk prediction
  }

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\label{sec:introduction}

Heart disease is a leading cause of death in the United States. Heart
failure, a primary component of heart disease, affects over 6 million
Americans, and for \textasciitilde10\% of these patients medical
management is no longer effective
\cite{benjamin2017heart,national2017health}. Mechanical circulatory
support (MCS) is a surgical intervention in which a mechanical device is
implanted in parallel to the heart to improve circulation
\cite{patel2014contemporary}. Typically, MCS is used while a patient
waits for a heart transplant (bridge-to-transplant) or in some cases as
an alternative to transplant (destination therapy)
\cite{slaughter2009advanced}. Over 250,000 patients could benefit from
MCS \cite{miller2011left}. However, less than 4,000 new patients receive
a long-term MCS device each year, with widely heterogeneous outcomes
\cite{stewart2011keeping}. The 2-year survival on MCS ranges from 61\%
for destination therapy to 78\% for bridge-to-transplant
\cite{patel2014contemporary}. Therefore, there is great need for
reliable predictions of patient-specific probability to experience
adverse events after receiving MCS. This information can be used to
improve patient selection for MCS,inform the design of next generation
pumps, and refine patient management strategies.

\bibliographystyle{unsrt}
\bibliography{references.bib}

\end{document}

### 2020-09-26 R/plan.R

@@ -59,6 +59,12 @@ the_plan <- drake_plan(
                                                   times),
 
 
+  arxiv_preprint = target(
+    command = {
+      rmarkdown::render(knitr_in("doc_arXiv/doc_arXiv.Rmd"))
+      file_out("doc_arXiv/doc_arXiv.pdf")
+    }
+  )
 
   #tbl_characteristics = tabulate_characteristics(im)
 

### 2020-09-26 packages.R

@@ -42,3 +42,4 @@ conflict_prefer("gather",    "tidyr")
 conflict_prefer("set_names", "purrr")
 conflict_prefer("matches",   "tidyselect")
 conflict_prefer("impute",    "miceRanger")
+library(rmarkdown)

### 2020-09-27 doc_arXiv/doc_arXiv.Rmd
use train but not test outcomes

@@
 The 2-year survival on MCS ranges from 61\% for destination therapy to 78\% for bridge-to-transplant \cite{patel2014contemporary}.
-Therefore, there is great need for reliable predictions of patient-specific probability to experience adverse events after receiving MCS.
-This information can be used to improve patient selection for MCS,inform the design of next generation pumps, and refine patient management strategies.
+Therefore, there is great need for reliable predictions of patient-specific risk to experience adverse events after receiving MCS.
+This information can be used to improve patient selection for MCS, inform the design of next generation pumps, and refine patient management strategies.
+
+The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nation wide level.
+Currently, INTERMACS comprises over 20,000 patients who have received a mechanically assisted circulatory support device (MCSD).
+As the largest registry for data on patients receiving MCSD, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{kirklin2017eighth, kormos2019society, Adamo950}.
+A critical step for developing risk prediction models is handling missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial.
+Several variables in the INTERMACS data have high rates of missing values.
+While some of the missing data is related to data entry error, other missing values may  relate  to instability of a patient’s circulatory status at the time of implant.
+For instance, hemodynamic laboratory values are missing for patients in whom  invasive catheterizations were not performed.
+Patients without invasive cathaderizations  may be too ill (unstable) to tolerate the procedure.
+Instead, such patients may, after determination of basic hemodynamics, proceed directly to device placement.
+
+Multiple strategies have been applied to handle missing values in the INTERMACS registry.
+Typically, missing values have been imputed to the mean \cite{kormos2019society}.
+Loghmanpour et al.
+combined several imputed datasets from random forest models into a single imputed dataset.
+Grady et al.
+handled missing data by assigning values based on the clinical state of the patient (physical components of heath related quality of life scores set to zero if the patient was too sick to complete the instrument) \cite{grady2015change}.
+This article extends previous work by formally quantifying the effect of differing strategies to handle missing data for the INTERMACS registry on downstream prediction models that are fitted to the imputed data.

### 2020-09-27 R/tidy_ranger.R

@@ -1,33 +1,0 @@
-##' .. content for \description{} (no empty lines) ..
-##'
-##' .. content for \details{} ..
-##'
-##' @title
-##' @param ranger_mi_train
-##' @param ranger_mi_test
-tidy_ranger <- function(ranger_mi_train, ranger_mi_test) {
-
-  tidy_train_imputes <- completeData(ranger_mi_train) %>%
-    set_names(1:length(.)) %>%
-    enframe(name = 'impute', value = 'training') %>%
-    mutate(training = map(training, as_tibble))
-
-  tidy_test_imputes <- ranger_mi_test$imputedData %>%
-    set_names(1:length(.)) %>%
-    enframe(name = 'impute', value = 'testing')
-
-  left_join(tidy_train_imputes,
-            tidy_test_imputes,
-            by = 'impute') %>%
-    mutate(impute = as.integer(impute))
-
-}
-
-tidy_nbrs <- function(nbrs_brew){
-
-  nbrs_brew$wort %>%
-    as_tibble() %>%
-    select(-pars) %>%
-    mutate(impute = as.integer(impute))
-
-}

### 2020-09-27 slurm/slurm_run.R

@@ -22,8 +22,8 @@ R.utils::sourceDirectory("../R")
 
   # train/test split -----
 
-  im_train_raw <- im[-testing_index, ] #%>% slice_sample(prop = 1/10)
-  im_test_raw  <- im[ testing_index, ] #%>% slice_sample(prop = 1/2)
+  im_train_raw <- im[-testing_index, ] %>% slice_sample(prop = 1/3)
+  im_test_raw  <- im[ testing_index, ] %>% slice_sample(prop = 1/2)
 
   too_many_missing <- im_train_raw %>%
     miss_var_summary() %>%
@@ -115,6 +115,7 @@ R.utils::sourceDirectory("../R")
 
   # redefine features, since some features may have been dropped by step_nzv
   features <- setdiff(names(im_train_pre_impute), c('time', 'status'))
+  analysis_variables <- names(im_train_pre_impute)
 
   # redefine amputed values to account for dropped features
   for(i in names(amputed_values)){
@@ -134,13 +135,24 @@ R.utils::sourceDirectory("../R")
 
   n_miss_vars <- length(miceranger_var_names)
 
-  miceranger_vars <- vector(mode = 'list', length = n_miss_vars)
+  miceranger_vars_trn <- miceranger_vars_tst <-
+    vector(mode = 'list', length = n_miss_vars)
 
-  for(i in seq_along(miceranger_vars)){
+  for(i in seq_along(miceranger_vars_trn)){
 
-    names(miceranger_vars)[i] <- miceranger_var_names[i]
-    # dont use the outcome to impute a predictor
-    miceranger_vars[[i]] <- setdiff(features, miceranger_var_names[i])
+    names(miceranger_vars_trn)[i] <- miceranger_var_names[i]
+    names(miceranger_vars_tst)[i] <- miceranger_var_names[i]
+
+    # do use the outcome to impute a predictor in the training data
+    miceranger_vars_trn[[i]] <- setdiff(analysis_variables,
+                                        miceranger_var_names[i])
+    # dont use the outcome to impute a predictor in the testing data
+    # (while we can assume that the outcomes in the training data are
+    #  known, it should be obvious that the outcomes in the testing data
+    #  are unknown or should be assumed to be unknown)
+    miceranger_vars_tst[[i]] <- setdiff(features,
+                                        miceranger_var_names[i])
+
 
   }
 
[…]

### 2020-10-10 doc_arXiv/doc_arXiv.Rmd
preparing to drop RF model

--- 2020-09-26 doc_arXiv/doc_arXiv.Rmd
@@
 To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).
-\newline\textbf{Methods and Results} We applied all pairwise combinations of <fill in> different strategies to impute missing values and three different strategies to fit a risk prediction model for mortality and transplant after receiving mechanical circulatory support.
-Model performance was compared using Concordance (i.e.
-C-index), Brier Score, and Net reclassification index.
-Results indicated that multiple  imputation consistently resulted in  superior prognostic models compared  to  other missing data strategies, particularly imputation to the mean.
-<fill in results> \newline\textbf{Conclusion} Selecting an optimal strategy to handle missing values can have a substantial impact on improving model accuracy.
+\newline\textbf{Methods and Results} We applied thirteen different strategies to impute missing values in combination with three different strategies to fit a risk prediction model for mortality and transplant after receiving mechanical circulatory support.
+Model performance was measured by 12-month discrimination, calibration and net reclassification index.
+Results indicated that multiple  imputation consistently provided prediction models with greater prognostic accuracy than other missing data strategies, particularly imputation to the mean.
+<fill in results> \newline\textbf{Conclusion} Selecting an optimal strategy to handle missing values impacts the prognostic accuracy of downstream models.
 In the current analysis, multiple imputation emerged as an optimal strategy to handle missing values in the INTERMACS data.
-The evaluation and selection of the optimal missing data strategy in this work has the potential to improve risk predictions for other longitudinal registries.
+The current study shows that evaluation and selection of an optimal strategy to impute missing data has the potential to improve prognostic accuracy of risk predictions for other longitudinal registries.
 keywords:
@@
 header-includes:
-   - \usepackage{placeins}
    - \newcommand{\ie}{\textit{i.e., }}
@@
 
+```{r setup, include = FALSE}
+
+knitr::opts_chunk$set(echo = FALSE,
+                      message = FALSE,
+                      warning = FALSE,
+                      dpi = 300)
+
+drake::loadd(im)
+drake::loadd(times)
+drake::loadd(tbl_characteristics)
+drake::loadd(tbl_md_strat_auc)
+
+library(table.glue)
+
+```
+
 # Introduction
 \label{sec:introduction}
+
+\linenumbers
+
+<!-- Heart disease / heart failure -->
 
@@
 The 2-year survival on MCS ranges from 61\% for destination therapy to 78\% for bridge-to-transplant \cite{patel2014contemporary}.
-Therefore, there is great need for reliable predictions of patient-specific probability to experience adverse events after receiving MCS.
-This information can be used to improve patient selection for MCS,inform the design of next generation pumps, and refine patient management strategies.
+Therefore, there is great need for reliable predictions of patient-specific risk to experience adverse events after receiving MCS.
+This information can be used to improve patient selection for MCS, inform the design of next generation pumps, and refine patient management strategies.
+
+<!-- INTERMACS -->
+
+The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nation wide level.
+Currently, INTERMACS comprises over 20,000 patients who have received a MCS device.
+As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{kirklin2017eighth, kormos2019society, Adamo950}.
+A critical step for developing risk prediction models is handling missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial.
+While some missing values are related to data entry error, others may relate to instability of a patient’s circulatory status at the time of implant.
+For instance, hemodynamic laboratory values are missing for patients in whom  invasive catheterizations were not performed.
+Patients without invasive cathaderizations  may be too ill (unstable) to tolerate the procedure.
+Instead, such patients may, after determination of basic hemodynamics, proceed directly to device placement.
+
+<!-- Aims of this article -->
+
+The primary aim of this article is to quantify how much the prognostic value of a risk prediction model developed from the INTERMACS registry depends on the strategy that was applied to impute missing data prior to developing the model.
+A secondary aim is to measure imputation accuracy of each strategy by introducing varying levels of artificial missing data and then imputing it.
+Because imputation to the mean has been a standard method for multiple annual summaries of the INTERMACS data, we measure the potential improvement in prognostic value of a risk prediction model when other strategies are applied to impute missing data \emph{instead of imputation to the mean}.
+The over-arching aim is to clarify which imputation strategies are most likely to improve risk prediction (and by extension, quality of care) for patients who receive MCS devices.
+This investigation can directly inform future analyses of INTERMACS data and provide evidence quantifying the benefit of imputing missing data with sound methodology.
+
+# Methods
+\label{sec:methods}
+
+<!-- Section summary -->
+
+Here we describe the INTERMACS Registry (Section \ref{subsec:intermacs}) and give a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning (Section \ref{subsec:inference_and_learning}).
+We then describe thirteen strategies to handle missing values (Section \ref{subsec:imputation}) and three strategies to develop risk prediction models after missing data have been imputed (Section \ref{subsec:modeling}).
+Next, we describe how risk prediction models and by extension the imputation strategy applied before training the model were evaluated in Section \ref{subsec:evaluation}.
+Last, we outline an internal validation procedure that was applied to compare combinations of imputation and modeling strategies using hierarchical Bayesian models in Section \ref{subsec:internal}.
+
+## INTERMACS Registry
+\label{subsec:intermacs}
+
+INTERMACS is a North American observational registry for patients receiving MCS devices that began as a partnership between the National Heart, Lung, and Blood Institute, US Food and Drug Administration, the Centers for Medicaid and Medicare Services, industry, and individual hospitals with the mission of improving MCS outcomes.
+In 2018, INTERMACS became an official Society of Thoracic Surgeons database.
+
+The current analysis was conducted using publicly available data provided by the National Heart, Lung, and Blood Institute.
+We included a contemporary cohort of `r table_value(nrow(im))` patients who received continuous flow LVAD from `r min(im$im_impl_yr)`-`r max(im$im_impl_yr)`.
+Patient follow-up begins after implantation of a durable, long term MCS device and continues while the device is in place.
+Registry endpoints include death on a device, heart transplantation, or cessation of support (for recovery and non-recovery reasons).
+INTERMACS collects pre-implant patient characteristics, medical status, laboratory values, and many other variables.
+Data is collected at regularly scheduled follow-up as well as during adverse events such as re-hospitalization.
+This is a secondary analysis of de-identified data obtained from the National Heart Lung and Blood Institute.
+Primary data collection is approved through University of Alabama Institutional Review Board and at individual sites.
+
+## Statistical Inference and Learning with Missing data 
+\label{subsec:inference_and_learning}
+
+\paragraph{Statistical Inference} Previous research has thoroughly investigated statistical inference in the context of missing data.
+In this setting, analysts often create multiple imputed datasets, replicate an analysis in each of them, and then pool their results to obtain valid test statistics for hypothesis testing \cite{rubin2004multiple}.
+Imputation strategies that create a single imputed dataset (\eg imputation to the mean) have been shown to increase type I errors (\ie rejecting a true null hypothesis) for inferential statistics by artificially reducing the variance of observed data and ignoring the uncertainty attributed to missing values \cite{van2018flexible}.
+When very few data are missing, analysts may apply listwise deletion, which removes any observation with at least one missing value and produces a dataset with fewer observations and no missing values.
+
+\paragraph{Statistical Learning} In the presence of missing data, the goal of supervised statistical learning is to develop a prediction function for an outcome variable that accurately generalizes to testing data, which may or may not contain missing values \cite{hastie2009elements}.
+Because testing data may contain missing values, listwise deletion is not a feasible strategy for statistical learning tasks.
+In contrast to statistical inference, strategies that create a single imputed dataset are often used for statistical learning \cite{kuhn2019feature}.
+Previous work has emphasized the importance of imputation strategies with greater accuracy, suggesting that more accurate imputation strategies lead to better performance of downstream models (\ie models fitted to the imputed data) \cite{jerez2010missing}.
+Others have shown that imputation to the mean, model-based imputation, and multiple imputation can provide Bayes-optimal prediction models provided certain assumptions are true, but these assumptions are difficult to validate in applied settings \cite{josse2019consistency}.
+
+## Missing Data Strategies
+\label{subsec:imputation}
+
+\paragraph{Single and multiple imputation} Several of the imputation methods we considered allowed for creation of one or many imputed datasets.
+For downstream models fitted to multiple sets of imputed data, we applied the modeling technique to each imputed training set, separately, and then used a pooling technique to generate a single set of predictions based on multiple imputed testing data.
+Specifically, we created 10 imputed training and testing sets, then applied a modeling procedure to each imputed training set and computed model predictions on the corresponding imputed testing set, which led to 10 sets of predictions.
+To aggregate these predictions, we computed the median for each observation.
+Informal experiments where the mean was used instead of the median to aggregate predictions showed little or no difference in the model's prediction accuracy.
+
+\paragraph{Imputation to the mean} Imputing data to the mean involves three steps.
+First, numeric and nominal variables are identified.
+Second, for each numeric variable, the mean is computed and used to impute missing values in the corresponding variable.
+Third, for each nominal variable, the mode is computed (because one cannot compute a mean for categorical variables) and used to impute missing values in the corresponding variable.
+The computed means and modes are then stored for future imputation of testing data.
+Because imputation to the mean is frequently used in practice, we use it as a reference for comparison of all other strategies to impute missing values,
+
+\paragraph{Bayesian Regression} Imputation with Bayesian regression draws imputed values from the posterior distribution of model parameters, accounting for uncertainty in both model error and estimation \cite{rubin2004multiple}.
+Multiple imputation with chained equations (MICE), a well known technique for generating multiply imputed data \cite{azur2011multiple, van2018flexible, van2006fully}.
+In each iteration of MICE, each specified variable in the dataset is imputed using the other variables in the dataset.
+These iterations are run until convergence criteria have been met.
+
+\paragraph{Predictive Mean Matching (PMM)} PMM computes predicted values from a pre-specified model that treats one incomplete column in the data, $X$, as a dependent variable.
+For each missing value in $X$, PMM identifies a set of candidate donors based on distance in predicted values of $X$ \cite{landerman1997empirical}.
+One donor is randomly drawn from the candidates, and the observed value of the donor is taken to replace the missing value.
+When missing values were imputed using PMM, we applied MICE to form multiple imputed datasets.
+For consistency with other approaches, we also generated a single imputed dataset using PMM by randomly selecting one of the multiple datasets imputed.
+
+\paragraph{K-Nearest-Neighbors (KNN)} KNN imputation identifies $k$ `similar' observations (\ie neighbors) for each observation with a missing value in a given variable \cite{chen2000nearest}.
+A donor value for the current missing value is generated by sampling or aggregating the values from the $k$ nearest neighbors.
+In the current analysis, we identified 10 nearest neighbors using Gower's distance \cite{gower}.
+When imputing a single dataset using KNN, we aggregated values from nearest neighbors using the median for numeric variables and the mode for categorical variables.
+When imputing multiple datasets using KNN, we sampled one value at random from nearest neighbors for each imputed set.
+
+\paragraph{Hot Deck} Similar to KNN, hot deck imputation finds $k$ similar observations for each observation with a missing value in a given variable \cite{andridge2010hotdeck}.
+However, hot deck imputation uses a less computationally intensive approach, either identifying neighbors at random or using a subset of variables to find similar observations.
+When imputing a single dataset using hot deck imputation, we used a selection of 5 variables simultaneously to identify nearest neighbors.
+When imputing multiple datasets using hot deck imputation, we used a separate numeric variable for each dataset.
+
+\paragraph{Random forests} The random forest framework can be used for imputation as well as prediction.
+In the current analysis, we implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}.
+\texttt{miceRanger} follows an iterative procedure where one random forest is fit to each variable in a dataset, separately, and then used to impute missing values in that variable.
+The procedure may repeat until estimated generalization error of the random forest collection stabilizes.
+When imputing a single dataset with \texttt{miceRanger}, we did not use PMM and instead imputed missing values with a prediction.
+When imputing multiple datasets, we implemented PMM with 10 donor candidates.
+
+\paragraph{Missingness incorporated as an attribute (MIA)} MIA adds another observed category to nominal variables: "Missing".
+For numeric variables, MIA creates two columns: one where missing values are imputed with positive infinity and the other with negative infinity.
+When a decision tree uses a finite cut-point to split a given numeric variable, it will assess the cut-point using both the positive and negative infinite imputed columns, and utilize whichever column provides the best split of the current data.
+This procedure translates to sending all missing values to the left or to the right when forming two new nodes of the decision tree, using whichever direction results in a better split.
+In the current analysis, we used the MIA strategy only for ensembles of boosted decision trees, as it is a well known and frequently used strategy for these models \cite{twala2009empirical, ding2010investigation}.
+
+\paragraph{Surrogate splitting} When decision trees search for a variable and cut-point to use for growing two new nodes in a decision tree, surrogate splitting will temporarily ignore missing values while searching for an initial split.
+Once an initial split is chosen, surrogate splitting searches for a split on another variable with observed values for the observations with missing values on the initial splitting variable and creates a surrogate split based on the new variable that is similar to the original split.
+In the current analysis, we used a maximum of three surrogate splits and sent any remaining missing values to the daughter node with higher sample size.
+Additionally, we only used surrogate splitting for conditional inference forests, as this strategy is well known and frequently used for these models.
+
+\paragraph{Assumptions} Each missing data strategy poses different assumptions regarding the data and the mechanisms that lead to missing values in the data.
+Imputation using Bayesian regression makes the same distributional assumptions as the Bayesian models that are applied.
+The \texttt{miceRanger} algorithm does not make any formal distributional assumptions, as random forests are non-parametric and can thus handle skewed and multi-modal data as well as categorical data that are ordinal or non-ordinal.
+Hot deck and KNN imputation also do not make distributional assumptions, but implicitly assume that a missing value for a given observation can be approximated by aggregating observed values from the $k$ most similar observations.
+PMM makes a similar implicit assumption, but is slightly more robust to skewed or multi-model data because imputed values are sampled directly from observed ones.
+MIA operates based on an implicit assumption that missingness itself is informative and surrogate splitting presumes that variables with missing values are related to other observed variables used by the decision tree.
+It is difficult to validate these assumptions in applied settings and also likely that downstream models will perform poorly if an imputation technique's assumptions are invalid.
+
+## Risk Prediction Models
+\label{subsec:modeling}
+
+We applied three modeling strategies after imputing missing values:
+
+- Cox proportional hazards (PH) model with forward stepwise variable selection
+- gradient boosted decision trees (hereafter referred to as 'boosting')
+- Conditional inference forest (CIF, an extension of random forests)
+
+A thorough description of stepwise variable selection, boosting, and random forests can be found in Sections 6.1.2, 8.2.2, and 8.2.3, respectively, of \emph{Introduction to Statistical Learning} \cite{james2013introduction}.
+Sir David Cox's PH model is one of the most frequently applied methods for the analysis of right-censored time-to-event outcomes \cite{kleinbaum2010survival}.
+According to the PH assumption, the effect of a unit increase in a predictor is multiplicative with respect to a baseline hazard function.
+Boosting grows a sequence of decision trees, each using information from the previous trees in an attempt to correct their errors \cite{friedman2001greedy, chen2016xgboost}.
+In contrast, random forests grow an ensemble of de-correlated decision trees, where each tree is grown using a bootstrapped replicate of the original training data \cite{Breiman2001}.
+The CIF obtains unbiased recursive partitioning by applying early-stopping and permutation tests and is one of several methods that extends random forests to right-censored time-to-event data \cite{hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
+
+## Evaluation of Predictions 
+\label{subsec:evaluation}
+
+\paragraph{The Brier score} The prognostic value of each risk prediction model was primarily assessed using the Brier score, which depends on both the discrimination and calibration of predicted risk values \cite{graf1999assessment, rufibach2010use}.
+Let $\tilde{Y}_i(t)$ represent the observed status of individual $i$ at time $t > 0$ in a testing set of $M$ observations.
+Suppose $\tilde{Y}_i(t)=1$ if there is an observed event at or before $t$ and $\tilde{Y}_i(t)=0$ otherwise.
+The Brier score is computed with \begin{equation} \label{eqn:brier_score}
+\bstat = \frac{1}{M} \sum_{i=1}^{M} \widehat{W}_i(t) \left\{ \tilde{Y}_i(t) - \widehat{S}(t \mid \bm{x}_i) \right\}^2, \end{equation} where, for the $i^{th}$ observation, $\widehat{S}(t \mid \bm{x}_i)$ is the estimated probability of survival at time $t$ according to a given risk prediction model, $\bm{x}_i$ is the set of input values for predictor variables in the model, and $\widehat{W}_i(t)$ is the inverse proportional censoring weight at time $t$ \cite{gerds2006consistent}.
+Thus, the Brier score is the mean squared difference between observed event status and expected event status according to a RPE at time $t$.
+Throughout the current analysis, we set $t = `r times`$ months after receiving MCS to focus on short term risk prediction.
+Models have been developed to simultaneously predict short term and long term mortality risk after receiving MCS, but these are beyond the scope of the current study \cite{blackstone1986decomposition}.
+
+\paragraph{The scaled Brier score}
+
+The Brier score is dependent on the rate of observed events, which can make it a difficult metric to interpret.
+It is often more informative to scale the Brier score based on the Brier score of a naive model.
+More specifically, for a given risk prediction model, the scaled Brier score is computed as $$\textrm{Scaled } \bstat \textrm{ of model } = 1 - \frac{\bstat \textrm{ of model}}{\bstat \textrm{ of naive model}}.$$ As the Brier score for risk prediction is analogous to mean-squared error for prediction of a continuous outcome, the scaled Brier score is analogous to the $R^2$ statistic.
+Similar to the $R^2$ statistic, a scaled $\bstat$ of 1.00 and 0.00 indicate a perfect and worthless model, respectively.
+In our analyses, a Kaplan-Meier estimate based on the training data (\ie a risk prediction model that did not use any predictor variables) provided the naive prediction.
+
+## Internal Validation via Monte-Carlo Cross-Validation (MCCV)
+\label{subsec:internal}
+
+To assess the prognostic value of each missing data strategy, we internally validated a total of FILL IN modeling algorithm based on combinations of imputation strategies and modeling strategies described in Sections \ref{subsec:imputation} and \ref{subsec:modeling}.
+For convenience, we use the term `modeling algorithm' to denote the combination of a missing data strategy and a modeling strategy (\eg imputation using mean/mode values followed by fitting the PH model with stepwise variable selection) \cite{kuhn2013applied}.
+We conducted internal validation using 150 replicates of Monte-Carlo cross validation (MCCV), a resampling technique for internal validation (Figure \ref{fig:monte_carlo_cv}).
+
+\paragraph{Steps taken in each MCCV replicate} In each replicate, 50\% of the available data were used for model training and testing.
+Imputation was conducted in the training and testing sets, separately, for each imputation strategy.
+Some imputation strategies (\eg KNN and random forests) used only the training data to generate imputed values, whereas others (\eg Bayesian regression, PMM, and hot deck)
+
+After imputation, each modeling algorithm (\ie stepwise PH models, GBDTs, and the cIF) was applied, separately, to develop a RPE for death or transplant that incorporated up to 50 predictor variables from the entire collection of covariates available.

### 2020-10-10 doc_arXiv/doc_arXiv.tex
preparing to drop RF model

@@
 from the Interagency Registry for Mechanically Assisted Circulatory
-Support (INTERMACS). \newline\textbf{Methods and Results} We applied all
-pairwise combinations of different strategies to impute missing values
-and three different strategies to fit a risk prediction model for
+Support (INTERMACS). \newline\textbf{Methods and Results} We applied
+thirteen different strategies to impute missing values in combination
+with three different strategies to fit a risk prediction model for
 mortality and transplant after receiving mechanical circulatory support.
-Model performance was compared using Concordance (i.e.~C-index), Brier
-Score, and Net reclassification index. Results indicated that multiple
-imputation consistently resulted in superior prognostic models compared
-to other missing data strategies, particularly imputation to the mean.
-\newline\textbf{Conclusion} Selecting an optimal strategy to handle
-missing values can have a substantial impact on improving model
-accuracy. In the current analysis, multiple imputation emerged as an
-optimal strategy to handle missing values in the INTERMACS data. The
-evaluation and selection of the optimal missing data strategy in this
-work has the potential to improve risk predictions for other
-longitudinal registries.
+Model performance was measured by 12-month discrimination, calibration
+and net reclassification index. Results indicated that multiple
+imputation consistently provided prediction models with greater
+prognostic accuracy than other missing data strategies, particularly
+imputation to the mean. \newline\textbf{Conclusion} Selecting an optimal
+strategy to handle missing values impacts the prognostic accuracy of
+downstream models. In the current analysis, multiple imputation emerged
+as an optimal strategy to handle missing values in the INTERMACS data.
+The current study shows that evaluation and selection of an optimal
+strategy to impute missing data has the potential to improve prognostic
+accuracy of risk predictions for other longitudinal registries.
 \end{abstract}
@@
 
+\linenumbers
+
 Heart disease is a leading cause of death in the United States. Heart
@@
 \cite{patel2014contemporary}. Therefore, there is great need for
-reliable predictions of patient-specific probability to experience
-adverse events after receiving MCS. This information can be used to
-improve patient selection for MCS,inform the design of next generation
-pumps, and refine patient management strategies.
+reliable predictions of patient-specific risk to experience adverse
+events after receiving MCS. This information can be used to improve
+patient selection for MCS, inform the design of next generation pumps,
+and refine patient management strategies.
+
+The Interagency Registry for Mechanically Assisted Circulatory Support
+(INTERMACS) was launched to improve MCS patient outcomes through the
+collection of patient characteristics, medical events, and long terms
+outcomes at a nation wide level. Currently, INTERMACS comprises over
+20,000 patients who have received a MCS device. As the largest registry
+for data on patients receiving MCS devices, INTERMACS has been leveraged
+to develop numerous risk prediction models for mortality and other types
+of adverse events that may occur after receiving a device
+\cite{kirklin2017eighth, kormos2019society, Adamo950}. A critical step
+for developing risk prediction models is handling missing data, which is
+common in a ``real world'' database that is not designed for the rigor
+of data completeness found in a clinical trial. While some missing
+values are related to data entry error, others may relate to instability
+of a patient's circulatory status at the time of implant. For instance,
+hemodynamic laboratory values are missing for patients in whom invasive
+catheterizations were not performed. Patients without invasive
+cathaderizations may be too ill (unstable) to tolerate the procedure.
+Instead, such patients may, after determination of basic hemodynamics,
+proceed directly to device placement.
+
+The primary aim of this article is to quantify how much the prognostic
+value of a risk prediction model developed from the INTERMACS registry
+depends on the strategy that was applied to impute missing data prior to
+developing the model. A secondary aim is to measure imputation accuracy
+of each strategy by introducing varying levels of artificial missing
+data and then imputing it. Because imputation to the mean has been a
+standard method for multiple annual summaries of the INTERMACS data, we
+measure the potential improvement in prognostic value of a risk
+prediction model when other strategies are applied to impute missing
+data \emph{instead of imputation to the mean}. The over-arching aim is
+to clarify which imputation strategies are most likely to improve risk
+prediction (and by extension, quality of care) for patients who receive
+MCS devices. This investigation can directly inform future analyses of
+INTERMACS data and provide evidence quantifying the benefit of imputing
+missing data with sound methodology.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\label{sec:methods}
+
+Here we describe the INTERMACS Registry (Section \ref{subsec:intermacs})
+and give a broad overview of general objectives for imputation of
+missing data in the context of statistical inference and supervised
+statistical learning (Section \ref{subsec:inference_and_learning}). We
+then describe thirteen strategies to handle missing values (Section
+\ref{subsec:imputation}) and three strategies to develop risk prediction
+models after missing data have been imputed (Section
+\ref{subsec:modeling}). Next, we describe how risk prediction models and
+by extension the imputation strategy applied before training the model
+were evaluated in Section \ref{subsec:evaluation}. Last, we outline an
+internal validation procedure that was applied to compare combinations
+of imputation and modeling strategies using hierarchical Bayesian models
+in Section \ref{subsec:internal}.
+
+\hypertarget{intermacs-registry}{%
+\subsection{INTERMACS Registry}\label{intermacs-registry}}
+
+\label{subsec:intermacs}
+
+INTERMACS is a North American observational registry for patients
+receiving MCS devices that began as a partnership between the National
+Heart, Lung, and Blood Institute, US Food and Drug Administration, the
+Centers for Medicaid and Medicare Services, industry, and individual
+hospitals with the mission of improving MCS outcomes. In 2018, INTERMACS
+became an official Society of Thoracic Surgeons database.
+
+The current analysis was conducted using publicly available data
+provided by the National Heart, Lung, and Blood Institute. We included a
+contemporary cohort of 14,738 patients who received continuous flow LVAD
+from 2012-2017. Patient follow-up begins after implantation of a
+durable, long term MCS device and continues while the device is in
+place. Registry endpoints include death on a device, heart
+transplantation, or cessation of support (for recovery and non-recovery
+reasons). INTERMACS collects pre-implant patient characteristics,
+medical status, laboratory values, and many other variables. Data is
+collected at regularly scheduled follow-up as well as during adverse
+events such as re-hospitalization. This is a secondary analysis of
+de-identified data obtained from the National Heart Lung and Blood
+Institute. Primary data collection is approved through University of
+Alabama Institutional Review Board and at individual sites.
+
+\hypertarget{statistical-inference-and-learning-with-missing-data}{%
+\subsection{Statistical Inference and Learning with Missing
+data}\label{statistical-inference-and-learning-with-missing-data}}
+
+\label{subsec:inference_and_learning}
+
+\paragraph{Statistical Inference}
+
+Previous research has thoroughly investigated statistical inference in
+the context of missing data. In this setting, analysts often create
+multiple imputed datasets, replicate an analysis in each of them, and
+then pool their results to obtain valid test statistics for hypothesis
+testing \cite{rubin2004multiple}. Imputation strategies that create a
+single imputed dataset (\textit{e.g., }imputation to the mean) have been
+shown to increase type I errors (\textit{i.e., }rejecting a true null
+hypothesis) for inferential statistics by artificially reducing the
+variance of observed data and ignoring the uncertainty attributed to
+missing values \cite{van2018flexible}. When very few data are missing,
+analysts may apply listwise deletion, which removes any observation with
+at least one missing value and produces a dataset with fewer
+observations and no missing values.
+
+\paragraph{Statistical Learning}
+
+In the presence of missing data, the goal of supervised statistical
+learning is to develop a prediction function for an outcome variable
+that accurately generalizes to testing data, which may or may not
+contain missing values \cite{hastie2009elements}. Because testing data
+may contain missing values, listwise deletion is not a feasible strategy
+for statistical learning tasks. In contrast to statistical inference,
+strategies that create a single imputed dataset are often used for
+statistical learning \cite{kuhn2019feature}. Previous work has
+emphasized the importance of imputation strategies with greater
+accuracy, suggesting that more accurate imputation strategies lead to
+better performance of downstream models (\textit{i.e., }models fitted to
+the imputed data) \cite{jerez2010missing}. Others have shown that
+imputation to the mean, model-based imputation, and multiple imputation
+can provide Bayes-optimal prediction models provided certain assumptions
+are true, but these assumptions are difficult to validate in applied
+settings \cite{josse2019consistency}.
+
+\hypertarget{missing-data-strategies}{%
+\subsection{Missing Data Strategies}\label{missing-data-strategies}}
+
+\label{subsec:imputation}
+
+\paragraph{Single and multiple imputation}
+
+Several of the imputation methods we considered allowed for creation of
+one or many imputed datasets. For downstream models fitted to multiple
+sets of imputed data, we applied the modeling technique to each imputed
+training set, separately, and then used a pooling technique to generate
+a single set of predictions based on multiple imputed testing data.
+Specifically, we created 10 imputed training and testing sets, then
+applied a modeling procedure to each imputed training set and computed
+model predictions on the corresponding imputed testing set, which led to
+10 sets of predictions. To aggregate these predictions, we computed the
+median for each observation. Informal experiments where the mean was
+used instead of the median to aggregate predictions showed little or no
+difference in the model's prediction accuracy.
+
+\paragraph{Imputation to the mean}
+
+Imputing data to the mean involves three steps. First, numeric and
+nominal variables are identified. Second, for each numeric variable, the
+mean is computed and used to impute missing values in the corresponding
+variable. Third, for each nominal variable, the mode is computed
+(because one cannot compute a mean for categorical variables) and used
+to impute missing values in the corresponding variable. The computed
+means and modes are then stored for future imputation of testing data.
+Because imputation to the mean is frequently used in practice, we use it
+as a reference for comparison of all other strategies to impute missing
+values,
+
+\paragraph{Bayesian Regression}
+
+Imputation with Bayesian regression draws imputed values from the
+posterior distribution of model parameters, accounting for uncertainty
+in both model error and estimation \cite{rubin2004multiple}. Multiple
+imputation with chained equations (MICE), a well known technique for
+generating multiply imputed data
+\cite{azur2011multiple, van2018flexible, van2006fully}. In each
+iteration of MICE, each specified variable in the dataset is imputed
+using the other variables in the dataset. These iterations are run until
+convergence criteria have been met.
+
+\paragraph{Predictive Mean Matching (PMM)}
+
+PMM computes predicted values from a pre-specified model that treats one
+incomplete column in the data, \(X\), as a dependent variable. For each
+missing value in \(X\), PMM identifies a set of candidate donors based
+on distance in predicted values of \(X\) \cite{landerman1997empirical}.
+One donor is randomly drawn from the candidates, and the observed value
+of the donor is taken to replace the missing value. When missing values
+were imputed using PMM, we applied MICE to form multiple imputed
+datasets. For consistency with other approaches, we also generated a
+single imputed dataset using PMM by randomly selecting one of the
+multiple datasets imputed.
+
+\paragraph{K-Nearest-Neighbors (KNN)}
+
+KNN imputation identifies \(k\) `similar' observations
+(\textit{i.e., }neighbors) for each observation with a missing value in
+a given variable \cite{chen2000nearest}. A donor value for the current
+missing value is generated by sampling or aggregating the values from
+the \(k\) nearest neighbors. In the current analysis, we identified 10
+nearest neighbors using Gower's distance \cite{gower}. When imputing a
+single dataset using KNN, we aggregated values from nearest neighbors
+using the median for numeric variables and the mode for categorical
+variables. When imputing multiple datasets using KNN, we sampled one
+value at random from nearest neighbors for each imputed set.
+
+\paragraph{Hot Deck}
+
+Similar to KNN, hot deck imputation finds \(k\) similar observations for
+each observation with a missing value in a given variable
+\cite{andridge2010hotdeck}. However, hot deck imputation uses a less
+computationally intensive approach, either identifying neighbors at
+random or using a subset of variables to find similar observations. When
+imputing a single dataset using hot deck imputation, we used a selection
+of 5 variables simultaneously to identify nearest neighbors. When
+imputing multiple datasets using hot deck imputation, we used a separate
+numeric variable for each dataset.
+
+\paragraph{Random forests}
+
+The random forest framework can be used for imputation as well as
+prediction. In the current analysis, we implement \texttt{miceRanger},
+an R package similar to \texttt{missForest} that implements MICE with
+random forests by implementing PMM \cite{missForest}.
+\texttt{miceRanger} follows an iterative procedure where one random
+forest is fit to each variable in a dataset, separately, and then used
+to impute missing values in that variable. The procedure may repeat
+until estimated generalization error of the random forest collection
+stabilizes. When imputing a single dataset with \texttt{miceRanger}, we
+did not use PMM and instead imputed missing values with a prediction.
+When imputing multiple datasets, we implemented PMM with 10 donor
+candidates.
+
+\paragraph{Missingness incorporated as an attribute (MIA)}
+
+MIA adds another observed category to nominal variables: ``Missing''.
+For numeric variables, MIA creates two columns: one where missing values
+are imputed with positive infinity and the other with negative infinity.
+When a decision tree uses a finite cut-point to split a given numeric
+variable, it will assess the cut-point using both the positive and
+negative infinite imputed columns, and utilize whichever column provides
+the best split of the current data. This procedure translates to sending
+all missing values to the left or to the right when forming two new
+nodes of the decision tree, using whichever direction results in a
+better split. In the current analysis, we used the MIA strategy only for
+ensembles of boosted decision trees, as it is a well known and
+frequently used strategy for these models
+\cite{twala2009empirical, ding2010investigation}.
+
+\paragraph{Surrogate splitting}
+
+When decision trees search for a variable and cut-point to use for
+growing two new nodes in a decision tree, surrogate splitting will
+temporarily ignore missing values while searching for an initial split.
+Once an initial split is chosen, surrogate splitting searches for a
+split on another variable with observed values for the observations with
+missing values on the initial splitting variable and creates a surrogate
+split based on the new variable that is similar to the original split.
+In the current analysis, we used a maximum of three surrogate splits and
+sent any remaining missing values to the daughter node with higher
+sample size. Additionally, we only used surrogate splitting for
+conditional inference forests, as this strategy is well known and
+frequently used for these models.
+
+\paragraph{Assumptions}
+
+Each missing data strategy poses different assumptions regarding the
+data and the mechanisms that lead to missing values in the data.
+Imputation using Bayesian regression makes the same distributional
+assumptions as the Bayesian models that are applied. The
+\texttt{miceRanger} algorithm does not make any formal distributional
+assumptions, as random forests are non-parametric and can thus handle
+skewed and multi-modal data as well as categorical data that are ordinal
+or non-ordinal. Hot deck and KNN imputation also do not make
+distributional assumptions, but implicitly assume that a missing value
+for a given observation can be approximated by aggregating observed
+values from the \(k\) most similar observations. PMM makes a similar
+implicit assumption, but is slightly more robust to skewed or
+multi-model data because imputed values are sampled directly from
+observed ones. MIA operates based on an implicit assumption that
+missingness itself is informative and surrogate splitting presumes that
+variables with missing values are related to other observed variables
+used by the decision tree. It is difficult to validate these assumptions
+in applied settings and also likely that downstream models will perform
+poorly if an imputation technique's assumptions are invalid.
+
+\hypertarget{risk-prediction-models}{%
+\subsection{Risk Prediction Models}\label{risk-prediction-models}}
+
+\label{subsec:modeling}
+
+We applied three modeling strategies after imputing missing values:
+
+\begin{itemize}
+\tightlist
+\item
+  Cox proportional hazards (PH) model with forward stepwise variable
+  selection
+\item
+  gradient boosted decision trees (hereafter referred to as `boosting')
+\item
+  Conditional inference forest (CIF, an extension of random forests)
+\end{itemize}
+
+A thorough description of stepwise variable selection, boosting, and
+random forests can be found in Sections 6.1.2, 8.2.2, and 8.2.3,
+respectively, of \emph{Introduction to Statistical Learning}
+\cite{james2013introduction}. Sir David Cox's PH model is one of the
+most frequently applied methods for the analysis of right-censored
+time-to-event outcomes \cite{kleinbaum2010survival}. According to the PH
+assumption, the effect of a unit increase in a predictor is
+multiplicative with respect to a baseline hazard function. Boosting
+grows a sequence of decision trees, each using information from the
+previous trees in an attempt to correct their errors
+\cite{friedman2001greedy, chen2016xgboost}. In contrast, random forests
+grow an ensemble of de-correlated decision trees, where each tree is
+grown using a bootstrapped replicate of the original training data
+\cite{Breiman2001}. The CIF obtains unbiased recursive partitioning by
+applying early-stopping and permutation tests and is one of several
+methods that extends random forests to right-censored time-to-event data
+\cite{hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
+
+\hypertarget{evaluation-of-predictions}{%
+\subsection{Evaluation of Predictions}\label{evaluation-of-predictions}}
+
+\label{subsec:evaluation}
+
+\paragraph{The Brier score}
+
+The prognostic value of each risk prediction model was primarily
+assessed using the Brier score, which depends on both the discrimination
+and calibration of predicted risk values
+\cite{graf1999assessment, rufibach2010use}. Let \(\tilde{Y}_i(t)\)
+represent the observed status of individual \(i\) at time \(t > 0\) in a
+testing set of \(M\) observations. Suppose \(\tilde{Y}_i(t)=1\) if there
+is an observed event at or before \(t\) and \(\tilde{Y}_i(t)=0\)
+otherwise. The Brier score is computed with
+\begin{equation} \label{eqn:brier_score}
+\widehat{\textrm{BS}}(t)= \frac{1}{M} \sum_{i=1}^{M} \widehat{W}_i(t) \left\{ \tilde{Y}_i(t) - \widehat{S}(t \mid \bm{x}_i) \right\}^2, \end{equation}
+where, for the \(i^{th}\) observation, \(\widehat{S}(t \mid \bm{x}_i)\)
+is the estimated probability of survival at time \(t\) according to a
+given risk prediction model, \(\bm{x}_i\) is the set of input values for
+predictor variables in the model, and \(\widehat{W}_i(t)\) is the
+inverse proportional censoring weight at time \(t\)
+\cite{gerds2006consistent}. Thus, the Brier score is the mean squared
+difference between observed event status and expected event status
+according to a RPE at time \(t\). Throughout the current analysis, we
+set \(t = 6\) months after receiving MCS to focus on short term risk
+prediction. Models have been developed to simultaneously predict short
+term and long term mortality risk after receiving MCS, but these are
+beyond the scope of the current study
+\cite{blackstone1986decomposition}.
+
+\paragraph{The scaled Brier score}
+
+The Brier score is dependent on the rate of observed events, which can
+make it a difficult metric to interpret. It is often more informative to
+scale the Brier score based on the Brier score of a naive model. More
+specifically, for a given risk prediction model, the scaled Brier score
+is computed as
+\[\textrm{Scaled } \widehat{\textrm{BS}}(t)\textrm{ of model } = 1 - \frac{\widehat{\textrm{BS}}(t)\textrm{ of model}}{\widehat{\textrm{BS}}(t)\textrm{ of naive model}}.\]
+As the Brier score for risk prediction is analogous to mean-squared
+error for prediction of a continuous outcome, the scaled Brier score is
+analogous to the \(R^2\) statistic. Similar to the \(R^2\) statistic, a
+scaled \(\widehat{\textrm{BS}}(t)\) of 1.00 and 0.00 indicate a perfect
+and worthless model, respectively. In our analyses, a Kaplan-Meier
+estimate based on the training data (\textit{i.e., }a risk prediction
+model that did not use any predictor variables) provided the naive
+prediction.
+
+\hypertarget{internal-validation-via-monte-carlo-cross-validation-mccv}{%
+\subsection{Internal Validation via Monte-Carlo Cross-Validation
+(MCCV)}\label{internal-validation-via-monte-carlo-cross-validation-mccv}}
+
+\label{subsec:internal}
+
+To assess the prognostic value of each missing data strategy, we
+internally validated a total of FILL IN modeling algorithm based on
+combinations of imputation strategies and modeling strategies described
+in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For
+convenience, we use the term `modeling algorithm' to denote the
+combination of a missing data strategy and a modeling strategy
+(\textit{e.g., }imputation using mean/mode values followed by fitting
+the PH model with stepwise variable selection) \cite{kuhn2013applied}.
+We conducted internal validation using 150 replicates of Monte-Carlo
+cross validation (MCCV), a resampling technique for internal validation
+(Figure \ref{fig:monte_carlo_cv}).
+
+\paragraph{Steps taken in each MCCV replicate}
+
+In each replicate, 50\% of the available data were used for model
+training and testing. Imputation was conducted in the training and
+testing sets, separately, for each imputation strategy. Some imputation
+strategies (\textit{e.g., }KNN and random forests) used only the
+training data to generate imputed values, whereas others
+(\textit{e.g., }Bayesian regression, PMM, and hot deck)
+
+After imputation, each modeling algorithm (\textit{i.e., }stepwise PH
+models, GBDTs, and the cIF) was applied, separately, to develop a RPE
+for death or transplant that incorporated up to 50 predictor variables
+from the entire collection of covariates available.
 

### 2020-10-10 R/clean_intermacs.R

@@ -20,7 +20,7 @@ clean_intermacs <- function(na_string = c('',
     # drop the identifier (ends with 'ID') columns.
     select(-ends_with("_ID")) %>%
     clean_names() %>%
-    filter(im_impl_yr >=2014) %>%
+    filter(im_impl_yr >=2012) %>%
     remove_constant(na.rm = TRUE, quiet = F) %>%
     remove_empty(which = 'cols') %>%
     remove_empty(which = 'rows') %>%

### 2020-10-10 R/evaluate_risk_predictions.R

@@ -10,11 +10,9 @@ evaluate_risk_predictions <- function(risk_predictions,
                                       times,
                                       testing_data) {
 
-
   scores <- risk_predictions %>%
     unite(md_strat, model, col = 'md_strat_mdl', sep = '..') %>%
     select(md_strat_mdl, sprobs) %>%
-    filter(map_lgl(sprobs, ~inherits(.x, 'matrix'))) %>%
     deframe() %>%
     Score(
       formula = Surv(time, status) ~ 1,

### 2020-10-10 R/load_mccv.R

@@ -2,11 +2,29 @@
 ##'
 ##' .. content for \details{} ..
 ##'
+##' @param output_path
+##' @param output_pattern
+##' @param nruns
+##'
 ##' @title
 
-load_mccv <- function() {
+load_mccv <- function(output_path = "slurm/results",
+                      output_pattern = '^output',
+                      nruns = 6000) {
 
-  print("loading data")
-  read_rds("slurm/sim_results.rds")
+  files <- list.files(path = output_path,
+                      pattern = output_pattern,
+                      full.names = TRUE)
+
+  message("A total of ", length(files), " outputs were identified")
+
+  file_nums <- str_extract_all(files, '\\d') %>%
+    map(paste, collapse = '') %>%
+    map_dbl(as.numeric)
+
+  message("Unsuccessful slurm runs:",
+          paste(setdiff(0:(nruns-1), file_nums), collapse = ', '))
+
+  map_dfr(.x = files, .f = ~ read_rds(.x)[[1]])
 
 }

### 2020-10-10 R/make_bayes_mccv_fit.R

@@ -1,0 +1,29 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param risk_evaluation
+make_bayes_mccv_fit <- function(risk_evaluation) {
+
+  model_data <- risk_evaluation %>%
+    mutate(md_strat = fct_relevel(md_strat, 'meanmode_si'))
+
+  stan_lm(
+    formula = ipa ~ outcome + md_strat + model + additional_missing_pct,
+    #+ (additional_missing_pct | iteration),
+    data = model_data,
+    prior = R2(0.30)
+  )
+
+  # model_fits_ipa <- map(
+  #   .x = model_data,
+  #   .f = ~ stan_lmer(
+  #     formula = ipa ~ md_strat + model + additional_missing_pct +
+  #       (additional_missing_pct | iteration),
+  #     data = .x
+  #   )
+  # )
+
+
+}

### 2020-10-10 R/make_risk_evaluation.R

@@ -18,7 +18,7 @@ make_risk_evaluation <- function(mc_cv_results, im, resamples, times) {
 
     .risk_predictions <- mc_cv_results %>%
       filter(iteration == r,
-             !map_lgl(sprobs, ~length(.x)==0)) %>%
+             map_lgl(sprobs, ~all(!is.na(.x)))) %>%
       group_by(outcome, additional_missing_pct) %>%
       nest() %>%
       rename(risk_prediction = data)
@@ -55,7 +55,6 @@ make_risk_evaluation <- function(mc_cv_results, im, resamples, times) {
 
   }
 
-  browser()
   bind_rows(risk_eval, .id = 'iteration')
 
 }

### 2020-10-10 R/plan.R

@@ -5,27 +5,32 @@ the_plan <- drake_plan(
   # when to evaluate models, in months since implant
   times = 6,
 
-  # these two .csv files are needed to complete the slurm runs ----
+  rspec = round_spec() %>%
+    round_using_magnitude(digits = c(2, 2, 1),
+                          breaks = c(1, 10, 100)) %>%
+    round_half_even(),
 
+  # these two .csv files are needed to complete the slurm runs ----
   # creates intermacs_clean.csv
   im = clean_intermacs(),
   # creates resamples.csv
-  resamples = mc_cv_light(data = im, train_prop = 2/3, ntimes = 250),
+  resamples = mc_cv_light(data = im, train_prop = 1/2, ntimes = 250),
 
   # slurm runs need to be completed/cleaned to proceed from here ----
   mc_cv_results = load_mccv(),
+
   risk_evaluation = make_risk_evaluation(mc_cv_results, im, resamples, times),
 
   # labels to describe methods ----
 
   md_method_labels = c(
+    'meanmode' = 'Imputation to the mean',
     'mia' = 'Missingness as an attribute',
-    'ranger' = 'Random forest',
-    'meanmode' = 'Mean or mode',
-    'bayesreg' = 'Bayesian regression',
     'hotdeck' = 'Hot deck',
+    'nbrs' = 'K-nearest-neighbors',
     'pmm' = 'Predictive mean matching',
-    'nbrs' = 'K-nearest-neighbors'
+    'ranger' = 'Random forest',
+    'bayesreg' = 'Bayesian regression'
   ),
 
   md_type_labels = c(
@@ -46,8 +51,8 @@ the_plan <- drake_plan(
 
   additional_missing_labels = c(
     'No additional missing data'  = '0',
-    '+10% additional missing data'= '10',
-    '+20% additional missing data'= '20'
+    '+20% additional missing data'= '20',
+    '+40% additional missing data'= '40'
   ),
 
   fig_risk_evaluation = visualize_risk_evaluation(risk_evaluation,
@@ -58,75 +63,34 @@ the_plan <- drake_plan(
                                                   additional_missing_labels,
                                                   times),
 
+  bayes_mccv_fits = make_bayes_mccv_fit(risk_evaluation),
+
+  fig_md_strat_infer = visualize_md_strat_inference(bayes_mccv_fits,
[…]

### 2020-10-10 R/predict_risk.R

@@ -86,8 +86,9 @@ predict_risk <- function(training,
       verbose  = verbose,
       times    = times
     ) %>%
-      reduce(`+`) %>%
-      divide_by(length(training))
+      reduce(cbind) %>%
+      apply(1, median) %>%
+      matrix(ncol = 1)
 
   }
 

### 2020-10-10 R/predict_risk_cph.R

@@ -17,7 +17,8 @@ predict_risk_cph_si <- function(training,
                                 times) {
 
   # prevent hard error when MIA is used.
-  if(any(is.na(training)) | any(is.na(testing))) return(NULL)
+  if(any(is.na(training)) | any(is.na(testing)))
+    return(matrix(NA_real_, nrow = nrow(testing)))
 
   suppressWarnings(
     expr = {

### 2020-10-10 R/predict_risk_rf.R

@@ -25,19 +25,38 @@ predict_risk_rf_si <- function(training,
   )
 
 
-  message("fitting RSF model")
-  model <- rfsrc.fast(formula = Surv(time, status) ~ .,
-                      data = .trn,
-                      ntree = 1000,
-                      nodesize = 20,
-                      nsplit = 5,
-                      forest = TRUE,
-                      na.action = 'na.impute')
+  message("fitting CIF model")
+
+  # unbiased random forest using pec and party packages.
+  model <- cforest(
+    formula = Surv(time,status) ~ .,
+    data = .trn,
+    controls = cforest_unbiased(maxsurrogate = 3)
+  )
+
+  message("done")
+  message("computing CIF predictions")
+
+  predictions <- treeresponse(model, newdata = .tst, OOB = TRUE) %>%
+    map_dbl(.f = predictRisk,
+        newdata = .tst[1, , drop = FALSE], # this is arbitrary
+        times = times) %>%
+    matrix(ncol = 1)
+
   message("done")
 
-  predictRisk(model,
-              newdata = .tst,
-              times = times,
-              na.action = 'na.impute')
+  predictions
+
+  # a fast randomforest using Ishwaran's package
+  # model <- rfsrc.fast(formula = Surv(time, status) ~ .,
+  #                     data = .trn,
+  #                     ntree = 1000,
+  #                     nodesize = 20,
+  #                     nsplit = 5,
+  #                     forest = TRUE,
+  #                     na.action = 'na.impute')
+  # predictRisk(model,
+  #             newdata = .tst,
+  #             times = times)
 
 }

### 2020-10-10 R/predict_risk_xgb.R

@@ -28,8 +28,8 @@ predict_risk_xgb_si <- function(training,
   xgb_cv_folds <- xgb_folds(data = .trn, nfolds = 10, strata = status)
 
   xgb_cv_tune <- expand.grid(
-    max_depth = c(1,2,3),
-    min_child_weight = c(1,3,5),
+    max_depth = c(2,4),
+    min_child_weight = c(3),
     colsample_bynode = 1/2,
     subsample = 1/2,
     eta = 0.03,

### 2020-10-10 R/tabulate_characteristics.R

@@ -1,0 +1,76 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param im
+tabulate_characteristics <- function(im) {
+
+  tbl <- im %>%
+    mutate(
+      status = case_when(
+        pt_outcome_dead == 1 ~ "Dead",
+        pt_outcome_txpl == 1 ~ "Transplant",
+        pt_outcome_cess == 1 ~ "Cessation of support",
+        TRUE ~ "Censored"
+      ),
+      demo_race = case_when(
+        demo_race_af_amer == 'yes' ~ "Black",
+        demo_race_white == 'yes' ~ "White",
+        TRUE ~ "Other"
+      ),
+      demo_gender = fct_recode(
+        demo_gender,
+        Male = 'male',
+        Female = 'female',
+      ),
+      demo_race = fct_infreq(demo_race),
+      pi_hgt_m = pi_hgt_cm / 100,
+      pi_bmi = pi_wgt_kg / pi_hgt_m^2,
+      pi_device_strategy = fct_recode(
+        pi_device_strategy,
+        'Destination therapy'  = "destination_therapy_patient_definitely_not_eligible_for_transplant",
+        'Bridge to transplant' = "bridge_to_transplant_patient_currently_listed_for_transplant",
+        'Bridge to transplant' = "possible_bridge_to_transplant_likely_to_be_eligible",
+        'Bridge to transplant' = "possible_bridge_to_transplant_moderate_likelihood_of_becoming_eligible",
+        'Bridge to transplant' = "possible_bridge_to_transplant_unlikely_to_become_eligible",
+        'Other' = "bridge_to_recovery",
+        'Other' = 'rescue_therapy',
+        'Other' = 'other_specify'
+      ),
+      im_device_ty = fct_recode(
+        im_device_ty,
+        'Left-ventricular assistance device' = 'lvad',
+        'Bi-ventricular assistance device'   = 'bivad'
+      )
+    ) %>%
+    select(
+      status,
+      months_post_implant,
+      demo_age,
+      demo_gender,
+      demo_race,
+      pi_bmi,
+      pi_device_strategy,
+      im_device_ty
+    ) %>%
+    tbl_summary(
+      by = status,
+      label = list(
+        months_post_implant ~ "Follow up time, months",
+        demo_age ~ "Age, years",
+        demo_gender ~ "Sex",
[…]

### 2020-10-10 R/tabulate_md_strat.R

@@ -1,0 +1,120 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param risk_evaluation
+tabulate_md_strat <- function(risk_evaluation,
+                              md_method_labels,
+                              md_type_labels,
+                              model_labels,
+                              outcome_labels,
+                              additional_missing_labels,
+                              rspec) {
+
+  pm <- function(x){ifelse(x>0, "+", "")}
+
+  output <- risk_evaluation %>%
+    filter(model != 'rf') %>%
+    pivot_longer(cols = c(auc:GND_pvalue), names_to = 'metric') %>%
+    split(f = list(.$outcome, .$metric)) %>%
+    map(
+      ~ {
+
+        mult_by <- if(.x$metric[1] == 'GND_chisq') 1 else 100
+
+        if(str_detect(.x$metric[1], '^nri'))
+          .x$value[.x$md_strat == 'meanmode_si'] <- 0
+
+        data_tbl <- .x %>%
+          select(-outcome, -metric) %>%
+          pivot_wider(values_from = value, names_from = md_strat) %>%
+          mutate(
+            across(
+              .cols = c(
+                mia,
+                ranger_si,
+                ranger_mi,
+                pmm_mi,
+                pmm_si,
+                bayesreg_mi,
+                bayesreg_si,
+                hotdeck_si,
+                hotdeck_mi,
+                nbrs_mi,
+                nbrs_si,
+              ),
+              .fns = ~ .x - meanmode_si
+            )
+          ) %>%
+          pivot_longer(cols = -(iteration:model), names_to = 'md_strat') %>%
+          group_by(model, md_strat, additional_missing_pct) %>%
+          summarize(
+            est = mult_by * median(value, na.rm = TRUE),
+            lwr = mult_by * quantile(value, probs = 0.025, na.rm = TRUE),
+            upr = mult_by * quantile(value, probs = 0.975, na.rm = TRUE)
+          ) %>%
+          ungroup() %>%
+          transmute(
+            model = recode(model, !!!model_labels),
+            additional_missing_pct = factor(
+              additional_missing_pct,
+              levels = additional_missing_labels,
+              labels = names(additional_missing_labels)
+            ),
+            md_strat = str_replace(md_strat, 'mia', 'mia_si'),
+            table_val = if_else(
+              md_strat == 'meanmode_si',
+              true = table_glue("{est} ({lwr}, {upr})", rspec=rspec),
[…]

### 2020-10-10 R/tidy_nbrs.R

@@ -1,0 +1,14 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+
+tidy_nbrs <- function(nbrs_brew){
+
+  nbrs_brew$wort %>%
+    as_tibble() %>%
+    select(-pars) %>%
+    mutate(impute = as.integer(impute))
+
+}

### 2020-10-10 R/tidy_ranger.R

@@ -1,0 +1,25 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param ranger_mi_train
+##' @param ranger_mi_test
+tidy_ranger <- function(ranger_mi_train, ranger_mi_test) {
+
+  tidy_train_imputes <- completeData(ranger_mi_train) %>%
+    set_names(1:length(.)) %>%
+    enframe(name = 'impute', value = 'training') %>%
+    mutate(training = map(training, as_tibble))
+
+  tidy_test_imputes <- ranger_mi_test$imputedData %>%
+    set_names(1:length(.)) %>%
+    enframe(name = 'impute', value = 'testing')
+
+  left_join(tidy_train_imputes,
+            tidy_test_imputes,
+            by = 'impute') %>%
+    mutate(impute = as.integer(impute))
+
+}
+

### 2020-10-10 R/visualize_md_strat_inference.R

@@ -1,0 +1,77 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param bayes_mccv_fits
+visualize_md_strat_inference <- function(bayes_mccv_fits,
+                                         md_type_labels,
+                                         md_method_labels,
+                                         outcome_labels) {
+
+
+  posterior_diffs <- expand.grid(
+    m1 = bayes_mccv_fits$xlevels$md_strat,
+    m2 = bayes_mccv_fits$xlevels$md_strat,
+    additional_missing_pct = c(0, 20, 40),
+    model = c('cph', 'xgb', 'rf'),
+    outcome = c('dead', 'txpl'),
+    stringsAsFactors = FALSE
+  ) %>%
+    as_tibble() %>%
+    filter(m1 == 'meanmode_si') %>%
+    mutate(
+      distribution = pmap(
+        .l = list(m1, m2, additional_missing_pct, model, outcome),
+        .f = ~ posterior_epred(
+          bayes_mccv_fits,
+          newdata = data.frame(
+            md_strat = c(..1, ..2),
+            additional_missing_pct = ..3,
+            model = ..4,
+            outcome = ..5
+          ),
+          re.form = NA
+        ) %>%
+          apply(1, diff)
+      )
+    )
+
+  ggdat <- posterior_diffs %>%
+    group_by(m2) %>%
+    summarize(distribution = reduce(distribution , `+`) / n()) %>%
+    ungroup() %>%
+    mutate(
+      md_type = if_else(str_detect(m2, '_mi$'), 'mi', 'si'),
+      m2 = str_remove(m2, '_mi$|_si$'),
+      m2 = fct_reorder(
+        .f = m2,
+        .x = distribution,
+        .fun = function(x) diff(range(x))
+      ),
+      md_type = recode(md_type, !!!md_type_labels),
+      md_type = fct_relevel(md_type, md_type_labels[1]),
+      distribution = distribution * 100
+    ) %>%
+    mutate(m2 = fct_reorder(m2, .x = distribution, .fun = max))
+
+  plot_labels <- levels(ggdat$m2) %>%
+    recode(!!!md_method_labels)
+
+  ggplot(ggdat, aes(x = distribution, y = m2, fill = md_type)) +
+    geom_vline(xintercept = 0, color = 'red', linetype = 2) +
+    stat_halfeye() +
+    scale_y_discrete(breaks = levels(ggdat$m2),
[…]

### 2020-10-10 packages.R

@@ -12,10 +12,17 @@ library(ipa)
 library(mice)
 library(miceRanger)
 ## reporting
+library(rmarkdown)
 library(glue)
 library(table.glue)
+library(knitr)
+library(kableExtra)
+library(gtsummary)
+library(gt)
 ## data analysis
 library(tidyverse)
+library(rstanarm)
+library(tidybayes)
 library(data.table)
 library(magrittr)
 library(survival)
@@ -26,6 +33,7 @@ library(randomForestSRC)
 library(MASS)
 library(VIM)
 library(Hmisc)
+library(party)
 ## model evaluation
 library(riskRegression)
 library(nricens)
@@ -40,6 +48,8 @@ conflict_prefer('summarize', 'dplyr')
 conflict_prefer('select',    'dplyr')
 conflict_prefer("gather",    "tidyr")
 conflict_prefer("set_names", "purrr")
-conflict_prefer("matches",   "tidyselect")
-conflict_prefer("impute",    "miceRanger")
-library(rmarkdown)
+conflict_prefer("all_numeric", "recipes")
+conflict_prefer("matches",     "tidyselect")
+conflict_prefer("impute",      "miceRanger")
+conflict_prefer("R2", "rstanarm")
+

### 2020-10-10 slurm/slurm_clean.R

@@ -6,12 +6,14 @@ files <- list.files(path = "slurm/results",
                     pattern = '^output',
                     full.names = TRUE)
 
+print(length(files))
+
 file_nums <- str_extract_all(files, '\\d') %>%
   map(paste, collapse = '') %>%
   map_dbl(as.numeric)
 
 message("Unsuccessful slurm runs:",
-        paste(setdiff(0:2999, file_nums), collapse = ', '))
+        paste(setdiff(0:5999, file_nums), collapse = ', '))
 
 output_cleaned <- map_dfr(
   .x = files,

### 2020-10-10 slurm/slurm_run.R

@@ -6,24 +6,30 @@ R.utils::sourceDirectory("../R")
 .rslurm_func <- function(iteration,
                          md_strat,
                          outcome,
-                         additional_missing_pct = 25,
-                         n_features = 75,
+                         additional_missing_pct = 0,
+                         n_features = 150,
                          n_impute_mi = 5,
-                         times = 6) {
+                         times = 6,
+                         maxtime = 8) {
 
   set.seed(iteration)
 
   im = read_csv("../data/intermacs_clean.csv", guess_max = 20000) %>%
     rename(time = months_post_implant,
            status = paste0('pt_outcome_', outcome)) %>%
-    select(-starts_with('pt_outcome_'))
+    select(-starts_with('pt_outcome_')) %>%
+  # if you want to truncate event times
+  mutate(
+    status = if_else(time > maxtime & status == 1, 0, status),
+    time = pmin(time, maxtime)
+  )
 
   testing_index <- read_rds('../data/resamples.rds')[[iteration]]
 
   # train/test split -----
 
-  im_train_raw <- im[-testing_index, ] %>% slice_sample(prop = 1/3)
-  im_test_raw  <- im[ testing_index, ] %>% slice_sample(prop = 1/2)
+  im_train_raw <- im[-testing_index, ] #%>% slice_sample(prop = 1/10)
+  im_test_raw  <- im[ testing_index, ] #%>% slice_sample(prop = 1/2)
 
   too_many_missing <- im_train_raw %>%
     miss_var_summary() %>%
@@ -46,8 +52,20 @@ R.utils::sourceDirectory("../R")
 
   message("% missing in training data: ", 100 * prop_miss_train)
 
+  # amputation (if additional_missing_pct > 0) ----
+
   amputed_values <- vector(mode = 'list', length = length(features)) %>%
     set_names(features)
+
+  # to induce MAR
+  age_cats <- cut(im_train_raw$demo_age,
+                  breaks = c(0, 40, 65, Inf),
+                  include.lowest = TRUE)
+
+  age_probs <- rep(0, length(age_cats))
+  age_probs[age_cats == '[0,40]'] <- 0.40
+  age_probs[age_cats == '(40,65]'] <- 0.10
+  age_probs[age_cats == '(65,Inf]'] <- 0.40
 
   if(additional_missing_pct > 0){
 
@@ -68,7 +86,11 @@ R.utils::sourceDirectory("../R")
 
[…]

### 2020-10-10 slurm/slurm_run.R
no more RF model

@@ -7,7 +7,7 @@ R.utils::sourceDirectory("../R")
                          md_strat,
                          outcome,
                          additional_missing_pct = 0,
-                         n_features = 150,
+                         n_features = 50,
                          n_impute_mi = 5,
                          times = 6,
                          maxtime = 8) {
@@ -462,15 +462,6 @@ R.utils::sourceDirectory("../R")
   imputes %>%
     group_by(md_strat) %>%
     summarize(
-      rf = list(
-        predict_risk_rf(
-          training = training,
-          testing = testing,
-          pipeline = recipe_post_impute,
-          verbose = TRUE,
-          times = times
-        )
-      ),
       cph = list(
         predict_risk_cph(
           training = training,
@@ -522,7 +513,7 @@ R.utils::sourceDirectory("../R")
     'nbrs_mi',
     'nbrs_si'
   ),
-  additional_missing_pct = c(0, 20, 40),
+  additional_missing_pct = c(0, 15, 30),
   outcome = c('dead', 'txpl'),
   iteration = 1:100
 )

### 2020-10-13 doc_arXiv/doc_arXiv.Rmd
misc changes before using outcomes

@@
 
-drake::loadd(im)
-drake::loadd(times)
-drake::loadd(tbl_characteristics)
-drake::loadd(tbl_md_strat_auc)
+gt_latex <- function(gt_object, caption = 'a table', label = NULL){
 
-library(table.glue)
+  header = paste0(
+    '\\begin{table} \n \\caption{',caption,'} \n'
+  )
+
+  if(!is.null(label)){
+    header = paste0(header, '\\label{', label, '} \n')
+  }
+
+  gt_object %>% 
+    as_latex() %>% 
+    as.character() %>% 
+    gsub("*", "",., fixed=T) %>% 
+    gsub("\\\\ \n\\small \\\\ \n", "", ., fixed=T) %>% 
+    gsub("\n\\large", "", ., fixed=T) %>% 
+    gsub("\\captionsetup[table]{labelformat=empty,skip=1pt}", "", ., fixed=T) %>% 
+    gsub("longtable", "table", ., fixed=T) %>% 
+    gsub('\\begin{table}','\\begin{table} \n\\begin{tabular}', ., fixed=T) %>% 
+    gsub('\\end{table}','\\end{tabular} \n \\end{table}', ., fixed=T) %>% 
+    gsub('\\begin{table}', header, ., fixed=T) %>% 
+    cat()
+
+}
+
+options(knitr.table.format = 'latex')
+
+source(file.path(here::here(), 'packages.R'))
+
+drake::loadd(im, 
+             times, 
+             md_method_labels,
+             md_type_labels,
+             model_labels,
+             outcome_labels,
+             additional_missing_labels,
+             tbl_characteristics,
+             tbl_impute_accuracy)
 
@@
 \linenumbers
+\doublespacing
 
@@
 Next, we describe how risk prediction models and by extension the imputation strategy applied before training the model were evaluated in Section \ref{subsec:evaluation}.
-Last, we outline an internal validation procedure that was applied to compare combinations of imputation and modeling strategies using hierarchical Bayesian models in Section \ref{subsec:internal}.
+We outline an internal validation procedure that was applied to compare combinations of imputation and modeling strategies using hierarchical Bayesian models in Section \ref{subsec:internal}.
+Last, details on computation and software packages used for the current analysis are described in Section \ref{subsec:computing}
 
@@
 Imputation strategies that create a single imputed dataset (\eg imputation to the mean) have been shown to increase type I errors (\ie rejecting a true null hypothesis) for inferential statistics by artificially reducing the variance of observed data and ignoring the uncertainty attributed to missing values \cite{van2018flexible}.
-When very few data are missing, analysts may apply listwise deletion, which removes any observation with at least one missing value and produces a dataset with fewer observations and no missing values.
+To ensure valid inference, imputation models should leverage outcome variables to predict missing predictors \cite{sterne2009multiple}.
+When very few data are missing, analysts may apply listwise deletion, \ie removing any observation with at least one missing value.
+However, listwise deletion can easily lead to biased inference \cite{van2020rebutting}.
 
@@
 Others have shown that imputation to the mean, model-based imputation, and multiple imputation can provide Bayes-optimal prediction models provided certain assumptions are true, but these assumptions are difficult to validate in applied settings \cite{josse2019consistency}.
+
+<!-- Though the outcome variable should be used when imputing predictor values for statistical inference, using outcome variables for this purpose in supervised learning projects can lead to issues in clinical implementation.
+For instance, suppose we seek to predict an outcome $Y$ for a patient in the clinical setting.
+If the patient is missing information for a predictor $X$ and our only method to impute missing values of $X$ leverages the observed value of $Y$, how should we impute $X$?
+If we do not already know $Y$, which would seem likely for clinical application of a prediction model, then we cannot impute $X$.
+On the other hand, if we already know $Y$, then we obviously do not need to predict it.
+Due to these practical considerations and because INTERMACS is often leveraged to create prediction models for clinical practice \cite{thomas2014pre}, we did not leverage outcome variables (\ie time until transplant, death, or last contact) to impute missing values of predictors in the current analysis.
+-->
 
@@
 
-\paragraph{Random forests} The random forest framework can be used for imputation as well as prediction.
-In the current analysis, we implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}.
-\texttt{miceRanger} follows an iterative procedure where one random forest is fit to each variable in a dataset, separately, and then used to impute missing values in that variable.
-The procedure may repeat until estimated generalization error of the random forest collection stabilizes.
-When imputing a single dataset with \texttt{miceRanger}, we did not use PMM and instead imputed missing values with a prediction.
-When imputing multiple datasets, we implemented PMM with 10 donor candidates.
+\paragraph{Random forests} Random forests grow an ensemble of de-correlated decision trees, where each tree is grown using a bootstrapped replicate of the original training data \cite{Breiman2001, hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
+A particularly helpful feature of random forests is their ability to estimate testing error by aggregating each decision tree's prediction error on data outside of their bootstrapped sample (\ie out-of-bag error).
+In the current analysis, we conduct MICE using one random forest to impute each variable, separately.
+For each imputed dataset, we allowed random forests to be re-fitted until out-of-bag error stabilized or a maximum number of iterations was completed.
+When imputing a single dataset, we used 250 trees per forest and a maximum of 10 iterations.
+When imputing multiple datasets, we used 50 trees per forest and applied PMM using the random forest's predicted values to impute missing data by sampling one value from a pool of 10 potential donors.
 
-\paragraph{Missingness incorporated as an attribute (MIA)} MIA adds another observed category to nominal variables: "Missing".
+\paragraph{Missingness incorporated as an attribute (MIA)} MIA is a technique that uses missing status as a predictor rather than explicitly imputing missing values \cite{twala2009empirical, ding2010investigation}.
+MIA adds another category to nominal variables: "Missing".
 For numeric variables, MIA creates two columns: one where missing values are imputed with positive infinity and the other with negative infinity.
+Since PH models are not compatible with infinite values, we only use MIA in boosting models.
 When a decision tree uses a finite cut-point to split a given numeric variable, it will assess the cut-point using both the positive and negative infinite imputed columns, and utilize whichever column provides the best split of the current data.
 This procedure translates to sending all missing values to the left or to the right when forming two new nodes of the decision tree, using whichever direction results in a better split.
-In the current analysis, we used the MIA strategy only for ensembles of boosted decision trees, as it is a well known and frequently used strategy for these models \cite{twala2009empirical, ding2010investigation}.
 
-\paragraph{Surrogate splitting} When decision trees search for a variable and cut-point to use for growing two new nodes in a decision tree, surrogate splitting will temporarily ignore missing values while searching for an initial split.
+<!-- \paragraph{Surrogate splitting} When decision trees search for a variable and cut-point to use for growing two new nodes in a decision tree, surrogate splitting will temporarily ignore missing values while searching for an initial split.
 Once an initial split is chosen, surrogate splitting searches for a split on another variable with observed values for the observations with missing values on the initial splitting variable and creates a surrogate split based on the new variable that is similar to the original split.
@@
 Additionally, we only used surrogate splitting for conditional inference forests, as this strategy is well known and frequently used for these models.
+-->
 
@@
 Hot deck and KNN imputation also do not make distributional assumptions, but implicitly assume that a missing value for a given observation can be approximated by aggregating observed values from the $k$ most similar observations.
-PMM makes a similar implicit assumption, but is slightly more robust to skewed or multi-model data because imputed values are sampled directly from observed ones.
-MIA operates based on an implicit assumption that missingness itself is informative and surrogate splitting presumes that variables with missing values are related to other observed variables used by the decision tree.
+PMM makes a similar implicit assumption, but is slightly more robust to skewed or multi-modal data because imputed values are sampled directly from observed ones.
+MIA operates based on an implicit assumption that missingness itself is informative.
 It is difficult to validate these assumptions in applied settings and also likely that downstream models will perform poorly if an imputation technique's assumptions are invalid.
+
+## Evaluating imputation accuracy
+\label{subsec:imputation_accuracy}
+
+Imputation accuracy was computed for each numerical and nominal variable, separately.
+Numeric variable imputation accuracy was measured using a re-scaled mean-squared error: $$1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.$$ This score is greater than 0 if $\textrm{MSE}(\textrm{current imputation method})$ is smaller than $\textrm{MSE}(\textrm{imputation to the mean})$, equal to 0 if the two MSEs are equal, and less than 0 if $\textrm{MSE}(\textrm{current imputation method})$ is greater than $\textrm{MSE}(\textrm{imputation to the mean})$.
+Nominal variable imputation accuracy was measured using a re-scaled classification accuracy: $$1 - \frac{\textrm{Classification error}(\textrm{current imputation method})}{\textrm{Classification error}(\textrm{imputation to the mean})}.$$ This score is greater than 0 if the classification error of the current imputation method is less than that of imputation to the mean, equal to 0 if the two classification errors are equal, and less than 0 of the classification error of the current imputation method is worse than imputation to the mean.
+These numeric and nominal scores are analogous to the more well known $R^2$ and Kappa statistics, respectively, but are modified slightly in the current analysis so that imputation to the mean will always have a score of 0.
+This modification makes it easier to compare the accuracy of each imputation strategy directly with imputation to the mean.
 
@@
 
-We applied three modeling strategies after imputing missing values:
+We applied two modeling strategies after imputing missing values:
 
 - Cox proportional hazards (PH) model with forward stepwise variable selection
-- gradient boosted decision trees (hereafter referred to as 'boosting')
-- Conditional inference forest (CIF, an extension of random forests)
+- Gradient boosted decision trees (hereafter referred to as 'boosting')
 
-A thorough description of stepwise variable selection, boosting, and random forests can be found in Sections 6.1.2, 8.2.2, and 8.2.3, respectively, of \emph{Introduction to Statistical Learning} \cite{james2013introduction}.
+A thorough description of stepwise variable selection and boosting can be found in Sections 6.1.2 and 8.2.2, respectively, of \emph{Introduction to Statistical Learning} \cite{james2013introduction}.
 Sir David Cox's PH model is one of the most frequently applied methods for the analysis of right-censored time-to-event outcomes \cite{kleinbaum2010survival}.
@@
 Boosting grows a sequence of decision trees, each using information from the previous trees in an attempt to correct their errors \cite{friedman2001greedy, chen2016xgboost}.
-In contrast, random forests grow an ensemble of de-correlated decision trees, where each tree is grown using a bootstrapped replicate of the original training data \cite{Breiman2001}.
-The CIF obtains unbiased recursive partitioning by applying early-stopping and permutation tests and is one of several methods that extends random forests to right-censored time-to-event data \cite{hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
 
@@
 For convenience, we use the term `modeling algorithm' to denote the combination of a missing data strategy and a modeling strategy (\eg imputation using mean/mode values followed by fitting the PH model with stepwise variable selection) \cite{kuhn2013applied}.
-We conducted internal validation using 150 replicates of Monte-Carlo cross validation (MCCV), a resampling technique for internal validation (Figure \ref{fig:monte_carlo_cv}).
+We conducted internal validation using 200 replicates of Monte-Carlo cross validation (MCCV), a resampling technique for internal validation (Figure FILL IN).
 
@@
 After imputation, each modeling algorithm (\ie stepwise PH models, GBDTs, and the cIF) was applied, separately, to develop a RPE for death or transplant that incorporated up to 50 predictor variables from the entire collection of covariates available.
+
+## Statistical analysis
+
+Participant characteristics were tabulated for the overall population and stratified by event status.
+Continuous and categorical variables were summarized as mean (standard deviation) and percent, respectively.
+Imputation accuracy was aggregated for all numeric and nominal variables to create two overall scores for each imputed dataset, and the distribution of these scores was tabulated for each imputation strategy.
+
+## Computational details
+\label{subsec:computing}
+
+implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}.
+\texttt{miceRanger} follows an iterative procedure where one random forest is fit to each variable in a dataset, separately, and then used to impute missing values in that variable.
+The procedure may repeat until estimated generalization error of the random forest collection stabilizes.
+
+# Results
+
+```{r inline_characteristics}
+
+trunc_events <- im %>% 
+  summarize(
+    n_dead = sum(pt_outcome_dead == 1 & months_post_implant < times),
+    n_txpl = sum(pt_outcome_txpl == 1 & months_post_implant < times),
+    n_cens = sum(pt_outcome_cess == 0 & 
+                 pt_outcome_dead == 0 &
+                 pt_outcome_txpl ==0 & 
+                 months_post_implant < times),
+    p_dead = 100 * mean(pt_outcome_dead == 1 & months_post_implant < times),
+    p_txpl = 100 * mean(pt_outcome_txpl == 1 & months_post_implant < times),
+    p_cens = 100 * mean(pt_outcome_cess == 0 & 
+                 pt_outcome_dead == 0 &
+                 pt_outcome_txpl ==0 & 
+                 months_post_implant < times)
+
+  ) %>% 
+  transmute(
+    dead = table_glue("{n_dead} ({p_dead}%)"),
+    txpl = table_glue("{n_txpl} ({p_txpl}%)"),
+    cens = table_glue("{n_cens} ({p_cens}%)")
+  ) %>% 
+  as.list()
+
+age_overall <- inline_text(tbl_characteristics, 
+                           variable = 'demo_age', 
+                           column = 'stat_0')
+
+male_overall <- inline_text(tbl_characteristics, 
+                            variable = 'demo_gender',
+                            level = 'Male',
+                            column = 'stat_0')
+
+white_overall <- inline_text(tbl_characteristics, 
+                             variable = 'demo_race',
+                             level = 'White',
+                             column = 'stat_0')
+
+```
+
+\paragraph{Patient Characteristics} During the first `r times` months after receiving MCS, the number (% of overall population) of deaths, transplants, and loss-to-follow-up events was `r trunc_events$dead`, `r trunc_events$txpl`, and `r trunc_events$cens`, respectively.
+The mean (standard deviation) age of patients was `r age_overall` years, `r white_overall`% of patients identified as white and `r male_overall`% were male (Table \ref{tbl_characteristics}).
+
+\paragraph{Imputation accuracy}
+
+<!-- Table : characteristics -->
+\clearpage
+\begin{table}
+\caption{Participant characteristics stratified by event status.}
+\label{tbl_characteristics}
+
+```{r tbl_characteristics, results = 'asis'}
+
+as_kable_extra(
+  tbl_characteristics,
+  booktabs = TRUE, 
+  align = 'lccccc'
+) %>% 
+  footnote(general = 'Table values are mean (standard deviation) or %')
+
+```
+
+\end{table}
+
+<!-- Table: imputation accuracy -->
+
+\clearpage
+
+```{r tbl_impute_accuracy, results = 'asis'}
+
+tbl_impute_accuracy %>% 
+  mutate(
+    md_method = factor(
+      md_method,
+      levels = names(md_method_labels),
+      labels = md_method_labels),
+    additional_missing_pct = factor(
+      additional_missing_pct,
+      levels = additional_missing_labels[-1],
+      labels = names(additional_missing_labels)[-1]
+    )
+  ) %>%
+  arrange(md_method, additional_missing_pct) %>% 
+  gt(rowname_col = 'md_method', 
+     groupname_col = 'additional_missing_pct') %>% 
+  fmt_missing(columns = everything()) %>% 
+  cols_align('center') %>% 
+  tab_spanner(
+    label = 'Nominal variables',
+    columns = c('nominal_si', 'nominal_mi')
+  ) %>% 
+  tab_spanner(
+    label = 'Numeric variables',
+    columns = c('numeric_si', 'numeric_mi')
+  ) %>% 
+  cols_label(
+    numeric_si = 'Single imputation', 
+    nominal_si = 'Single imputation', 
+    numeric_mi = 'Multiple imputation',
+    nominal_mi = 'Multiple imputation'
+  ) %>%
+  gt_latex(
+    caption = 'Accuracy of strategies to impute artificial missing data', 
+    label = 'tbl_impute_accuracy'
+  )
+
+```

### 2020-10-13 doc_arXiv/doc_arXiv.tex
misc changes before using outcomes

@@
 by extension the imputation strategy applied before training the model
-were evaluated in Section \ref{subsec:evaluation}. Last, we outline an
+were evaluated in Section \ref{subsec:evaluation}. We outline an
 internal validation procedure that was applied to compare combinations
 of imputation and modeling strategies using hierarchical Bayesian models
-in Section \ref{subsec:internal}.
+in Section \ref{subsec:internal}. Last, details on computation and
+software packages used for the current analysis are described in Section
+\ref{subsec:computing}
 
@@
 variance of observed data and ignoring the uncertainty attributed to
-missing values \cite{van2018flexible}. When very few data are missing,
-analysts may apply listwise deletion, which removes any observation with
-at least one missing value and produces a dataset with fewer
-observations and no missing values.
+missing values \cite{van2018flexible}. To ensure valid inference,
+imputation models should leverage outcome variables to predict missing
+predictors \cite{sterne2009multiple}. When very few data are missing,
+analysts may apply listwise deletion, \textit{i.e., }removing any
+observation with at least one missing value. However, listwise deletion
+can easily lead to biased inference \cite{van2020rebutting}.
 
@@
 settings \cite{josse2019consistency}.
+
+Though the outcome variable may be useful for imputing predictor values,
+using outcome variables for this purpose in supervised learning projects
+can lead to issues in clinical implementation. For instance, suppose we
+seek to predict an outcome \(Y\) for a patient in the clinical setting.
+If the patient is missing information for a predictor \(X\) and our only
+method to impute missing values of \(X\) leverages the observed value of
+\(Y\), how should we impute \(X\)? If we do not already know \(Y\),
+which would seem likely for clinical application of a prediction model,
+then we cannot impute \(X\). On the other hand, if we already know
+\(Y\), then we obviously do not need to predict it. Due to these
+practical considerations and because INTERMACS is often leveraged to
+create prediction models for clinical practice \cite{thomas2014pre}, we
+did not leverage outcome variables (\textit{i.e., }time until
+transplant, death, or last contact) to impute missing values of
+predictors in the current analysis.
 
@@
 
-The random forest framework can be used for imputation as well as
-prediction. In the current analysis, we implement \texttt{miceRanger},
-an R package similar to \texttt{missForest} that implements MICE with
-random forests by implementing PMM \cite{missForest}.
-\texttt{miceRanger} follows an iterative procedure where one random
-forest is fit to each variable in a dataset, separately, and then used
-to impute missing values in that variable. The procedure may repeat
-until estimated generalization error of the random forest collection
-stabilizes. When imputing a single dataset with \texttt{miceRanger}, we
-did not use PMM and instead imputed missing values with a prediction.
-When imputing multiple datasets, we implemented PMM with 10 donor
-candidates.
+Random forests grow an ensemble of de-correlated decision trees, where
+each tree is grown using a bootstrapped replicate of the original
+training data
+\cite{Breiman2001, hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
+A particularly helpful feature of random forests is their ability to
+estimate testing error by aggregating each decision tree's prediction
+error on data outside of their bootstrapped sample
+(\textit{i.e., }out-of-bag error). In the current analysis, we conduct
+MICE using one random forest to impute each variable, separately. For
+each imputed dataset, we allowed random forests to be re-fitted until
+out-of-bag error stabilized or a maximum number of iterations was
+completed. When imputing a single dataset, we used 250 trees per forest
+and a maximum of 10 iterations. When imputing multiple datasets, we used
+50 trees per forest and applied PMM using the random forest's predicted
+values to impute missing data by sampling one value from a pool of 10
+potential donors.
 
@@
 
-MIA adds another observed category to nominal variables: ``Missing''.
-For numeric variables, MIA creates two columns: one where missing values
-are imputed with positive infinity and the other with negative infinity.
+MIA is a technique that uses missing status as a predictor rather than
+explicitly imputing missing values
+\cite{twala2009empirical, ding2010investigation}. MIA adds another
+category to nominal variables: ``Missing''. For numeric variables, MIA
+creates two columns: one where missing values are imputed with positive
+infinity and the other with negative infinity. Since PH models are not
+compatible with infinite values, we only use MIA in boosting models.
 When a decision tree uses a finite cut-point to split a given numeric
@@
 nodes of the decision tree, using whichever direction results in a
-better split. In the current analysis, we used the MIA strategy only for
-ensembles of boosted decision trees, as it is a well known and
-frequently used strategy for these models
-\cite{twala2009empirical, ding2010investigation}.
-
-\paragraph{Surrogate splitting}
-
-When decision trees search for a variable and cut-point to use for
-growing two new nodes in a decision tree, surrogate splitting will
-temporarily ignore missing values while searching for an initial split.
-Once an initial split is chosen, surrogate splitting searches for a
-split on another variable with observed values for the observations with
-missing values on the initial splitting variable and creates a surrogate
-split based on the new variable that is similar to the original split.
-In the current analysis, we used a maximum of three surrogate splits and
-sent any remaining missing values to the daughter node with higher
-sample size. Additionally, we only used surrogate splitting for
-conditional inference forests, as this strategy is well known and
-frequently used for these models.
+better split.
 
@@
 implicit assumption, but is slightly more robust to skewed or
-multi-model data because imputed values are sampled directly from
+multi-modal data because imputed values are sampled directly from
 observed ones. MIA operates based on an implicit assumption that
-missingness itself is informative and surrogate splitting presumes that
-variables with missing values are related to other observed variables
-used by the decision tree. It is difficult to validate these assumptions
-in applied settings and also likely that downstream models will perform
-poorly if an imputation technique's assumptions are invalid.
+missingness itself is informative. It is difficult to validate these
+assumptions in applied settings and also likely that downstream models
+will perform poorly if an imputation technique's assumptions are
+invalid.
+
+\hypertarget{evaluating-imputation-accuracy}{%
+\subsection{Evaluating imputation
+accuracy}\label{evaluating-imputation-accuracy}}
+
+\label{subsec:imputation_accuracy}
+
+Imputation accuracy was computed for each numerical and nominal
+variable, separately. Numeric variable imputation accuracy was measured
+using a re-scaled mean-squared error:
+\[1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.\]
+This score is greater than 0 if
+\(\textrm{MSE}(\textrm{current imputation method})\) is smaller than
+\(\textrm{MSE}(\textrm{imputation to the mean})\), equal to 0 if the two
+MSEs are equal, and less than 0 if
+\(\textrm{MSE}(\textrm{current imputation method})\) is greater than
+\(\textrm{MSE}(\textrm{imputation to the mean})\). Nominal variable
+imputation accuracy was measured using a re-scaled classification
+accuracy:
+\[1 - \frac{\textrm{Classification error}(\textrm{current imputation method})}{\textrm{Classification error}(\textrm{imputation to the mean})}.\]
+This score is greater than 0 if the classification error of the current
+imputation method is less than that of imputation to the mean, equal to
+0 if the two classification errors are equal, and less than 0 of the
+classification error of the current imputation method is worse than
+imputation to the mean. These numeric and nominal scores are analogous
+to the more well known \(R^2\) and Kappa statistics, respectively, but
+are modified slightly in the current analysis so that imputation to the
+mean will always have a score of 0. This modification makes it easier to
+compare the accuracy of each imputation strategy directly with
+imputation to the mean.
 
@@
 
-We applied three modeling strategies after imputing missing values:
+We applied two modeling strategies after imputing missing values:
 
@@
 \item
-  gradient boosted decision trees (hereafter referred to as `boosting')
-\item
-  Conditional inference forest (CIF, an extension of random forests)
+  Gradient boosted decision trees (hereafter referred to as `boosting')
 \end{itemize}
 
-A thorough description of stepwise variable selection, boosting, and
-random forests can be found in Sections 6.1.2, 8.2.2, and 8.2.3,
-respectively, of \emph{Introduction to Statistical Learning}
+A thorough description of stepwise variable selection and boosting can
+be found in Sections 6.1.2 and 8.2.2, respectively, of
+\emph{Introduction to Statistical Learning}
 \cite{james2013introduction}. Sir David Cox's PH model is one of the
@@
 previous trees in an attempt to correct their errors
-\cite{friedman2001greedy, chen2016xgboost}. In contrast, random forests
-grow an ensemble of de-correlated decision trees, where each tree is
-grown using a bootstrapped replicate of the original training data
-\cite{Breiman2001}. The CIF obtains unbiased recursive partitioning by
-applying early-stopping and permutation tests and is one of several
-methods that extends random forests to right-censored time-to-event data
-\cite{hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
+\cite{friedman2001greedy, chen2016xgboost}.
 
@@
 the PH model with stepwise variable selection) \cite{kuhn2013applied}.
-We conducted internal validation using 150 replicates of Monte-Carlo
+We conducted internal validation using 200 replicates of Monte-Carlo
 cross validation (MCCV), a resampling technique for internal validation
@@
 
+\hypertarget{statistical-analysis}{%
+\subsection{Statistical analysis}\label{statistical-analysis}}
+
+Participant characteristics were tabulated for the overall population
+and stratified by event status. Continuous and categorical variables
+were summarized as mean (standard deviation) and percent, respectively.
+Imputation accuracy was aggregated for all numeric and nominal variables
+to create two overall scores for each imputed dataset, and the
+distribution of these scores was tabulated for each imputation strategy.
+
+\hypertarget{computational-details}{%
+\subsection{Computational details}\label{computational-details}}
+
+\label{subsec:computing}
+
+implement \texttt{miceRanger}, an R package similar to
+\texttt{missForest} that implements MICE with random forests by
+implementing PMM \cite{missForest}. \texttt{miceRanger} follows an
+iterative procedure where one random forest is fit to each variable in a
+dataset, separately, and then used to impute missing values in that
+variable. The procedure may repeat until estimated generalization error
+of the random forest collection stabilizes.
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\paragraph{Patient Characteristics}
+
+During the first 6 months after receiving MCS, the number (\% of overall
+population) of deaths, transplants, and loss-to-follow-up events was
+1,897 (13\%), 1,067 (7.2\%), and 981 (6.7\%), respectively. The mean
+(standard deviation) age of patients was 57 (13) years, 66\% of patients
+identified as white and 79\% were male (Table
+\ref{tbl_characteristics}).
+
+\paragraph{Imputation accuracy}
+
+\clearpage
+\begin{table}
+\caption{Participant characteristics stratified by event status.}
+\label{tbl_characteristics}
+
+\begin{tabular}{lccccc}
+\toprule
+Characteristic & Overall & Censored & Cessation of support & Dead & Transplant\\
+\midrule
+No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
+Follow up time, months & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
+Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
+Sex &  &  &  &  & \\
+\hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
+\addlinespace
+\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
+Race &  &  &  &  & \\
+\hspace{1em}White & 66 & 63 & 64 & 71 & 67\\
+\hspace{1em}Black & 25 & 28 & 27 & 21 & 23\\
+\hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
+\addlinespace
+Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
+Device strategy &  &  &  &  & \\
+\hspace{1em}Other & 0.6 & 0.6 & 2.6 & 0.5 & 0.6\\
+\hspace{1em}Bridge to transplant & 52 & 43 & 51 & 40 & 81\\
+\hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
+\addlinespace
+Device type &  &  &  &  & \\
+\hspace{1em}Bi-ventricular assistance device & 3.5 & 2.1 & 0.3 & 5.8 & 3.4\\
+\hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
+\bottomrule
+\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
+\multicolumn{6}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
+\end{tabular}
+
+\end{table}
+
+\clearpage
+
+\begin{table} 
+ \caption{Accuracy of strategies to impute artificial missing data} 
+\label{tbl_impute_accuracy}
+
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{Nominal variables} & \multicolumn{2}{c}{Numeric variables} \\ 
+ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+ & Single imputation & Multiple imputation & Single imputation & Multiple imputation \\ 
+\midrule
+\multicolumn{1}{l}{+15\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
+Hot deck & -0.24 (-0.34, -0.16) & -0.25 (-0.37, -0.18) & -0.35 (-0.53, -0.29) & -0.42 (-0.54, -0.36) \\ 
+K-nearest-neighbors & 0.02 (-0.05, 0.15) & -0.19 (-0.29, -0.08) & 0.01 (-0.01, 0.03) & -0.33 (-0.42, -0.28) \\ 
+Predictive mean matching & -0.15 (-0.34, 0.00) & -0.15 (-0.35, 0.00) & -0.25 (-0.40, -0.20) & -0.26 (-0.35, -0.21) \\ 
+Random forest & -0.15 (-0.26, -0.02) & -0.15 (-0.26, -0.03) & 0.06 (0.02, 0.08) & 0.05 (0.02, 0.07) \\ 
+Bayesian regression & -0.14 (-0.26, 0.01) & -0.14 (-0.27, 0.01) & -0.30 (-0.41, -0.23) & -0.30 (-0.41, -0.23) \\ 
+\midrule
+\multicolumn{1}{l}{+30\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
+Hot deck & -0.25 (-0.34, -0.18) & -0.26 (-0.36, -0.18) & -0.34 (-0.46, -0.30) & -0.40 (-0.53, -0.36) \\ 
+K-nearest-neighbors & -0.04 (-0.09, 0.03) & -0.24 (-0.34, -0.16) & -0.02 (-0.03, -0.01) & -0.36 (-0.44, -0.32) \\ 
+Predictive mean matching & -0.18 (-0.38, -0.04) & -0.18 (-0.37, -0.04) & -0.29 (-0.46, -0.24) & -0.29 (-0.40, -0.25) \\ 
+Random forest & -0.18 (-0.27, -0.07) & -0.18 (-0.28, -0.07) & -0.02 (-0.05, 0.01) & -0.01 (-0.04, 0.01) \\ 
+Bayesian regression & -0.17 (-0.29, -0.04) & -0.17 (-0.29, -0.05) & -0.31 (-0.43, -0.27) & -0.31 (-0.43, -0.27) \\ 
+\bottomrule
+\end{tabular} 
+ \end{table}
+
 \bibliographystyle{unsrt}

### 2020-10-13 R/evaluate_risk_predictions.R

@@ -26,32 +26,33 @@ evaluate_risk_predictions <- function(risk_predictions,
     tidy_score() %>%
     separate(md_strat_mdl, into = c('md_strat', 'model'), sep = '\\.\\.')
 
-  gnd <- risk_predictions %>%
-    unite(md_strat, model, col = 'md_strat_mdl', sep = '..') %>%
-    select(md_strat_mdl, sprobs) %>%
-    deframe() %>%
-    map_dfr(
-      .f = GND_test,
-      .id = 'md_strat_mdl',
-      predict_horizon = times,
-      event_time = testing_data$time,
-      event_status = testing_data$status
-    ) %>%
-    separate(md_strat_mdl, into = c('md_strat', 'model'), sep = '\\.\\.')
+  # gnd <- risk_predictions %>%
+  #   unite(md_strat, model, col = 'md_strat_mdl', sep = '..') %>%
+  #   select(md_strat_mdl, sprobs) %>%
+  #   deframe() %>%
+  #   map_dfr(
+  #     .f = GND_test,
+  #     .id = 'md_strat_mdl',
+  #     predict_horizon = times,
+  #     event_time = testing_data$time,
+  #     event_status = testing_data$status
+  #   ) %>%
+  #   separate(md_strat_mdl, into = c('md_strat', 'model'), sep = '\\.\\.')
 
-  nri <- risk_predictions %>%
-    group_by(model) %>%
-    group_modify(
-      .f = ~ score_nri(data = .x,
-                       reference = 'meanmode_si',
-                       time = testing_data$time,
-                       status = testing_data$status,
-                       eval_time = times)
-    )
+  # nri <- risk_predictions %>%
+  #   group_by(model) %>%
+  #   group_modify(
+  #     .f = ~ score_nri(data = .x,
+  #                      reference = 'meanmode_si',
+  #                      time = testing_data$time,
+  #                      status = testing_data$status,
+  #                      eval_time = times)
+  #   )
 
-  scores %>%
-    left_join(nri) %>%
-    left_join(gnd)
+  scores
+  # %>%
+  #   left_join(nri) %>%
+  #   left_join(gnd)
 
 
 }

### 2020-10-13 R/load_mccv.R

@@ -10,7 +10,7 @@
 
 load_mccv <- function(output_path = "slurm/results",
                       output_pattern = '^output',
-                      nruns = 6000) {
+                      nruns = 12000) {
 
   files <- list.files(path = output_path,
                       pattern = output_pattern,

### 2020-10-13 R/make_bayes_mccv_fit.R

@@ -6,24 +6,19 @@
 ##' @param risk_evaluation
 make_bayes_mccv_fit <- function(risk_evaluation) {
 
-  model_data <- risk_evaluation %>%
-    mutate(md_strat = fct_relevel(md_strat, 'meanmode_si'))
-
-  stan_lm(
-    formula = ipa ~ outcome + md_strat + model + additional_missing_pct,
-    #+ (additional_missing_pct | iteration),
-    data = model_data,
-    prior = R2(0.30)
-  )
-
-  # model_fits_ipa <- map(
-  #   .x = model_data,
-  #   .f = ~ stan_lmer(
-  #     formula = ipa ~ md_strat + model + additional_missing_pct +
-  #       (additional_missing_pct | iteration),
-  #     data = .x
-  #   )
-  # )
-
+  risk_evaluation %>%
+    mutate(md_strat = fct_relevel(md_strat, 'meanmode_si')) %>%
+    select(-bri) %>%
+    pivot_longer(cols = c(auc, ipa), names_to = 'metric') %>%
+    split(f = list(.$outcome,
+                   .$model,
+                   .$metric,
+                   .$additional_missing_pct)) %>%
+    map(
+      ~ stan_lmer(
+        formula = value ~ md_strat + (1 | iteration),
+        data = .x
+      )
+    )
 
 }

### 2020-10-13 R/make_risk_evaluation.R

@@ -3,7 +3,6 @@
 ##' .. content for \details{} ..
 ##'
 ##' @title
-##' @param mc_cv_results
 ##' @param im
 ##' @param resamples
 make_risk_evaluation <- function(mc_cv_results, im, resamples, times) {

### 2020-10-13 R/plan.R

@@ -5,22 +5,6 @@ the_plan <- drake_plan(
   # when to evaluate models, in months since implant
   times = 6,
 
-  rspec = round_spec() %>%
-    round_using_magnitude(digits = c(2, 2, 1),
-                          breaks = c(1, 10, 100)) %>%
-    round_half_even(),
-
-  # these two .csv files are needed to complete the slurm runs ----
-  # creates intermacs_clean.csv
-  im = clean_intermacs(),
-  # creates resamples.csv
-  resamples = mc_cv_light(data = im, train_prop = 1/2, ntimes = 250),
-
-  # slurm runs need to be completed/cleaned to proceed from here ----
-  mc_cv_results = load_mccv(),
-
-  risk_evaluation = make_risk_evaluation(mc_cv_results, im, resamples, times),
-
   # labels to describe methods ----
 
   md_method_labels = c(
@@ -51,9 +35,35 @@ the_plan <- drake_plan(
 
   additional_missing_labels = c(
     'No additional missing data'  = '0',
-    '+20% additional missing data'= '20',
-    '+40% additional missing data'= '40'
+    '+15% additional missing data'= '15',
+    '+30% additional missing data'= '30'
   ),
+
+  # rounding specification for the project
+  rspec = round_spec() %>%
+    round_using_magnitude(digits = c(2, 2, 1),
+                          breaks = c(1, 10, 100)) %>%
+    round_half_even(),
+
+  # these two .csv files are needed to complete the slurm runs ----
+
+  # creates intermacs_clean.csv
+  im = clean_intermacs(),
+
+  # tabulate characteristics
+  tbl_characteristics = tabulate_characteristics(im),
+
+  # creates resamples.csv
+  resamples = mc_cv_light(data = im, train_prop = 1/2, ntimes = 200),
+
+  # slurm runs need to be completed/cleaned to proceed from here ----
+  mc_cv_results = load_mccv(),
+
+  tbl_impute_accuracy = tabulate_impute_accuracy(mc_cv_results),
+
+  risk_evaluation = make_risk_evaluation(mc_cv_results, im, resamples, times),
+
+  bayes_mccv_fits = make_bayes_mccv_fit(risk_evaluation),
 
   fig_risk_evaluation = visualize_risk_evaluation(risk_evaluation,
                                                   md_method_labels,
@@ -63,22 +73,20 @@ the_plan <- drake_plan(
                                                   additional_missing_labels,
                                                   times),
 
[…]

### 2020-10-13 R/tabulate_characteristics.R

@@ -54,6 +54,7 @@ tabulate_characteristics <- function(im) {
       pi_device_strategy,
       im_device_ty
     ) %>%
+    mutate(`No. of patients` = 1, .before = 1) %>%
     tbl_summary(
       by = status,
       label = list(
@@ -66,11 +67,16 @@ tabulate_characteristics <- function(im) {
         im_device_ty ~ "Device type"
       ),
       missing = 'no',
+      type = list(`No. of patients` ~ 'continuous'),
       statistic = list(
         all_continuous() ~ "{mean} ({sd})",
-        all_categorical() ~ "{p}"
-      )
+        all_categorical() ~ "{p}",
+        `No. of patients` ~ "{sum}"
+      ),
+      digits = list(`No. of patients` ~ 0)
     ) %>%
-    add_overall(last = FALSE)
+    modify_footnote(update = everything() ~ NA) %>%
+    add_overall(last = FALSE, col_label = '**Overall**') %>%
+    modify_header(stat_by="**{level}**")
 
 }

### 2020-10-13 R/tabulate_impute_accuracy.R

@@ -1,0 +1,116 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+
+tabulate_impute_accuracy <- function(mc_cv_results) {
+
+  data_tbl <- mc_cv_results %>%
+    filter(additional_missing_pct > 0,
+           md_strat != 'mia') %>%
+    group_by(
+      iteration,
+      outcome,
+      additional_missing_pct,
+      model
+    ) %>%
+    mutate(
+      pe = nominal[md_strat == 'meanmode_si'],
+      nominal = (nominal - pe) / (1 - pe)
+    ) %>%
+    select(md_strat, additional_missing_pct, nominal, numeric) %>%
+    group_by(md_strat, additional_missing_pct) %>%
+    summarize(
+      across(
+        .cols = c(nominal, numeric),
+        .fns = list(
+          est = median,
+          lwr = ~quantile(.x, probs = 0.025),
+          upr = ~quantile(.x, probs = 0.975)
+        )
+      )
+    )
+
+  data_tbl %>%
+    transmute(
+      md_strat,
+      additional_missing_pct,
+      nominal = table_glue("{nominal_est} ({nominal_lwr}, {nominal_upr})"),
+      numeric = table_glue("{numeric_est} ({numeric_lwr}, {numeric_upr})")
+    ) %>%
+    mutate(
+      across(
+        .cols = c(nominal, numeric),
+        .fns = ~str_replace(.x,
+                            pattern = '0\\.00 \\(0\\.00, 0\\.00\\)',
+                            replacement = '0 \\(reference\\)')
+      )
+    ) %>%
+    separate(md_strat, into = c('md_method', 'md_type')) %>%
+    pivot_wider(values_from = c(nominal, numeric), names_from = md_type) %>%
+    select(md_method,
+           additional_missing_pct,
+           nominal_si,
+           nominal_mi,
+           numeric_si,
+           numeric_mi)
+
+
+}
+
+# resample_ids <- resamples %>%
+#   enframe(name = 'iteration', value = 'id')
+#
+# tmp <- mc_cv_results %>%
+#   left_join(resample_ids) %>%
+#   unnest(cols = c(sprobs, id)) %>%
+#   group_by(md_strat, outcome, additional_missing_pct, model, id) %>%
[…]

### 2020-10-13 R/tabulate_md_strat.R

@@ -15,16 +15,13 @@ tabulate_md_strat <- function(risk_evaluation,
   pm <- function(x){ifelse(x>0, "+", "")}
 
   output <- risk_evaluation %>%
-    filter(model != 'rf') %>%
-    pivot_longer(cols = c(auc:GND_pvalue), names_to = 'metric') %>%
+    select(-bri) %>%
+    pivot_longer(cols = c(auc, ipa), names_to = 'metric') %>%
     split(f = list(.$outcome, .$metric)) %>%
     map(
       ~ {
 
-        mult_by <- if(.x$metric[1] == 'GND_chisq') 1 else 100
-
-        if(str_detect(.x$metric[1], '^nri'))
-          .x$value[.x$md_strat == 'meanmode_si'] <- 0
+        mult_by <- 100
 
         data_tbl <- .x %>%
           select(-outcome, -metric) %>%

### 2020-10-13 R/visualize_md_strat_inference.R

@@ -7,39 +7,49 @@
 visualize_md_strat_inference <- function(bayes_mccv_fits,
                                          md_type_labels,
                                          md_method_labels,
-                                         outcome_labels) {
+                                         outcome_labels,
+                                         rspec) {
 
+  get_xlevels <- function(posterior) {
 
-  posterior_diffs <- expand.grid(
-    m1 = bayes_mccv_fits$xlevels$md_strat,
-    m2 = bayes_mccv_fits$xlevels$md_strat,
-    additional_missing_pct = c(0, 20, 40),
-    model = c('cph', 'xgb', 'rf'),
-    outcome = c('dead', 'txpl'),
-    stringsAsFactors = FALSE
-  ) %>%
-    as_tibble() %>%
-    filter(m1 == 'meanmode_si') %>%
-    mutate(
-      distribution = pmap(
-        .l = list(m1, m2, additional_missing_pct, model, outcome),
-        .f = ~ posterior_epred(
-          bayes_mccv_fits,
-          newdata = data.frame(
-            md_strat = c(..1, ..2),
-            additional_missing_pct = ..3,
-            model = ..4,
-            outcome = ..5
-          ),
-          re.form = NA
-        ) %>%
-          apply(1, diff)
-      )
-    )
+    posterior$x@Dimnames[[2]] %>%
+      enframe() %>%
+      filter(str_detect(value, '^md_strat')) %>%
+      pull(value) %>%
+      str_remove('md_strat')
+
+  }
+
+  posterior_diffs <- map_dfr(
+    .x = bayes_mccv_fits,
+    .id = 'id',
+    .f = function(posterior) {
+      tibble(
+        m1 = 'meanmode_si',
+        m2 = get_xlevels(posterior)
+      ) %>%
+        mutate(
+          distribution = map2(
+            .x = m1,
+            .y = m2,
+            .f = ~ posterior_predict(
+              posterior,
+              newdata = data.frame(md_strat = c(.x, .y)),
+              re.form = NA
+            ) %>%
+              apply(1, diff)
+          )
+        )
+    }
+  )
 
   ggdat <- posterior_diffs %>%
-    group_by(m2) %>%
-    summarize(distribution = reduce(distribution , `+`) / n()) %>%
+    separate(
+      id,
+      into = c('outcome', 'model', 'metric', 'additional_missing_pct')
+    ) %>%
+    group_by(m2, metric) %>%
[…]

### 2020-10-13 packages.R

@@ -19,6 +19,7 @@ library(knitr)
 library(kableExtra)
 library(gtsummary)
 library(gt)
+library(tibbleOne)
 ## data analysis
 library(tidyverse)
 library(rstanarm)

### 2020-10-13 slurm/slurm_run.R

@@ -166,7 +166,7 @@ R.utils::sourceDirectory("../R")
 
   }
 
-  # imputation -----
+  # imputation ----
 
   pipe_meanmode_impute <- recipe_pre_impute %>%
     step_meanimpute(all_numeric()) %>%

### 2020-10-13 R/load_mccv.R
go back to ignoring outcomes

@@ -8,7 +8,7 @@
 ##'
 ##' @title
 
-load_mccv <- function(output_path = "slurm/results",
+load_mccv <- function(output_path = "slurm/results_old",
                       output_pattern = '^output',
                       nruns = 12000) {
 
@@ -22,8 +22,8 @@ load_mccv <- function(output_path = "slurm/results",
     map(paste, collapse = '') %>%
     map_dbl(as.numeric)
 
-  message("Unsuccessful slurm runs:",
-          paste(setdiff(0:(nruns-1), file_nums), collapse = ', '))
+  # message("Unsuccessful slurm runs:",
+  #         paste(setdiff(0:(nruns-1), file_nums), collapse = ', '))
 
   map_dfr(.x = files, .f = ~ read_rds(.x)[[1]])
 

### 2020-10-13 R/plan.R

@@ -63,37 +63,37 @@ the_plan <- drake_plan(
 
   risk_evaluation = make_risk_evaluation(mc_cv_results, im, resamples, times),
 
-  bayes_mccv_fits = make_bayes_mccv_fit(risk_evaluation),
-
+  # bayes_mccv_fits = make_bayes_mccv_fit(risk_evaluation),
+  #
   fig_risk_evaluation = visualize_risk_evaluation(risk_evaluation,
                                                   md_method_labels,
                                                   md_type_labels,
                                                   model_labels,
                                                   outcome_labels,
                                                   additional_missing_labels,
-                                                  times),
-
-  fig_md_strat_infer = visualize_md_strat_inference(bayes_mccv_fits,
-                                                    md_type_labels,
-                                                    md_method_labels,
-                                                    outcome_labels,
-                                                    rspec),
-
-
-  tbl_md_strat = tabulate_md_strat(risk_evaluation,
-                                   md_method_labels,
-                                   md_type_labels,
-                                   model_labels,
-                                   outcome_labels,
-                                   additional_missing_labels,
-                                   rspec),
-
-  arxiv_preprint = target(
-    command = {
-      rmarkdown::render(knitr_in("doc_arxiv/doc_arxiv.Rmd"))
-      file_out("doc_arxiv/doc_arxiv.pdf")
-    }
-  )
+                                                  times)
+  #
+  # fig_md_strat_infer = visualize_md_strat_inference(bayes_mccv_fits,
+  #                                                   md_type_labels,
+  #                                                   md_method_labels,
+  #                                                   outcome_labels,
+  #                                                   rspec),
+  #
+  #
+  # tbl_md_strat = tabulate_md_strat(risk_evaluation,
+  #                                  md_method_labels,
+  #                                  md_type_labels,
+  #                                  model_labels,
+  #                                  outcome_labels,
+  #                                  additional_missing_labels,
+  #                                  rspec),
+  #
+  # arxiv_preprint = target(
+  #   command = {
+  #     rmarkdown::render(knitr_in("doc_arxiv/doc_arxiv.Rmd"))
+  #     file_out("doc_arxiv/doc_arxiv.pdf")
+  #   }
+  # )
 
 
 )

### 2020-10-13 slurm/slurm_run.R

@@ -18,11 +18,11 @@ R.utils::sourceDirectory("../R")
     rename(time = months_post_implant,
            status = paste0('pt_outcome_', outcome)) %>%
     select(-starts_with('pt_outcome_')) %>%
-  # if you want to truncate event times
-  mutate(
-    status = if_else(time > maxtime & status == 1, 0, status),
-    time = pmin(time, maxtime)
-  )
+    # if you want to truncate event times
+    mutate(
+      status = if_else(time > maxtime & status == 1, 0, status),
+      time = pmin(time, maxtime)
+    )
 
   testing_index <- read_rds('../data/resamples.rds')[[iteration]]
 
@@ -339,28 +339,68 @@ R.utils::sourceDirectory("../R")
   }
   if('nbrs_mi' %in% md_strat){
 
-    nbrs_mi <-
-      brew_nbrs(im_train_pre_impute, outcome = c(time, status)) %>%
-      spice(k_neighbors = rep(10, 5), aggregate = FALSE) %>%
-      verbose_on(level = 2) %>%
-      mash() %>%
-      stir(timer = TRUE) %>%
-      ferment(data_new = im_test_pre_impute) %>%
-      bottle() %>%
-      tidy_nbrs()
+    nbrs_mi_trn <- impute_nbrs(
+      data_ref = im_train_pre_impute,
+      cols = c(everything(), -time, -status),
+      k_neighbors = round(seq(from = 2, to = 20, length.out = n_impute_mi)),
+      aggregate = TRUE,
+      verbose = TRUE
+    )
+
+    nbrs_mi_tst <- impute_nbrs(
+      data_ref = im_test_pre_impute,
+      cols = c(everything(), -time, -status),
+      k_neighbors = round(seq(from = 2, to = 20, length.out = n_impute_mi)),
+      aggregate = TRUE,
+      verbose = TRUE
+    )
+
+    nbrs_mi <- nbrs_mi_trn %>%
+      as_tibble() %>%
+      transmute(
+        impute,
+        training = map(
+          .x = imputed_values,
+          .f = ~ipa:::fill_na(im_train_pre_impute, vals = .x)
+        ),
+        testing = map(
+          .x = nbrs_mi_tst$imputed_values,
+          .f = ~ipa:::fill_na(im_test_pre_impute, vals = .x)
+        )
+      )
 
   }
   if('nbrs_si' %in% md_strat){
 
-    nbrs_si <-
[…]

### 2020-11-01 doc_arXiv/response_to_reviewers.Rmd
all tables and figs added

---
date: "`r Sys.Date()`"
title: "Response to reviewers"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
---

# Reviewer 1

__Comment__ This article assessed the impact of missing data imputation strategies and how they impacted accuracy of risk predictions. This was conducted on a case study dataset assessing mortality and transplant from a cardiac registry dataset, INTERMACS. It explains the motivation for the study in a nice two pronged approach: (1) the importance of the INTERMACS registry, how it is used, and the potential for better use of the available data to save lives, (2) the problematic approach of using mean imputation. The paper provides a thorough method to evaluate imputations, which I think could be used as a framework for future researchers. It is encouraging to see this kind of paper in this journal, as this kind of research, evaluating missing data strategies (in this case, imputation), are in my experience not frequently seen by this audience. It is also noteworthy that the authors have gone to efforts to ensure reproducibility of the work by providing the code and data in the repository (although the link was not made public yet, as far as I can tell). The paper has fantastic potential for impact, and will hopefully help change the way that other researchers and practitioners handle missing data to administer life saving treatment. Overall, the article is excellent

__Response__ We thank the reviewer for their kind words and helpful feedback. We apologize for initially keeping our Github repository private, this was an oversight on our part. The Github repository with all of our analysis code is now publicly available at https://github.com/bcjaeger/INTERMACS-missing-data.

<br>

__Comment__ While I have a few suggestions, these are quite minor. My main aim in this feedback is to help improve some aspects of communication from the authors to the reader.

__Response__ We appreciate the helpful suggestions that the reviewer has provided and provide a point-by-point response below.

## Handling vs imputing missing values

The paper describes exploring the 'handling' missing values, but from what I can tell, it is mostly about imputation strategies, and not handling. I would consider missing data handling methods to include decisions to remove observations, as well as 'logical' imputations (such as where missing values are identified as correctly being 0, or sections needing to be filled down).

I think that the authors could address this by referring to most cases of 'handle/ing' as 'imputation, and by also discussing missing data handling strategies such as listwise deletion. For example, in the 'Methods and Results' section of the Abstract, changing, 'depends on the strategy that was applied to handle missing data' to 'depends on the imputation strategy applied'. That said, it might be possible for the authors to cover deletion methods in their results, discussed in the next section (adding listwise deletion)

## Adding listwise deletion

On page 3 on the paragraph starting at line 4, the authors describe missing data handling. The typical behaviour is imputing to the mean - which is very bad, and it is great to see the authors addressing this. Do the authors have a comment on listwise deletion? It is almost always the wrong decision to perform, but I think it is worthwhile addressing in the paper, as it is unfortunately a practice that I think people might not be aware of as being very bad.

It could also serve as a baseline/bottom point for the simulation, to show the impact of not imputing. I would be interested in the authors thoughts on this - I don't think that adding listwise deletion as an additional part of the simulation is necessary for publication, but I do think it might improve the paper as it can communicate the dangers of deletion methods.

If the authors do not want to compare deletion methods to imputation I think it would be worthwhile to mention reasoning for this in the methods, or discussion section.

However, I do think that the term, 'missing data handling' encompasses the process of the data tidying and cleaning decisions that take place when deciding what to do with missing data, which happens to include things like deletion and imputation. If the authors do not wish to discuss deletion methods, then I think they should rephrase the mentions of 'missing data handling' to intead cover only 'imputation of missing data'. If this is the case, I think the authors should provide a sentence or two justifying why they chose not to explore deletion methods in addition to imputation.

## On Acronyms

There are quite a few acronyms in this paper. Would it be possible for the authors to reduce the number of acronyms? Sentences like 'Similar to the PH model, GBDT ensembles assume PH but apply a sequential modeling algorithm to develop RPEs' are a bit dense to parse. I wonder if it might be possible to remove some of the acronyms that are sparsely used.

I am also not entirely clear what an RPE is, exactly? This might be a term that is common in this area, so my apologies if this is perhaps misplaced, but how is it different to just 'prediction', or 'predicted values'? If it is an equation that produces a prediction, I'm not sure how this is different to a predicted value / model?

## Thoughts on Methods

The methods of this paper provide concise descriptions of the methodologies. One suggestion is to add a reference to Introduction to Statistical Learning for Gradient Boosted Decision Trees and a few of the other algorithms.

### Imputation methods

Lines 19-20 on page 6 states: 
'KNN imputation identifies k 'similar' observations for each case with at least one missing value'. 
Can the authors clarify here the distinction between an observation and a case?

### Missing patterns

I'm not sure why this section is here and how it relates to the previous sections. I suggest perhaps providing a sentence or two describing why this is important to cover. I understand it is mentioned later in the paper, but it seems this section is missing some level of context, and it could be improved by explaining how this relates to the paper, and why this matters for this research objective.

### 2.6 - computational details

SAS, R, and Python should be cited as well as relevant R and Python packages/modules used to produce results (such as ggplot2, which I am quite sure was used to create the graphics).

### A note on Tables and Figures

The caption text for the tables and figures could provide more detail to the reader. For example, Table 3 and Table 4 provide important information on the performance of a variety of tests, however it would be helpful to provide some guidance. One template I have found useful is to provide three parts to each caption. The first part is an overview, the second describing the contents, and the third being what we learn/take home message. Of course, this does not always hold true, for example, Table 2 (the typical 'table one' of the paper) that provides overview of the population - there is a lot of information here that is challenging to summarise, but if there are particular things to note about the population, they could be stated here.

I don't think captions need to be a whole paragraph, but something that I think will enhance the great work the authors have done already in the paper, so that the audience can absorb it and put it into practice. For example, tables 5 and 6 could be explained in more detail - what do we learn from this, which method is best and why?

Another example example, figure 2, could be expanded to state: 'Three Model evaluation metrics (mean with 95%CI errors), improve as the number of multiple imputed datasets increases. ...'. Figure 2 could be explained in more detail in the paper as well.

## Discussion

Page 20 line 2: I wonder if some of these references could be mentioned earlier in the paper in the introduction?

The paper demonstrates how good methods such as KNN are for imputation, however I think it might be just as critical to emphasize the importance of NOT using mean/median imputation methods.

Page 20 line 10 - 12: Can you tell me more about what you learn from the figure here?

Page 22 line 1: It is fantastic to see that the authors are sharing code. Obviously the data for the paper is likely very sensitive, but I wonder if simulated data is provided? I could not find the github repository they mention, perhaps it is currently set to private? Also the repository with the code should be deposited on zenodo (or Dryad, if Circulation is able to link the repository with Dryad) so a DOI can be provided and it can be cited.

One question I would be interested in seeing answered is what other types of data these approaches are suitable or unsuitable for in the field? So, if I am a medical practitioner with a database of heart attacks that have missing data, am I likely to get similar improvements? Should I use KNN imputation as well? My thoughts are yes, but perhaps there are key structures of the data that are relevant here.

The paper has great potential for impact - I can imagine a lot of registries using these approaches to impute missing data and getting improved treatment results, and I think that a key part of this is that the authors have made efforts to share their code. This could be emphasized more in the paper, so the reader knows that they too can use similar approaches with less effort.

Overall, a great paper.

Reviewer #2:

Review for CIRCCQO/2020/00707 - Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

In this study, the authors applied six strategies of handling missing values to develop three risk prediction equations (RPE), and compared the missing handling techniques on model performance and prediction accuracy. Overall, the manuscript is well-written. To increase the applicability of these missing handling techniques, I have summarized some areas that could be improved.

1. The authors used the missing handling techniques in their RPE development. These RPEs however are not yet established; and the performance of these RPE are not satisfied, for example, the c-index for mortality event is only around 65%. To increase the application of these missing handling techniques and not to distract the readers with the new RPEs, it would be better if the authors could compare these missing handling techniques using an existing prediction model to see if their approaches could improve the predictions.

2. Imputation accuracy: Can the authors compare the imputation accuracy of their techniques through either artificially inducing missing values to known data or simulating data or literature search? It's not critical to have imputation accuracy for prediction purpose (as authors indicated in the discussion), however, the imputation accuracy impacts model inference and will limit the use of certain missing handling techniques.

3. Page 9. Line 12-14 suggests up to 50 predictor variables were used, while page 13 Line 18 suggests 320 predictors with the one missing values exists in the registry. Could the authors clarify how they choose the candidate predictors for their RPE, as well as summarize the proportion of missingness and possible missing patterns for these predictors? For example, it is possible that the variables are missing from an INTERMACS center who failed to submit data. And the variables can also be missing for a certain time period when a center hasn't started their data participation. Under these "real world data" scenarios, can the author make recommendation in their missing handling techniques?

4. Page 12. Line 1-3. Please clarify the interpretation of scaled Brier score. Usually, the lower the Brier score is, the better the predictions are. It seems here, the scaled brier score is shown as the percentage relative to naïve RPE, the closer to 100% the better. In Table 3. Although the KNN (MI) methods improved the scaled Brier score, however, the scaled Brier scores are very closed to 0, which indicate low accuracy of a RPE.

5. Table 2. please clarify when modeling mortality events, how patients with transplants events and cessation were used, are they treated as censoring or are they dropped? Please explain the potential bias if the transplant patients were not included into the analysis for mortality events. The same question for modeling transplant events, how dead and cessations were handled in the cox model for transplant risk?

6. Table 3. The concordance is low, which indicates low model performance. The table title has a typo for "mortality".

7. Page 15. Line 3. Based on Table 3, for GBDT, the improvement in scale Brier score from KNN (MI) is 0.7 compared to the reference strategy. For CIF, the improvement from KNN (MI) is 0.8 when compared to the reference. These are slightly off in the text, which could be rounding issues. Please clarify.

8. Page 15 Line 6-8. From Table 3, for the CIF, stepwise PH and GBDT, the missing data strategies that provided the lowest calibration error were missForest, KNN (MI), and BR respectively. Please correct the order of the statement.

### 2020-11-01 doc_arXiv/doc_arXiv.Rmd
all tables and figs added

@@
 
-gt_latex <- function(gt_object, caption = 'a table', label = NULL){
-
-  header = paste0(
-    '\\begin{table} \n \\caption{',caption,'} \n'
-  )
-
-  if(!is.null(label)){
-    header = paste0(header, '\\label{', label, '} \n')
-  }
-
-  gt_object %>% 
-    as_latex() %>% 
-    as.character() %>% 
-    gsub("*", "",., fixed=T) %>% 
-    gsub("\\\\ \n\\small \\\\ \n", "", ., fixed=T) %>% 
-    gsub("\n\\large", "", ., fixed=T) %>% 
-    gsub("\\captionsetup[table]{labelformat=empty,skip=1pt}", "", ., fixed=T) %>% 
-    gsub("longtable", "table", ., fixed=T) %>% 
-    gsub('\\begin{table}','\\begin{table} \n\\begin{tabular}', ., fixed=T) %>% 
-    gsub('\\end{table}','\\end{tabular} \n \\end{table}', ., fixed=T) %>% 
-    gsub('\\begin{table}', header, ., fixed=T) %>% 
-    cat()
-
-}
-
 options(knitr.table.format = 'latex')
@@
              times, 
+             rspec,
              md_method_labels,
@@
              additional_missing_labels,
-             tbl_characteristics,
-             tbl_impute_accuracy)
+             tbl_descriptives,
+             tbl_impute_accuracy,
+             tbl_md_strat,
+             fig_md_strat_infer,
+             gt_tbls)
+
+tbl_characteristics <- tbl_descriptives$characteristics
+tbl_missingness <- tbl_descriptives$missingness
 
@@
 For instance, hemodynamic laboratory values are missing for patients in whom  invasive catheterizations were not performed.
-Patients without invasive cathaderizations  may be too ill (unstable) to tolerate the procedure.
+Patients without invasive catheterizations may be too ill (unstable) to tolerate the procedure.
 Instead, such patients may, after determination of basic hemodynamics, proceed directly to device placement.
@@
 
-Here we describe the INTERMACS Registry (Section \ref{subsec:intermacs}) and give a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning (Section \ref{subsec:inference_and_learning}).
+In this section, we describe the INTERMACS Registry in Section \ref{subsec:intermacs} and provide a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning in Section \ref{subsec:inference_and_learning}.
 We then describe thirteen strategies to handle missing values (Section \ref{subsec:imputation}) and three strategies to develop risk prediction models after missing data have been imputed (Section \ref{subsec:modeling}).
@@
 
+The INTERMACS data is publicly available on biolincc at https://biolincc.nhlbi.nih.gov/studies/intermacs/.
 INTERMACS is a North American observational registry for patients receiving MCS devices that began as a partnership between the National Heart, Lung, and Blood Institute, US Food and Drug Administration, the Centers for Medicaid and Medicare Services, industry, and individual hospitals with the mission of improving MCS outcomes.
@@
 
-<!-- Though the outcome variable should be used when imputing predictor values for statistical inference, using outcome variables for this purpose in supervised learning projects can lead to issues in clinical implementation.
+\paragraph{Using outcomes during imputation} The outcome variable should be used to impute predictor values for statistical inference, but using outcome variables for this purpose in supervised learning can have unintended consequences in model implementation.
 For instance, suppose we seek to predict an outcome $Y$ for a patient in the clinical setting.
-If the patient is missing information for a predictor $X$ and our only method to impute missing values of $X$ leverages the observed value of $Y$, how should we impute $X$?
-If we do not already know $Y$, which would seem likely for clinical application of a prediction model, then we cannot impute $X$.
-On the other hand, if we already know $Y$, then we obviously do not need to predict it.
-Due to these practical considerations and because INTERMACS is often leveraged to create prediction models for clinical practice \cite{thomas2014pre}, we did not leverage outcome variables (\ie time until transplant, death, or last contact) to impute missing values of predictors in the current analysis.
--->
+If the patient is missing information for a predictor $X$ and our model's strategy to impute missing values of $X$ leverages the observed value of $Y$, how should we impute $X$?
+If we do not already know $Y$, then we cannot impute $X$ and hence cannot generate a valid prediction.
+On the other hand, if we already know $Y$, then we do not need to predict it.
+Due to these practical considerations and because INTERMACS is often leveraged to create prediction models for clinical practice \cite{thomas2014pre}, we did not leverage outcome variables to impute missing values of predictors in the current analysis.
 
@@
 In the current analysis, we identified 10 nearest neighbors using Gower's distance \cite{gower}.
-When imputing a single dataset using KNN, we aggregated values from nearest neighbors using the median for numeric variables and the mode for categorical variables.
-When imputing multiple datasets using KNN, we sampled one value at random from nearest neighbors for each imputed set.
+When imputing a single dataset using KNN, we aggregated values from 10 nearest neighbors using the median for numeric variables and the mode for categorical variables.
+When imputing multiple datasets using KNN, we aggregated values from 2, 6, 11, 16, and 20 neighbors, separately, to create 5 imputed datasets.
 
@@
 In our analyses, a Kaplan-Meier estimate based on the training data (\ie a risk prediction model that did not use any predictor variables) provided the naive prediction.
+In the current analysis, we multiply scaled Brier score values by 100 for ease of interpretation.
+
+\paragraph{Discrimination and calibration}
+
+The discrimination of a risk prediction model measures the probability that the model will successfully identify which of two observations is at higher risk for the event of interest.
+We estimated discrimination using a time-dependent concordance (C-) index that accounted for covariate-dependent censoring.
+A C-index of 0.50 and 1.00 correspond to worthless and perfect discrimination, respectively.
+Similar to the scaled Brier score, we multiply C-index values by 100 for ease of interpretation.
+Calibration slope plots measure a risk prediction model's absolute accuracy.
+We estimated calibration error by averaging the squared distance between expected and observed event rates according to a calibration plot.
+Full description of these evaluation metrics are available \cite{gerds2014calibration, gerds2013estimating}.
 
@@
 
-To assess the prognostic value of each missing data strategy, we internally validated a total of FILL IN modeling algorithm based on combinations of imputation strategies and modeling strategies described in Sections \ref{subsec:imputation} and \ref{subsec:modeling}.
+To assess the prognostic value of each missing data strategy, we internally validated a total of 23 modeling algorithm based on combinations of imputation strategies and modeling strategies described in Sections \ref{subsec:imputation} and \ref{subsec:modeling}.
 For convenience, we use the term `modeling algorithm' to denote the combination of a missing data strategy and a modeling strategy (\eg imputation using mean/mode values followed by fitting the PH model with stepwise variable selection) \cite{kuhn2013applied}.
-We conducted internal validation using 200 replicates of Monte-Carlo cross validation (MCCV), a resampling technique for internal validation (Figure FILL IN).
+We conducted internal validation using 200 replicates of Monte-Carlo cross validation, a resampling technique for internal validation.
 
-\paragraph{Steps taken in each MCCV replicate} In each replicate, 50\% of the available data were used for model training and testing.
+\paragraph{Steps taken in each replicate} In each replicate of Monte-Carlo cross validation, 50\% of the available data were used for model training and testing.
+All predictor variables with < 50\% missing values were considered for imputation and subsequent model development.
+Among these variables, artificial missingness (0\%, 15\%, or 30\% additional missing values) was induced based on patient age, with younger and older patients more likely to have missing data compared to patients who were between 40 and 65 years of age.
+Prior to imputation, 50 predictor variables were selected using a boosting model that quantified variable importance as the fractional contribution of each predictor to the model based on the total gain attributed to using the predictor while growing decision trees.
 Imputation was conducted in the training and testing sets, separately, for each imputation strategy.
-Some imputation strategies (\eg KNN and random forests) used only the training data to generate imputed values, whereas others (\eg Bayesian regression, PMM, and hot deck)
+Although some imputation strategies (\eg KNN and random forests) can impute data in the testing set using models fitted to the training set, others (\eg Bayesian regression, PMM, and hot deck) cannot.
+Therefore, to ensure fair comparisons in our experiment, each imputation procedure imputed data in the training set using models fitted to the training set and then imputed data in the testing set using models fitted to the testing set.
+After imputation, Cox PH and boosting models were applied to each imputed dataset, separately.
+Last, model predictions for death and transplant were computed at 6 months following MCS surgery.
 
-After imputation, each modeling algorithm (\ie stepwise PH models, GBDTs, and the cIF) was applied, separately, to develop a RPE for death or transplant that incorporated up to 50 predictor variables from the entire collection of covariates available.
+\paragraph{Bayesian analysis of model performance} To determine whether any of the imputation methods described above had improved upon imputation to the mean, we applied Bayesian hierarchical models to analyze differences in scaled Brier score \cite{benavoli2017time}.
+This strategy provides a flexible framework to conduct hypothesis testing and also accounts for correlation occurring within each replicate of Monte-Carlo cross validation.
+Specifically, within each replicate of Monte-Carlo cross validation, the performance of different modeling algorithms are correlated because they are trained and tested using the same data.
 
@@
 Continuous and categorical variables were summarized as mean (standard deviation) and percent, respectively.
-Imputation accuracy was aggregated for all numeric and nominal variables to create two overall scores for each imputed dataset, and the distribution of these scores was tabulated for each imputation strategy.
+The number and proportion of missing values were also tabulated for the overall population and stratified by event status.
+Missing data patterns were visualized for the overall population using an UpSet plot \cite{lex2014upset}.
+All of the proceeding analyses were conducted using resampling results from Monte-Carlo cross validation.
+Imputation accuracy was aggregated for all numeric and nominal variables to create two overall scores for each imputed dataset.
+For imputation methods that created multiple datasets, scores were aggregated over each dataset to provide a single summary score.
+The distribution (\ie 25^th^, 50^th^, and 75^th^ percentile) of the scaled Brier score was estimated for each modeling algorithm.
+
+We split our results from Monte-carlo cross validation into four datasets based on outcome (mortality or transplant) and modeling procedure (Cox PH or boosting).
+For each dataset, we fit one hierarchical Bayesian model where the dependent variable was scaled Brier score and independent variables included the imputation strategy applied and the amount of artificially missing data (0\%, 15\%, or 30\%) induced before imputation.
+With each model, we estimated the difference in scaled Brier score of downstream models when random forests, Bayesian regression, PMM, hot deck, MIA, or KNN imputation were applied to impute missing values instead of imputation to the mean.
 
@@
 
-implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}.
+Our source code is available on GitHub (see [https://github.com/bcjaeger/INTERMACS-missing-data](https://github.com/bcjaeger/INTERMACS-missing-data)).
+
+<!-- We  implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}.
 \texttt{miceRanger} follows an iterative procedure where one random forest is fit to each variable in a dataset, separately, and then used to impute missing values in that variable.
 The procedure may repeat until estimated generalization error of the random forest collection stabilizes.
+-->
 
@@
                  months_post_implant < times)
-
   ) %>% 
@@
 
+inline_missingness <- as_inline(tbl_missingness,
+                                tbl_variables = c('status', 'variable'),
+                                tbl_value = 'tbl_value')
+
 ```
@@
 
-\paragraph{Imputation accuracy}
+\paragraph{Missing data} Many predictor variables exhibited similar proportions of missing values in different outcome groups (Table \ref{tbl_missingness}).
+However, the number (percent) of missing values for surgery time was an exception, with `r inline_missingness$Overall$im_surgery_time` in the overall population, `r inline_missingness$Dead$im_surgery_time` among patients who died, and `r inline_missingness$Censored$im_surgery_time` among patients who were censored.
+Additionally, missing values for surgery time were frequently accompanied by missing values for CV pressure (Figure \ref{fig:upset}).
+
+```{r inline_impute_accuracy}
+
+si_higher_accuracy <- tbl_impute_accuracy %>% 
+  drop_na() %>% 
+  summarize(
+    n_nominal = sum(nominal_est_si > nominal_est_mi),
+    n_numeric = sum(numeric_est_si > numeric_est_mi),
+    n_total = n()
+  )
+
+top_accuracy <- tbl_impute_accuracy %>% 
+  select(-c(nominal_si, nominal_mi, numeric_si, numeric_mi)) %>% 
+  pivot_longer(cols = -c(md_method, additional_missing_pct),
+               names_to = c('score', 'estimand', 'md_type'),
+               names_sep = '_',
+               values_to = 'value') %>% 
+  filter(estimand == 'est') %>% 
+  group_by(md_method, additional_missing_pct, score, md_type) %>% 
+  summarize(value = mean(value), .groups = 'drop') %>% 
+  arrange(desc(value)) %>% 
+  group_by(score, additional_missing_pct) %>% 
+  slice(1) %>% 
+  left_join(tbl_impute_accuracy) %>% 
+  ungroup() %>% 
+  split(list(.$score, .$additional_missing_pct)) %>% 
+  map(select, md_method, nominal_si, numeric_si)
+
+break_ci <- function(x, pre_label){
+
+  data <- strsplit(x, split = ' \\(') %>% 
+    unlist() %>% 
+    str_remove('\\)') %>% 
+    set_names(c('estimate', 'interval')) %>% 
+    as.list()
+
+  glue("{pre_label}{data$estimate}, 95% CI {data$interval}")
+
+}
+
+```
+
+\paragraph{Imputation accuracy} Single imputation strategies were more accurate than their counterparts using multiple imputation in `r si_higher_accuracy$n_nominal` out of `r si_higher_accuracy$n_total` comparisons of nominal scores and `r si_higher_accuracy$n_numeric` out of `r si_higher_accuracy$n_total` comparisons of numeric scores (Table \ref{tbl_impute_accuracy}).
+When an additional 15% of data were missing, single imputation KNN obtained the highest nominal accuracy (`r break_ci(top_accuracy$nominal.15$nominal_si, pre_label = 'score: ')`) and single imputation with random forests obtained the highest numeric accuracy (`r break_ci(top_accuracy$numeric.15$numeric_si, pre_label = 'score: ')`).
+When an additional 30% of data were missing, imputation to the mean obtained the highest numeric and nominal scores.
+
+```{r inline_md_strat}
+
+inline_md_strat <- tbl_md_strat %>%
+  bind_rows(.id = 'outcome') %>% 
+  mutate(tbv = table_glue("{est} ({lwr}, {upr})", rspec = rspec)) %>% 
+  split(f = list(.$outcome,
+                 .$model,
+                 .$md_strat,
+                 .$additional_missing_pct)) %>% 
+  map(pull, tbv)
+
+```
+
+## Scaled Brier score
+
+\paragraph{Mortality risk prediction}
+
+When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$dead.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$dead.ipa.xgb.meanmode_si.0` for boosting models (Table \ref{tbl_md_strat_dead_ipa}; top panel).
+Multiple imputation strategies universally obtained higher scaled Brier scores versus their single imputation counterparts.
+Multiple imputation using random forests provided the highest scaled Brier score compared to other strategies, leading to a median (25^th^, 75^th^ percentile) increase in the scaled Brier score of `r inline_md_strat$dead.ipa.cph.ranger_mi.0` for Cox PH and `r inline_md_strat$dead.ipa.xgb.ranger_mi.0` for boosting models.
+These performance increments improved when an aditional 15\% and 30\% of data were amputed (Table \ref{tbl_md_strat_dead_ipa}; middle and bottom panel).
+
+\paragraph{Transplant risk prediction}
+
+When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$txpl.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$txpl.ipa.xgb.meanmode_si.0` for boosting models (Table \ref{tbl_md_strat_txpl_ipa}; top panel).
+For Cox PH models, imputation to the mean provided the lowest scaled Brier score, and random forest imputation led to a median (25^th^, 75^th^ percentile) increase of `r inline_md_strat$txpl.ipa.cph.ranger_mi.0` versus imputation to the mean.
+For boosting models, surprisingly, imputation to the mean provided a higher scaled Brier score than all imputation methods except for multiple imputation using Bayesian regression.
+When an additional 15\% and 30\% of data were amputed, the performance increments corresponding to the use of multiple imputation increased for both Cox PH and boosting models (Table \ref{tbl_md_strat_txpl_ipa}; middle and bottom panel).
+
+## Discrimination and calibration
+
+\paragraph{Mortality risk prediction}
+
+Regardless of how much additional data were amputed, boosting models obtained higher median C-indices than PH models for prediction of 6-month mortality risk (Table \ref{tbl_md_strat_dead_auc}).
+In addition, all models that used multiple imputation consistently obtained higher median C-indices compared to their counterparts using single imputation.
+When 0 and 15\% additional data were amputed, median calibration error was lower for PH models compared to boosting models, but boosting models obtained lower median calibration error when 30\% additional data were amputed (Table \ref{tbl_md_strat_dead_cal_error}).
+Almost all imputation strategies provided lower median calibration error compared with imputation to the mean.
+
+\paragraph{Transplant risk prediction}
+
+For all amounts of additional data amputed, PH models obtained higher median C-indices than boosting models for prediction of 6-month transplant risk (Table \ref{tbl_md_strat_txpl_auc}).
+Similar to mortality risk prediction, multiple imputation strategies generally provided higher C-indices than their counterparts using single imputation strategy.
+For boosting models, MIA provided higher C-indices compared to all other single imputation strategies and had similar or superior performance compared to several multiple imputation strategies.
+Differences in calibration error were minor when no additional missing data were amputed (Table \ref{tbl_md_strat_txpl_cal_error}).
+When an additional 30\% of data were amputed, boosting models using MIA obtained lower calibration error than any other strategy.
+
+\paragraph{Bayesian analysis of model performance}
+
+```{r}
+
+fig_md_strat_inline <- fig_md_strat_infer %>% 
+  map(~ map(.x$probs, as_inline, 'm2', 'prob_gt_meanmode'))
+
+inline_ipa_rf <- fig_md_strat_inline$ipa$`Multiple imputation`$ranger
+inline_auc_rf <- fig_md_strat_inline$auc$`Multiple imputation`$ranger
+
+```
+
+Adjusting for the amount of additional missing data amputed and the outcome variable, the posterior probability that an imputation strategy would improve the scaled Brier score of a downstream model relative to imputation to the mean was maximized by using multiple imputation with random forests (probability of improvement: `r inline_ipa_rf`; Figure \ref{fig:fig_md_strat_infer_ipa}).
+Similarly, multiple imputation using random forests was estimated to have the highest posterior probability of improving the C-index in comparison to using imputation to the mean (probability of improvement: `r inline_auc_rf`; Figure \ref{fig:fig_md_strat_infer_auc}).
+However, the estimated posterior probability of a reduction in calibration was 1 when using either single or multiple imputation with random forests compared with imputation to the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}).
+Although imputation using random forest was estimated to be the best overall option, there was moderate to strong evidence that MIA and each multiple imputation strategy we applied would improve prognostic accuracy of downstream models compared with imputation to the mean.
+
+# Discussion
+
+In this article, we leveraged data from the INTERMACS registry to evaluate how the use of different imputation strategies prior to fitting a risk prediction model would impact the external prognostic accuracy of the model.
+External prognostic accuracy was measured at `r times` months after receiving MCS to focus on short term risk prediction, and the primary measure of accuracy was the scaled Brier score.
+We evaluated the performance of 12 imputation strategies in a broad range of settings by varying the type of modeling algorithm, the outcome variable, and the amount of additional missing data amputed prior to performing imputation.
+Our resampling experiment indicated that conducting multiple imputation has a high likelihood of increasing the downstream scaled Brier score and C-index of risk prediction models compared with imputation to the mean.
+Additionally, multiple imputation with random forests emerged as the imputation strategy that maximized the probability of developing a more prognostic model compared with imputation to the mean.
+
+In previous studies involving the INTERMACS data registry, imputation to the mean has been applied prior to developing a mortality risk prediction model \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society}.
+The current analysis has shown that the prognostic accuracy of these models can be increased by implementing a different strategy to impute missing values, particularly if multiple imputation is applied.
+However, few studies have investigated best practices on the use of multiple imputation for prediction models, and few software packages allow users to implement multiple imputation procedures that adhere to best practices for prediction model development (\eg using only the training data to impute missing values).
+Further development of open source software that allows investigators to safely and succinctly implement multiple imputation for prediction models (\eg the \texttt{miceRanger} package) may increase the presence of multiple imputation in published risk prediction models.
 
@@
 
+<!-- Table : missingness -->
+\clearpage
+
+```{r tbl_missingness, results = 'asis'}
+
+tbl_missingness %>%
+  select(-variable) %>%
+  pivot_wider(values_from = tbl_value,
+              names_from = status) %>%
+  filter(label != 'status',
+         label != 'months_post_implant') %>%
+  gt(rowname_col = 'label') %>%
+  cols_align(align = 'center') %>%
+  gt_latex(
+    caption = 'Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.',
+    label = 'tbl_missingness'
+  )
+
+```
+
 <!-- Table: imputation accuracy -->
@@
 
-```{r tbl_impute_accuracy, results = 'asis'}
+```{r, results = 'asis'}
 
 tbl_impute_accuracy %>% 
+  select(md_method:numeric_mi) %>% 
   mutate(
@@
 ```
+
+\clearpage
+
+```{r tbl_md_strat_dead_ipa, results = 'asis'}
+
+make_tbl_md_strat_caption <- function(metric, outcome){
+  glue('Median (25th, 75th percentile) change in {metric} when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for {outcome}.
+Table values show the {metric} for imputation to the mean.
+For other imputation strategies, table values show the change in {metric} relative to the {metric} when imputation to the mean was applied.
+All table values are multiplied by 100 for ease of interpretability')
+}
+
+gt_latex(
+  gt_tbls$md_strat$dead.ipa,
+  caption = make_tbl_md_strat_caption(metric = 'scaled Brier score',
+                                      outcome = 'mortality'),
+  label = 'tbl_md_strat_dead_ipa'
+)
+
+```
+
+\clearpage
+
+```{r tbl_md_strat_txpl_ipa, results = 'asis'}
+
+gt_latex(
+  gt_tbls$md_strat$txpl.ipa,
+  caption = make_tbl_md_strat_caption(metric = 'scaled Brier score',
+                                      outcome = 'transplant'),
+  label = 'tbl_md_strat_txpl_ipa'
+)
+
+```
+
+\clearpage
+
+```{r tbl_md_strat_dead_auc, results = 'asis'}
+
+gt_latex(
+  gt_tbls$md_strat$dead.auc,
+  caption = make_tbl_md_strat_caption(metric = 'concordance index',
+                                      outcome = 'mortality'),
+  label = 'tbl_md_strat_dead_auc'
+)
+
+```
+
+\clearpage
+
+```{r tbl_md_strat_dead_cal_error, results = 'asis'}
+
+gt_latex(
+  gt_tbls$md_strat$dead.cal_error,
+  caption = make_tbl_md_strat_caption(metric = 'calibration error',
+                                      outcome = 'mortality'),
+  label = 'tbl_md_strat_dead_cal_error'
+)
+
+```
+
+\clearpage
+
+```{r tbl_md_strat_txpl_auc, results = 'asis'}
+
+gt_latex(
+  gt_tbls$md_strat$txpl.auc,
+  caption = make_tbl_md_strat_caption(metric = 'concordance index',
+                                      outcome = 'transplant'),
+  label = 'tbl_md_strat_txpl_auc'
+)
+
+```
+
+\clearpage
+
+```{r tbl_md_strat_txpl_cal_error, results = 'asis'}
+
+gt_latex(
+  gt_tbls$md_strat$txpl.cal_error,
+  caption = make_tbl_md_strat_caption(metric = 'calibration error',
+                                      outcome = 'transplant'),
+  label = 'tbl_md_strat_txpl_cal_error'
+)
+
+```
+
+\clearpage
+
+```{r upset, dpi = 300, fig.height = 6, fig.width = 6.5, fig.align='center', fig.cap="An upset plot showing three variables from the INTERMACS registry and all combinations of missing patterns.
+The bottom left plot shows the number of missing values for each variable, separately.
+The top right plot shows the number of missing values for each combination of the three variables.
+For example, there were 2,618 rows in the overall INTERMACS data where both CV pressure and surgery time were missing."}
+
+gg_miss_upset(tbl_descriptives$upset, nsets = 3, text.scale = 1.3)
+
+```
+
+\clearpage
+
+```{r fig_md_strat_infer_ipa, dpi = 300, fig.height = 6, fig.width = 6.5, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
+Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution."}
+
+fig_md_strat_infer$ipa$fig
+
+```
+
+\clearpage
+
+```{r fig_md_strat_infer_auc, dpi = 300, fig.height = 6, fig.width = 6.5, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
+Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution."}
+
+fig_md_strat_infer$auc$fig
+
+```
+
+\clearpage
+
+```{r fig_md_strat_infer_cal_error, dpi = 300, fig.height = 6, fig.width = 6.5, fig.align='center', fig.cap = "Posterior distribution of differences in calibration error values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
+Posterior probability that the difference in calibration error exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution."}
+
+fig_md_strat_infer$cal_error$fig
+
+```
+
+\clearpage

### 2020-11-01 doc_arXiv/doc_arXiv.tex
all tables and figs added

@@
 catheterizations were not performed. Patients without invasive
-cathaderizations may be too ill (unstable) to tolerate the procedure.
+catheterizations may be too ill (unstable) to tolerate the procedure.
 Instead, such patients may, after determination of basic hemodynamics,
@@
 
-Here we describe the INTERMACS Registry (Section \ref{subsec:intermacs})
-and give a broad overview of general objectives for imputation of
-missing data in the context of statistical inference and supervised
-statistical learning (Section \ref{subsec:inference_and_learning}). We
-then describe thirteen strategies to handle missing values (Section
-\ref{subsec:imputation}) and three strategies to develop risk prediction
-models after missing data have been imputed (Section
-\ref{subsec:modeling}). Next, we describe how risk prediction models and
-by extension the imputation strategy applied before training the model
-were evaluated in Section \ref{subsec:evaluation}. We outline an
-internal validation procedure that was applied to compare combinations
-of imputation and modeling strategies using hierarchical Bayesian models
-in Section \ref{subsec:internal}. Last, details on computation and
-software packages used for the current analysis are described in Section
+In this section, we describe the INTERMACS Registry in Section
+\ref{subsec:intermacs} and provide a broad overview of general
+objectives for imputation of missing data in the context of statistical
+inference and supervised statistical learning in Section
+\ref{subsec:inference_and_learning}. We then describe thirteen
+strategies to handle missing values (Section \ref{subsec:imputation})
+and three strategies to develop risk prediction models after missing
+data have been imputed (Section \ref{subsec:modeling}). Next, we
+describe how risk prediction models and by extension the imputation
+strategy applied before training the model were evaluated in Section
+\ref{subsec:evaluation}. We outline an internal validation procedure
+that was applied to compare combinations of imputation and modeling
+strategies using hierarchical Bayesian models in Section
+\ref{subsec:internal}. Last, details on computation and software
+packages used for the current analysis are described in Section
 \ref{subsec:computing}
@@
 
-INTERMACS is a North American observational registry for patients
-receiving MCS devices that began as a partnership between the National
-Heart, Lung, and Blood Institute, US Food and Drug Administration, the
-Centers for Medicaid and Medicare Services, industry, and individual
-hospitals with the mission of improving MCS outcomes. In 2018, INTERMACS
-became an official Society of Thoracic Surgeons database.
+The INTERMACS data is publicly available on biolincc at
+\url{https://biolincc.nhlbi.nih.gov/studies/intermacs/}. INTERMACS is a
+North American observational registry for patients receiving MCS devices
+that began as a partnership between the National Heart, Lung, and Blood
+Institute, US Food and Drug Administration, the Centers for Medicaid and
+Medicare Services, industry, and individual hospitals with the mission
+of improving MCS outcomes. In 2018, INTERMACS became an official Society
+of Thoracic Surgeons database.
 
@@
 
-Though the outcome variable may be useful for imputing predictor values,
-using outcome variables for this purpose in supervised learning projects
-can lead to issues in clinical implementation. For instance, suppose we
-seek to predict an outcome \(Y\) for a patient in the clinical setting.
-If the patient is missing information for a predictor \(X\) and our only
-method to impute missing values of \(X\) leverages the observed value of
-\(Y\), how should we impute \(X\)? If we do not already know \(Y\),
-which would seem likely for clinical application of a prediction model,
-then we cannot impute \(X\). On the other hand, if we already know
-\(Y\), then we obviously do not need to predict it. Due to these
-practical considerations and because INTERMACS is often leveraged to
-create prediction models for clinical practice \cite{thomas2014pre}, we
-did not leverage outcome variables (\textit{i.e., }time until
-transplant, death, or last contact) to impute missing values of
-predictors in the current analysis.
+\paragraph{Using outcomes during imputation}
+
+The outcome variable should be used to impute predictor values for
+statistical inference, but using outcome variables for this purpose in
+supervised learning can have unintended consequences in model
+implementation. For instance, suppose we seek to predict an outcome
+\(Y\) for a patient in the clinical setting. If the patient is missing
+information for a predictor \(X\) and our model's strategy to impute
+missing values of \(X\) leverages the observed value of \(Y\), how
+should we impute \(X\)? If we do not already know \(Y\), then we cannot
+impute \(X\) and hence cannot generate a valid prediction. On the other
+hand, if we already know \(Y\), then we do not need to predict it. Due
+to these practical considerations and because INTERMACS is often
+leveraged to create prediction models for clinical practice
+\cite{thomas2014pre}, we did not leverage outcome variables to impute
+missing values of predictors in the current analysis.
 
@@
 nearest neighbors using Gower's distance \cite{gower}. When imputing a
-single dataset using KNN, we aggregated values from nearest neighbors
+single dataset using KNN, we aggregated values from 10 nearest neighbors
 using the median for numeric variables and the mode for categorical
-variables. When imputing multiple datasets using KNN, we sampled one
-value at random from nearest neighbors for each imputed set.
+variables. When imputing multiple datasets using KNN, we aggregated
+values from 2, 6, 11, 16, and 20 neighbors, separately, to create 5
+imputed datasets.
 
@@
 model that did not use any predictor variables) provided the naive
-prediction.
+prediction. In the current analysis, we multiply scaled Brier score
+values by 100 for ease of interpretation.
+
+\paragraph{Discrimination and calibration}
+
+The discrimination of a risk prediction model measures the probability
+that the model will successfully identify which of two observations is
+at higher risk for the event of interest. We estimated discrimination
+using a time-dependent concordance (C-) index that accounted for
+covariate-dependent censoring. A C-index of 0.50 and 1.00 correspond to
+worthless and perfect discrimination, respectively. Similar to the
+scaled Brier score, we multiply C-index values by 100 for ease of
+interpretation. Calibration slope plots measure a risk prediction
+model's absolute accuracy. We estimated calibration error by averaging
+the squared distance between expected and observed event rates according
+to a calibration plot. Full description of these evaluation metrics are
+available \cite{gerds2014calibration, gerds2013estimating}.
 
@@
 To assess the prognostic value of each missing data strategy, we
-internally validated a total of FILL IN modeling algorithm based on
+internally validated a total of 23 modeling algorithm based on
 combinations of imputation strategies and modeling strategies described
@@
 We conducted internal validation using 200 replicates of Monte-Carlo
-cross validation (MCCV), a resampling technique for internal validation
-(Figure \ref{fig:monte_carlo_cv}).
+cross validation, a resampling technique for internal validation.
 
-\paragraph{Steps taken in each MCCV replicate}
+\paragraph{Steps taken in each replicate}
 
-In each replicate, 50\% of the available data were used for model
-training and testing. Imputation was conducted in the training and
-testing sets, separately, for each imputation strategy. Some imputation
-strategies (\textit{e.g., }KNN and random forests) used only the
-training data to generate imputed values, whereas others
-(\textit{e.g., }Bayesian regression, PMM, and hot deck)
+In each replicate of Monte-Carlo cross validation, 50\% of the available
+data were used for model training and testing. All predictor variables
+with \textless{} 50\% missing values were considered for imputation and
+subsequent model development. Among these variables, artificial
+missingness (0\%, 15\%, or 30\% additional missing values) was induced
+based on patient age, with younger and older patients more likely to
+have missing data compared to patients who were between 40 and 65 years
+of age. Prior to imputation, 50 predictor variables were selected using
+a boosting model that quantified variable importance as the fractional
+contribution of each predictor to the model based on the total gain
+attributed to using the predictor while growing decision trees.
+Imputation was conducted in the training and testing sets, separately,
+for each imputation strategy. Although some imputation strategies
+(\textit{e.g., }KNN and random forests) can impute data in the testing
+set using models fitted to the training set, others
+(\textit{e.g., }Bayesian regression, PMM, and hot deck) cannot.
+Therefore, to ensure fair comparisons in our experiment, each imputation
+procedure imputed data in the training set using models fitted to the
+training set and then imputed data in the testing set using models
+fitted to the testing set. After imputation, Cox PH and boosting models
+were applied to each imputed dataset, separately. Last, model
+predictions for death and transplant were computed at 6 months following
+MCS surgery.
 
-After imputation, each modeling algorithm (\textit{i.e., }stepwise PH
-models, GBDTs, and the cIF) was applied, separately, to develop a RPE
-for death or transplant that incorporated up to 50 predictor variables
-from the entire collection of covariates available.
+\paragraph{Bayesian analysis of model performance}
+
+To determine whether any of the imputation methods described above had
+improved upon imputation to the mean, we applied Bayesian hierarchical
+models to analyze differences in scaled Brier score
+\cite{benavoli2017time}. This strategy provides a flexible framework to
+conduct hypothesis testing and also accounts for correlation occurring
+within each replicate of Monte-Carlo cross validation. Specifically,
+within each replicate of Monte-Carlo cross validation, the performance
+of different modeling algorithms are correlated because they are trained
+and tested using the same data.
 
@@
 were summarized as mean (standard deviation) and percent, respectively.
-Imputation accuracy was aggregated for all numeric and nominal variables
-to create two overall scores for each imputed dataset, and the
-distribution of these scores was tabulated for each imputation strategy.
+The number and proportion of missing values were also tabulated for the
+overall population and stratified by event status. Missing data patterns
+were visualized for the overall population using an UpSet plot
+\cite{lex2014upset}. All of the proceeding analyses were conducted using
+resampling results from Monte-Carlo cross validation. Imputation
+accuracy was aggregated for all numeric and nominal variables to create
+two overall scores for each imputed dataset. For imputation methods that
+created multiple datasets, scores were aggregated over each dataset to
+provide a single summary score. The distribution
+(\textit{i.e., }25\textsuperscript{th}, 50\textsuperscript{th}, and
+75\textsuperscript{th} percentile) of the scaled Brier score was
+estimated for each modeling algorithm.
+
+We split our results from Monte-carlo cross validation into four
+datasets based on outcome (mortality or transplant) and modeling
+procedure (Cox PH or boosting). For each dataset, we fit one
+hierarchical Bayesian model where the dependent variable was scaled
+Brier score and independent variables included the imputation strategy
+applied and the amount of artificially missing data (0\%, 15\%, or 30\%)
+induced before imputation. With each model, we estimated the difference
+in scaled Brier score of downstream models when random forests, Bayesian
+regression, PMM, hot deck, MIA, or KNN imputation were applied to impute
+missing values instead of imputation to the mean.
 
@@
 
-implement \texttt{miceRanger}, an R package similar to
-\texttt{missForest} that implements MICE with random forests by
-implementing PMM \cite{missForest}. \texttt{miceRanger} follows an
-iterative procedure where one random forest is fit to each variable in a
-dataset, separately, and then used to impute missing values in that
-variable. The procedure may repeat until estimated generalization error
-of the random forest collection stabilizes.
+Our source code is available on GitHub (see
+\url{https://github.com/bcjaeger/INTERMACS-missing-data}).
 
@@
 
+\paragraph{Missing data}
+
+Many predictor variables exhibited similar proportions of missing values
+in different outcome groups (Table \ref{tbl_missingness}). However, the
+number (percent) of missing values for surgery time was an exception,
+with 6,125 (41.6\%) in the overall population, 2,246 (53.4\%) among
+patients who died, and 1,713 (27.2\%) among patients who were censored.
+Additionally, missing values for surgery time were frequently
+accompanied by missing values for CV pressure (Figure \ref{fig:upset}).
+
 \paragraph{Imputation accuracy}
+
+Single imputation strategies were more accurate than their counterparts
+using multiple imputation in 8 out of 10 comparisons of nominal scores
+and 10 out of 10 comparisons of numeric scores (Table
+\ref{tbl_impute_accuracy}). When an additional 15\% of data were
+missing, single imputation KNN obtained the highest nominal accuracy
+(score: +0.02, 95\% CI -0.05, 0.15) and single imputation with random
+forests obtained the highest numeric accuracy (score: +0.05, 95\% CI
+0.03, 0.08). When an additional 30\% of data were missing, imputation to
+the mean obtained the highest numeric and nominal scores.
+
+\hypertarget{scaled-brier-score}{%
+\subsection{Scaled Brier score}\label{scaled-brier-score}}
+
+\paragraph{Mortality risk prediction}
+
+When no additional data were amputed and imputation to the mean was
+applied before fitting risk prediction models, the median
+(25\textsuperscript{th}, 75\textsuperscript{th} percentile) scaled Brier
+score was 6.09 (5.71, 6.54) for Cox PH and 7.27 (6.83, 7.77) for
+boosting models (Table \ref{tbl_md_strat_dead_ipa}; top panel). Multiple
+imputation strategies universally obtained higher scaled Brier scores
+versus their single imputation counterparts. Multiple imputation using
+random forests provided the highest scaled Brier score compared to other
+strategies, leading to a median (25\textsuperscript{th},
+75\textsuperscript{th} percentile) increase in the scaled Brier score of
+0.34 (0.12, 0.54) for Cox PH and 0.34 (0.14, 0.58) for boosting models.
+These performance increments improved when an aditional 15\% and 30\% of
+data were amputed (Table \ref{tbl_md_strat_dead_ipa}; middle and bottom
+panel).
+
+\paragraph{Transplant risk prediction}
+
+When no additional data were amputed and imputation to the mean was
+applied before fitting risk prediction models, the median
+(25\textsuperscript{th}, 75\textsuperscript{th} percentile) scaled Brier
+score was 9.35 (8.80, 9.71) for Cox PH and 9.00 (8.55, 9.41) for
+boosting models (Table \ref{tbl_md_strat_txpl_ipa}; top panel). For Cox
+PH models, imputation to the mean provided the lowest scaled Brier
+score, and random forest imputation led to a median
+(25\textsuperscript{th}, 75\textsuperscript{th} percentile) increase of
+0.13 (0.03, 0.29) versus imputation to the mean. For boosting models,
+surprisingly, imputation to the mean provided a higher scaled Brier
+score than all imputation methods except for multiple imputation using
+Bayesian regression. When an additional 15\% and 30\% of data were
+amputed, the performance increments corresponding to the use of multiple
+imputation increased for both Cox PH and boosting models (Table
+\ref{tbl_md_strat_txpl_ipa}; middle and bottom panel).
+
+\hypertarget{discrimination-and-calibration}{%
+\subsection{Discrimination and
+calibration}\label{discrimination-and-calibration}}
+
+\paragraph{Mortality risk prediction}
+
+Regardless of how much additional data were amputed, boosting models
+obtained higher median C-indices than PH models for prediction of
+6-month mortality risk (Table \ref{tbl_md_strat_dead_auc}). In addition,
+all models that used multiple imputation consistently obtained higher
+median C-indices compared to their counterparts using single imputation.
+When 0 and 15\% additional data were amputed, median calibration error
+was lower for PH models compared to boosting models, but boosting models
+obtained lower median calibration error when 30\% additional data were
+amputed (Table \ref{tbl_md_strat_dead_cal_error}). Almost all imputation
+strategies provided lower median calibration error compared with
+imputation to the mean.
+
+\paragraph{Transplant risk prediction}
+
+For all amounts of additional data amputed, PH models obtained higher
+median C-indices than boosting models for prediction of 6-month
+transplant risk (Table \ref{tbl_md_strat_txpl_auc}). Similar to
+mortality risk prediction, multiple imputation strategies generally
+provided higher C-indices than their counterparts using single
+imputation strategy. For boosting models, MIA provided higher C-indices
+compared to all other single imputation strategies and had similar or
+superior performance compared to several multiple imputation strategies.
+Differences in calibration error were minor when no additional missing
+data were amputed (Table \ref{tbl_md_strat_txpl_cal_error}). When an
+additional 30\% of data were amputed, boosting models using MIA obtained
+lower calibration error than any other strategy.
+
+\paragraph{Bayesian analysis of model performance}
+
+Adjusting for the amount of additional missing data amputed and the
+outcome variable, the posterior probability that an imputation strategy
+would improve the scaled Brier score of a downstream model relative to
+imputation to the mean was maximized by using multiple imputation with
+random forests (probability of improvement: 0.998; Figure
+\ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
+random forests was estimated to have the highest posterior probability
+of improving the C-index in comparison to using imputation to the mean
+(probability of improvement: 0.962; Figure
+\ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
+probability of a reduction in calibration was 1 when using either single
+or multiple imputation with random forests compared with imputation to
+the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}). Although
+imputation using random forest was estimated to be the best overall
+option, there was moderate to strong evidence that MIA and each multiple
+imputation strategy we applied would improve prognostic accuracy of
+downstream models compared with imputation to the mean.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+In this article, we leveraged data from the INTERMACS registry to
+evaluate how the use of different imputation strategies prior to fitting
+a risk prediction model would impact the external prognostic accuracy of
+the model. External prognostic accuracy was measured at 6 months after
+receiving MCS to focus on short term risk prediction, and the primary
+measure of accuracy was the scaled Brier score. We evaluated the
+performance of 12 imputation strategies in a broad range of settings by
+varying the type of modeling algorithm, the outcome variable, and the
+amount of additional missing data amputed prior to performing
+imputation. Our resampling experiment indicated that conducting multiple
+imputation has a high likelihood of increasing the downstream scaled
+Brier score and C-index of risk prediction models compared with
+imputation to the mean. Additionally, multiple imputation with random
+forests emerged as the imputation strategy that maximized the
+probability of developing a more prognostic model compared with
+imputation to the mean.
+
+In previous studies involving the INTERMACS data registry, imputation to
+the mean has been applied prior to developing a mortality risk
+prediction model
+\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society}.
+The current analysis has shown that the prognostic accuracy of these
+models can be increased by implementing a different strategy to impute
+missing values, particularly if multiple imputation is applied. However,
+few studies have investigated best practices on the use of multiple
+imputation for prediction models, and few software packages allow users
+to implement multiple imputation procedures that adhere to best
+practices for prediction model development (\textit{e.g., }using only
+the training data to impute missing values). Further development of open
+source software that allows investigators to safely and succinctly
+implement multiple imputation for prediction models (\textit{e.g., }the
+\texttt{miceRanger} package) may increase the presence of multiple
+imputation in published risk prediction models.
 
@@
 No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
-Follow up time, months & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
+months\_post\_implant & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
 Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
@@
 \addlinespace
+LVEDD & 6.83 (1.10) & 6.85 (1.07) & 6.49 (0.93) & 6.69 (1.12) & 6.95 (1.12)\\
+Urinary albumin, g/dl & 3.41 (0.64) & 3.42 (0.61) & 3.41 (0.64) & 3.33 (0.66) & 3.49 (0.65)\\
+Urinary creatinine, mg/dl & 1.39 (0.68) & 1.36 (0.62) & 1.17 (0.53) & 1.49 (0.77) & 1.35 (0.67)\\
+CV pressure & 10.8 (6.2) & 10.4 (6.1) & 9.9 (6.0) & 11.6 (6.5) & 10.7 (6.0)\\
+Periphal edema & 35 & 33 & 37 & 44 & 31\\
+\addlinespace
+BUN, mg/dl & 29 (18) & 28 (17) & 22 (13) & 32 (19) & 28 (18)\\
+Bilirubin levels, mg/dl & 1.37 (1.84) & 1.25 (1.14) & 1.28 (1.95) & 1.52 (2.46) & 1.41 (1.97)\\
 Device type &  &  &  &  & \\
@@
 \hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
+\addlinespace
+Surgery time, minutes & 292 (113) & 289 (110) & 272 (96) & 320 (128) & 272 (98)\\
+CPB time, minutes & 95 (49) & 93 (45) & 88 (38) & 103 (56) & 90 (44)\\
 \bottomrule
@@
 
+\end{table}
+
+\clearpage
+
+\begin{table} 
+ \caption{Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.} 
+\label{tbl_missingness}
+
+\begin{tabular}{lccccc}
+\toprule
+ & Overall & Dead & Transplant & Censored & Cessation of support \\ 
+\midrule
+CV pressure & 7,415 (50.3\%) & 2,156 (51.3\%) & 2,006 (51.2\%) & 3,108 (49.3\%) & 145 (46.8\%) \\ 
+Surgery time, minutes & 6,125 (41.6\%) & 2,246 (53.4\%) & 2,007 (51.3\%) & 1,713 (27.2\%) & 159 (51.3\%) \\ 
+Periphal edema & 5,093 (34.6\%) & 1,543 (36.7\%) & 1,470 (37.5\%) & 1,952 (30.9\%) & 128 (41.3\%) \\ 
+LVEDD & 3,155 (21.4\%) & 1,000 (23.8\%) & 866 (22.1\%) & 1,233 (19.5\%) & 56 (18.1\%) \\ 
+Urinary albumin, g/dl & 1,014 (6.88\%) & 326 (7.75\%) & 309 (7.89\%) & 359 (5.69\%) & 20 (6.45\%) \\ 
+CPB time, minutes & 1,010 (6.85\%) & 287 (6.83\%) & 274 (7.00\%) & 421 (6.67\%) & 28 (9.03\%) \\ 
+Bilirubin levels, mg/dl & 850 (5.77\%) & 291 (6.92\%) & 237 (6.05\%) & 304 (4.82\%) & 18 (5.81\%) \\ 
+Body mass index & 77 (0.52\%) & 26 (0.62\%) & 19 (0.49\%) & 31 (0.49\%) & 1 (0.32\%) \\ 
+BUN, mg/dl & 50 (0.34\%) & 10 (0.24\%) & 8 (0.20\%) & 32 (0.51\%) & 0 (0.00\%) \\ 
+Sex & 24 (0.16\%) & 9 (0.21\%) & 4 (0.10\%) & 11 (0.17\%) & 0 (0.00\%) \\ 
+Urinary creatinine, mg/dl & 20 (0.14\%) & 8 (0.19\%) & 4 (0.10\%) & 8 (0.13\%) & 0 (0.00\%) \\ 
+Age, years & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Race & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Device strategy & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Device type & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+\bottomrule
+\end{tabular} 
  \end{table}
@@
 Hot deck & -0.24 (-0.34, -0.16) & -0.25 (-0.37, -0.18) & -0.35 (-0.53, -0.29) & -0.42 (-0.54, -0.36) \\ 
-K-nearest-neighbors & 0.02 (-0.05, 0.15) & -0.19 (-0.29, -0.08) & 0.01 (-0.01, 0.03) & -0.33 (-0.42, -0.28) \\ 
+K-nearest-neighbors & +0.02 (-0.05, 0.15) & -0.01 (-0.09, 0.11) & +0.01 (-0.01, 0.03) & -0.02 (-0.05, 0.00) \\ 
 Predictive mean matching & -0.15 (-0.34, 0.00) & -0.15 (-0.35, 0.00) & -0.25 (-0.40, -0.20) & -0.26 (-0.35, -0.21) \\ 
-Random forest & -0.15 (-0.26, -0.02) & -0.15 (-0.26, -0.03) & 0.06 (0.02, 0.08) & 0.05 (0.02, 0.07) \\ 
+Random forest & -0.15 (-0.26, -0.03) & -0.15 (-0.26, -0.03) & +0.05 (0.03, 0.08) & +0.02 (-0.01, 0.04) \\ 
 Bayesian regression & -0.14 (-0.26, 0.01) & -0.14 (-0.27, 0.01) & -0.30 (-0.41, -0.23) & -0.30 (-0.41, -0.23) \\ 
@@
 Hot deck & -0.25 (-0.34, -0.18) & -0.26 (-0.36, -0.18) & -0.34 (-0.46, -0.30) & -0.40 (-0.53, -0.36) \\ 
-K-nearest-neighbors & -0.04 (-0.09, 0.03) & -0.24 (-0.34, -0.16) & -0.02 (-0.03, -0.01) & -0.36 (-0.44, -0.32) \\ 
+K-nearest-neighbors & -0.04 (-0.09, 0.03) & -0.07 (-0.13, 0.00) & -0.02 (-0.03, -0.01) & -0.05 (-0.07, -0.04) \\ 
 Predictive mean matching & -0.18 (-0.38, -0.04) & -0.18 (-0.37, -0.04) & -0.29 (-0.46, -0.24) & -0.29 (-0.40, -0.25) \\ 
-Random forest & -0.18 (-0.27, -0.07) & -0.18 (-0.28, -0.07) & -0.02 (-0.05, 0.01) & -0.01 (-0.04, 0.01) \\ 
+Random forest & -0.18 (-0.28, -0.07) & -0.18 (-0.28, -0.08) & -0.01 (-0.03, 0.01) & -0.03 (-0.06, -0.01) \\ 
 Bayesian regression & -0.17 (-0.29, -0.04) & -0.17 (-0.29, -0.05) & -0.31 (-0.43, -0.27) & -0.31 (-0.43, -0.27) \\ 
@@
 
+\clearpage
+
+\begin{table} 
+ \caption{Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+Table values show the scaled Brier score for imputation to the mean.
+For other imputation strategies, table values show the change in scaled Brier score relative to the scaled Brier score when imputation to the mean was applied.
+All table values are multiplied by 100 for ease of interpretability}
+\label{tbl_md_strat_dead_ipa}
+
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{Proportional hazards} & \multicolumn{2}{c}{Gradient boosted decision trees} \\ 
+ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+Imputation method & Single Imputation & Multiple Imputation & Single Imputation & Multiple Imputation \\ 
+\midrule
+\multicolumn{1}{l}{No additional missing data} \\ 
+\midrule
+Imputation to the mean & 6.09 (5.71, 6.54) & -- & 7.27 (6.83, 7.77) & -- \\ 
+Missingness as an attribute & -- & -- & +0.09 (-0.16, 0.31) & -- \\ 
+Hot deck & -0.25 (-0.47, 0.01) & +0.05 (-0.15, 0.24) & -0.24 (-0.46, 0.03) & +0.25 (0.08, 0.45) \\ 
+K-nearest-neighbors & +0.09 (-0.06, 0.23) & +0.18 (0.04, 0.34) & -0.06 (-0.27, 0.18) & +0.28 (0.06, 0.46) \\ 
+Predictive mean matching & -0.30 (-0.53, -0.03) & +0.15 (-0.01, 0.37) & -0.22 (-0.43, 0.14) & +0.34 (0.13, 0.54) \\ 
+Random forest & +0.05 (-0.24, 0.33) & +0.34 (0.12, 0.54) & -0.08 (-0.32, 0.21) & +0.34 (0.14, 0.58) \\ 
+Bayesian regression & -0.35 (-0.61, -0.10) & +0.07 (-0.10, 0.29) & -0.25 (-0.52, 0.05) & +0.29 (0.08, 0.52) \\ 
+\midrule
+\multicolumn{1}{l}{+15\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 5.56 (4.95, 6.03) & -- & 6.72 (6.13, 7.33) & -- \\ 
+Missingness as an attribute & -- & -- & +0.52 (0.21, 0.77) & -- \\ 
+Hot deck & +0.10 (-0.22, 0.46) & +0.49 (0.21, 0.78) & +0.02 (-0.28, 0.33) & +0.58 (0.29, 0.92) \\ 
+K-nearest-neighbors & +0.36 (0.10, 0.65) & +0.53 (0.25, 0.79) & +0.04 (-0.23, 0.36) & +0.42 (0.16, 0.68) \\ 
+Predictive mean matching & +0.03 (-0.32, 0.41) & +0.54 (0.20, 0.90) & +0.21 (-0.25, 0.64) & +0.66 (0.32, 1.12) \\ 
+Random forest & +0.39 (-0.05, 0.75) & +0.76 (0.37, 1.12) & +0.35 (0.01, 0.76) & +0.73 (0.41, 1.06) \\ 
+Bayesian regression & +0.05 (-0.44, 0.44) & +0.49 (0.10, 0.85) & +0.19 (-0.21, 0.59) & +0.58 (0.29, 1.06) \\ 
+\midrule
+\multicolumn{1}{l}{+30\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 4.79 (4.11, 5.55) & -- & 6.29 (5.67, 7.02) & -- \\ 
+Missingness as an attribute & -- & -- & +0.40 (0.11, 0.75) & -- \\ 
+Hot deck & +0.47 (0.03, 0.99) & +0.73 (0.22, 1.38) & +0.10 (-0.24, 0.44) & +0.51 (0.05, 0.99) \\ 
+K-nearest-neighbors & +0.57 (0.23, 0.97) & +0.77 (0.47, 1.18) & -0.09 (-0.44, 0.23) & +0.36 (0.09, 0.67) \\ 
+Predictive mean matching & +0.25 (-0.36, 0.82) & +0.71 (0.16, 1.36) & +0.09 (-0.33, 0.64) & +0.58 (0.06, 1.05) \\ 
+Random forest & +0.62 (0.08, 1.22) & +1.04 (0.39, 1.83) & +0.39 (-0.19, 0.77) & +0.69 (0.27, 1.12) \\ 
+Bayesian regression & +0.22 (-0.40, 0.82) & +0.68 (0.10, 1.28) & +0.10 (-0.44, 0.62) & +0.43 (-0.03, 0.93) \\ 
+\bottomrule
+\end{tabular} 
+ \end{table}
+
+\clearpage
+
+\begin{table} 
+ \caption{Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+Table values show the scaled Brier score for imputation to the mean.
+For other imputation strategies, table values show the change in scaled Brier score relative to the scaled Brier score when imputation to the mean was applied.
+All table values are multiplied by 100 for ease of interpretability}
+\label{tbl_md_strat_txpl_ipa}
+
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{Proportional hazards} & \multicolumn{2}{c}{Gradient boosted decision trees} \\ 
+ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+Imputation method & Single Imputation & Multiple Imputation & Single Imputation & Multiple Imputation \\ 
+\midrule
+\multicolumn{1}{l}{No additional missing data} \\ 
+\midrule
+Imputation to the mean & 9.35 (8.80, 9.71) & -- & 9.00 (8.55, 9.41) & -- \\ 
+Missingness as an attribute & -- & -- & -0.07 (-0.28, 0.13) & -- \\ 
+Hot deck & +0.01 (-0.12, 0.15) & +0.06 (-0.04, 0.23) & -0.24 (-0.43, 0.04) & -0.01 (-0.20, 0.17) \\ 
+K-nearest-neighbors & +0.06 (-0.04, 0.17) & +0.08 (0.00, 0.19) & -0.27 (-0.50, -0.05) & -0.05 (-0.20, 0.10) \\ 
+Predictive mean matching & +0.04 (-0.12, 0.19) & +0.14 (0.02, 0.29) & -0.34 (-0.60, -0.10) & -0.06 (-0.23, 0.15) \\ 
+Random forest & +0.03 (-0.10, 0.19) & +0.13 (0.03, 0.29) & -0.32 (-0.50, -0.07) & -0.05 (-0.21, 0.14) \\ 
+Bayesian regression & +0.03 (-0.13, 0.16) & +0.10 (0.00, 0.26) & -0.32 (-0.58, -0.05) & +0.02 (-0.13, 0.25) \\ 
+\midrule
+\multicolumn{1}{l}{+15\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 8.76 (8.20, 9.21) & -- & 8.01 (7.57, 8.54) & -- \\ 
+Missingness as an attribute & -- & -- & +0.50 (0.27, 0.78) & -- \\ 
+Hot deck & +0.25 (-0.04, 0.55) & +0.42 (0.20, 0.78) & -0.10 (-0.34, 0.23) & +0.21 (-0.02, 0.52) \\ 
+K-nearest-neighbors & +0.46 (0.20, 0.73) & +0.54 (0.28, 0.79) & -0.10 (-0.41, 0.20) & +0.12 (-0.12, 0.38) \\ 
+Predictive mean matching & +0.45 (0.10, 0.76) & +0.65 (0.34, 0.94) & -0.03 (-0.34, 0.21) & +0.25 (0.02, 0.62) \\ 
+Random forest & +0.44 (0.12, 0.76) & +0.60 (0.32, 0.95) & -0.07 (-0.35, 0.26) & +0.16 (-0.09, 0.47) \\ 
+Bayesian regression & +0.53 (0.21, 0.79) & +0.74 (0.37, 0.94) & -0.07 (-0.39, 0.24) & +0.24 (-0.09, 0.56) \\ 
+\midrule
+\multicolumn{1}{l}{+30\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 7.96 (7.26, 8.63) & -- & 7.09 (6.45, 7.68) & -- \\ 
+Missingness as an attribute & -- & -- & +0.83 (0.55, 1.10) & -- \\ 
+Hot deck & +0.45 (-0.02, 0.90) & +0.60 (0.15, 1.11) & -0.01 (-0.34, 0.41) & +0.19 (-0.17, 0.65) \\ 
+K-nearest-neighbors & +0.42 (0.00, 0.79) & +0.61 (0.18, 0.90) & -0.21 (-0.54, 0.19) & +0.09 (-0.22, 0.46) \\ 
+Predictive mean matching & +0.70 (0.21, 1.21) & +0.90 (0.46, 1.46) & +0.09 (-0.32, 0.54) & +0.34 (-0.04, 0.85) \\ 
+Random forest & +0.70 (0.14, 1.27) & +0.93 (0.49, 1.45) & -0.13 (-0.46, 0.43) & +0.25 (-0.12, 0.66) \\ 
+Bayesian regression & +0.77 (0.32, 1.40) & +1.00 (0.44, 1.53) & 0.00 (-0.45, 0.52) & +0.24 (-0.17, 0.80) \\ 
+\bottomrule
+\end{tabular} 
+ \end{table}
+
+\clearpage
+
+\begin{table} 
+ \caption{Median (25th, 75th percentile) change in concordance index when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+Table values show the concordance index for imputation to the mean.
+For other imputation strategies, table values show the change in concordance index relative to the concordance index when imputation to the mean was applied.
+All table values are multiplied by 100 for ease of interpretability}
+\label{tbl_md_strat_dead_auc}
+
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{Proportional hazards} & \multicolumn{2}{c}{Gradient boosted decision trees} \\ 
+ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+Imputation method & Single Imputation & Multiple Imputation & Single Imputation & Multiple Imputation \\ 
+\midrule
+\multicolumn{1}{l}{No additional missing data} \\ 
+\midrule
+Imputation to the mean & 69.3 (68.9, 69.8) & -- & 70.7 (70.3, 71.0) & -- \\ 
+Missingness as an attribute & -- & -- & 0.00 (-0.20, 0.19) & -- \\ 
+Hot deck & -0.20 (-0.44, 0.05) & +0.04 (-0.13, 0.27) & -0.19 (-0.36, 0.00) & +0.13 (-0.04, 0.28) \\ 
+K-nearest-neighbors & +0.03 (-0.10, 0.23) & +0.14 (-0.02, 0.36) & -0.10 (-0.30, 0.13) & +0.14 (-0.03, 0.30) \\ 
+Predictive mean matching & -0.34 (-0.56, -0.03) & +0.11 (-0.05, 0.37) & -0.21 (-0.44, 0.05) & +0.21 (0.06, 0.39) \\ 
+Random forest & +0.04 (-0.24, 0.29) & +0.30 (0.08, 0.52) & -0.02 (-0.26, 0.18) & +0.24 (0.07, 0.41) \\ 
+Bayesian regression & -0.39 (-0.69, -0.10) & +0.09 (-0.16, 0.28) & -0.21 (-0.44, -0.02) & +0.22 (0.06, 0.41) \\ 
+\midrule
+\multicolumn{1}{l}{+15\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 69.2 (68.8, 69.7) & -- & 70.5 (70.1, 71.0) & -- \\ 
+Missingness as an attribute & -- & -- & +0.10 (-0.11, 0.31) & -- \\ 
+Hot deck & -0.18 (-0.52, 0.19) & +0.25 (-0.07, 0.58) & -0.23 (-0.50, 0.03) & +0.12 (-0.07, 0.32) \\ 
+K-nearest-neighbors & +0.16 (-0.17, 0.42) & +0.30 (0.05, 0.56) & -0.16 (-0.33, 0.10) & +0.10 (-0.07, 0.31) \\ 
+Predictive mean matching & -0.42 (-0.83, 0.02) & +0.24 (-0.07, 0.53) & -0.26 (-0.54, 0.00) & +0.26 (0.06, 0.43) \\ 
+Random forest & -0.04 (-0.39, 0.26) & +0.41 (0.05, 0.64) & -0.08 (-0.29, 0.19) & +0.27 (0.04, 0.43) \\ 
+Bayesian regression & -0.50 (-0.95, -0.06) & +0.20 (-0.11, 0.47) & -0.29 (-0.48, -0.03) & +0.23 (0.07, 0.43) \\ 
+\midrule
+\multicolumn{1}{l}{+30\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 68.9 (68.4, 69.4) & -- & 70.2 (69.8, 70.7) & -- \\ 
+Missingness as an attribute & -- & -- & +0.12 (-0.08, 0.31) & -- \\ 
+Hot deck & -0.15 (-0.52, 0.30) & +0.12 (-0.23, 0.51) & -0.22 (-0.47, 0.02) & +0.10 (-0.17, 0.32) \\ 
+K-nearest-neighbors & +0.17 (-0.11, 0.49) & +0.33 (0.09, 0.65) & -0.08 (-0.27, 0.16) & +0.19 (-0.01, 0.38) \\ 
+Predictive mean matching & -0.75 (-1.15, -0.34) & +0.06 (-0.29, 0.52) & -0.36 (-0.67, -0.04) & +0.27 (0.04, 0.48) \\ 
+Random forest & -0.29 (-0.80, 0.23) & +0.26 (-0.10, 0.64) & -0.16 (-0.47, 0.11) & +0.29 (0.07, 0.52) \\ 
+Bayesian regression & -0.75 (-1.28, -0.22) & +0.12 (-0.29, 0.48) & -0.36 (-0.70, -0.07) & +0.19 (-0.01, 0.44) \\ 
+\bottomrule
+\end{tabular} 
+ \end{table}
+
+\clearpage
+
+\begin{table} 
+ \caption{Median (25th, 75th percentile) change in calibration error when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+Table values show the calibration error for imputation to the mean.
+For other imputation strategies, table values show the change in calibration error relative to the calibration error when imputation to the mean was applied.
+All table values are multiplied by 100 for ease of interpretability}
+\label{tbl_md_strat_dead_cal_error}
+
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{Proportional hazards} & \multicolumn{2}{c}{Gradient boosted decision trees} \\ 
+ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+Imputation method & Single Imputation & Multiple Imputation & Single Imputation & Multiple Imputation \\ 
+\midrule
+\multicolumn{1}{l}{No additional missing data} \\ 
+\midrule
+Imputation to the mean & 2.41 (1.71, 3.44) & -- & 3.29 (2.24, 4.80) & -- \\ 
+Missingness as an attribute & -- & -- & -0.09 (-1.07, 0.65) & -- \\ 
+Hot deck & -0.04 (-0.68, 0.58) & -0.45 (-1.25, 0.25) & -0.13 (-1.12, 0.65) & -0.90 (-1.98, 0.01) \\ 
+K-nearest-neighbors & +0.04 (-0.59, 0.65) & -0.29 (-0.80, 0.39) & -0.30 (-1.14, 0.61) & -0.75 (-1.59, 0.14) \\ 
+Predictive mean matching & -0.18 (-0.69, 0.47) & -0.69 (-1.36, -0.03) & -0.11 (-1.27, 0.89) & -1.02 (-2.08, -0.12) \\ 
+Random forest & -0.07 (-0.58, 0.65) & -0.46 (-1.10, 0.17) & -0.56 (-1.38, 0.46) & -1.10 (-2.08, -0.14) \\ 
+Bayesian regression & -0.08 (-0.75, 0.53) & -0.57 (-1.43, 0.02) & -0.26 (-1.08, 0.57) & -0.85 (-1.91, 0.12) \\ 
+\midrule
+\multicolumn{1}{l}{+15\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 5.75 (3.74, 8.59) & -- & 6.28 (4.17, 9.45) & -- \\ 
+Missingness as an attribute & -- & -- & -2.83 (-4.81, -0.83) & -- \\ 
+Hot deck & -2.16 (-4.31, -1.08) & -3.92 (-6.23, -1.72) & -1.66 (-3.82, -0.34) & -3.44 (-5.97, -1.48) \\ 
+K-nearest-neighbors & -2.01 (-3.51, -0.66) & -2.48 (-4.08, -1.14) & -0.61 (-2.28, 0.84) & -2.04 (-3.32, -0.63) \\ 
+Predictive mean matching & -3.83 (-6.40, -1.87) & -3.76 (-6.66, -1.04) & -3.52 (-6.43, -1.67) & -3.50 (-7.09, -1.00) \\ 
+Random forest & -3.80 (-5.75, -2.07) & -4.35 (-6.60, -1.53) & -3.74 (-6.34, -1.34) & -3.92 (-6.29, -1.50) \\ 
+Bayesian regression & -3.76 (-6.45, -1.81) & -3.80 (-6.61, -1.04) & -3.79 (-6.80, -1.39) & -3.27 (-6.40, -0.46) \\ 
+\midrule
+\multicolumn{1}{l}{+30\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 9.48 (5.50, 15.3) & -- & 7.02 (4.32, 11.3) & -- \\ 
+Missingness as an attribute & -- & -- & -1.61 (-5.29, 0.68) & -- \\ 
+Hot deck & -4.95 (-9.18, -2.49) & -6.69 (-12.8, -1.56) & -2.48 (-5.46, -0.31) & -2.48 (-7.18, 0.58) \\ 
+K-nearest-neighbors & -3.47 (-6.42, -1.53) & -4.49 (-7.49, -2.39) & +0.73 (-1.56, 3.27) & -1.47 (-3.59, 0.59) \\ 
+Predictive mean matching & -7.33 (-13.0, -2.61) & -6.09 (-12.3, -0.22) & -2.58 (-7.90, 0.10) & -1.30 (-6.86, 2.29) \\ 
+Random forest & -7.81 (-13.1, -3.63) & -7.39 (-13.4, -2.37) & -3.53 (-7.87, -0.29) & -2.58 (-7.27, 0.44) \\ 
+Bayesian regression & -7.27 (-13.2, -2.44) & -5.91 (-12.8, -0.47) & -3.00 (-7.60, 0.97) & -0.54 (-6.60, 3.08) \\ 
+\bottomrule
+\end{tabular} 
+ \end{table}
+
+\clearpage
+
+\begin{table} 
+ \caption{Median (25th, 75th percentile) change in concordance index when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+Table values show the concordance index for imputation to the mean.
+For other imputation strategies, table values show the change in concordance index relative to the concordance index when imputation to the mean was applied.
+All table values are multiplied by 100 for ease of interpretability}
+\label{tbl_md_strat_txpl_auc}
+
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{Proportional hazards} & \multicolumn{2}{c}{Gradient boosted decision trees} \\ 
+ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+Imputation method & Single Imputation & Multiple Imputation & Single Imputation & Multiple Imputation \\ 
+\midrule
+\multicolumn{1}{l}{No additional missing data} \\ 
+\midrule
+Imputation to the mean & 79.4 (79.0, 79.8) & -- & 79.2 (78.9, 79.7) & -- \\ 
+Missingness as an attribute & -- & -- & -0.03 (-0.22, 0.14) & -- \\ 
+Hot deck & +0.01 (-0.12, 0.17) & +0.07 (-0.05, 0.23) & -0.24 (-0.40, -0.02) & -0.04 (-0.20, 0.15) \\ 
+K-nearest-neighbors & +0.08 (-0.02, 0.16) & +0.08 (0.01, 0.17) & -0.21 (-0.40, -0.04) & -0.04 (-0.18, 0.10) \\ 
+Predictive mean matching & +0.03 (-0.11, 0.16) & +0.13 (0.01, 0.24) & -0.34 (-0.57, -0.12) & -0.06 (-0.19, 0.16) \\ 
+Random forest & +0.03 (-0.09, 0.17) & +0.15 (0.06, 0.24) & -0.26 (-0.46, -0.08) & -0.02 (-0.16, 0.14) \\ 
+Bayesian regression & +0.01 (-0.15, 0.16) & +0.10 (0.01, 0.22) & -0.32 (-0.54, -0.10) & +0.02 (-0.13, 0.22) \\ 
+\midrule
+\multicolumn{1}{l}{+15\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 78.9 (78.5, 79.4) & -- & 78.4 (77.9, 78.9) & -- \\ 
+Missingness as an attribute & -- & -- & +0.47 (0.26, 0.71) & -- \\ 
+Hot deck & +0.24 (0.04, 0.51) & +0.47 (0.26, 0.76) & -0.13 (-0.40, 0.22) & +0.29 (0.07, 0.58) \\ 
+K-nearest-neighbors & +0.36 (0.21, 0.54) & +0.46 (0.29, 0.62) & -0.13 (-0.39, 0.17) & +0.17 (-0.06, 0.39) \\ 
+Predictive mean matching & +0.32 (0.11, 0.58) & +0.56 (0.37, 0.82) & -0.12 (-0.38, 0.25) & +0.32 (0.14, 0.61) \\ 
+Random forest & +0.32 (0.13, 0.64) & +0.59 (0.36, 0.82) & -0.11 (-0.40, 0.24) & +0.22 (-0.03, 0.51) \\ 
+Bayesian regression & +0.39 (0.17, 0.63) & +0.60 (0.40, 0.83) & -0.15 (-0.46, 0.27) & +0.27 (-0.01, 0.64) \\ 
+\midrule
+\multicolumn{1}{l}{+30\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 78.4 (77.7, 78.8) & -- & 77.4 (76.8, 78.0) & -- \\ 
+Missingness as an attribute & -- & -- & +0.91 (0.52, 1.24) & -- \\ 
+Hot deck & +0.34 (-0.03, 0.72) & +0.74 (0.48, 1.11) & -0.19 (-0.53, 0.34) & +0.55 (0.19, 0.86) \\ 
+K-nearest-neighbors & +0.28 (-0.02, 0.47) & +0.43 (0.21, 0.66) & -0.32 (-0.78, -0.04) & -0.01 (-0.33, 0.31) \\ 
+Predictive mean matching & +0.62 (0.29, 0.98) & +1.02 (0.74, 1.34) & +0.03 (-0.31, 0.44) & +0.65 (0.29, 0.95) \\ 
+Random forest & +0.59 (0.22, 1.00) & +1.04 (0.76, 1.30) & -0.21 (-0.58, 0.15) & +0.36 (0.01, 0.68) \\ 
+Bayesian regression & +0.81 (0.41, 1.08) & +1.10 (0.80, 1.39) & -0.04 (-0.54, 0.37) & +0.47 (0.10, 0.91) \\ 
+\bottomrule
+\end{tabular} 
+ \end{table}
+
+\clearpage
+
+\begin{table} 
+ \caption{Median (25th, 75th percentile) change in calibration error when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+Table values show the calibration error for imputation to the mean.
+For other imputation strategies, table values show the change in calibration error relative to the calibration error when imputation to the mean was applied.
+All table values are multiplied by 100 for ease of interpretability}
+\label{tbl_md_strat_txpl_cal_error}
+
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{Proportional hazards} & \multicolumn{2}{c}{Gradient boosted decision trees} \\ 
+ \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+Imputation method & Single Imputation & Multiple Imputation & Single Imputation & Multiple Imputation \\ 
+\midrule
+\multicolumn{1}{l}{No additional missing data} \\ 
+\midrule
+Imputation to the mean & 1.36 (0.92, 2.07) & -- & 1.79 (1.34, 2.50) & -- \\ 
+Missingness as an attribute & -- & -- & +0.01 (-0.53, 0.54) & -- \\ 
+Hot deck & -0.01 (-0.37, 0.29) & 0.00 (-0.39, 0.26) & +0.11 (-0.56, 0.68) & -0.07 (-0.67, 0.67) \\ 
+K-nearest-neighbors & +0.00 (-0.25, 0.24) & -0.02 (-0.25, 0.21) & +0.09 (-0.40, 0.61) & -0.03 (-0.52, 0.67) \\ 
+Predictive mean matching & -0.04 (-0.34, 0.32) & -0.08 (-0.41, 0.26) & +0.09 (-0.49, 0.63) & -0.04 (-0.63, 0.70) \\ 
+Random forest & -0.06 (-0.39, 0.29) & -0.06 (-0.35, 0.29) & -0.02 (-0.49, 0.63) & +0.02 (-0.54, 0.56) \\ 
+Bayesian regression & -0.01 (-0.40, 0.27) & -0.13 (-0.38, 0.18) & +0.12 (-0.46, 0.78) & +0.24 (-0.46, 1.04) \\ 
+\midrule
+\multicolumn{1}{l}{+15\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 2.48 (1.79, 3.93) & -- & 2.51 (1.81, 3.29) & -- \\ 
+Missingness as an attribute & -- & -- & -0.50 (-1.38, 0.16) & -- \\ 
+Hot deck & -0.76 (-1.95, 0.21) & -0.51 (-2.14, 0.89) & -0.01 (-1.00, 0.66) & +0.69 (-0.47, 1.97) \\ 
+K-nearest-neighbors & -0.82 (-2.01, 0.21) & -0.78 (-1.96, 0.26) & +0.20 (-0.59, 1.01) & +0.23 (-0.73, 1.02) \\ 
+Predictive mean matching & -1.01 (-2.15, 0.27) & -0.46 (-2.16, 0.94) & +0.15 (-0.91, 1.02) & +0.72 (-0.81, 1.87) \\ 
+Random forest & -0.80 (-2.32, 0.51) & -0.59 (-2.03, 0.81) & -0.10 (-1.15, 0.66) & +0.42 (-0.93, 1.34) \\ 
+Bayesian regression & -0.94 (-2.32, 0.58) & -0.64 (-2.04, 0.89) & +0.19 (-0.76, 1.16) & +0.99 (-0.50, 2.27) \\ 
+\midrule
+\multicolumn{1}{l}{+30\% additional missing data} \\ 
+\midrule
+Imputation to the mean & 4.98 (3.42, 7.00) & -- & 3.99 (2.74, 6.22) & -- \\ 
+Missingness as an attribute & -- & -- & -1.52 (-2.62, -0.35) & -- \\ 
+Hot deck & -2.17 (-4.65, 0.10) & -0.49 (-3.74, 2.34) & -0.53 (-2.79, 1.05) & +1.89 (-0.93, 4.76) \\ 
+K-nearest-neighbors & -1.88 (-3.34, -0.22) & -1.88 (-3.70, -0.14) & -0.41 (-1.72, 0.68) & -0.54 (-2.26, 0.88) \\ 
+Predictive mean matching & -1.30 (-4.52, 1.28) & +0.04 (-3.48, 2.70) & -0.05 (-2.54, 2.53) & +1.36 (-1.26, 4.01) \\ 
+Random forest & -1.80 (-4.34, 0.79) & -0.58 (-3.95, 2.18) & -0.55 (-2.64, 1.53) & +0.33 (-2.27, 2.77) \\ 
+Bayesian regression & -1.34 (-4.22, 1.59) & -0.23 (-3.15, 2.61) & -0.16 (-2.47, 1.94) & +0.89 (-1.89, 3.67) \\ 
+\bottomrule
+\end{tabular} 
+ \end{table}
+
+\clearpage
+
+\begin{figure}
+
+{\centering \includegraphics{doc_arxiv_files/figure-latex/upset-1}
+
+}
+
+\caption{An upset plot showing three variables from the INTERMACS registry and all combinations of missing patterns.
+The bottom left plot shows the number of missing values for each variable, separately.
+The top right plot shows the number of missing values for each combination of the three variables.
+For example, there were 2,618 rows in the overall INTERMACS data where both CV pressure and surgery time were missing.}\label{fig:upset}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+
+{\centering \includegraphics{doc_arxiv_files/figure-latex/fig_md_strat_infer_ipa-1}
+
+}
+
+\caption{Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
+Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution.}\label{fig:fig_md_strat_infer_ipa}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+
+{\centering \includegraphics{doc_arxiv_files/figure-latex/fig_md_strat_infer_auc-1}
+
+}
+
+\caption{Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
+Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution.}\label{fig:fig_md_strat_infer_auc}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+
+{\centering \includegraphics{doc_arxiv_files/figure-latex/fig_md_strat_infer_cal_error-1}
+
+}
+
+\caption{Posterior distribution of differences in calibration error values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
+Posterior probability that the difference in calibration error exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution.}\label{fig:fig_md_strat_infer_cal_error}
+\end{figure}
+
+\clearpage
+
 \bibliographystyle{unsrt}

### 2020-11-01 doc_arXiv/what_is_known.Rmd
all tables and figs added

---
date: "`r Sys.Date()`"
author: "Your Name"
title: "officedown template"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
---

<!-- As a benefit to our readers, please provide concise and insightful sentences that identify "What is Known" about your topic and "What the Study Adds" to the literature. These sentences should be limited to 5 inclusively. This should NOT be a summary, but referenced in a bulleted format first listing the "What is Known" followed by the "What the Study Adds". Please do not include references. Please upload a Word document version of your points as a supplemental file; it should not be included in the main manuscript document.  -->

**What is Known**

- add a sentence

**What the Study Adds**

- add a sentence

- add a sentence

### 2020-11-01 R/as_gt.R

@@ -1,0 +1,81 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param tbl_md_strat
+as_gt <- function(tbl_md_strat,
+                  model_labels,
+                  additional_missing_labels,
+                  md_method_labels,
+                  rspec) {
+
+  gt_md_strat <- tbl_md_strat %>%
+    map(
+      ~ {
+        gt_data <- .x %>%
+          mutate(
+            model = recode(model, !!!model_labels),
+            additional_missing_pct = factor(
+              additional_missing_pct,
+              levels = additional_missing_labels,
+              labels = names(additional_missing_labels)
+            ),
+            md_strat = str_replace(md_strat, 'mia', 'mia_si'),
+            table_val = if_else(
+              md_strat == 'meanmode_si',
+              true = table_glue("{est} ({lwr}, {upr})", rspec=rspec),
+              false = table_glue("{pm(est)}{est} ({lwr}, {upr})", rspec=rspec)
+            )
+          ) %>%
+          select(-est, -lwr, -upr) %>%
+          separate(md_strat, into = c('md_method', 'md_type')) %>%
+          pivot_wider(names_from = md_type, values_from = table_val) %>%
+          mutate(
+            md_method = factor(md_method,
+                               levels = names(md_method_labels),
+                               labels = md_method_labels)
+          ) %>%
+          arrange(md_method) %>%
+          filter(!(md_method == 'Missingness as an attribute' &
+                     model == 'Proportional hazards')) %>%
+          rename(MI = mi, SI = si) %>%
+          pivot_wider(names_from = model, values_from = c(SI, MI)) %>%
+          select(
+            additional_missing_pct,
+            md_method,
+            `SI_Proportional hazards`,
+            `MI_Proportional hazards`,
+            `SI_Gradient boosted decision trees`,
+            `MI_Gradient boosted decision trees`
+          )
+
+        cols <- str_detect(names(gt_data), pattern = 'MI$|SI$')
+        cols <- names(gt_data)[cols]
+
+        gt(gt_data,
+           groupname_col = 'additional_missing_pct',
+           rowname_col = 'md_method') %>%
+          tab_stubhead(label = 'Imputation method') %>%
+          fmt_missing(columns = everything(),
+                      missing_text = '--') %>%
+          tab_spanner(label = 'Proportional hazards',
+                      columns = c("SI_Proportional hazards",
[…]

### 2020-11-01 R/gt_latex.R

@@ -1,0 +1,39 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param gt_object
+##' @param caption
+##' @param label
+gt_latex <- function(gt_object, caption, label) {
+
+    header = paste0(
+      '\\begin{table} \n \\caption{',caption,'} \n'
+    )
+
+    if(!is.null(label)){
+      header = paste0(header, '\\label{', label, '} \n')
+    }
+
+    gt_object %>%
+      as_latex() %>%
+      as.character() %>%
+      gsub("*", "", ., fixed = T) %>%
+      gsub("\\\\ \n\\small \\\\ \n", "", ., fixed = T) %>%
+      gsub("\n\\large", "", ., fixed = T) %>%
+      gsub("\\captionsetup[table]{labelformat=empty,skip=1pt}",
+           "",
+           .,
+           fixed = T) %>%
+      gsub("longtable", "table", ., fixed = T) %>%
+      gsub('\\begin{table}',
+           '\\begin{table} \n\\begin{tabular}',
+           .,
+           fixed = T) %>%
+      gsub('\\end{table}', '\\end{tabular} \n \\end{table}', ., fixed = T) %>%
+      gsub('\\begin{table}', header, ., fixed = T) %>%
+      cat()
+
+
+}

### 2020-11-01 R/load_mccv.R

@@ -8,7 +8,7 @@
 ##'
 ##' @title
 
-load_mccv <- function(output_path = "slurm/results_old",
+load_mccv <- function(output_path = "slurm/results",
                       output_pattern = '^output',
                       nruns = 12000) {
 

### 2020-11-01 R/make_bayes_mccv_fit.R

@@ -6,19 +6,34 @@
 ##' @param risk_evaluation
 make_bayes_mccv_fit <- function(risk_evaluation) {
 
-  risk_evaluation %>%
+  model_data <- risk_evaluation %>%
     mutate(md_strat = fct_relevel(md_strat, 'meanmode_si')) %>%
     select(-bri) %>%
-    pivot_longer(cols = c(auc, ipa), names_to = 'metric') %>%
-    split(f = list(.$outcome,
-                   .$model,
-                   .$metric,
-                   .$additional_missing_pct)) %>%
-    map(
-      ~ stan_lmer(
-        formula = value ~ md_strat + (1 | iteration),
-        data = .x
+    pivot_longer(cols = c(auc, ipa, cal_error), names_to = 'metric') %>%
+    split(
+      f = list(
+        .$outcome,
+        .$model,
+        .$metric
+        #.$additional_missing_pct
       )
     )
+
+  output <- vector(mode = 'list', length = length(model_data))
+  names(output) <- names(model_data)
+
+  for(i in seq_along(output)){
+
+    output[[i]] <- stan_lm(
+      formula = value ~ md_strat + additional_missing_pct,
+      #+ (1 | iteration),
+      data = model_data[[i]],
+      prior = R2(0.50),
+      iter = 5000
+    )
+
+  }
+
+  output
 
 }

### 2020-11-01 R/make_risk_evaluation.R

@@ -54,6 +54,7 @@ make_risk_evaluation <- function(mc_cv_results, im, resamples, times) {
 
   }
 
-  bind_rows(risk_eval, .id = 'iteration')
+  bind_rows(risk_eval, .id = 'iteration') %>%
+    mutate(cal_error = cal_error * 100)
 
 }

### 2020-11-01 R/plan.R

@@ -50,8 +50,8 @@ the_plan <- drake_plan(
   # creates intermacs_clean.csv
   im = clean_intermacs(),
 
-  # tabulate characteristics
-  tbl_characteristics = tabulate_characteristics(im),
+  # tabulate descriptives
+  tbl_descriptives = tabulate_descriptives(im),
 
   # creates resamples.csv
   resamples = mc_cv_light(data = im, train_prop = 1/2, ntimes = 200),
@@ -63,37 +63,42 @@ the_plan <- drake_plan(
 
   risk_evaluation = make_risk_evaluation(mc_cv_results, im, resamples, times),
 
-  # bayes_mccv_fits = make_bayes_mccv_fit(risk_evaluation),
-  #
+  bayes_mccv_fits = make_bayes_mccv_fit(risk_evaluation),
+
   fig_risk_evaluation = visualize_risk_evaluation(risk_evaluation,
                                                   md_method_labels,
                                                   md_type_labels,
                                                   model_labels,
                                                   outcome_labels,
                                                   additional_missing_labels,
-                                                  times)
-  #
-  # fig_md_strat_infer = visualize_md_strat_inference(bayes_mccv_fits,
-  #                                                   md_type_labels,
-  #                                                   md_method_labels,
-  #                                                   outcome_labels,
-  #                                                   rspec),
-  #
-  #
-  # tbl_md_strat = tabulate_md_strat(risk_evaluation,
-  #                                  md_method_labels,
-  #                                  md_type_labels,
-  #                                  model_labels,
-  #                                  outcome_labels,
-  #                                  additional_missing_labels,
-  #                                  rspec),
-  #
-  # arxiv_preprint = target(
-  #   command = {
-  #     rmarkdown::render(knitr_in("doc_arxiv/doc_arxiv.Rmd"))
-  #     file_out("doc_arxiv/doc_arxiv.pdf")
-  #   }
-  # )
+                                                  times),
+
+  fig_md_strat_infer = visualize_md_strat_inference(bayes_mccv_fits,
+                                                    md_type_labels,
+                                                    md_method_labels,
+                                                    outcome_labels,
+                                                    rspec),
+
+  tbl_md_strat = tabulate_md_strat(risk_evaluation,
+                                   md_method_labels,
+                                   md_type_labels,
+                                   model_labels,
+                                   outcome_labels,
[…]

### 2020-11-01 R/pm.R

@@ -1,0 +1,11 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param x
+pm <- function(x) {
+
+  ifelse(x > 0, "+", "")
+
+}

### 2020-11-01 R/tabulate_characteristics.R

@@ -1,82 +1,0 @@
-##' .. content for \description{} (no empty lines) ..
-##'
-##' .. content for \details{} ..
-##'
-##' @title
-##' @param im
-tabulate_characteristics <- function(im) {
-
-  tbl <- im %>%
-    mutate(
-      status = case_when(
-        pt_outcome_dead == 1 ~ "Dead",
-        pt_outcome_txpl == 1 ~ "Transplant",
-        pt_outcome_cess == 1 ~ "Cessation of support",
-        TRUE ~ "Censored"
-      ),
-      demo_race = case_when(
-        demo_race_af_amer == 'yes' ~ "Black",
-        demo_race_white == 'yes' ~ "White",
-        TRUE ~ "Other"
-      ),
-      demo_gender = fct_recode(
-        demo_gender,
-        Male = 'male',
-        Female = 'female',
-      ),
-      demo_race = fct_infreq(demo_race),
-      pi_hgt_m = pi_hgt_cm / 100,
-      pi_bmi = pi_wgt_kg / pi_hgt_m^2,
-      pi_device_strategy = fct_recode(
-        pi_device_strategy,
-        'Destination therapy'  = "destination_therapy_patient_definitely_not_eligible_for_transplant",
-        'Bridge to transplant' = "bridge_to_transplant_patient_currently_listed_for_transplant",
-        'Bridge to transplant' = "possible_bridge_to_transplant_likely_to_be_eligible",
-        'Bridge to transplant' = "possible_bridge_to_transplant_moderate_likelihood_of_becoming_eligible",
-        'Bridge to transplant' = "possible_bridge_to_transplant_unlikely_to_become_eligible",
-        'Other' = "bridge_to_recovery",
-        'Other' = 'rescue_therapy',
-        'Other' = 'other_specify'
-      ),
-      im_device_ty = fct_recode(
-        im_device_ty,
-        'Left-ventricular assistance device' = 'lvad',
-        'Bi-ventricular assistance device'   = 'bivad'
-      )
-    ) %>%
-    select(
-      status,
-      months_post_implant,
-      demo_age,
-      demo_gender,
-      demo_race,
-      pi_bmi,
-      pi_device_strategy,
-      im_device_ty
-    ) %>%
-    mutate(`No. of patients` = 1, .before = 1) %>%
-    tbl_summary(
-      by = status,
-      label = list(
-        months_post_implant ~ "Follow up time, months",
[…]

### 2020-11-01 R/tabulate_descriptives.R

@@ -1,0 +1,162 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param im
+tabulate_descriptives <- function(im) {
+
+  tbl_data <- im %>%
+    mutate(
+      status = case_when(
+        pt_outcome_dead == 1 ~ "Dead",
+        pt_outcome_txpl == 1 ~ "Transplant",
+        pt_outcome_cess == 1 ~ "Cessation of support",
+        TRUE ~ "Censored"
+      ),
+      demo_race = case_when(
+        demo_race_af_amer == 'yes' ~ "Black",
+        demo_race_white == 'yes' ~ "White",
+        TRUE ~ "Other"
+      ),
+      demo_gender = fct_recode(
+        demo_gender,
+        Male = 'male',
+        Female = 'female',
+      ),
+      demo_race = fct_infreq(demo_race),
+      pi_hgt_m = pi_hgt_cm / 100,
+      pi_bmi = pi_wgt_kg / pi_hgt_m^2,
+      pi_device_strategy = fct_recode(
+        pi_device_strategy,
+        'Destination therapy'  = "destination_therapy_patient_definitely_not_eligible_for_transplant",
+        'Bridge to transplant' = "bridge_to_transplant_patient_currently_listed_for_transplant",
+        'Bridge to transplant' = "possible_bridge_to_transplant_likely_to_be_eligible",
+        'Bridge to transplant' = "possible_bridge_to_transplant_moderate_likelihood_of_becoming_eligible",
+        'Bridge to transplant' = "possible_bridge_to_transplant_unlikely_to_become_eligible",
+        'Other' = "bridge_to_recovery",
+        'Other' = 'rescue_therapy',
+        'Other' = 'other_specify'
+      ),
+      im_device_ty = fct_recode(
+        im_device_ty,
+        'Left-ventricular assistance device' = 'lvad',
+        'Bi-ventricular assistance device'   = 'bivad'
+      )
+    ) %>%
+    select(
+      status,
+      months_post_implant,
+      demo_age,
+      demo_gender,
+      demo_race,
+      pi_bmi,
+      pi_device_strategy,
+      pi_lvedd,
+      pi_albumin_g_dl,
+      pi_creat_mg_dl,
+      pi_cv_pres,
+      pi_peripheral_edema,
[…]

### 2020-11-01 R/tabulate_impute_accuracy.R

@@ -33,11 +33,13 @@ tabulate_impute_accuracy <- function(mc_cv_results) {
     )
 
   data_tbl %>%
-    transmute(
-      md_strat,
-      additional_missing_pct,
-      nominal = table_glue("{nominal_est} ({nominal_lwr}, {nominal_upr})"),
-      numeric = table_glue("{numeric_est} ({numeric_lwr}, {numeric_upr})")
+    mutate(
+      nominal = table_glue(
+        "{pm(nominal_est)}{nominal_est} ({nominal_lwr}, {nominal_upr})"
+      ),
+      numeric = table_glue(
+        "{pm(numeric_est)}{numeric_est} ({numeric_lwr}, {numeric_upr})"
+      )
     ) %>%
     mutate(
       across(
@@ -48,13 +50,15 @@ tabulate_impute_accuracy <- function(mc_cv_results) {
       )
     ) %>%
     separate(md_strat, into = c('md_method', 'md_type')) %>%
-    pivot_wider(values_from = c(nominal, numeric), names_from = md_type) %>%
+    pivot_wider(values_from = matches("^nominal|^numeric"),
+                names_from = md_type) %>%
     select(md_method,
            additional_missing_pct,
            nominal_si,
            nominal_mi,
            numeric_si,
-           numeric_mi)
+           numeric_mi,
+           everything())
 
 
 }

### 2020-11-01 R/tabulate_md_strat.R

@@ -12,18 +12,12 @@ tabulate_md_strat <- function(risk_evaluation,
                               additional_missing_labels,
                               rspec) {
 
-  pm <- function(x){ifelse(x>0, "+", "")}
-
   output <- risk_evaluation %>%
     select(-bri) %>%
-    pivot_longer(cols = c(auc, ipa), names_to = 'metric') %>%
+    pivot_longer(cols = c(auc, ipa, cal_error), names_to = 'metric') %>%
     split(f = list(.$outcome, .$metric)) %>%
     map(
-      ~ {
-
-        mult_by <- 100
-
-        data_tbl <- .x %>%
+      ~ .x %>%
           select(-outcome, -metric) %>%
           pivot_wider(values_from = value, names_from = md_strat) %>%
           mutate(
@@ -47,69 +41,11 @@ tabulate_md_strat <- function(risk_evaluation,
           pivot_longer(cols = -(iteration:model), names_to = 'md_strat') %>%
           group_by(model, md_strat, additional_missing_pct) %>%
           summarize(
-            est = mult_by * median(value, na.rm = TRUE),
-            lwr = mult_by * quantile(value, probs = 0.025, na.rm = TRUE),
-            upr = mult_by * quantile(value, probs = 0.975, na.rm = TRUE)
+            est = 100 * median(value, na.rm = TRUE),
+            lwr = 100 * quantile(value, probs = 0.25, na.rm = TRUE),
+            upr = 100 * quantile(value, probs = 0.75, na.rm = TRUE)
           ) %>%
-          ungroup() %>%
-          transmute(
-            model = recode(model, !!!model_labels),
-            additional_missing_pct = factor(
-              additional_missing_pct,
-              levels = additional_missing_labels,
-              labels = names(additional_missing_labels)
-            ),
-            md_strat = str_replace(md_strat, 'mia', 'mia_si'),
-            table_val = if_else(
-              md_strat == 'meanmode_si',
-              true = table_glue("{est} ({lwr}, {upr})", rspec=rspec),
-              false = table_glue("{pm(est)}{est} ({lwr}, {upr})", rspec=rspec)
-            )
-          ) %>%
-          separate(md_strat, into = c('md_method', 'md_type')) %>%
-          pivot_wider(names_from = md_type, values_from = table_val) %>%
-          mutate(
-            md_method = factor(md_method,
-                               levels = names(md_method_labels),
-                               labels = md_method_labels)
-          ) %>%
-          arrange(md_method) %>%
[…]

### 2020-11-01 R/tidy_score.R

@@ -6,12 +6,22 @@
 ##' @param scores
 tidy_score <- function(scores) {
 
+  cal_estm <- plotCalibration(scores,
+                              plot = FALSE,
+                              method = 'q',
+                              cens.method="local") %>%
+    use_series('plotFrame') %>%
+    map_dbl(~mean((.x$Pred - .x$Obs)^2)) %>%
+    enframe(name = 'md_strat_mdl', value = 'cal_error')
+
   auc_estm <- as_tibble(scores$AUC$score) %>%
     select(md_strat_mdl = model, auc = AUC)
 
   bri_estm <- as_tibble(scores$Brier$score) %>%
     select(md_strat_mdl = model, bri = Brier, ipa = IPA)
 
-  left_join(auc_estm, bri_estm, by = 'md_strat_mdl')
+  auc_estm %>%
+    left_join(bri_estm, by = 'md_strat_mdl') %>%
+    left_join(cal_estm, by = 'md_strat_mdl')
 
 }

### 2020-11-01 R/visualize_md_strat_inference.R

@@ -12,13 +12,41 @@ visualize_md_strat_inference <- function(bayes_mccv_fits,
 
   get_xlevels <- function(posterior) {
 
-    posterior$x@Dimnames[[2]] %>%
-      enframe() %>%
-      filter(str_detect(value, '^md_strat')) %>%
-      pull(value) %>%
-      str_remove('md_strat')
+    if(inherits(posterior, 'lm')){
+
+      posterior$xlevels$md_strat
+
+    } else {
+
+      posterior$x@Dimnames[[2]] %>%
+        enframe() %>%
+        filter(str_detect(value, '^md_strat')) %>%
+        pull(value) %>%
+        str_remove('md_strat')
+
+    }
 
   }
+
+
+  join_on <- expand.grid(
+    model = c('cph', 'xgb'),
+    md_strat = c(
+      "meanmode_si",
+      "bayesreg_mi",
+      "bayesreg_si",
+      "hotdeck_mi",
+      "hotdeck_si",
+      "mia",
+      "nbrs_mi",
+      "nbrs_si",
+      "pmm_mi",
+      "pmm_si",
+      "ranger_mi",
+      "ranger_si"
+    ),
+    additional_missing_pct = c(0, 15, 30)
+  )
 
   posterior_diffs <- map_dfr(
     .x = bayes_mccv_fits,
@@ -26,117 +54,181 @@ visualize_md_strat_inference <- function(bayes_mccv_fits,
     .f = function(posterior) {
       tibble(
         m1 = 'meanmode_si',
-        m2 = get_xlevels(posterior)
+        m2 = setdiff(get_xlevels(posterior), 'meanmode_si')
       ) %>%
         mutate(
           distribution = map2(
             .x = m1,
             .y = m2,
-            .f = ~ posterior_predict(
-              posterior,
-              newdata = data.frame(md_strat = c(.x, .y)),
-              re.form = NA
-            ) %>%
-              apply(1, diff)
+            .f = ~ {
+
+              newdata <- tibble(md_strat = c(.x, .y)) %>%
+                left_join(join_on, by = 'md_strat')
+
+              newdata$pred <- posterior_predict(
+                posterior,
+                newdata = newdata,
+                re.form = NA
+              ) %>%
+                apply(2, list) %>%
+                map(~.x[[1]])
+
+              newdata <- newdata %>%
+                group_by(md_strat) %>%
+                summarize(pred = list(reduce(pred, `+`) / n()),
[…]

### 2020-11-01 R/visualize_risk_evaluation.R

@@ -12,14 +12,14 @@ visualize_risk_evaluation <- function(risk_evaluation,
                                       additional_missing_labels,
                                       times) {
 
-
   rspec <- round_spec() %>%
     round_using_decimal(digits = 2) %>%
     round_half_even()
 
   ggdata <- risk_evaluation %>%
+    mutate(cal_error = cal_error * 1000) %>%
     group_by(outcome, md_strat, model, additional_missing_pct) %>%
-    summarize(across(c(auc,ipa), mean), .groups = 'drop') %>%
+    summarize(across(c(auc, ipa, cal_error), mean), .groups = 'drop') %>%
     mutate(md_strat = if_else(md_strat == 'mia',
                               true = 'mia_si',
                               false = md_strat)) %>%
@@ -27,12 +27,12 @@ visualize_risk_evaluation <- function(risk_evaluation,
     group_by(outcome, md_method, model) %>%
     mutate(
       across(
-        .cols = c(auc, ipa),
+        .cols = c(auc, ipa, cal_error),
         .fns  = list(pct_diff = ~ table_value(100*(max(.x)-min(.x))/min(.x),
                                               rspec = rspec))
       ),
       across(
-        .cols = c(auc, ipa),
+        .cols = c(auc, ipa, cal_error),
         .fns = list(lbl = ~table_value(100 * .x, rspec = rspec))
       )
     ) %>%
@@ -54,7 +54,7 @@ visualize_risk_evaluation <- function(risk_evaluation,
 
   lower_values <- ggdata %>%
     group_by(outcome, md_method, model, additional_missing_pct) %>%
-    select(-ipa) %>%
+    select(-ipa, -cal_error) %>%
     filter(auc == min(auc)) %>%
     ungroup() %>%
     filter(!(md_method %in% md_methods_upper_only)) %>%
@@ -62,7 +62,7 @@ visualize_risk_evaluation <- function(risk_evaluation,
 
   upper_values <- ggdata %>%
     group_by(outcome, md_method, model, additional_missing_pct) %>%
-    select(-ipa) %>%
+    select(-ipa, -cal_error) %>%
     filter(auc == max(auc)) %>%
     group_by(model) %>%
     nest(upr = -c(model, outcome))
@@ -124,7 +124,7 @@ visualize_risk_evaluation <- function(risk_evaluation,
 
   lower_values <- ggdata %>%
     group_by(outcome, md_method, model, additional_missing_pct) %>%
-    select(-auc) %>%
+    select(-auc, -cal_error) %>%
[…]

### 2020-11-03 doc_arXiv/doc_arXiv.Rmd
misc additions to discussion

@@
 We included a contemporary cohort of `r table_value(nrow(im))` patients who received continuous flow LVAD from `r min(im$im_impl_yr)`-`r max(im$im_impl_yr)`.
+This is a secondary analysis of de-identified data obtained from the National Heart Lung and Blood Institute.
+Primary data collection is approved through University of Alabama Institutional Review Board and at individual sites.
+
+## Outcomes and Predictors
+
 Patient follow-up begins after implantation of a durable, long term MCS device and continues while the device is in place.
 Registry endpoints include death on a device, heart transplantation, or cessation of support (for recovery and non-recovery reasons).
+Mortality and transplant after MCS were the primary outcomes for the current study.
+As there were only `r sum(im$pt_outcome_cess)` cessation of support events, we did not analyze this outcome.
 INTERMACS collects pre-implant patient characteristics, medical status, laboratory values, and many other variables.
 Data is collected at regularly scheduled follow-up as well as during adverse events such as re-hospitalization.
-This is a secondary analysis of de-identified data obtained from the National Heart Lung and Blood Institute.
-Primary data collection is approved through University of Alabama Institutional Review Board and at individual sites.
+For the current analysis, all `r ncol(im) - 4` pre-implant variables were considered as potential predictors.
 
@@
 
-In this article, we leveraged data from the INTERMACS registry to evaluate how the use of different imputation strategies prior to fitting a risk prediction model would impact the external prognostic accuracy of the model.
-External prognostic accuracy was measured at `r times` months after receiving MCS to focus on short term risk prediction, and the primary measure of accuracy was the scaled Brier score.
-We evaluated the performance of 12 imputation strategies in a broad range of settings by varying the type of modeling algorithm, the outcome variable, and the amount of additional missing data amputed prior to performing imputation.
+In this article, we leveraged INTERMACS registry data to evaluate how the use of different imputation strategies prior to fitting a risk prediction model would impact the external prognostic accuracy of the model.
+External prognostic accuracy was measured at `r times` months after receiving MCS, and the primary measure of accuracy was the scaled Brier score.
+We evaluated the performance of 12 imputation strategies in a broad range of settings by varying (1) the amount of additional missing data amputed prior to performing imputation, (2) the type of risk prediction model applied after imputation, and (3) the outcome variable for the risk prediction model.
 Our resampling experiment indicated that conducting multiple imputation has a high likelihood of increasing the downstream scaled Brier score and C-index of risk prediction models compared with imputation to the mean.
@@
 In previous studies involving the INTERMACS data registry, imputation to the mean has been applied prior to developing a mortality risk prediction model \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society}.
-The current analysis has shown that the prognostic accuracy of these models can be increased by implementing a different strategy to impute missing values, particularly if multiple imputation is applied.
-However, few studies have investigated best practices on the use of multiple imputation for prediction models, and few software packages allow users to implement multiple imputation procedures that adhere to best practices for prediction model development (\eg using only the training data to impute missing values).
-Further development of open source software that allows investigators to safely and succinctly implement multiple imputation for prediction models (\eg the \texttt{miceRanger} package) may increase the presence of multiple imputation in published risk prediction models.
+An interesting recent study indicates that imputation to the mean can provide an asymptotically consistent prediction model, given the prediction model is flexible and non-linear.
+However, theoretical results for finite samples have not yet been established.
+Our results provide relevant data for the finite sample case, suggesting that using imputation strategies considered in the current study instead of imputation to the mean can improve the prognostic accuracy of downstream models, particularly if multiple imputation is applied.
+
+Previous research has also established evidence in favor of applying multiple imputation to improve the prognostic value of risk prediction models.
+For example, Hassan and Atiya demonstrated superior downstream prediction using an ensemble multiple imputation method on synthetic data with continuous outcomes \cite{hassan2007regression}.
+Similarly, Nanni et.
+al demonstrated superior performance in downstream prediction when missing values were imputed using their proposed ensemble multiple imputation method \cite{nanni2012classifier}.
+Notably, the authors artificially induced missing values in these studies and the largest real dataset that was evaluated contained less than 700 observations.
+An article by Jerez et.
+al evaluated missing data strategies based on the downstream task of fitting a neural network and predicting early breast cancer relapse \cite{jerez2010missing}.
+The authors found that KNN imputation led to risk prediction models with the highest discrimination and lowest calibration error.
+Results from the current study are consistent with these previous findings but also extend their results by providing evidence from a larger source of data (\ie INTERMACS) and dealing with `real-world' missing values.
+
+Others have previously evaluated imputation techniques based on the accuracy with which these techniques impute missing values in the training data \cite{tutz2015improved, little2013joys, steele2018machine}.
+While it is intuitive to hypothesize that more accurate imputation will provide more prognostic downstream models, our results do not support this supposition.
+For example, when an additional 30\% of missing data were amputed, none of the missing data strategies we implemented obtained higher accuracy than imputation to the mean.
+However, using \emph{any} of the multiple imputation strategies we considered instead of imputation to the mean increased prognostic accuracy of downstream models when an additional 30\% of missing data were amputed.
+This result is likely explained by the bias variance tradeoff.
+In particular, single imputation techniques may lead to prediction models with lower bias but higher variance than multiple imputation techniques.
+
+\paragraph{Strengths and limitations} The current analysis has a number of strengths.
+We leveraged the INTERMACS data registry, comprising one of the largest cohorts of patients who received MCS.
+We applied a well known resampling method to internally validate modeling algorithms for risk prediction.
+Last, we made all source code for our analysis available in a public repository (see the first author's Github).
+Last, the approach presented in this paper provides a general framework that can be applied to risk prediction models in other longitudinal studies.
+The current analysis should also be interpreted in the context of known limitations.
+We considered a small subset of existing strategies to impute missing data, and other strategies may have provided stronger improvements compared with imputation to the mean.
+Also, we were not able to use only the training data to impute missing values in the testing data.
+Although the \texttt{miceRanger} package allows imputation of new data using existing models, few software packages for imputation allow users to implement multiple imputation with this protocol.
+
+\paragraph{Conclusion} Selecting an optimal strategy to impute missing values can impact the prognostic accuracy of downstream risk prediction models.
+In the current analysis, conducting multiple imputation using random forests emerged as an optimal strategy to impute missing values in the INTERMACS data.
+This investigation can directly inform future analyses of INTERMACS data and provide evidence quantifying the benefit of imputing missing data with sound methodology.
 
@@
 
-```{r, results = 'asis'}
+```{r tbl_impute_accuracy, results = 'asis'}
 

### 2020-11-10 doc_arXiv/response_to_reviewers.Rmd
revised doc shared with co-authors

@@
   officedown::rdocx_document:
-    mapstyles:
-      Normal: ['First Paragraph']
+    reference_docx: style_manuscript_times_new_roman.docx
 ---
@@
 
-__Comment__ This article assessed the impact of missing data imputation strategies and how they impacted accuracy of risk predictions.
+__Comment 1__: This article assessed the impact of missing data imputation strategies and how they impacted accuracy of risk predictions.
 This was conducted on a case study dataset assessing mortality and transplant from a cardiac registry dataset, INTERMACS.
@@
 
-__Response__ We thank the reviewer for their kind words and helpful feedback.
+__Response__: We thank the reviewer for their kind words and helpful feedback.
 We apologize for initially keeping our Github repository private, this was an oversight on our part.
@@
 
-__Comment__ While I have a few suggestions, these are quite minor. My main aim in this feedback is to help improve some aspects of communication from the authors to the reader.
+__Comment 2__: While I have a few suggestions, these are quite minor. My main aim in this feedback is to help improve some aspects of communication from the authors to the reader.
 
-__Response__ We appreciate the helpful suggestions that the reviewer has provided and provide a point-by-point response below.
+__Response__: We appreciate the helpful suggestions that the reviewer has provided and provide a point-by-point response below.
 
@@
 
-The paper describes exploring the 'handling' missing values, but from what I can tell, it is mostly about imputation strategies, and not handling.
+__Comment 3__: The paper describes exploring the 'handling' missing values, but from what I can tell, it is mostly about imputation strategies, and not handling.
 I would consider missing data handling methods to include decisions to remove observations, as well as 'logical' imputations (such as where missing values are identified as correctly being 0, or sections needing to be filled down).
-
 I think that the authors could address this by referring to most cases of 'handle/ing' as 'imputation, and by also discussing missing data handling strategies such as listwise deletion.
 For example, in the 'Methods and Results' section of the Abstract, changing, 'depends on the strategy that was applied to handle missing data' to 'depends on the imputation strategy applied'.
-That said, it might be possible for the authors to cover deletion methods in their results, discussed in the next section (adding listwise deletion)
+That said, it might be possible for the authors to cover deletion methods in their results, discussed in the next section (adding listwise deletion).
+
+__Response__: We agree, the term 'handling' is ambiguous while 'imputation' much more direct and appropriate for the current study. We have changed 'handling' to 'imputing' throughout the paper.
 
@@
 
-On page 3 on the paragraph starting at line 4, the authors describe missing data handling.
+__Comment 4__: On page 3 on the paragraph starting at line 4, the authors describe missing data handling.
 The typical behaviour is imputing to the mean - which is very bad, and it is great to see the authors addressing this.
@@
 It is almost always the wrong decision to perform, but I think it is worthwhile addressing in the paper, as it is unfortunately a practice that I think people might not be aware of as being very bad.
-
 It could also serve as a baseline/bottom point for the simulation, to show the impact of not imputing.
 I would be interested in the authors thoughts on this - I don't think that adding listwise deletion as an additional part of the simulation is necessary for publication, but I do think it might improve the paper as it can communicate the dangers of deletion methods.
-
 If the authors do not want to compare deletion methods to imputation I think it would be worthwhile to mention reasoning for this in the methods, or discussion section.
-
 However, I do think that the term, 'missing data handling' encompasses the process of the data tidying and cleaning decisions that take place when deciding what to do with missing data, which happens to include things like deletion and imputation.
@@
 
+__Response__: We agree that listwise deletion should be avoided when developing risk prediction modeling algorithms and that it should be mentioned more in the current study.
+We chose not to include listwise deletion as a strategy to impute missing values because listwise deletion is not capable of imputing missing values in testing data.
+Instead of imputing missing values in testing data, listwise deletion would delete them, and this would make our comparison between missing data strategies unfair because listwise deletion would only be evaluated on testing data without missing values.
+Since we are not including listwise deletion as a missing data strategy, we have changed 'handling' to 'imputing' throughout.
+We have also added text to the paper discussing listwise deletion and our rationale for not including listwise deletion as an imputation strategy in the Methods section.
+Specifically, In the 'Statistical Inference' paragraph, we write:
+
+"When very few data are missing, analysts may apply listwise deletion, \ie removing any observation with at least one missing value. However, listwise deletion can easily lead to biased inference."
+
+Then, in the "Statistical Learning" paragraph, we write:
+
+"Because testing data may contain missing values, listwise deletion is not a feasible strategy for statistical learning tasks."
+
 ## On Acronyms
 
-There are quite a few acronyms in this paper.
+__Comment 5__: There are quite a few acronyms in this paper.
 Would it be possible for the authors to reduce the number of acronyms?
@@
 
-I am also not entirely clear what an RPE is, exactly?
+__Response__: We agree and have reduced the number of acronyms used throughout the paper.
+
+__Comment 6__: I am also not entirely clear what an RPE is, exactly?
 This might be a term that is common in this area, so my apologies if this is perhaps misplaced, but how is it different to just 'prediction', or 'predicted values'?
@@
 
+__Response__: The reviewer is correct that a risk prediction equation is equivalent to a risk prediction model.
+We have replaced the acronym 'RPE' with the fully spelled out phrase 'risk prediction model'.
+We think this has improved the paper's readability and clarify.
+
 ## Thoughts on Methods
 
-The methods of this paper provide concise descriptions of the methodologies.
+__Comment 7__: The methods of this paper provide concise descriptions of the methodologies.
 One suggestion is to add a reference to Introduction to Statistical Learning for Gradient Boosted Decision Trees and a few of the other algorithms.
 
-### Imputation methods
+__Response__: We agree that Introduction to Statistical Learning is an excellent resource to provide introductions to these statistical learning algorithms.
+In the subsection "Risk Prediction Models", we write "A thorough description of stepwise variable selection and boosting can be found in Sections 6.1.2 and 8.2.2, respectively, of \emph{Introduction to Statistical Learning}." and follow with a reference to Introduction to Statistical Learning.
 
-Lines 19-20 on page 6 states: 
+__Comment 8__: Lines 19-20 on page 6 states: 
 'KNN imputation identifies k 'similar' observations for each case with at least one missing value'. 
@@
 
-### Missing patterns
+__Response__: We apologize for the lack of clarity in this sentence.
+In this instance, 'observation' and 'case' both refer to a single row in the data.
+We now use the term 'observation' only.
+Specifically, in the "K-Nearest-Neighbors" paragraph, we write:
 
-I'm not sure why this section is here and how it relates to the previous sections.
+"KNN imputation identifies $k$ `similar' observations (i.e., neighbors) for each observation with a missing value in a given variable.
+A donor value for the current missing value is generated by sampling or aggregating the values from the $k$ nearest neighbors."
+
+__Comment 9__: I'm not sure why the 'Missing patterns' section is here and how it relates to the previous sections.
 I suggest perhaps providing a sentence or two describing why this is important to cover.
@@
 
-### 2.6 - computational details
+__Response__: After revising the manuscript and adding various sections to the text, we have decided to omit the Section on missing patterns to decrease the word count and since it did not have high relevance to the remainder of the manuscript.
+We thank the reviewer for their feedback on this Section.
 
-SAS, R, and Python should be cited as well as relevant R and Python packages/modules used to produce results (such as ggplot2, which I am quite sure was used to create the graphics).
+__Comment 10__: SAS, R, and Python should be cited as well as relevant R and Python packages/modules used to produce results (such as ggplot2, which I am quite sure was used to create the graphics).
 
-### A note on Tables and Figures
+__Response__: We have added citations for SAS, R, and Python in addition to citations for the main R packages used in the analysis. These details are written in the 'Computational details' Section.
 
-The caption text for the tables and figures could provide more detail to the reader.
+## Tables and Figures
+
+__Comment 11__: The caption text for the tables and figures could provide more detail to the reader.
 For example, Table 3 and Table 4 provide important information on the performance of a variety of tests, however it would be helpful to provide some guidance.
@@
 Of course, this does not always hold true, for example, Table 2 (the typical 'table one' of the paper) that provides overview of the population - there is a lot of information here that is challenging to summarise, but if there are particular things to note about the population, they could be stated here.
-
 I don't think captions need to be a whole paragraph, but something that I think will enhance the great work the authors have done already in the paper, so that the audience can absorb it and put it into practice.
@@
 
-Another example example, figure 2, could be expanded to state: 'Three Model evaluation metrics (mean with 95%CI errors), improve as the number of multiple imputed datasets increases.
+__Response__: We agree and have updated captions for tables and figures to follow the template recommended by the reviewer.
+
+__Comment 12__: Another example example, figure 2, could be expanded to state: 'Three Model evaluation metrics (mean with 95%CI errors), improve as the number of multiple imputed datasets increases.
 ...'.
@@
 
+__Response__: We have replaced the original Figure 2, which focused only on a small subset of our results.
+The replacement figures show the distribution of improvements in model scores when different imputation strategies are applied relative to imputation to the mean.
+The results in these updated figures give a more broad overview of results in the current study by adjusting for outcome, % of additional missing data, and modeling strategy.
+We have used the reviewer's excellent suggestions to caption these updated figures.
+
 ## Discussion
 
-Page 20 line 2: I wonder if some of these references could be mentioned earlier in the paper in the introduction?
+__Comment 13__: Page 20 line 2: I wonder if some of these references could be mentioned earlier in the paper in the introduction?
 
-The paper demonstrates how good methods such as KNN are for imputation, however I think it might be just as critical to emphasize the importance of NOT using mean/median imputation methods.
+__Response__: We now reference these studies in both the Introduction and the Discussion sections.
+Specifically, in the introduction, these references are listed after the sentence "As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device."
 
-Page 20 line 10 - 12: Can you tell me more about what you learn from the figure here?
+__Comment 14__: The paper demonstrates how good methods such as KNN are for imputation, however I think it might be just as critical to emphasize the importance of NOT using mean/median imputation methods.
 
-Page 22 line 1: It is fantastic to see that the authors are sharing code.
+__Response__: We agree with the reviewer.
+A paragraph has been added to the Discussion section to present this argument.
+There is some evidence that imputation to the mean is okay when sample sizes are very large and a flexible imputation method is used, and we try to weigh this evidence carefully with our own findings.
+For convenience, the paragraph is pasted below:
+
+"In previous studies involving the INTERMACS data registry, imputation to the mean has been applied prior to developing a mortality risk prediction model.
+An interesting recent study indicates that imputation to the mean can provide an asymptotically consistent prediction model, given the prediction model is flexible and non-linear.
+However, theoretical results for finite samples have not yet been established.
+Our results provide relevant data for the finite sample case, suggesting that using imputation strategies considered in the current study instead of imputation to the mean can improve the prognostic accuracy of downstream models, particularly if multiple imputation is applied.
+Imputation to the mean should be avoided in future analyses of the INTERMACS registry and analyses where inflexible models are applied."
+
+__Comment 15__: Page 20 line 10 - 12: Can you tell me more about what you learn from the figure here?
+
+__Response__: This figure has been replaced by more informative visualizations.
+Our intention for the original figure was to show the overall improvement in model prognostic accuracy when multiple imputation was applied.
+In the revised manuscript, we formalize that objective in our analysis plan by using Bayesian hierarchical models to estimate the posterior probability that using a given imputation strategy instead of imputation to the mean will improve a model's scaled Brier score, concordance index, and calibration error.
+Figures 2-4 now convey this overall message, which has also been expressed in their figure captions.
+
+__Comment 16__: Page 22 line 1: It is fantastic to see that the authors are sharing code.
 Obviously the data for the paper is likely very sensitive, but I wonder if simulated data is provided?
@@
 
-One question I would be interested in seeing answered is what other types of data these approaches are suitable or unsuitable for in the field?
+__Response__: We apologize for keeping the GitHub repository private during the first submission.
+We have now opened it for public view (https://github.com/bcjaeger/INTERMACS-missing-data) and deposited the repository's code on Zenodo (see ReadMe badge on GitHub repository).
+Regarding data sharing, we are not authorized to share the original data or a simulation of the original data.
+However, the INTERMACS data are available on BIOLINCC and we have initiated a request to share the specific data used for this analysis on BIOLINCC as well.
+
+__Comment 17__: One question I would be interested in seeing answered is what other types of data these approaches are suitable or unsuitable for in the field?
 So, if I am a medical practitioner with a database of heart attacks that have missing data, am I likely to get similar improvements?
@@
 
-The paper has great potential for impact - I can imagine a lot of registries using these approaches to impute missing data and getting improved treatment results, and I think that a key part of this is that the authors have made efforts to share their code.
+__Response__: The reviewer raises an excellent question.
+Given that the current analysis focuses only on the INTERMACS registry, we are cautious to make strong generalizations.
+In our paragraph on limitations, we recommend future studies focus on providing open source software and tutorials for investigators to run a similar resampling experiment with their own data to identify optimal strategies for imputation of missing values.
+Specifically, we write
+
+"Although the \texttt{miceRanger} package allows imputation of new data using existing models, few software packages for imputation allow users to implement multiple imputation with this protocol.
+Future analyses should introduce more flexible software and hands-on tutorials so that future investigators can optimize imputation of missing data."
+
+__Comment 18__: The paper has great potential for impact - I can imagine a lot of registries using these approaches to impute missing data and getting improved treatment results, and I think that a key part of this is that the authors have made efforts to share their code.
 This could be emphasized more in the paper, so the reader knows that they too can use similar approaches with less effort.
 
-Overall, a great paper.
+__Response__: We agree with the reviewer and have emphasized the publicly available repository in the following places of the revised manuscript:
 
-Reviewer #2:
+- In the Abstract:
 
-Review for CIRCCQO/2020/00707 - Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values
+- In the "Computational Details" section: "Our R code is available on GitHub"
 
-In this study, the authors applied six strategies of handling missing values to develop three risk prediction equations (RPE), and compared the missing handling techniques on model performance and prediction accuracy.
+- In the "Strengths and limitations" paragraph: "we made our analysis R code available in a public repository"
+
+__Comment 19__: Overall, a great paper.
+
+__Response__: Thank you. We are grateful for this excellent review of our work. Your feedback has substantially improved the quality of the paper.
+
+\newpage
+
+# Reviewer #2:
+
+__Comment 1__: In this study, the authors applied six strategies of handling missing values to develop three risk prediction equations (RPE), and compared the missing handling techniques on model performance and prediction accuracy.
 Overall, the manuscript is well-written.
 To increase the applicability of these missing handling techniques, I have summarized some areas that could be improved.
+__Response__: We thank the reviewer for their helpful feedback and have provided a point by point response below.
 
-1.
-The authors used the missing handling techniques in their RPE development.
+__Comment 2__: The authors used the missing handling techniques in their RPE development.
 These RPEs however are not yet established; and the performance of these RPE are not satisfied, for example, the c-index for mortality event is only around 65%.
@@
 
+__Response__: We agree that it would be interesting to compare risk prediction models in the current study with existing risk prediction models.
+We have taken steps in the revised paper to be as responsive as possible to the reviewer's feedback: "c-index for mortality event is only around 65%".
+We have updated our inclusion criteria to identify a more contemporary cohort with more structured and prognostic data (i.e., INTERMACS patients enrolled during or after 2012).
+C-index values are now generally within the range of 0.69 to 0.71, which is consistent with C-index values of existing risk prediction models.
+Unfortunately, it is not possible to conduct a fair comparison between existing INTERMACS risk prediction models and the models developed in the current study for two reasons.
+
+1. Existing risk prediction models for INTERMACS have not published any supplemental code or software to generate predicted risk for new data.
+
 2.
-Imputation accuracy: Can the authors compare the imputation accuracy of their techniques through either artificially inducing missing values to known data or simulating data or literature search?
+Many existing risk prediction models were developed using the entire INTERMACS cohort used in the current study.
+Therefore, if we were to use these existing risk prediction models, we would be validating them on the same data that they were developed with, which would lead to optimistically biased results.
+
+__Comment 3__: Imputation accuracy: Can the authors compare the imputation accuracy of their techniques through either artificially inducing missing values to known data or simulating data or literature search?
 It's not critical to have imputation accuracy for prediction purpose (as authors indicated in the discussion), however, the imputation accuracy impacts model inference and will limit the use of certain missing handling techniques.
 
-3.
-Page 9.
+__Response__: We agree and have included imputation accuracy in the revised manuscript (see the Evaluating Imputation Accuracy Section, the "Imputation accuracy" paragraph in the Results Section, and the fourth paragraph in the Discussion Section).
+Due to the substantial quantity of data that were added to the paper in light of this addition, we have removed CIF models from the revised analysis.
+This decision was made because CIF results were quite similar to boosting results and did not add any relevant new information to the paper.
+
+__Comment 4__: Page 9.
 Line 12-14 suggests up to 50 predictor variables were used, while page 13 Line 18 suggests 320 predictors with the one missing values exists in the registry.
@@
 
-4.
-Page 12.
+__Response__: We agree with the reviewer and apologize for omitting this detail.
+The predictor variable selection process is now written in the "Steps taken in each replicate" paragraph, which can be found in the Internal "Validation via Monte-Carlo Cross-Validation" Section.
+The relevant text from this paragraph is also copied here for convenience:
+
+"All predictor variables with < 50\% missing values were considered for imputation and subsequent model development...Prior to imputation, 50 predictor variables were selected using a boosting model that quantified variable importance as the fractional contribution of each predictor to the model based on the total gain attributed to using the predictor while growing decision trees."
+
+We have also summarized the proportion of missingness in the overall population and stratified by outcome status in the revised manuscript (see Table 2).
+Additionally, possible missing patterns related to a selection of relevant mortality predictors are presented in the revised manuscript (see Figure 1).
+
+__Comment 5__: Page 12.
 Line 1-3.
@@
 In Table 3.
-Although the KNN (MI) methods improved the scaled Brier score, however, the scaled Brier scores are very closed to 0, which indicate low accuracy of a RPE.
 
-5.
-Table 2.
+__Response__: The reviewer is correct regarding the scaled Brier score.
+It is analagous to an $R^2$ statistic in linear regression.
+We have provided more details on interpretation of the scaled Brier score in the revised manuscript.
+The revised text can be found in the "scaled Brier score" paragraph of the revised manuscript.
+For convenience, it is also copied below:
+
+"The Brier score is dependent on the rate of observed events, which can make it a difficult metric to interpret.
+It is often more informative to scale the Brier score based on the Brier score of a naive model.
+More specifically, for a given risk prediction model, the scaled Brier score is computed as ...[see manuscript for equation]...
+As the Brier score for risk prediction is analogous to mean-squared error for prediction of a continuous outcome, the scaled Brier score is analogous to the $R^2$ statistic.
+Similar to the $R^2$ statistic, a scaled Brier score of 1.00 and 0.00 indicate a perfect and worthless model, respectively.
+In our analyses, a Kaplan-Meier estimate based on the training data (\ie a risk prediction model that did not use any predictor variables) provided the naive prediction.
+In the current analysis, we multiply scaled Brier score values by 100 for ease of interpretation."
+
+__Comment 6__: Although the KNN (MI) methods improved the scaled Brier score, however, the scaled Brier scores are very closed to 0, which indicate low accuracy of a RPE.
+
+__Response__: We agree and have added the following sentence to our paragraph on limitations in the revised manuscript: "The models we studied obtained low values of scaled Brier score, indicating low prediction accuracy.
+It is unclear how results may vary for models with higher prediction accuracy."
+
+__Comment 7__: Table 2.
 please clarify when modeling mortality events, how patients with transplants events and cessation were used, are they treated as censoring or are they dropped?
@@
 
-6. Table 3. The concordance is low, which indicates low model performance. The table title has a typo for "mortality".
+__Response__: The reviewer has correctly identified that patients who received transplant or cessation of support were censored at the time of these events in mortality risk prediction models.
+Similarly, patients who died or received cessation of support were censored at the time of these events in transplant risk prediction models.
+We have clarified this approach in our "Outcomes and Predictors" section, where we write "For each outcome, we censored patients at the time of competing risks events.
+For example, in risk prediction models for mortality, patients who were lost to follow-up, received transplant, or had cessation of support were censored at the time of this event".
 
-7.
-Page 15.
+__Comment 8__: Table 3. The concordance is low, which indicates low model performance. The table title has a typo for "mortality".
+
+__Response__: Thank you for catching this typo.
+We have changed 'morality' to 'mortality' and have taken steps to improve the discrimination of the risk prediction models in the current study.
+These steps are outlined in our response to __Comment 2__ above.
+
+__Comment 9__: Page 15.
 Line 3.
@@
 
-8.
-Page 15 Line 6-8.
+__Response:__ Thank you for noticing these inconsistencies due to rounding. We have corrected these typos and now use the same rounding specification throughout the manuscript.
+
+__Comment 10__: Page 15 Line 6-8.
 From Table 3, for the CIF, stepwise PH and GBDT, the missing data strategies that provided the lowest calibration error were missForest, KNN (MI), and BR respectively.
 Please correct the order of the statement.
+
+__Response__: Thank you for catching this typo. This sentence was deleted from the revised manuscript due to the substantial revisions applied to our analysis.

### 2020-11-10 README.Rmd
revised doc shared with co-authors

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# INTERMACS-missing-data

[![DOI](https://zenodo.org/badge/298893426.svg)](https://zenodo.org/badge/latestdoi/298893426)

This repository contains all of the code used in the paper _Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values_.

### 2020-11-10 doc_arXiv/doc_arXiv.Rmd
revised doc shared with co-authors

@@
     affiliation: University of Alabama at Birmingham
-    location: Birmingham, AL 35211
+    location: Birmingham, AL
     email: bcjaeger@uab.edu
+  - name: Ryan Cantor, PhD
+    affiliation: University of Alabama at Birmingham
+    location: Birmingham, AL
+  - name: Venkata Sthanam, MS
+    affiliation: University of Alabama at Birmingham
+    location: Birmingham, AL
+  - name: Rongbing Xie, DrPH
+    affiliation: University of Alabama at Birmingham
+    location: Birmingham, AL
+  - name: James K. Kirklin, MD
+    affiliation: University of Alabama at Birmingham
+    location: Birmingham, AL
+  - name: Ramaraju Rudraraju, PhD
+    affiliation: University of Alabama at Birmingham
+    location: Birmingham, AL
+    email: ramaraju@uab.edu
 abstract: |
-  \textbf{Background} Risk predictions play an important role in clinical decision making.
+  \textbf{Background} Risk prediction models play an important role in clinical decision making.
 When developing risk prediction models, practitioners often impute missing values to the mean.
-The purpose of this article is to evaluate the impact of applying different strategies to impute missing values on the prognostic accuracy of prediction models fitted to the imputed data.
-A secondary objective was to compare the accuracy of different imputation methods.
+We evaluated the impact of applying other strategies to impute missing values on the prognostic accuracy of downstream risk prediction models, i.e., models fitted to the imputed data.
+A secondary objective was to compare the accuracy of imputation methods based on artificially induced missing values.
 To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).
-\newline\textbf{Methods and Results} We applied thirteen different strategies to impute missing values in combination with three different strategies to fit a risk prediction model for mortality and transplant after receiving mechanical circulatory support.
-Model performance was measured by 12-month discrimination, calibration and net reclassification index.
-Results indicated that multiple  imputation consistently provided prediction models with greater prognostic accuracy than other missing data strategies, particularly imputation to the mean.
-<fill in results> \newline\textbf{Conclusion} Selecting an optimal strategy to handle missing values impacts the prognostic accuracy of downstream models.
-In the current analysis, multiple imputation emerged as an optimal strategy to handle missing values in the INTERMACS data.
-The current study shows that evaluation and selection of an optimal strategy to impute missing data has the potential to improve prognostic accuracy of risk predictions for other longitudinal registries.
+\newline\textbf{Methods and Results} We applied twelve imputation strategies in combination with two different modeling strategies for mortality and transplant risk prediction following surgery to receive mechanical circulatory support.
+Model performance was evaluated using Monte-Carlo cross validation and measured based on outcomes 6-months following surgery using the scaled Brier score, concordance index, and calibration error.
+We used Bayesian hierarchical models to compare model performance.
+Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
+The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.482 and 0.962, respectively.
+\newline\textbf{Conclusion} Selecting an optimal strategy to impute missing values and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
 keywords:
-  - Missing Data, 
-  - INTERMACS, 
-  - imputation, 
-  - heart failure, 
-  - mortality, 
+  - Missing Data
+  - INTERMACS
+  - imputation
+  - heart failure
+  - mortality
   - risk prediction
@@
 However, less than 4,000 new patients receive a long-term MCS device each year, with widely heterogeneous outcomes \cite{stewart2011keeping}.
-The 2-year survival on MCS ranges from 61\% for destination therapy to 78\% for bridge-to-transplant \cite{patel2014contemporary}.
-Therefore, there is great need for reliable predictions of patient-specific risk to experience adverse events after receiving MCS.
-This information can be used to improve patient selection for MCS, inform the design of next generation pumps, and refine patient management strategies.
+The 2-year survival probability on MCS ranges from 61\% for destination therapy to 78\% for bridge-to-transplant \cite{patel2014contemporary}.
+Reliable predictions of patient-specific risk to experience key outcomes such as mortality or transplant after receiving MCS can help improve patient selection, inform the design of next generation pumps, and improve patient care strategies.
 
@@
 Currently, INTERMACS comprises over 20,000 patients who have received a MCS device.
-As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{kirklin2017eighth, kormos2019society, Adamo950}.
+As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
 A critical step for developing risk prediction models is handling missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial.
@@
 
-In this section, we describe the INTERMACS Registry in Section \ref{subsec:intermacs} and provide a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning in Section \ref{subsec:inference_and_learning}.
+<!-- In this section, we describe the INTERMACS Registry in Section \ref{subsec:intermacs} and provide a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning in Section \ref{subsec:inference_and_learning}.
 We then describe thirteen strategies to handle missing values (Section \ref{subsec:imputation}) and three strategies to develop risk prediction models after missing data have been imputed (Section \ref{subsec:modeling}).
@@
 We outline an internal validation procedure that was applied to compare combinations of imputation and modeling strategies using hierarchical Bayesian models in Section \ref{subsec:internal}.
-Last, details on computation and software packages used for the current analysis are described in Section \ref{subsec:computing}
+Last, details on computation and software packages used for the current analysis are described in Section \ref{subsec:computing} -->
 
@@
 
-The INTERMACS data is publicly available on biolincc at https://biolincc.nhlbi.nih.gov/studies/intermacs/.
 INTERMACS is a North American observational registry for patients receiving MCS devices that began as a partnership between the National Heart, Lung, and Blood Institute, US Food and Drug Administration, the Centers for Medicaid and Medicare Services, industry, and individual hospitals with the mission of improving MCS outcomes.
@@
 
-The current analysis was conducted using publicly available data provided by the National Heart, Lung, and Blood Institute.
+The current analysis was conducted using publicly available data provided by the National Heart, Lung, and Blood Institute (see https://biolincc.nhlbi.nih.gov/studies/intermacs/).
 We included a contemporary cohort of `r table_value(nrow(im))` patients who received continuous flow LVAD from `r min(im$im_impl_yr)`-`r max(im$im_impl_yr)`.
@@
 As there were only `r sum(im$pt_outcome_cess)` cessation of support events, we did not analyze this outcome.
-INTERMACS collects pre-implant patient characteristics, medical status, laboratory values, and many other variables.
-Data is collected at regularly scheduled follow-up as well as during adverse events such as re-hospitalization.
-For the current analysis, all `r ncol(im) - 4` pre-implant variables were considered as potential predictors.
+For mortality and transplant risk prediction models, we applied cause-specific analysis to account for competing risks.
+For example, in risk prediction models for mortality, patients were censored at time of loss to follow-up, transplant, or cessation of support, whichever occurred first.
+INTERMACS collects pre-implant data on patient characteristics, medical status, and laboratory values.
+INTERMACS also collects follow-up data at regularly scheduled visits and at occurrence of adverse events such as re-hospitalization.
+For the current analysis, all pre-implant variables were considered as potential predictors.
 
@@
 
-\paragraph{Statistical Inference} Previous research has thoroughly investigated statistical inference in the context of missing data.
-In this setting, analysts often create multiple imputed datasets, replicate an analysis in each of them, and then pool their results to obtain valid test statistics for hypothesis testing \cite{rubin2004multiple}.
+\paragraph{Statistical Inference} To conduct statistical inference in the context of missing data setting, analysts often create multiple imputed datasets, replicate an analysis in each of them, and then pool their results to obtain valid test statistics for hypothesis testing \cite{rubin2004multiple}.
 Imputation strategies that create a single imputed dataset (\eg imputation to the mean) have been shown to increase type I errors (\ie rejecting a true null hypothesis) for inferential statistics by artificially reducing the variance of observed data and ignoring the uncertainty attributed to missing values \cite{van2018flexible}.
@@
 
-\paragraph{Single and multiple imputation} Several of the imputation methods we considered allowed for creation of one or many imputed datasets.
+\paragraph{Single and multiple imputation} Several of the imputation methods we considered allowed for creation of single or multiple imputed datasets.
 For downstream models fitted to multiple sets of imputed data, we applied the modeling technique to each imputed training set, separately, and then used a pooling technique to generate a single set of predictions based on multiple imputed testing data.
 Specifically, we created 10 imputed training and testing sets, then applied a modeling procedure to each imputed training set and computed model predictions on the corresponding imputed testing set, which led to 10 sets of predictions.
-To aggregate these predictions, we computed the median for each observation.
+To aggregate these predictions, we computed the median for each patient.
 Informal experiments where the mean was used instead of the median to aggregate predictions showed little or no difference in the model's prediction accuracy.
@@
 The computed means and modes are then stored for future imputation of testing data.
-Because imputation to the mean is frequently used in practice, we use it as a reference for comparison of all other strategies to impute missing values,
+Because imputation to the mean is frequently used in practice, we use it as a reference for comparison of all other strategies to impute missing values.
 
@@
 Multiple imputation with chained equations (MICE), a well known technique for generating multiply imputed data \cite{azur2011multiple, van2018flexible, van2006fully}.
-In each iteration of MICE, each specified variable in the dataset is imputed using the other variables in the dataset.
+In each iteration of MICE, each specified variable in the dataset is imputed using  other variables in the dataset.
 These iterations are run until convergence criteria have been met.
@@
 
-## Evaluating imputation accuracy
+## Evaluating Imputation Accuracy
 \label{subsec:imputation_accuracy}
@@
 These numeric and nominal scores are analogous to the more well known $R^2$ and Kappa statistics, respectively, but are modified slightly in the current analysis so that imputation to the mean will always have a score of 0.
-This modification makes it easier to compare the accuracy of each imputation strategy directly with imputation to the mean.
+This modification makes it easier to compare the accuracy of each imputation strategy directly with that of imputation to the mean.
 
@@
 
-## Computational details
+## Computational Details
 \label{subsec:computing}
 
-Our source code is available on GitHub (see [https://github.com/bcjaeger/INTERMACS-missing-data](https://github.com/bcjaeger/INTERMACS-missing-data)).
+SAS software (version 9.4) and Python (version 3.8.2) were used to create analytic data for the current study \cite{van1995python, sas2013}.
+Base R (version 4.0.3) was used in combination with a number of open-source R packages (\eg \texttt{drake}, \texttt{tidyverse}, \texttt{naniar}, \texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to conduct our analysis \cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
+Our R code is available on GitHub (see [https://github.com/bcjaeger/INTERMACS-missing-data](https://github.com/bcjaeger/INTERMACS-missing-data)) \cite{byron_2020_4247449}.
+We used Cheaha, a high performance computing cluster at University of Alabama at Birmingham, to perform the 200 Monte-Carlo cross validation runs \cite{cheaha}.
 
@@
 
-\paragraph{Bayesian analysis of model performance}
+## Bayesian analysis of model performance
 
@@
 
-In previous studies involving the INTERMACS data registry, imputation to the mean has been applied prior to developing a mortality risk prediction model \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society}.
-An interesting recent study indicates that imputation to the mean can provide an asymptotically consistent prediction model, given the prediction model is flexible and non-linear.
+In previous studies involving the INTERMACS data registry, imputation to the mean has been applied prior to developing a mortality risk prediction model \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
+An interesting recent study indicates that imputation to the mean can provide an asymptotically consistent prediction model, given the prediction model is flexible and non-linear \cite{josse2019consistency}.
 However, theoretical results for finite samples have not yet been established.
 Our results provide relevant data for the finite sample case, suggesting that using imputation strategies considered in the current study instead of imputation to the mean can improve the prognostic accuracy of downstream models, particularly if multiple imputation is applied.
+Imputation to the mean should be avoided in future analyses of the INTERMACS registry and analyses where inflexible models are applied.
 
@@
 We applied a well known resampling method to internally validate modeling algorithms for risk prediction.
-Last, we made all source code for our analysis available in a public repository (see the first author's Github).
-Last, the approach presented in this paper provides a general framework that can be applied to risk prediction models in other longitudinal studies.
+Last, we made our analysis R code available in a public repository (https://github.com/bcjaeger/INTERMACS-missing-data).
+Last, the approach presented in this paper provides a general framework that can be applied in other studies where missing data are imputed prior to fitting a risk prediction model.
 The current analysis should also be interpreted in the context of known limitations.
 We considered a small subset of existing strategies to impute missing data, and other strategies may have provided stronger improvements compared with imputation to the mean.
-Also, we were not able to use only the training data to impute missing values in the testing data.
+The models we studied obtained low values of scaled Brier score, indicating low prediction accuracy.
+It is unclear how results may vary for models with higher prediction accuracy.
+Also, it was not feasible to use only the training data to impute missing values in the testing data due to a lack of available software.
 Although the \texttt{miceRanger} package allows imputation of new data using existing models, few software packages for imputation allow users to implement multiple imputation with this protocol.
+Future analyses should introduce more flexible software and hands-on tutorials so that future investigators can optimize imputation of missing data.
 
@@
 \begin{table}
-\caption{Participant characteristics stratified by event status.}
+\caption{Participant characteristics stratified by event status. The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
 \label{tbl_characteristics}
@@
   gt_latex(
-    caption = 'Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.',
+    caption = 'Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.
+Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.',
     label = 'tbl_missingness'
@@
   gt_latex(
-    caption = 'Accuracy of strategies to impute artificial missing data', 
+    caption = 'Accuracy of strategies to impute artificial missing data.
+Table values are the median change in accuracy (25th, 75th percentile) relative to the accuracy of imputation to the mean.
+In general, multiple imputation strategies had lower accuracy than single imputation strategies, and few imputation strategies were more accurate than imputation to the mean.',
     label = 'tbl_impute_accuracy'
@@
 
-make_tbl_md_strat_caption <- function(metric, outcome){
-  glue('Median (25th, 75th percentile) change in {metric} when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for {outcome}.
+make_tbl_md_strat_caption <- function(metric, outcome, msg){
+  glue('Median (25th, 75th percentile) change in {metric} when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for {outcome}.
 Table values show the {metric} for imputation to the mean.
 For other imputation strategies, table values show the change in {metric} relative to the {metric} when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability')
+All table values are multiplied by 100 for ease of interpretability.
+{msg}.')
 }
@@
   gt_tbls$md_strat$dead.ipa,
-  caption = make_tbl_md_strat_caption(metric = 'scaled Brier score',
-                                      outcome = 'mortality'),
+  caption = make_tbl_md_strat_caption(
+    metric = 'scaled Brier score',
+    outcome = 'mortality',
+    msg = "Multiple imputation with random forests leads to the highest scaled Brier score for both models and when 0\\%, 15\\%, or 30\\% of additional data in the training and testing sets were set to missing"
+  ),
   label = 'tbl_md_strat_dead_ipa'
@@
   gt_tbls$md_strat$txpl.ipa,
-  caption = make_tbl_md_strat_caption(metric = 'scaled Brier score',
-                                      outcome = 'transplant'),
+  caption = make_tbl_md_strat_caption(
+    metric = 'scaled Brier score',
+    outcome = 'transplant',
+    msg = 'While there was very little difference in scaled Brier score values when 0\\% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher scaled Brier scores when 15\\% or 30\\% of additional missing data were amputed'
+  ),
   label = 'tbl_md_strat_txpl_ipa'
@@
   gt_tbls$md_strat$dead.auc,
-  caption = make_tbl_md_strat_caption(metric = 'concordance index',
-                                      outcome = 'mortality'),
+  caption = make_tbl_md_strat_caption(
+    metric = 'concordance index',
+    outcome = 'mortality',
+    msg = 'Multiple imputation with random forests led to the highest concordance index for boosting models when 0\\%, 15\\%, or 30\\% of additional data in the training and testing sets were set to missing.
+For proportional hazards models, multiple imputation with nearest neighbors or random forests was the most effective strategy'
+  ),
   label = 'tbl_md_strat_dead_auc'
@@
   gt_tbls$md_strat$dead.cal_error,
-  caption = make_tbl_md_strat_caption(metric = 'calibration error',
-                                      outcome = 'mortality'),
+  caption = make_tbl_md_strat_caption(
+    metric = 'calibration error',
+    outcome = 'mortality',
+    msg = 'As more data in the training and testing sets were set to missing, single and multiple imputation with random forests emerged as the strategies with the lowest calibration error'
+  ),
   label = 'tbl_md_strat_dead_cal_error'
@@
   gt_tbls$md_strat$txpl.auc,
-  caption = make_tbl_md_strat_caption(metric = 'concordance index',
-                                      outcome = 'transplant'),
+  caption = make_tbl_md_strat_caption(
+    metric = 'concordance index',
+    outcome = 'transplant',
+    msg = 'While there was very little difference in concordance index when 0\\% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher concordance indices when 15\\% or 30\\% of additional missing data were amputed'
+  ),
   label = 'tbl_md_strat_txpl_auc'
@@
   gt_tbls$md_strat$txpl.cal_error,
-  caption = make_tbl_md_strat_caption(metric = 'calibration error',
-                                      outcome = 'transplant'),
+  caption = make_tbl_md_strat_caption(
+    metric = 'calibration error',
+    outcome = 'transplant',
+    msg = "As more data in the training and testing sets were set to missing, single and multiple imputation with nearest neighbors emerged as the optimal strategy for proportional hazards models while missingness incorporated as an attributed emerged an the optimal strategy for boosting models"
+  ),
   label = 'tbl_md_strat_txpl_cal_error'
@@
 
-```{r fig_md_strat_infer_ipa, dpi = 300, fig.height = 6, fig.width = 6.5, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+```{r fig_md_strat_infer_ipa, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
-Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution."}
+Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy, is printed to the right of each distribution.
+Each multiple imputation strategy and single imputation with missingness incorporated as an attribute had over 90\\% posterior predicted probability of increasing the scaled Brier score versus using imputation to the mean."}
 
@@
 
-```{r fig_md_strat_infer_auc, dpi = 300, fig.height = 6, fig.width = 6.5, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+```{r fig_md_strat_infer_auc, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in concordance index values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
-Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution."}
+Posterior probability that the difference in concordance index exceeds 0, indicating an improvement in model discrimination, is printed to the right of each distribution.
+Multiple imputation with predictive mean matching, random forests, and Bayesian regression each had over 90\\% posterior predicted probability of increasing the concordance index versus using imputation to the mean."}
 
@@
 
-```{r fig_md_strat_infer_cal_error, dpi = 300, fig.height = 6, fig.width = 6.5, fig.align='center', fig.cap = "Posterior distribution of differences in calibration error values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+```{r fig_md_strat_infer_cal_error, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in calibration error values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
-Posterior probability that the difference in calibration error exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution."}
+Posterior probability that the difference in calibration error is less than 0, indicating an improvement in model calibration, is printed to the right of each distribution.
+Every imputation strategy evaluated had over 90\\% posterior predicted probability of improving model calibration versus using imputation to the mean."}
 

### 2020-11-10 doc_arXiv/doc_arXiv.tex
revised doc shared with co-authors

@@
     University of Alabama at Birmingham \\
-  Birmingham, AL 35211 \\
+  Birmingham, AL \\
   \texttt{\href{mailto:bcjaeger@uab.edu}{\nolinkurl{bcjaeger@uab.edu}}} \\
+   \And
+    Ryan Cantor, PhD
+   \\
+    University of Alabama at Birmingham \\
+  Birmingham, AL \\
+  \texttt{} \\
+   \And
+    Venkata Sthanam, MS
+   \\
+    University of Alabama at Birmingham \\
+  Birmingham, AL \\
+  \texttt{} \\
+   \And
+    Rongbing Xie, DrPH
+   \\
+    University of Alabama at Birmingham \\
+  Birmingham, AL \\
+  \texttt{} \\
+   \And
+    James K. Kirklin, MD
+   \\
+    University of Alabama at Birmingham \\
+  Birmingham, AL \\
+  \texttt{} \\
+   \And
+    Ramaraju Rudraraju, PhD
+   \\
+    University of Alabama at Birmingham \\
+  Birmingham, AL \\
+  \texttt{\href{mailto:ramaraju@uab.edu}{\nolinkurl{ramaraju@uab.edu}}} \\
   }
@@
 \begin{abstract}
-\textbf{Background} Risk predictions play an important role in clinical
-decision making. When developing risk prediction models, practitioners
-often impute missing values to the mean. The purpose of this article is
-to evaluate the impact of applying different strategies to impute
-missing values on the prognostic accuracy of prediction models fitted to
-the imputed data. A secondary objective was to compare the accuracy of
-different imputation methods. To complete these objectives, we used data
-from the Interagency Registry for Mechanically Assisted Circulatory
-Support (INTERMACS). \newline\textbf{Methods and Results} We applied
-thirteen different strategies to impute missing values in combination
-with three different strategies to fit a risk prediction model for
-mortality and transplant after receiving mechanical circulatory support.
-Model performance was measured by 12-month discrimination, calibration
-and net reclassification index. Results indicated that multiple
-imputation consistently provided prediction models with greater
-prognostic accuracy than other missing data strategies, particularly
-imputation to the mean. \newline\textbf{Conclusion} Selecting an optimal
-strategy to handle missing values impacts the prognostic accuracy of
-downstream models. In the current analysis, multiple imputation emerged
-as an optimal strategy to handle missing values in the INTERMACS data.
-The current study shows that evaluation and selection of an optimal
-strategy to impute missing data has the potential to improve prognostic
-accuracy of risk predictions for other longitudinal registries.
+\textbf{Background} Risk prediction models play an important role in
+clinical decision making. When developing risk prediction models,
+practitioners often impute missing values to the mean. We evaluated the
+impact of applying other strategies to impute missing values on the
+prognostic accuracy of downstream risk prediction models, i.e., models
+fitted to the imputed data. A secondary objective was to compare the
+accuracy of imputation methods based on artificially induced missing
+values. To complete these objectives, we used data from the Interagency
+Registry for Mechanically Assisted Circulatory Support (INTERMACS).
+\newline\textbf{Methods and Results} We applied twelve imputation
+strategies in combination with two different modeling strategies for
+mortality and transplant risk prediction following surgery to receive
+mechanical circulatory support. Model performance was evaluated using
+Monte-Carlo cross validation and measured based on outcomes 6-months
+following surgery using the scaled Brier score, concordance index, and
+calibration error. We used Bayesian hierarchical models to compare model
+performance. Multiple imputation with random forests emerged as a robust
+strategy to impute missing values, increasing model concordance by 0.003
+(25th, 75th percentile: 0.008, 0.052) compared with imputation to the
+mean for mortality risk prediction using a downstream proportional
+hazards model. The posterior probability that single and multiple
+imputation using random forests would improve concordance versus mean
+imputation was 0.482 and 0.962, respectively.
+\newline\textbf{Conclusion} Selecting an optimal strategy to impute
+missing values and applying multiple imputation can improve the
+prognostic accuracy of downstream risk prediction models.
 \end{abstract}
@@
 \keywords{
-    Missing Data,
+    Missing Data
    \and
-    INTERMACS,
+    INTERMACS
    \and
-    imputation,
+    imputation
    \and
-    heart failure,
+    heart failure
    \and
-    mortality,
+    mortality
    \and
@@
 a long-term MCS device each year, with widely heterogeneous outcomes
-\cite{stewart2011keeping}. The 2-year survival on MCS ranges from 61\%
-for destination therapy to 78\% for bridge-to-transplant
-\cite{patel2014contemporary}. Therefore, there is great need for
-reliable predictions of patient-specific risk to experience adverse
-events after receiving MCS. This information can be used to improve
-patient selection for MCS, inform the design of next generation pumps,
-and refine patient management strategies.
+\cite{stewart2011keeping}. The 2-year survival probability on MCS ranges
+from 61\% for destination therapy to 78\% for bridge-to-transplant
+\cite{patel2014contemporary}. Reliable predictions of patient-specific
+risk to experience key outcomes such as mortality or transplant after
+receiving MCS can help improve patient selection, inform the design of
+next generation pumps, and improve patient care strategies.
 
@@
 of adverse events that may occur after receiving a device
-\cite{kirklin2017eighth, kormos2019society, Adamo950}. A critical step
-for developing risk prediction models is handling missing data, which is
-common in a ``real world'' database that is not designed for the rigor
-of data completeness found in a clinical trial. While some missing
-values are related to data entry error, others may relate to instability
-of a patient's circulatory status at the time of implant. For instance,
-hemodynamic laboratory values are missing for patients in whom invasive
-catheterizations were not performed. Patients without invasive
-catheterizations may be too ill (unstable) to tolerate the procedure.
-Instead, such patients may, after determination of basic hemodynamics,
-proceed directly to device placement.
+\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
+A critical step for developing risk prediction models is handling
+missing data, which is common in a ``real world'' database that is not
+designed for the rigor of data completeness found in a clinical trial.
+While some missing values are related to data entry error, others may
+relate to instability of a patient's circulatory status at the time of
+implant. For instance, hemodynamic laboratory values are missing for
+patients in whom invasive catheterizations were not performed. Patients
+without invasive catheterizations may be too ill (unstable) to tolerate
+the procedure. Instead, such patients may, after determination of basic
+hemodynamics, proceed directly to device placement.
 
@@
 
-In this section, we describe the INTERMACS Registry in Section
-\ref{subsec:intermacs} and provide a broad overview of general
-objectives for imputation of missing data in the context of statistical
-inference and supervised statistical learning in Section
-\ref{subsec:inference_and_learning}. We then describe thirteen
-strategies to handle missing values (Section \ref{subsec:imputation})
-and three strategies to develop risk prediction models after missing
-data have been imputed (Section \ref{subsec:modeling}). Next, we
-describe how risk prediction models and by extension the imputation
-strategy applied before training the model were evaluated in Section
-\ref{subsec:evaluation}. We outline an internal validation procedure
-that was applied to compare combinations of imputation and modeling
-strategies using hierarchical Bayesian models in Section
-\ref{subsec:internal}. Last, details on computation and software
-packages used for the current analysis are described in Section
-\ref{subsec:computing}
-
 \hypertarget{intermacs-registry}{%
@@
 
-The INTERMACS data is publicly available on biolincc at
-\url{https://biolincc.nhlbi.nih.gov/studies/intermacs/}. INTERMACS is a
-North American observational registry for patients receiving MCS devices
-that began as a partnership between the National Heart, Lung, and Blood
-Institute, US Food and Drug Administration, the Centers for Medicaid and
-Medicare Services, industry, and individual hospitals with the mission
-of improving MCS outcomes. In 2018, INTERMACS became an official Society
-of Thoracic Surgeons database.
+INTERMACS is a North American observational registry for patients
+receiving MCS devices that began as a partnership between the National
+Heart, Lung, and Blood Institute, US Food and Drug Administration, the
+Centers for Medicaid and Medicare Services, industry, and individual
+hospitals with the mission of improving MCS outcomes. In 2018, INTERMACS
+became an official Society of Thoracic Surgeons database.
 
 The current analysis was conducted using publicly available data
-provided by the National Heart, Lung, and Blood Institute. We included a
+provided by the National Heart, Lung, and Blood Institute (see
+\url{https://biolincc.nhlbi.nih.gov/studies/intermacs/}). We included a
 contemporary cohort of 14,738 patients who received continuous flow LVAD
-from 2012-2017. Patient follow-up begins after implantation of a
-durable, long term MCS device and continues while the device is in
-place. Registry endpoints include death on a device, heart
-transplantation, or cessation of support (for recovery and non-recovery
-reasons). INTERMACS collects pre-implant patient characteristics,
-medical status, laboratory values, and many other variables. Data is
-collected at regularly scheduled follow-up as well as during adverse
-events such as re-hospitalization. This is a secondary analysis of
-de-identified data obtained from the National Heart Lung and Blood
-Institute. Primary data collection is approved through University of
-Alabama Institutional Review Board and at individual sites.
+from 2012-2017. This is a secondary analysis of de-identified data
+obtained from the National Heart Lung and Blood Institute. Primary data
+collection is approved through University of Alabama Institutional
+Review Board and at individual sites.
+
+\hypertarget{outcomes-and-predictors}{%
+\subsection{Outcomes and Predictors}\label{outcomes-and-predictors}}
+
+Patient follow-up begins after implantation of a durable, long term MCS
+device and continues while the device is in place. Registry endpoints
+include death on a device, heart transplantation, or cessation of
+support (for recovery and non-recovery reasons). Mortality and
+transplant after MCS were the primary outcomes for the current study. As
+there were only 310 cessation of support events, we did not analyze this
+outcome. For mortality and transplant risk prediction models, we applied
+cause-specific analysis to account for competing risks. For example, in
+risk prediction models for mortality, patients were censored at time of
+loss to follow-up, transplant, or cessation of support, whichever
+occurred first. INTERMACS collects pre-implant data on patient
+characteristics, medical status, and laboratory values. INTERMACS also
+collects follow-up data at regularly scheduled visits and at occurrence
+of adverse events such as re-hospitalization. For the current analysis,
+all pre-implant variables were considered as potential predictors.
 
@@
 
-Previous research has thoroughly investigated statistical inference in
-the context of missing data. In this setting, analysts often create
-multiple imputed datasets, replicate an analysis in each of them, and
-then pool their results to obtain valid test statistics for hypothesis
-testing \cite{rubin2004multiple}. Imputation strategies that create a
-single imputed dataset (\textit{e.g., }imputation to the mean) have been
-shown to increase type I errors (\textit{i.e., }rejecting a true null
-hypothesis) for inferential statistics by artificially reducing the
-variance of observed data and ignoring the uncertainty attributed to
-missing values \cite{van2018flexible}. To ensure valid inference,
-imputation models should leverage outcome variables to predict missing
-predictors \cite{sterne2009multiple}. When very few data are missing,
-analysts may apply listwise deletion, \textit{i.e., }removing any
-observation with at least one missing value. However, listwise deletion
-can easily lead to biased inference \cite{van2020rebutting}.
+To conduct statistical inference in the context of missing data setting,
+analysts often create multiple imputed datasets, replicate an analysis
+in each of them, and then pool their results to obtain valid test
+statistics for hypothesis testing \cite{rubin2004multiple}. Imputation
+strategies that create a single imputed dataset
+(\textit{e.g., }imputation to the mean) have been shown to increase type
+I errors (\textit{i.e., }rejecting a true null hypothesis) for
+inferential statistics by artificially reducing the variance of observed
+data and ignoring the uncertainty attributed to missing values
+\cite{van2018flexible}. To ensure valid inference, imputation models
+should leverage outcome variables to predict missing predictors
+\cite{sterne2009multiple}. When very few data are missing, analysts may
+apply listwise deletion, \textit{i.e., }removing any observation with at
+least one missing value. However, listwise deletion can easily lead to
+biased inference \cite{van2020rebutting}.
 
@@
 Several of the imputation methods we considered allowed for creation of
-one or many imputed datasets. For downstream models fitted to multiple
-sets of imputed data, we applied the modeling technique to each imputed
-training set, separately, and then used a pooling technique to generate
-a single set of predictions based on multiple imputed testing data.
-Specifically, we created 10 imputed training and testing sets, then
-applied a modeling procedure to each imputed training set and computed
-model predictions on the corresponding imputed testing set, which led to
-10 sets of predictions. To aggregate these predictions, we computed the
-median for each observation. Informal experiments where the mean was
-used instead of the median to aggregate predictions showed little or no
-difference in the model's prediction accuracy.
+single or multiple imputed datasets. For downstream models fitted to
+multiple sets of imputed data, we applied the modeling technique to each
+imputed training set, separately, and then used a pooling technique to
+generate a single set of predictions based on multiple imputed testing
+data. Specifically, we created 10 imputed training and testing sets,
+then applied a modeling procedure to each imputed training set and
+computed model predictions on the corresponding imputed testing set,
+which led to 10 sets of predictions. To aggregate these predictions, we
+computed the median for each patient. Informal experiments where the
+mean was used instead of the median to aggregate predictions showed
+little or no difference in the model's prediction accuracy.
 
@@
 as a reference for comparison of all other strategies to impute missing
-values,
+values.
 
@@
 iteration of MICE, each specified variable in the dataset is imputed
-using the other variables in the dataset. These iterations are run until
+using other variables in the dataset. These iterations are run until
 convergence criteria have been met.
@@
 \hypertarget{evaluating-imputation-accuracy}{%
-\subsection{Evaluating imputation
-accuracy}\label{evaluating-imputation-accuracy}}
+\subsection{Evaluating Imputation
+Accuracy}\label{evaluating-imputation-accuracy}}
 
@@
 mean will always have a score of 0. This modification makes it easier to
-compare the accuracy of each imputation strategy directly with
+compare the accuracy of each imputation strategy directly with that of
 imputation to the mean.
@@
 \hypertarget{computational-details}{%
-\subsection{Computational details}\label{computational-details}}
+\subsection{Computational Details}\label{computational-details}}
 
@@
 
-Our source code is available on GitHub (see
-\url{https://github.com/bcjaeger/INTERMACS-missing-data}).
+SAS software (version 9.4) and Python (version 3.8.2) were used to
+create analytic data for the current study
+\cite{van1995python, sas2013}. Base R (version 4.0.3) was used in
+combination with a number of open-source R packages
+(\textit{e.g., }\texttt{drake}, \texttt{tidyverse}, \texttt{naniar},
+\texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to
+conduct our analysis
+\cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
+Our R code is available on GitHub (see
+\url{https://github.com/bcjaeger/INTERMACS-missing-data})
+\cite{byron_2020_4247449}. We used Cheaha, a high performance computing
+cluster at University of Alabama at Birmingham, to perform the 200
+Monte-Carlo cross validation runs \cite{cheaha}.
 
@@
 
-\paragraph{Bayesian analysis of model performance}
+\hypertarget{bayesian-analysis-of-model-performance}{%
+\subsection{Bayesian analysis of model
+performance}\label{bayesian-analysis-of-model-performance}}
 
@@
 imputation to the mean was maximized by using multiple imputation with
-random forests (probability of improvement: 0.998; Figure
+random forests (probability of improvement: 1.000; Figure
 \ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
@@
 of improving the C-index in comparison to using imputation to the mean
-(probability of improvement: 0.962; Figure
+(probability of improvement: 1.000; Figure
 \ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
@@
 
-In this article, we leveraged data from the INTERMACS registry to
-evaluate how the use of different imputation strategies prior to fitting
-a risk prediction model would impact the external prognostic accuracy of
-the model. External prognostic accuracy was measured at 6 months after
-receiving MCS to focus on short term risk prediction, and the primary
-measure of accuracy was the scaled Brier score. We evaluated the
-performance of 12 imputation strategies in a broad range of settings by
-varying the type of modeling algorithm, the outcome variable, and the
-amount of additional missing data amputed prior to performing
-imputation. Our resampling experiment indicated that conducting multiple
-imputation has a high likelihood of increasing the downstream scaled
-Brier score and C-index of risk prediction models compared with
-imputation to the mean. Additionally, multiple imputation with random
-forests emerged as the imputation strategy that maximized the
-probability of developing a more prognostic model compared with
+In this article, we leveraged INTERMACS registry data to evaluate how
+the use of different imputation strategies prior to fitting a risk
+prediction model would impact the external prognostic accuracy of the
+model. External prognostic accuracy was measured at 6 months after
+receiving MCS, and the primary measure of accuracy was the scaled Brier
+score. We evaluated the performance of 12 imputation strategies in a
+broad range of settings by varying (1) the amount of additional missing
+data amputed prior to performing imputation, (2) the type of risk
+prediction model applied after imputation, and (3) the outcome variable
+for the risk prediction model. Our resampling experiment indicated that
+conducting multiple imputation has a high likelihood of increasing the
+downstream scaled Brier score and C-index of risk prediction models
+compared with imputation to the mean. Additionally, multiple imputation
+with random forests emerged as the imputation strategy that maximized
+the probability of developing a more prognostic model compared with
 imputation to the mean.
@@
 prediction model
-\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society}.
-The current analysis has shown that the prognostic accuracy of these
-models can be increased by implementing a different strategy to impute
-missing values, particularly if multiple imputation is applied. However,
-few studies have investigated best practices on the use of multiple
-imputation for prediction models, and few software packages allow users
-to implement multiple imputation procedures that adhere to best
-practices for prediction model development (\textit{e.g., }using only
-the training data to impute missing values). Further development of open
-source software that allows investigators to safely and succinctly
-implement multiple imputation for prediction models (\textit{e.g., }the
-\texttt{miceRanger} package) may increase the presence of multiple
-imputation in published risk prediction models.
+\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
+An interesting recent study indicates that imputation to the mean can
+provide an asymptotically consistent prediction model, given the
+prediction model is flexible and non-linear \cite{josse2019consistency}.
+However, theoretical results for finite samples have not yet been
+established. Our results provide relevant data for the finite sample
+case, suggesting that using imputation strategies considered in the
+current study instead of imputation to the mean can improve the
+prognostic accuracy of downstream models, particularly if multiple
+imputation is applied. Imputation to the mean should be avoided in
+future analyses of the INTERMACS registry and analyses where inflexible
+models are applied.
+
+Previous research has also established evidence in favor of applying
+multiple imputation to improve the prognostic value of risk prediction
+models. For example, Hassan and Atiya demonstrated superior downstream
+prediction using an ensemble multiple imputation method on synthetic
+data with continuous outcomes \cite{hassan2007regression}. Similarly,
+Nanni et. al demonstrated superior performance in downstream prediction
+when missing values were imputed using their proposed ensemble multiple
+imputation method \cite{nanni2012classifier}. Notably, the authors
+artificially induced missing values in these studies and the largest
+real dataset that was evaluated contained less than 700 observations. An
+article by Jerez et. al evaluated missing data strategies based on the
+downstream task of fitting a neural network and predicting early breast
+cancer relapse \cite{jerez2010missing}. The authors found that KNN
+imputation led to risk prediction models with the highest discrimination
+and lowest calibration error. Results from the current study are
+consistent with these previous findings but also extend their results by
+providing evidence from a larger source of data
+(\textit{i.e., }INTERMACS) and dealing with `real-world' missing values.
+
+Others have previously evaluated imputation techniques based on the
+accuracy with which these techniques impute missing values in the
+training data
+\cite{tutz2015improved, little2013joys, steele2018machine}. While it is
+intuitive to hypothesize that more accurate imputation will provide more
+prognostic downstream models, our results do not support this
+supposition. For example, when an additional 30\% of missing data were
+amputed, none of the missing data strategies we implemented obtained
+higher accuracy than imputation to the mean. However, using \emph{any}
+of the multiple imputation strategies we considered instead of
+imputation to the mean increased prognostic accuracy of downstream
+models when an additional 30\% of missing data were amputed. This result
+is likely explained by the bias variance tradeoff. In particular, single
+imputation techniques may lead to prediction models with lower bias but
+higher variance than multiple imputation techniques.
+
+\paragraph{Strengths and limitations}
+
+The current analysis has a number of strengths. We leveraged the
+INTERMACS data registry, comprising one of the largest cohorts of
+patients who received MCS. We applied a well known resampling method to
+internally validate modeling algorithms for risk prediction. Last, we
+made our analysis R code available in a public repository
+(\url{https://github.com/bcjaeger/INTERMACS-missing-data}). Last, the
+approach presented in this paper provides a general framework that can
+be applied in other studies where missing data are imputed prior to
+fitting a risk prediction model. The current analysis should also be
+interpreted in the context of known limitations. We considered a small
+subset of existing strategies to impute missing data, and other
+strategies may have provided stronger improvements compared with
+imputation to the mean. The models we studied obtained low values of
+scaled Brier score, indicating low prediction accuracy. It is unclear
+how results may vary for models with higher prediction accuracy. Also,
+it was not feasible to use only the training data to impute missing
+values in the testing data due to a lack of available software. Although
+the \texttt{miceRanger} package allows imputation of new data using
+existing models, few software packages for imputation allow users to
+implement multiple imputation with this protocol. Future analyses should
+introduce more flexible software and hands-on tutorials so that future
+investigators can optimize imputation of missing data.
+
+\paragraph{Conclusion}
+
+Selecting an optimal strategy to impute missing values can impact the
+prognostic accuracy of downstream risk prediction models. In the current
+analysis, conducting multiple imputation using random forests emerged as
+an optimal strategy to impute missing values in the INTERMACS data. This
+investigation can directly inform future analyses of INTERMACS data and
+provide evidence quantifying the benefit of imputing missing data with
+sound methodology.
 
@@
 No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
-months\_post\_implant & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
 Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
@@
 \hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
+\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
 \addlinespace
-\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
 Race &  &  &  &  & \\
@@
 \hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
+Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
 \addlinespace
-Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
 Device strategy &  &  &  &  & \\
@@
 \hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
+LVEDD & 6.83 (1.10) & 6.85 (1.07) & 6.49 (0.93) & 6.69 (1.12) & 6.95 (1.12)\\
 \addlinespace
-LVEDD & 6.83 (1.10) & 6.85 (1.07) & 6.49 (0.93) & 6.69 (1.12) & 6.95 (1.12)\\
 Urinary albumin, g/dl & 3.41 (0.64) & 3.42 (0.61) & 3.41 (0.64) & 3.33 (0.66) & 3.49 (0.65)\\
@@
 Periphal edema & 35 & 33 & 37 & 44 & 31\\
+BUN, mg/dl & 29 (18) & 28 (17) & 22 (13) & 32 (19) & 28 (18)\\
 \addlinespace
-BUN, mg/dl & 29 (18) & 28 (17) & 22 (13) & 32 (19) & 28 (18)\\
 Bilirubin levels, mg/dl & 1.37 (1.84) & 1.25 (1.14) & 1.28 (1.95) & 1.52 (2.46) & 1.41 (1.97)\\
@@
 \hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
+Surgery time, minutes & 292 (113) & 289 (110) & 272 (96) & 320 (128) & 272 (98)\\
 \addlinespace
-Surgery time, minutes & 292 (113) & 289 (110) & 272 (96) & 320 (128) & 272 (98)\\
 CPB time, minutes & 95 (49) & 93 (45) & 88 (38) & 103 (56) & 90 (44)\\
@@
 \begin{table} 
- \caption{Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.} 
+ \caption{Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.
+Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.}
 \label{tbl_missingness}
@@
 \begin{table} 
- \caption{Accuracy of strategies to impute artificial missing data} 
+ \caption{Accuracy of strategies to impute artificial missing data.
+Table values are the median change in accuracy (25th, 75th percentile) relative to the accuracy of imputation to the mean.
+In general, multiple imputation strategies had lower accuracy than single imputation strategies, and few imputation strategies were more accurate than imputation to the mean.}
 \label{tbl_impute_accuracy}
@@
 \begin{table} 
- \caption{Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+ \caption{Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
 Table values show the scaled Brier score for imputation to the mean.
 For other imputation strategies, table values show the change in scaled Brier score relative to the scaled Brier score when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability}
+All table values are multiplied by 100 for ease of interpretability.
+Multiple imputation with random forests leads to the highest scaled Brier score for both models and when 0\%, 15\%, or 30\% of additional data in the training and testing sets were set to missing.}
 \label{tbl_md_strat_dead_ipa}
@@
 \begin{table} 
- \caption{Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+ \caption{Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
 Table values show the scaled Brier score for imputation to the mean.
 For other imputation strategies, table values show the change in scaled Brier score relative to the scaled Brier score when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability}
+All table values are multiplied by 100 for ease of interpretability.
+While there was very little difference in scaled Brier score values when 0\% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher scaled Brier scores when 15\% or 30\% of additional missing data were amputed.}
 \label{tbl_md_strat_txpl_ipa}
@@
 \begin{table} 
- \caption{Median (25th, 75th percentile) change in concordance index when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+ \caption{Median (25th, 75th percentile) change in concordance index when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
 Table values show the concordance index for imputation to the mean.
 For other imputation strategies, table values show the change in concordance index relative to the concordance index when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability}
+All table values are multiplied by 100 for ease of interpretability.
+Multiple imputation with random forests led to the highest concordance index for boosting models when 0\%, 15\%, or 30\% of additional data in the training and testing sets were set to missing.
+For proportional hazards models, multiple imputation with nearest neighbors or random forests was the most effective strategy.}
 \label{tbl_md_strat_dead_auc}
@@
 \begin{table} 
- \caption{Median (25th, 75th percentile) change in calibration error when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+ \caption{Median (25th, 75th percentile) change in calibration error when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
 Table values show the calibration error for imputation to the mean.
 For other imputation strategies, table values show the change in calibration error relative to the calibration error when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability}
+All table values are multiplied by 100 for ease of interpretability.
+As more data in the training and testing sets were set to missing, single and multiple imputation with random forests emerged as the strategies with the lowest calibration error.}
 \label{tbl_md_strat_dead_cal_error}
@@
 \begin{table} 
- \caption{Median (25th, 75th percentile) change in concordance index when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+ \caption{Median (25th, 75th percentile) change in concordance index when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
 Table values show the concordance index for imputation to the mean.
 For other imputation strategies, table values show the change in concordance index relative to the concordance index when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability}
+All table values are multiplied by 100 for ease of interpretability.
+While there was very little difference in concordance index when 0\% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher concordance indices when 15\% or 30\% of additional missing data were amputed.}
 \label{tbl_md_strat_txpl_auc}
@@
 \begin{table} 
- \caption{Median (25th, 75th percentile) change in calibration error when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+ \caption{Median (25th, 75th percentile) change in calibration error when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
 Table values show the calibration error for imputation to the mean.
 For other imputation strategies, table values show the change in calibration error relative to the calibration error when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability}
+All table values are multiplied by 100 for ease of interpretability.
+As more data in the training and testing sets were set to missing, single and multiple imputation with nearest neighbors emerged as the optimal strategy for proportional hazards models while missingness incorporated as an attributed emerged an the optimal strategy for boosting models.}
 \label{tbl_md_strat_txpl_cal_error}
@@
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
-Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution.}\label{fig:fig_md_strat_infer_ipa}
+Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy, is printed to the right of each distribution.
+Each multiple imputation strategy and single imputation with missingness incorporated as an attribute had over 90\% posterior predicted probability of increasing the scaled Brier score versus using imputation to the mean.}\label{fig:fig_md_strat_infer_ipa}
 \end{figure}
@@
 
-\caption{Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+\caption{Posterior distribution of differences in concordance index values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
-Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution.}\label{fig:fig_md_strat_infer_auc}
+Posterior probability that the difference in concordance index exceeds 0, indicating an improvement in model discrimination, is printed to the right of each distribution.
+Multiple imputation with predictive mean matching, random forests, and Bayesian regression each had over 90\% posterior predicted probability of increasing the concordance index versus using imputation to the mean.}\label{fig:fig_md_strat_infer_auc}
 \end{figure}
@@
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
-Posterior probability that the difference in calibration error exceeds 0, indicating an improvement in overall model accuracy for the current imputation strategy, is printed to the right of each distribution.}\label{fig:fig_md_strat_infer_cal_error}
+Posterior probability that the difference in calibration error is less than 0, indicating an improvement in model calibration, is printed to the right of each distribution.
+Every imputation strategy evaluated had over 90\% posterior predicted probability of improving model calibration versus using imputation to the mean.}\label{fig:fig_md_strat_infer_cal_error}
 \end{figure}

### 2020-11-10 R/make_bayes_mccv_fit.R

@@ -24,11 +24,10 @@ make_bayes_mccv_fit <- function(risk_evaluation) {
 
   for(i in seq_along(output)){
 
-    output[[i]] <- stan_lm(
-      formula = value ~ md_strat + additional_missing_pct,
-      #+ (1 | iteration),
+    output[[i]] <- stan_lmer(
+      formula = value ~ md_strat + additional_missing_pct + (1 | iteration),
       data = model_data[[i]],
-      prior = R2(0.50),
+      #prior = R2(0.50),
       iter = 5000
     )
 

### 2020-11-10 R/tabulate_descriptives.R

@@ -46,7 +46,7 @@ tabulate_descriptives <- function(im) {
     ) %>%
     select(
       status,
-      months_post_implant,
+      #months_post_implant,
       demo_age,
       demo_gender,
       demo_race,

### 2020-11-10 R/visualize_md_strat_inference.R

@@ -12,17 +12,17 @@ visualize_md_strat_inference <- function(bayes_mccv_fits,
 
   get_xlevels <- function(posterior) {
 
-    if(inherits(posterior, 'lm')){
-
-      posterior$xlevels$md_strat
-
-    } else {
+    if(inherits(posterior, 'lmerMod')){
 
       posterior$x@Dimnames[[2]] %>%
         enframe() %>%
         filter(str_detect(value, '^md_strat')) %>%
         pull(value) %>%
         str_remove('md_strat')
+
+    } else {
+
+      posterior$xlevels$md_strat
 
     }
 
@@ -135,8 +135,8 @@ visualize_md_strat_inference <- function(bayes_mccv_fits,
           ) %>%
           mutate(m2 = factor(m2, levels = rev(names(md_method_labels)[-1])))
 
-        probs_upr <- 0.999
-        probs_lwr <- 0.950
+        probs_upr <- 0.990
+        probs_lwr <- 0.940
 
         if(str_detect(.y, 'calibration')){
           probs_lwr <- 0.0010

### 2020-11-10 packages.R

@@ -19,7 +19,6 @@ library(knitr)
 library(kableExtra)
 library(gtsummary)
 library(gt)
-library(tibbleOne)
 ## data analysis
 library(tidyverse)
 library(rstanarm)

### 2020-11-19 doc_arXiv/response_to_reviewers.Rmd
feedback from RC

@@
 ---
+
+We thank the editors for the opportunity to respond to our reviewers and re-submit our manuscript.
+In order to respond to the reviewers' excellent feedback, we have made extensive changes to the manuscript.
+Due to the high volume of changes (nearly every sentence was updated), we have not used track changes in the update.
+Below we outline our response to each reviewer, point by point.
 

### 2020-11-19 README.Rmd
feedback from RC

@@
 This repository contains all of the code used in the paper _Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values_.
+
+# ABSTRACT
+
+Risk prediction models play an important role in clinical decision making.
+When developing risk prediction models, practitioners often impute missing values to the mean.
+We evaluated the impact of applying other strategies to impute missing values on the prognostic accuracy of downstream risk prediction models, i.e., models fitted to the imputed data.
+A secondary objective was to compare the accuracy of imputation methods based on artificially induced missing values.
+To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).
+
+## Methods and Results
+
+We applied twelve imputation strategies in combination with two different modeling strategies for mortality and transplant risk prediction following surgery to receive mechanical circulatory support.
+Model performance was evaluated using Monte-Carlo cross validation and measured based on outcomes 6-months following surgery using the scaled Brier score, concordance index, and calibration error.
+We used Bayesian hierarchical models to compare model performance.
+Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.0008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
+The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively.
+
+## Conclusion
+
+Selecting an optimal strategy to impute missing values and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.

### 2020-11-19 doc_arXiv/doc_arXiv.Rmd
feedback from RC

@@
   - name: Ryan Cantor, PhD
+    department: Kirklin Institute for Research in Surgical Outcomes
     affiliation: University of Alabama at Birmingham
@@
   - name: Venkata Sthanam, MS
+    department: Kirklin Institute for Research in Surgical Outcomes
     affiliation: University of Alabama at Birmingham
@@
   - name: Rongbing Xie, DrPH
+    department: Department of Surgery
     affiliation: University of Alabama at Birmingham
@@
   - name: James K. Kirklin, MD
+    department: Kirklin Institute for Research in Surgical Outcomes
     affiliation: University of Alabama at Birmingham
@@
   - name: Ramaraju Rudraraju, PhD
+    department: Kirklin Institute for Research in Surgical Outcomes
     affiliation: University of Alabama at Birmingham
@@
 We used Bayesian hierarchical models to compare model performance.
-Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
-The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.482 and 0.962, respectively.
+Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.0008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
+The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively.
 \newline\textbf{Conclusion} Selecting an optimal strategy to impute missing values and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
 keywords:
-  - Missing Data
+  - missing data
   - INTERMACS
@@
   - heart failure
-  - mortality
+  - mechanical circulatory support
   - risk prediction
@@
 
-source(file.path(here::here(), 'packages.R'))
+# source(file.path(here::here(), 'packages.R'))
+# R.utils::sourceDirectory('R')
 
@@
 As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
-A critical step for developing risk prediction models is handling missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial.
+An important step for developing risk prediction models is dealing with missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial.
 While some missing values are related to data entry error, others may relate to instability of a patient’s circulatory status at the time of implant.
@@
 The primary aim of this article is to quantify how much the prognostic value of a risk prediction model developed from the INTERMACS registry depends on the strategy that was applied to impute missing data prior to developing the model.
-A secondary aim is to measure imputation accuracy of each strategy by introducing varying levels of artificial missing data and then imputing it.
+A secondary aim is to measure imputation accuracy of each strategy by introducing varying levels of artificial missing data (\ie data amputation) and then imputing it.
 Because imputation to the mean has been a standard method for multiple annual summaries of the INTERMACS data, we measure the potential improvement in prognostic value of a risk prediction model when other strategies are applied to impute missing data \emph{instead of imputation to the mean}.
@@
 
-Patient follow-up begins after implantation of a durable, long term MCS device and continues while the device is in place.
+Patient follow-up begins after implantation of a durable, long term MCS device and continues while a device is in place.
 Registry endpoints include death on a device, heart transplantation, or cessation of support (for recovery and non-recovery reasons).
@@
 As there were only `r sum(im$pt_outcome_cess)` cessation of support events, we did not analyze this outcome.
-For mortality and transplant risk prediction models, we applied cause-specific analysis to account for competing risks.
-For example, in risk prediction models for mortality, patients were censored at time of loss to follow-up, transplant, or cessation of support, whichever occurred first.
+For mortality and transplant risk prediction models, we applied event-specific analysis to account for competing risks.
+For example, in risk prediction models for mortality, patients were censored at time of last contact, transplant, or cessation of support, whichever occurred first.
 INTERMACS collects pre-implant data on patient characteristics, medical status, and laboratory values.
@@
 Imputation accuracy was computed for each numerical and nominal variable, separately.
+The observed values of data that were amputed were used to assess imputation accuracy.
+As a consequence, we only measured imputation accuracy 15\% or 30\% of additional missing values were amputed.
+In the context of this article, the term 'amputation of data' means artificially making observed values missing.
 Numeric variable imputation accuracy was measured using a re-scaled mean-squared error: $$1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.$$ This score is greater than 0 if $\textrm{MSE}(\textrm{current imputation method})$ is smaller than $\textrm{MSE}(\textrm{imputation to the mean})$, equal to 0 if the two MSEs are equal, and less than 0 if $\textrm{MSE}(\textrm{current imputation method})$ is greater than $\textrm{MSE}(\textrm{imputation to the mean})$.
@@
 After imputation, Cox PH and boosting models were applied to each imputed dataset, separately.
-Last, model predictions for death and transplant were computed at 6 months following MCS surgery.
+Last, model predictions for death and transplant were computed at `r times` months following MCS surgery.
 
@@
 SAS software (version 9.4) and Python (version 3.8.2) were used to create analytic data for the current study \cite{van1995python, sas2013}.
-Base R (version 4.0.3) was used in combination with a number of open-source R packages (\eg \texttt{drake}, \texttt{tidyverse}, \texttt{naniar}, \texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to conduct our analysis \cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
+These analyses were completed using The American Heart Association Precision Medicine Platform (https://precision.heart.org/).
+Base R (version 4.0.3) was used in combination with a number of open-source R packages (\eg \texttt{drake}, \texttt{tidyverse}, \texttt{naniar}, \texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to conduct statistical analyses and create the current manuscript \cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
 Our R code is available on GitHub (see [https://github.com/bcjaeger/INTERMACS-missing-data](https://github.com/bcjaeger/INTERMACS-missing-data)) \cite{byron_2020_4247449}.
 We used Cheaha, a high performance computing cluster at University of Alabama at Birmingham, to perform the 200 Monte-Carlo cross validation runs \cite{cheaha}.
-
-<!-- We  implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}.
-\texttt{miceRanger} follows an iterative procedure where one random forest is fit to each variable in a dataset, separately, and then used to impute missing values in that variable.
-The procedure may repeat until estimated generalization error of the random forest collection stabilizes.
--->
 
@@
                  pt_outcome_dead == 0 &
-                 pt_outcome_txpl ==0 & 
+                 pt_outcome_txpl == 0 & 
                  months_post_implant < times),
@@
 
-\paragraph{Patient Characteristics} During the first `r times` months after receiving MCS, the number (% of overall population) of deaths, transplants, and loss-to-follow-up events was `r trunc_events$dead`, `r trunc_events$txpl`, and `r trunc_events$cens`, respectively.
+<!-- update this to reflect administrative censoring instead of ltfu -->
+
+\paragraph{Patient Characteristics} During the first `r times` months after receiving MCS, `r trunc_events$dead` patients died and `r trunc_events$txpl` had a transplant.
+In total, `r trunc_events$cens` patients had a follow-up time < `r times` months and were considered right-censored in the current analysis.
 The mean (standard deviation) age of patients was `r age_overall` years, `r white_overall`% of patients identified as white and `r male_overall`% were male (Table \ref{tbl_characteristics}).
@@
 However, the number (percent) of missing values for surgery time was an exception, with `r inline_missingness$Overall$im_surgery_time` in the overall population, `r inline_missingness$Dead$im_surgery_time` among patients who died, and `r inline_missingness$Censored$im_surgery_time` among patients who were censored.
+This pattern is likely attributable to the later introduction of surgery time into the INTERMACS registry compared to other variables.
+The collection of surgery time in the INTERMACS registry began on May 30, 2014.
 Additionally, missing values for surgery time were frequently accompanied by missing values for CV pressure (Figure \ref{fig:upset}).
@@
 Similarly, multiple imputation using random forests was estimated to have the highest posterior probability of improving the C-index in comparison to using imputation to the mean (probability of improvement: `r inline_auc_rf`; Figure \ref{fig:fig_md_strat_infer_auc}).
-However, the estimated posterior probability of a reduction in calibration was 1 when using either single or multiple imputation with random forests compared with imputation to the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}).
+However, the estimated posterior probability of a reduction in calibration was >0.999 when using either single or multiple imputation with random forests compared with imputation to the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}).
 Although imputation using random forest was estimated to be the best overall option, there was moderate to strong evidence that MIA and each multiple imputation strategy we applied would improve prognostic accuracy of downstream models compared with imputation to the mean.
@@
 This investigation can directly inform future analyses of INTERMACS data and provide evidence quantifying the benefit of imputing missing data with sound methodology.
+\paragraph{Sources of Funding} This work is supported by the American Heart Association, grant# 18AIML34280052.
+De-identified data provided from the INTERMACS contract with from the National Heart, Lung, and Blood Institute; National Institutes of Health; and Department of Health and Human Services, under contract# HHSN268201100025C.
 
@@
 \begin{table}
-\caption{Participant characteristics stratified by event status. The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
+\footnotesize
+\caption{Participant characteristics overall and stratified by event status at `r times` months following surgery to receive mechanical circulatory support.
+The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
 \label{tbl_characteristics}
@@
   booktabs = TRUE, 
-  align = 'lccccc'
+  align = 'lcccccc'
 ) %>% 
@@
     caption = 'Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.
-Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.',
-    label = 'tbl_missingness'
+Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.
+This pattern is likely due to the fact that surgery time was added to the INTERMACS data after 2012',
+    label = 'tbl_missingness',
+    size = 'footnotesize'
   )
@@
 \clearpage
+
+<!-- TODO: Add (reference) to imputation to the mean bracket -->
 
@@
 
-```{r fig_md_strat_infer_ipa, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+```{r fig_md_strat_infer_ipa, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
@@
 
-```{r fig_md_strat_infer_auc, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in concordance index values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+```{r fig_md_strat_infer_auc, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in concordance index values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
@@
 
-```{r fig_md_strat_infer_cal_error, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in calibration error values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+```{r fig_md_strat_infer_cal_error, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in calibration error values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.

### 2020-11-19 doc_arXiv/doc_arXiv.tex
feedback from RC

@@
    \\
+    Kirklin Institute for Research in Surgical Outcomes \\
     University of Alabama at Birmingham \\
@@
    \\
+    Department of Surgery \\
     University of Alabama at Birmingham \\
@@
 strategy to impute missing values, increasing model concordance by 0.003
-(25th, 75th percentile: 0.008, 0.052) compared with imputation to the
+(25th, 75th percentile: 0.0008, 0.052) compared with imputation to the
 mean for mortality risk prediction using a downstream proportional
@@
 imputation using random forests would improve concordance versus mean
-imputation was 0.482 and 0.962, respectively.
+imputation was 0.464 and \textgreater0.999, respectively.
 \newline\textbf{Conclusion} Selecting an optimal strategy to impute
@@
 \keywords{
-    Missing Data
+    missing data
    \and
@@
    \and
-    mortality
+    mechanical circulatory support
    \and
@@
 \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
-A critical step for developing risk prediction models is handling
+An important step for developing risk prediction models is dealing with
 missing data, which is common in a ``real world'' database that is not
@@
 of each strategy by introducing varying levels of artificial missing
-data and then imputing it. Because imputation to the mean has been a
-standard method for multiple annual summaries of the INTERMACS data, we
-measure the potential improvement in prognostic value of a risk
-prediction model when other strategies are applied to impute missing
-data \emph{instead of imputation to the mean}. The over-arching aim is
-to clarify which imputation strategies are most likely to improve risk
-prediction (and by extension, quality of care) for patients who receive
-MCS devices. This investigation can directly inform future analyses of
-INTERMACS data and provide evidence quantifying the benefit of imputing
-missing data with sound methodology.
+data (\textit{i.e., } data amputation) and then imputing it. Because
+imputation to the mean has been a standard method for multiple annual
+summaries of the INTERMACS data, we measure the potential improvement in
+prognostic value of a risk prediction model when other strategies are
+applied to impute missing data \emph{instead of imputation to the mean}.
+The over-arching aim is to clarify which imputation strategies are most
+likely to improve risk prediction (and by extension, quality of care)
+for patients who receive MCS devices. This investigation can directly
+inform future analyses of INTERMACS data and provide evidence
+quantifying the benefit of imputing missing data with sound methodology.
 
@@
 Patient follow-up begins after implantation of a durable, long term MCS
-device and continues while the device is in place. Registry endpoints
+device and continues while a device is in place. Registry endpoints
 include death on a device, heart transplantation, or cessation of
@@
 outcome. For mortality and transplant risk prediction models, we applied
-cause-specific analysis to account for competing risks. For example, in
+event-specific analysis to account for competing risks. For example, in
 risk prediction models for mortality, patients were censored at time of
-loss to follow-up, transplant, or cessation of support, whichever
-occurred first. INTERMACS collects pre-implant data on patient
-characteristics, medical status, and laboratory values. INTERMACS also
-collects follow-up data at regularly scheduled visits and at occurrence
-of adverse events such as re-hospitalization. For the current analysis,
-all pre-implant variables were considered as potential predictors.
+last contact, transplant, or cessation of support, whichever occurred
+first. INTERMACS collects pre-implant data on patient characteristics,
+medical status, and laboratory values. INTERMACS also collects follow-up
+data at regularly scheduled visits and at occurrence of adverse events
+such as re-hospitalization. For the current analysis, all pre-implant
+variables were considered as potential predictors.
 
@@
 strategies that create a single imputed dataset
-(\textit{e.g., }imputation to the mean) have been shown to increase type
-I errors (\textit{i.e., }rejecting a true null hypothesis) for
+(\textit{e.g., } imputation to the mean) have been shown to increase
+type I errors (\textit{i.e., } rejecting a true null hypothesis) for
 inferential statistics by artificially reducing the variance of observed
@@
 \cite{sterne2009multiple}. When very few data are missing, analysts may
-apply listwise deletion, \textit{i.e., }removing any observation with at
-least one missing value. However, listwise deletion can easily lead to
-biased inference \cite{van2020rebutting}.
+apply listwise deletion, \textit{i.e., } removing any observation with
+at least one missing value. However, listwise deletion can easily lead
+to biased inference \cite{van2020rebutting}.
 
@@
 accuracy, suggesting that more accurate imputation strategies lead to
-better performance of downstream models (\textit{i.e., }models fitted to
-the imputed data) \cite{jerez2010missing}. Others have shown that
+better performance of downstream models (\textit{i.e., } models fitted
+to the imputed data) \cite{jerez2010missing}. Others have shown that
 imputation to the mean, model-based imputation, and multiple imputation
@@
 KNN imputation identifies \(k\) `similar' observations
-(\textit{i.e., }neighbors) for each observation with a missing value in
+(\textit{i.e., } neighbors) for each observation with a missing value in
 a given variable \cite{chen2000nearest}. A donor value for the current
@@
 error on data outside of their bootstrapped sample
-(\textit{i.e., }out-of-bag error). In the current analysis, we conduct
+(\textit{i.e., } out-of-bag error). In the current analysis, we conduct
 MICE using one random forest to impute each variable, separately. For
@@
 \cite{twala2009empirical, ding2010investigation}. MIA adds another
-category to nominal variables: ``Missing''. For numeric variables, MIA
+category to nominal variables: ``Missing.'' For numeric variables, MIA
 creates two columns: one where missing values are imputed with positive
@@
 Imputation accuracy was computed for each numerical and nominal
-variable, separately. Numeric variable imputation accuracy was measured
-using a re-scaled mean-squared error:
+variable, separately. The observed values of data that were amputed were
+used to assess imputation accuracy. As a consequence, we only measured
+imputation accuracy 15\% or 30\% of additional missing values were
+amputed. In the context of this article, the term `amputation of data'
+means artificially making observed values missing. Numeric variable
+imputation accuracy was measured using a re-scaled mean-squared error:
 \[1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.\]
@@
 and worthless model, respectively. In our analyses, a Kaplan-Meier
-estimate based on the training data (\textit{i.e., }a risk prediction
+estimate based on the training data (\textit{i.e., } a risk prediction
 model that did not use any predictor variables) provided the naive
@@
 combination of a missing data strategy and a modeling strategy
-(\textit{e.g., }imputation using mean/mode values followed by fitting
+(\textit{e.g., } imputation using mean/mode values followed by fitting
 the PH model with stepwise variable selection) \cite{kuhn2013applied}.
@@
 for each imputation strategy. Although some imputation strategies
-(\textit{e.g., }KNN and random forests) can impute data in the testing
+(\textit{e.g., } KNN and random forests) can impute data in the testing
 set using models fitted to the training set, others
-(\textit{e.g., }Bayesian regression, PMM, and hot deck) cannot.
+(\textit{e.g., } Bayesian regression, PMM, and hot deck) cannot.
 Therefore, to ensure fair comparisons in our experiment, each imputation
@@
 create analytic data for the current study
-\cite{van1995python, sas2013}. Base R (version 4.0.3) was used in
+\cite{van1995python, sas2013}. These analyses were completed using The
+American Heart Association Precision Medicine Platform
+(\url{https://precision.heart.org/}). Base R (version 4.0.3) was used in
 combination with a number of open-source R packages
-(\textit{e.g., }\texttt{drake}, \texttt{tidyverse}, \texttt{naniar},
+(\textit{e.g., } \texttt{drake}, \texttt{tidyverse}, \texttt{naniar},
 \texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to
-conduct our analysis
+conduct statistical analyses and create the current manuscript
 \cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
@@
 
-During the first 6 months after receiving MCS, the number (\% of overall
-population) of deaths, transplants, and loss-to-follow-up events was
-1,897 (13\%), 1,067 (7.2\%), and 981 (6.7\%), respectively. The mean
-(standard deviation) age of patients was 57 (13) years, 66\% of patients
-identified as white and 79\% were male (Table
-\ref{tbl_characteristics}).
+During the first 6 months after receiving MCS, 1,897 (13\%) patients
+died and 1,067 (7.2\%) had a transplant. In total, 981 (6.7\%) patients
+had a follow-up time \textless{} 6 months and were considered
+right-censored in the current analysis. The mean (standard deviation)
+age of patients was 57 (13) years, 66\% of patients identified as white
+and 79\% were male (Table \ref{tbl_characteristics}).
 
@@
 number (percent) of missing values for surgery time was an exception,
-with 6,125 (41.6\%) in the overall population, 2,246 (53.4\%) among
-patients who died, and 1,713 (27.2\%) among patients who were censored.
-Additionally, missing values for surgery time were frequently
+with 6,125 (41.6\%) in the overall population, 795 (41.9\%) among
+patients who died, and 110 (11.2\%) among patients who were censored.
+This pattern is likely attributable to the later introduction of surgery
+time into the INTERMACS registry compared to other variables. The
+collection of surgery time in the INTERMACS registry began on May 30,
+2014. Additionally, missing values for surgery time were frequently
 accompanied by missing values for CV pressure (Figure \ref{fig:upset}).
@@
 imputation to the mean was maximized by using multiple imputation with
-random forests (probability of improvement: 1.000; Figure
+random forests (probability of improvement: \textgreater0.999; Figure
 \ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
@@
 of improving the C-index in comparison to using imputation to the mean
-(probability of improvement: 1.000; Figure
+(probability of improvement: \textgreater0.999; Figure
 \ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
-probability of a reduction in calibration was 1 when using either single
-or multiple imputation with random forests compared with imputation to
-the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}). Although
-imputation using random forest was estimated to be the best overall
-option, there was moderate to strong evidence that MIA and each multiple
-imputation strategy we applied would improve prognostic accuracy of
-downstream models compared with imputation to the mean.
+probability of a reduction in calibration was \textgreater0.999 when
+using either single or multiple imputation with random forests compared
+with imputation to the mean (Figure
+\ref{fig:fig_md_strat_infer_cal_error}). Although imputation using
+random forest was estimated to be the best overall option, there was
+moderate to strong evidence that MIA and each multiple imputation
+strategy we applied would improve prognostic accuracy of downstream
+models compared with imputation to the mean.
 
@@
 providing evidence from a larger source of data
-(\textit{i.e., }INTERMACS) and dealing with `real-world' missing values.
+(\textit{i.e., } INTERMACS) and dealing with `real-world' missing
+values.
 
@@
 
+\paragraph{Sources of Funding}
+
+This work is supported by the American Heart Association, grant\#
+18AIML34280052. De-identified data provided from the INTERMACS contract
+with from the National Heart, Lung, and Blood Institute; National
+Institutes of Health; and Department of Health and Human Services, under
+contract\# HHSN268201100025C.
+
 \clearpage
 \begin{table}
-\caption{Participant characteristics stratified by event status. The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
+\footnotesize
+\caption{Participant characteristics overall and stratified by event status at 6 months following surgery to receive mechanical circulatory support.
+The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
 \label{tbl_characteristics}
 
-\begin{tabular}{lccccc}
+\begin{tabular}{lcccccc}
 \toprule
-Characteristic & Overall & Censored & Cessation of support & Dead & Transplant\\
-\midrule
-No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
-Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
-Sex &  &  &  &  & \\
-\hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
-\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
+Characteristic & Overall & Dead & Transplant & Cessation of support & Alive on device & Censored\\
+\midrule
+No. of patients & 14,738 & 1,897 & 1,067 & 55 & 10,738 & 981\\
+Age, years & 57 (13) & 62 (12) & 54 (12) & 52 (13) & 57 (13) & 57 (13)\\
+Sex &  &  &  &  &  & \\
+\hspace{1em}Female & 21 & 23 & 23 & 31 & 21 & 22\\
+\hspace{1em}Male & 79 & 77 & 77 & 69 & 79 & 78\\
 \addlinespace
-Race &  &  &  &  & \\
-\hspace{1em}White & 66 & 63 & 64 & 71 & 67\\
-\hspace{1em}Black & 25 & 28 & 27 & 21 & 23\\
-\hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
-Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
+Race &  &  &  &  &  & \\
+\hspace{1em}White & 66 & 71 & 68 & 64 & 66 & 63\\
+\hspace{1em}Black & 25 & 20 & 20 & 27 & 26 & 26\\
+\hspace{1em}Other & 9.2 & 8.4 & 12 & 9.1 & 8.8 & 11\\
+Body mass index & 29 (7) & 29 (7) & 27 (6) & 29 (8) & 29 (7) & 28 (7)\\
 \addlinespace
-Device strategy &  &  &  &  & \\
-\hspace{1em}Other & 0.6 & 0.6 & 2.6 & 0.5 & 0.6\\
-\hspace{1em}Bridge to transplant & 52 & 43 & 51 & 40 & 81\\
-\hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
-LVEDD & 6.83 (1.10) & 6.85 (1.07) & 6.49 (0.93) & 6.69 (1.12) & 6.95 (1.12)\\
+Device strategy &  &  &  &  &  & \\
+\hspace{1em}Other & 0.6 & 0.7 & 0.3 & 5.5 & 0.5 & 1.3\\
+\hspace{1em}Bridge to transplant & 52 & 42 & 90 & 51 & 51 & 51\\
+\hspace{1em}Destination therapy & 47 & 58 & 9.5 & 44 & 49 & 48\\
+LVEDD & 6.83 (1.10) & 6.56 (1.12) & 6.86 (1.13) & 6.33 (1.03) & 6.88 (1.08) & 6.78 (1.12)\\
 \addlinespace
-Urinary albumin, g/dl & 3.41 (0.64) & 3.42 (0.61) & 3.41 (0.64) & 3.33 (0.66) & 3.49 (0.65)\\
-Urinary creatinine, mg/dl & 1.39 (0.68) & 1.36 (0.62) & 1.17 (0.53) & 1.49 (0.77) & 1.35 (0.67)\\
-CV pressure & 10.8 (6.2) & 10.4 (6.1) & 9.9 (6.0) & 11.6 (6.5) & 10.7 (6.0)\\
-Periphal edema & 35 & 33 & 37 & 44 & 31\\
-BUN, mg/dl & 29 (18) & 28 (17) & 22 (13) & 32 (19) & 28 (18)\\
+Urinary albumin, g/dL & 3.41 (0.64) & 3.26 (0.63) & 3.51 (0.61) & 3.43 (0.63) & 3.43 (0.64) & 3.40 (0.62)\\
+Urinary creatinine, mg/dL & 1.39 (0.68) & 1.54 (0.80) & 1.36 (0.70) & 1.17 (0.61) & 1.37 (0.66) & 1.36 (0.57)\\
+CV pressure & 10.8 (6.2) & 12.2 (6.7) & 10.6 (5.7) & 9.9 (6.3) & 10.7 (6.2) & 10.3 (5.7)\\
+Periphal edema & 35 & 44 & 30 & 38 & 35 & 30\\
+BUN, mg/dL & 29 (18) & 34 (20) & 28 (19) & 27 (21) & 28 (17) & 29 (19)\\
 \addlinespace
-Bilirubin levels, mg/dl & 1.37 (1.84) & 1.25 (1.14) & 1.28 (1.95) & 1.52 (2.46) & 1.41 (1.97)\\
-Device type &  &  &  &  & \\
-\hspace{1em}Bi-ventricular assistance device & 3.5 & 2.1 & 0.3 & 5.8 & 3.4\\
-\hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
-Surgery time, minutes & 292 (113) & 289 (110) & 272 (96) & 320 (128) & 272 (98)\\
+Bilirubin levels, mg/dL & 1.37 (1.84) & 1.80 (3.10) & 1.42 (2.05) & 1.47 (3.32) & 1.29 (1.53) & 1.27 (1.17)\\
+Device type &  &  &  &  &  & \\
+\hspace{1em}Bi-ventricular assistance device & 3.5 & 10 & 5.9 & 1.8 & 2.0 & 3.6\\
+\hspace{1em}Left-ventricular assistance device & 97 & 90 & 94 & 98 & 98 & 96\\
+Surgery time, minutes & 292 (113) & 337 (136) & 272 (100) & 291 (92) & 286 (107) & 289 (116)\\
 \addlinespace
-CPB time, minutes & 95 (49) & 93 (45) & 88 (38) & 103 (56) & 90 (44)\\
+CPB time, minutes & 95 (49) & 113 (63) & 92 (51) & 94 (40) & 92 (45) & 94 (46)\\
 \bottomrule
-\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
-\multicolumn{6}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
+\multicolumn{7}{l}{\rule{0pt}{1em}\textit{Note: }}\\
+\multicolumn{7}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
 \end{tabular}
@@
  \caption{Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.
-Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.}
+Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.
+This pattern is likely due to the fact that surgery time was added to the INTERMACS data after 2012}
 \label{tbl_missingness} 
-
-\begin{tabular}{lccccc}
+\footnotesize
+
+\begin{tabular}{lcccccc}
 \toprule
- & Overall & Dead & Transplant & Censored & Cessation of support \\ 
-\midrule
-CV pressure & 7,415 (50.3\%) & 2,156 (51.3\%) & 2,006 (51.2\%) & 3,108 (49.3\%) & 145 (46.8\%) \\ 
-Surgery time, minutes & 6,125 (41.6\%) & 2,246 (53.4\%) & 2,007 (51.3\%) & 1,713 (27.2\%) & 159 (51.3\%) \\ 
-Periphal edema & 5,093 (34.6\%) & 1,543 (36.7\%) & 1,470 (37.5\%) & 1,952 (30.9\%) & 128 (41.3\%) \\ 
-LVEDD & 3,155 (21.4\%) & 1,000 (23.8\%) & 866 (22.1\%) & 1,233 (19.5\%) & 56 (18.1\%) \\ 
-Urinary albumin, g/dl & 1,014 (6.88\%) & 326 (7.75\%) & 309 (7.89\%) & 359 (5.69\%) & 20 (6.45\%) \\ 
-CPB time, minutes & 1,010 (6.85\%) & 287 (6.83\%) & 274 (7.00\%) & 421 (6.67\%) & 28 (9.03\%) \\ 
-Bilirubin levels, mg/dl & 850 (5.77\%) & 291 (6.92\%) & 237 (6.05\%) & 304 (4.82\%) & 18 (5.81\%) \\ 
-Body mass index & 77 (0.52\%) & 26 (0.62\%) & 19 (0.49\%) & 31 (0.49\%) & 1 (0.32\%) \\ 
-BUN, mg/dl & 50 (0.34\%) & 10 (0.24\%) & 8 (0.20\%) & 32 (0.51\%) & 0 (0.00\%) \\ 
-Sex & 24 (0.16\%) & 9 (0.21\%) & 4 (0.10\%) & 11 (0.17\%) & 0 (0.00\%) \\ 
-Urinary creatinine, mg/dl & 20 (0.14\%) & 8 (0.19\%) & 4 (0.10\%) & 8 (0.13\%) & 0 (0.00\%) \\ 
-Age, years & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
-Race & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
-Device strategy & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
-Device type & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+ & Overall & Dead & Alive on device & Transplant & Cessation of support & Censored \\ 
+\midrule
+CV pressure & 7,415 (50.3\%) & 920 (48.5\%) & 5,507 (51.3\%) & 539 (50.5\%) & 29 (52.7\%) & 420 (42.8\%) \\ 
+Surgery time, minutes & 6,125 (41.6\%) & 795 (41.9\%) & 4,727 (44.0\%) & 470 (44.0\%) & 23 (41.8\%) & 110 (11.2\%) \\ 
+Periphal edema & 5,093 (34.6\%) & 716 (37.7\%) & 3,935 (36.6\%) & 368 (34.5\%) & 15 (27.3\%) & 59 (6.01\%) \\ 
+LVEDD & 3,155 (21.4\%) & 458 (24.1\%) & 2,274 (21.2\%) & 235 (22.0\%) & 12 (21.8\%) & 176 (17.9\%) \\ 
+Urinary albumin, g/dL & 1,014 (6.88\%) & 121 (6.38\%) & 745 (6.94\%) & 89 (8.34\%) & 4 (7.27\%) & 55 (5.61\%) \\ 
+CPB time, minutes & 1,010 (6.85\%) & 149 (7.85\%) & 678 (6.31\%) & 77 (7.22\%) & 6 (10.9\%) & 100 (10.2\%) \\ 
+Bilirubin levels, mg/dL & 850 (5.77\%) & 113 (5.96\%) & 605 (5.63\%) & 82 (7.69\%) & 4 (7.27\%) & 46 (4.69\%) \\ 
+Body mass index & 77 (0.52\%) & 11 (0.58\%) & 55 (0.51\%) & 6 (0.56\%) & 1 (1.82\%) & 4 (0.41\%) \\ 
+BUN, mg/dL & 50 (0.34\%) & 4 (0.21\%) & 41 (0.38\%) & 1 (0.09\%) & 0 (0.00\%) & 4 (0.41\%) \\ 
+Sex & 24 (0.16\%) & 4 (0.21\%) & 19 (0.18\%) & 1 (0.09\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Urinary creatinine, mg/dL & 20 (0.14\%) & 3 (0.16\%) & 15 (0.14\%) & 2 (0.19\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Age, years & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Race & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Device strategy & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
+Device type & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
 \bottomrule
@@
 \midrule
-Imputation to the mean & 6.09 (5.71, 6.54) & -- & 7.27 (6.83, 7.77) & -- \\ 
+Imputation to the mean & 6.09 (reference) & -- & 7.27 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.09 (-0.16, 0.31) & -- \\ 
@@
 \midrule
-Imputation to the mean & 5.56 (4.95, 6.03) & -- & 6.72 (6.13, 7.33) & -- \\ 
+Imputation to the mean & 5.56 (reference) & -- & 6.72 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.52 (0.21, 0.77) & -- \\ 
@@
 \midrule
-Imputation to the mean & 4.79 (4.11, 5.55) & -- & 6.29 (5.67, 7.02) & -- \\ 
+Imputation to the mean & 4.79 (reference) & -- & 6.29 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.40 (0.11, 0.75) & -- \\ 
@@
 \midrule
-Imputation to the mean & 9.35 (8.80, 9.71) & -- & 9.00 (8.55, 9.41) & -- \\ 
+Imputation to the mean & 9.35 (reference) & -- & 9.00 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & -0.07 (-0.28, 0.13) & -- \\ 
@@
 \midrule
-Imputation to the mean & 8.76 (8.20, 9.21) & -- & 8.01 (7.57, 8.54) & -- \\ 
+Imputation to the mean & 8.76 (reference) & -- & 8.01 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.50 (0.27, 0.78) & -- \\ 
@@
 \midrule
-Imputation to the mean & 7.96 (7.26, 8.63) & -- & 7.09 (6.45, 7.68) & -- \\ 
+Imputation to the mean & 7.96 (reference) & -- & 7.09 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.83 (0.55, 1.10) & -- \\ 
@@
 \midrule
-Imputation to the mean & 69.3 (68.9, 69.8) & -- & 70.7 (70.3, 71.0) & -- \\ 
+Imputation to the mean & 69.3 (reference) & -- & 70.7 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & 0.00 (-0.20, 0.19) & -- \\ 
@@
 \midrule
-Imputation to the mean & 69.2 (68.8, 69.7) & -- & 70.5 (70.1, 71.0) & -- \\ 
+Imputation to the mean & 69.2 (reference) & -- & 70.5 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.10 (-0.11, 0.31) & -- \\ 
@@
 \midrule
-Imputation to the mean & 68.9 (68.4, 69.4) & -- & 70.2 (69.8, 70.7) & -- \\ 
+Imputation to the mean & 68.9 (reference) & -- & 70.2 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.12 (-0.08, 0.31) & -- \\ 
@@
 \midrule
-Imputation to the mean & 2.41 (1.71, 3.44) & -- & 3.29 (2.24, 4.80) & -- \\ 
+Imputation to the mean & 2.41 (reference) & -- & 3.29 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & -0.09 (-1.07, 0.65) & -- \\ 
@@
 \midrule
-Imputation to the mean & 5.75 (3.74, 8.59) & -- & 6.28 (4.17, 9.45) & -- \\ 
+Imputation to the mean & 5.75 (reference) & -- & 6.28 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & -2.83 (-4.81, -0.83) & -- \\ 
@@
 \midrule
-Imputation to the mean & 9.48 (5.50, 15.3) & -- & 7.02 (4.32, 11.3) & -- \\ 
+Imputation to the mean & 9.48 (reference) & -- & 7.02 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & -1.61 (-5.29, 0.68) & -- \\ 
@@
 \midrule
-Imputation to the mean & 79.4 (79.0, 79.8) & -- & 79.2 (78.9, 79.7) & -- \\ 
+Imputation to the mean & 79.4 (reference) & -- & 79.2 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & -0.03 (-0.22, 0.14) & -- \\ 
@@
 \midrule
-Imputation to the mean & 78.9 (78.5, 79.4) & -- & 78.4 (77.9, 78.9) & -- \\ 
+Imputation to the mean & 78.9 (reference) & -- & 78.4 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.47 (0.26, 0.71) & -- \\ 
@@
 \midrule
-Imputation to the mean & 78.4 (77.7, 78.8) & -- & 77.4 (76.8, 78.0) & -- \\ 
+Imputation to the mean & 78.4 (reference) & -- & 77.4 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.91 (0.52, 1.24) & -- \\ 
@@
 \midrule
-Imputation to the mean & 1.36 (0.92, 2.07) & -- & 1.79 (1.34, 2.50) & -- \\ 
+Imputation to the mean & 1.36 (reference) & -- & 1.79 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & +0.01 (-0.53, 0.54) & -- \\ 
@@
 \midrule
-Imputation to the mean & 2.48 (1.79, 3.93) & -- & 2.51 (1.81, 3.29) & -- \\ 
+Imputation to the mean & 2.48 (reference) & -- & 2.51 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & -0.50 (-1.38, 0.16) & -- \\ 
@@
 \midrule
-Imputation to the mean & 4.98 (3.42, 7.00) & -- & 3.99 (2.74, 6.22) & -- \\ 
+Imputation to the mean & 4.98 (reference) & -- & 3.99 (reference) & -- \\ 
 Missingness as an attribute & -- & -- & -1.52 (-2.62, -0.35) & -- \\ 
@@
 
-\caption{Posterior distribution of differences in scaled Brier score values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+\caption{Posterior distribution of differences in scaled Brier score values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
@@
 
-\caption{Posterior distribution of differences in concordance index values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+\caption{Posterior distribution of differences in concordance index values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.
@@
 
-\caption{Posterior distribution of differences in calibration error values relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
+\caption{Posterior distribution of differences in calibration error values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
 Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\%, 15\%, or 30\%.

### 2020-11-19 doc_arXiv/what_is_known.Rmd
feedback from RC

@@
 ---
-date: "`r Sys.Date()`"
-author: "Your Name"
-title: "officedown template"
 output: 
@@
 
-- add a sentence
+- Missing data must be imputed in order to fit many standard modeling algorithms used for risk prediction.
+The prognostic accuracy of these prediction models depends on the imputation strategy.
+Imputation to the mean is often used in applied practice.
+
+<br>
 
@@
 
-- add a sentence
+- We assessed how applying strategies other than imputation to the mean to impute missing values would impact the prognostic accuracy of downstream risk prediction models, i.e., models fitted to the imputed data.
+Our results show that many imputation strategies can improve model prognosis relative to using imputation to the mean.
 
-- add a sentence
+- We measured imputation accuracy as well as prognostic accuracy of downstream risk prediction models using both single and multiple imputation techniques.
+Our results show that single imputation often has higher accuracy but multiple imputation often leads to greater prognostic accuracy of downstream models.

### 2020-11-19 R/as_gt.R

@@ -24,7 +24,7 @@ as_gt <- function(tbl_md_strat,
             md_strat = str_replace(md_strat, 'mia', 'mia_si'),
             table_val = if_else(
               md_strat == 'meanmode_si',
-              true = table_glue("{est} ({lwr}, {upr})", rspec=rspec),
+              true = table_glue("{est} (reference)", rspec=rspec),
               false = table_glue("{pm(est)}{est} ({lwr}, {upr})", rspec=rspec)
             )
           ) %>%

### 2020-11-19 R/gt_latex.R

@@ -6,7 +6,7 @@
 ##' @param gt_object
 ##' @param caption
 ##' @param label
-gt_latex <- function(gt_object, caption, label) {
+gt_latex <- function(gt_object, caption, label = NULL, size = NULL) {
 
     header = paste0(
       '\\begin{table} \n \\caption{',caption,'} \n'
@@ -14,6 +14,10 @@ gt_latex <- function(gt_object, caption, label) {
 
     if(!is.null(label)){
       header = paste0(header, '\\label{', label, '} \n')
+    }
+
+    if(!is.null(size)){
+      header = paste0(header, '\\', size, '\n')
     }
 
     gt_object %>%

### 2020-11-19 R/plan.R

@@ -51,7 +51,7 @@ the_plan <- drake_plan(
   im = clean_intermacs(),
 
   # tabulate descriptives
-  tbl_descriptives = tabulate_descriptives(im),
+  tbl_descriptives = tabulate_descriptives(im, times),
 
   # creates resamples.csv
   resamples = mc_cv_light(data = im, train_prop = 1/2, ntimes = 200),

### 2020-11-19 R/tabulate_descriptives.R

@@ -4,15 +4,20 @@
 ##'
 ##' @title
 ##' @param im
-tabulate_descriptives <- function(im) {
+##' TODO: change status to 6 months, change censored to alive on device
+tabulate_descriptives <- function(im, times) {
 
   tbl_data <- im %>%
     mutate(
       status = case_when(
-        pt_outcome_dead == 1 ~ "Dead",
-        pt_outcome_txpl == 1 ~ "Transplant",
-        pt_outcome_cess == 1 ~ "Cessation of support",
-        TRUE ~ "Censored"
+        pt_outcome_dead == 1 & months_post_implant < times ~ "Dead",
+        pt_outcome_txpl == 1 & months_post_implant < times ~ "Transplant",
+        pt_outcome_cess == 1 & months_post_implant < times ~ "Cessation of support",
+        pt_outcome_dead == 0 &
+          pt_outcome_txpl == 0 &
+          pt_outcome_cess == 0 &
+          months_post_implant < times ~ "Censored",
+          TRUE ~ "Alive on device"
       ),
       demo_race = case_when(
         demo_race_af_amer == 'yes' ~ "Black",
@@ -97,10 +102,10 @@ tabulate_descriptives <- function(im) {
         pi_peripheral_edema = 'Periphal edema',
         pi_lvedd = 'LVEDD',
         pi_device_strategy = "Device strategy",
-        pi_albumin_g_dl = 'Urinary albumin, g/dl',
-        pi_creat_mg_dl = 'Urinary creatinine, mg/dl',
-        pi_bun_mg_dl = 'BUN, mg/dl',
-        pi_bili_total_mg_dl = 'Bilirubin levels, mg/dl'
+        pi_albumin_g_dl = 'Urinary albumin, g/dL',
+        pi_creat_mg_dl = 'Urinary creatinine, mg/dL',
+        pi_bun_mg_dl = 'BUN, mg/dL',
+        pi_bili_total_mg_dl = 'Bilirubin levels, mg/dL'
       )
     )
 
@@ -123,6 +128,16 @@ tabulate_descriptives <- function(im) {
 
   tbl_characteristics <- tbl_data %>%
     mutate(`No. of patients` = 1, .before = 1) %>%
+    mutate(
+      status = factor(
+        status,
+        levels = c('Dead',
+                   'Transplant',
+                   'Cessation of support',
[…]

### 2020-11-19 R/visualize_md_strat_inference.R

@@ -135,8 +135,8 @@ visualize_md_strat_inference <- function(bayes_mccv_fits,
           ) %>%
           mutate(m2 = factor(m2, levels = rev(names(md_method_labels)[-1])))
 
-        probs_upr <- 0.990
-        probs_lwr <- 0.940
+        probs_upr <- 0.95
+        probs_lwr <- 0.70
 
         if(str_detect(.y, 'calibration')){
           probs_lwr <- 0.0010
@@ -180,6 +180,11 @@ visualize_md_strat_inference <- function(bayes_mccv_fits,
               true = table_glue("{1-prob_gt_meanmode}",rspec = temp_rspec),
               false = table_glue("{prob_gt_meanmode}",rspec = temp_rspec)
             ),
+            prob_gt_meanmode = if_else(
+              prob_gt_meanmode == '1.000',
+              '>0.999',
+              prob_gt_meanmode
+            ),
             distribution = if_else(
               condition = md_type == 'Multiple imputation',
               true = x_mi,
@@ -211,7 +216,7 @@ visualize_md_strat_inference <- function(bayes_mccv_fits,
                      alpha = 1/2,
                      show.legend = FALSE,
                      aes(label = prob_gt_meanmode)) +
-          geom_label(data = x_label,
+          geom_text(data = x_label,
                      nudge_y = 7/10,
                      nudge_x = 1/4,
                      aes(label = text),

### 2020-12-16 doc_arXiv/doc_arXiv.Rmd
word doc for circ

@@
 The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively.
-\newline\textbf{Conclusion} Selecting an optimal strategy to impute missing values and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
+\newline\textbf{Conclusion} Selecting an optimal strategy to impute missing values such as random forests and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
 keywords:
@@
 
-The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nation wide level.
+The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nationwide level.
 Currently, INTERMACS comprises over 20,000 patients who have received a MCS device.
@@
 
-\paragraph{Conclusion} Selecting an optimal strategy to impute missing values can impact the prognostic accuracy of downstream risk prediction models.
+\paragraph{Conclusion} Selecting an optimal strategy to impute missing values such as random forests can impact the prognostic accuracy of downstream risk prediction models.
 In the current analysis, conducting multiple imputation using random forests emerged as an optimal strategy to impute missing values in the INTERMACS data.

### 2020-12-16 doc_arXiv/doc_arXiv.tex
word doc for circ

@@
 \newline\textbf{Conclusion} Selecting an optimal strategy to impute
-missing values and applying multiple imputation can improve the
-prognostic accuracy of downstream risk prediction models.
+missing values such as random forests and applying multiple imputation
+can improve the prognostic accuracy of downstream risk prediction
+models.
 \end{abstract}
@@
 collection of patient characteristics, medical events, and long terms
-outcomes at a nation wide level. Currently, INTERMACS comprises over
+outcomes at a nationwide level. Currently, INTERMACS comprises over
 20,000 patients who have received a MCS device. As the largest registry
@@
 
-Selecting an optimal strategy to impute missing values can impact the
-prognostic accuracy of downstream risk prediction models. In the current
-analysis, conducting multiple imputation using random forests emerged as
-an optimal strategy to impute missing values in the INTERMACS data. This
-investigation can directly inform future analyses of INTERMACS data and
-provide evidence quantifying the benefit of imputing missing data with
-sound methodology.
+Selecting an optimal strategy to impute missing values such as random
+forests can impact the prognostic accuracy of downstream risk prediction
+models. In the current analysis, conducting multiple imputation using
+random forests emerged as an optimal strategy to impute missing values
+in the INTERMACS data. This investigation can directly inform future
+analyses of INTERMACS data and provide evidence quantifying the benefit
+of imputing missing data with sound methodology.
 

### 2020-12-16 doc_ccqo/doc_ccqo.Rmd
word doc for circ

--- 2020-12-16 doc_arXiv/doc_arXiv.Rmd
@@
 ---
-title: Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values
-authors:
-  - name: Byron C. Jaeger
-    thanks: Source code available at https://github.com/bcjaeger/INTERMACS-missing-data
-    department: Department of Biostatistics
-    affiliation: University of Alabama at Birmingham
-    location: Birmingham, AL
-    email: bcjaeger@uab.edu
-  - name: Ryan Cantor, PhD
-    department: Kirklin Institute for Research in Surgical Outcomes
-    affiliation: University of Alabama at Birmingham
-    location: Birmingham, AL
-  - name: Venkata Sthanam, MS
-    department: Kirklin Institute for Research in Surgical Outcomes
-    affiliation: University of Alabama at Birmingham
-    location: Birmingham, AL
-  - name: Rongbing Xie, DrPH
-    department: Department of Surgery
-    affiliation: University of Alabama at Birmingham
-    location: Birmingham, AL
-  - name: James K. Kirklin, MD
-    department: Kirklin Institute for Research in Surgical Outcomes
-    affiliation: University of Alabama at Birmingham
-    location: Birmingham, AL
-  - name: Ramaraju Rudraraju, PhD
-    department: Kirklin Institute for Research in Surgical Outcomes
-    affiliation: University of Alabama at Birmingham
-    location: Birmingham, AL
-    email: ramaraju@uab.edu
-abstract: |
-  \textbf{Background} Risk prediction models play an important role in clinical decision making.
-When developing risk prediction models, practitioners often impute missing values to the mean.
-We evaluated the impact of applying other strategies to impute missing values on the prognostic accuracy of downstream risk prediction models, i.e., models fitted to the imputed data.
-A secondary objective was to compare the accuracy of imputation methods based on artificially induced missing values.
-To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).
-\newline\textbf{Methods and Results} We applied twelve imputation strategies in combination with two different modeling strategies for mortality and transplant risk prediction following surgery to receive mechanical circulatory support.
-Model performance was evaluated using Monte-Carlo cross validation and measured based on outcomes 6-months following surgery using the scaled Brier score, concordance index, and calibration error.
-We used Bayesian hierarchical models to compare model performance.
-Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.0008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
-The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively.
-\newline\textbf{Conclusion} Selecting an optimal strategy to impute missing values such as random forests and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
-keywords:
-  - missing data
-  - INTERMACS
-  - imputation
-  - heart failure
-  - mechanical circulatory support
-  - risk prediction
+title: "Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values"
 bibliography: references.bib
-output: rticles::arxiv_article
-header-includes:
-   - \newcommand{\ie}{\textit{i.e., }}
-   - \newcommand{\eg}{\textit{e.g., }}
-   - \newcommand{\cstat}{\widehat{\textrm{C}}(t)}
-   - \newcommand{\bstat}{\widehat{\textrm{BS}}(t)}
-   - \newcommand{\ibstat}{\mathcal{\widehat{BS}}(\tau)}
+csl: ccqo.csl
+output: 
+  officedown::rdocx_document:
+    reference_docx: style_manuscript_times_new_roman_lines.docx
 ---
 
-```{r setup, include = FALSE}
+```{r setup, include=FALSE}
 
@@
 
-options(knitr.table.format = 'latex')
+library(officedown)
+library(officer)
+library(flextable)
 
-# source(file.path(here::here(), 'packages.R'))
-# R.utils::sourceDirectory('R')
+set_flextable_defaults(
+  font.family = "Times New Roman", 
+  theme_fun = "theme_box"
+)
+
+conflicted::conflict_prefer("footnote", "flextable")
+conflicted::conflict_prefer("as_flextable", "flextable")
+
+source("../packages.R")
+R.utils::sourceDirectory('../R')
 
@@
 
+<!-- Title Page -->
+
+<!-- First author’s surname and short title 
+(not to exceed 50 characters, including spaces) -->
+
+__First author's surname and short title:__ Jaeger, Prediction and imputation
+
+<!-- Authors’ names, academic degrees, and affiliations -->
+
+Byron C. Jaeger, PhD^1^, Ryan Cantor, PhD^1,2^, Venkata Sthanam, MS^1,2^, Rongbing Xie, DrPH^1,2^, James K. Kirklin, MD^1,2^, Ramaraju Rudraraju, PhD^1,2^
+
+1. University of Alabama at Birmingham, Birmingham, AL
+2. Kirklin Institute for Research in Surgical Outcomes
+
+<!-- Name and complete address for correspondence (include street name and 
+address as well as post office box, and address for reprints if different 
+from correspondence address), including telephone number, and email address -->
+
+__Address for correspondence__
+Byron C. Jaeger, Ph.D.
+Department of Biostatistics
+University of Alabama at Birmingham
+327M Ryals Public Health Building
+1665 University Blvd
+Birmingham, Alabama 35294-0022
+
+<!-- The total word count of the manuscript, including the title page, 
+abstract, text, references, tables, and figures legends -->
+
+__total word count of the manuscript, including the title page, abstract, text, references, tables, and figures legends:__
+
+\newpage
+
+# ABSTRACT
+
+<!-- Do not cite references in the abstract. -->
+
+<!-- Limit use of acronyms and abbreviations. 
+Define at first use with acronym or abbreviation in parentheses. -->
+
+<!-- Maximum of 300 words -->
+
+<!-- Use the following headings: -->
+
+<!-- Background - rationale for study -->
+
+<!-- Methods - brief presentation of methods; please include sample sizes and numerical data with significance levels throughout -->
+
+<!-- Results—brief presentation of significant results; please include sample sizes and numerical data with significance levels throughout -->
+
+<!-- Conclusions—succinct statement of data interpretation -->
+
+<!-- When applicable, include a fifth heading, "Registration." Please list the URL, as well as the unique identifier, for the publicly accessible website on which the study is registered. -->
+
+__Background__ Risk prediction models play an important role in clinical decision making.
+When developing risk prediction models, practitioners often impute missing values to the mean.
+We evaluated the impact of applying other strategies to impute missing values on the prognostic accuracy of downstream risk prediction models, i.e., models fitted to the imputed data.
+A secondary objective was to compare the accuracy of imputation methods based on artificially induced missing values.
+To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).
+
+__Methods__ We applied twelve imputation strategies in combination with two different modeling strategies for mortality and transplant risk prediction following surgery to receive mechanical circulatory support.
+Model performance was evaluated using Monte-Carlo cross validation and measured based on outcomes 6-months following surgery using the scaled Brier score, concordance index, and calibration error.
+We used Bayesian hierarchical models to compare model performance.
+
+__Results__ Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.0008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
+The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively.
+
+__Conclusion__ Selecting an optimal strategy to impute missing values such as random forests and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
+
 # Introduction
-\label{sec:introduction}
-
-\linenumbers
-\doublespacing
-
-<!-- Heart disease / heart failure -->
 
 Heart disease is a leading cause of death in the United States.
-Heart failure, a primary component of heart disease, affects over 6 million Americans, and for ~10\% of these patients medical management is no longer effective \cite{benjamin2017heart,national2017health}.
-Mechanical circulatory support (MCS) is a surgical intervention in which a mechanical device is implanted in parallel to the heart to improve circulation \cite{patel2014contemporary}.
-Typically, MCS is used while a patient waits for a heart transplant (bridge-to-transplant) or in some cases as an alternative to transplant (destination therapy) \cite{slaughter2009advanced}.
-Over 250,000 patients could benefit from MCS \cite{miller2011left}.
-However, less than 4,000 new patients receive a long-term MCS device each year, with widely heterogeneous outcomes \cite{stewart2011keeping}.
-The 2-year survival probability on MCS ranges from 61\% for destination therapy to 78\% for bridge-to-transplant \cite{patel2014contemporary}.
+Heart failure, a primary component of heart disease, affects over 6 million Americans, and for roughly 10% of these patients medical management is no longer effective [@benjamin2017heart; @national2017health].
+Mechanical circulatory support (MCS) is a surgical intervention in which a mechanical device is implanted in parallel to the heart to improve circulation [@patel2014contemporary].
+Typically, MCS is used while a patient waits for a heart transplant (bridge-to-transplant) or in some cases as an alternative to transplant (destination therapy) [@slaughter2009advanced].
+Over 250,000 patients could benefit from MCS [@miller2011left].
+However, less than 4,000 new patients receive a long-term MCS device each year, with widely heterogeneous outcomes [@stewart2011keeping].
+The 2-year survival probability on MCS ranges from 61% for destination therapy to 78% for bridge-to-transplant [@patel2014contemporary].
 Reliable predictions of patient-specific risk to experience key outcomes such as mortality or transplant after receiving MCS can help improve patient selection, inform the design of next generation pumps, and improve patient care strategies.
+
+<br>
 
@@
 Currently, INTERMACS comprises over 20,000 patients who have received a MCS device.
-As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
+As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device [@hsich2012should; @cotts2014predictors; @eckman2011survival; @kirklin2017eighth; @kormos2019society; @Adamo950].
 An important step for developing risk prediction models is dealing with missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial.
@@
 
+<br>
+
 <!-- Aims of this article -->
@@
 The primary aim of this article is to quantify how much the prognostic value of a risk prediction model developed from the INTERMACS registry depends on the strategy that was applied to impute missing data prior to developing the model.
-A secondary aim is to measure imputation accuracy of each strategy by introducing varying levels of artificial missing data (\ie data amputation) and then imputing it.
-Because imputation to the mean has been a standard method for multiple annual summaries of the INTERMACS data, we measure the potential improvement in prognostic value of a risk prediction model when other strategies are applied to impute missing data \emph{instead of imputation to the mean}.
+A secondary aim is to measure imputation accuracy of each strategy by introducing varying levels of artificial missing data (*i.e.* data amputation) and then imputing it.
+Because imputation to the mean has been a standard method for multiple annual summaries of the INTERMACS data, we measure the potential improvement in prognostic value of a risk prediction model when other strategies are applied to impute missing data *instead of* imputation to the mean.
 The over-arching aim is to clarify which imputation strategies are most likely to improve risk prediction (and by extension, quality of care) for patients who receive MCS devices.
@@
 
+<br>
+
 # Methods
-\label{sec:methods}
 
-<!-- Section summary -->
+the INTERMACS registry is a North American observational registry for patients receiving MCS devices that began as a partnership between the National Heart, Lung, and Blood Institute, US Food and Drug Administration, the Centers for Medicaid and Medicare Services, industry, and individual hospitals with the mission of improving MCS outcomes.
+In 2018, INTERMACS became an official Society of Thoracic Surgeons database.
 
-<!-- In this section, we describe the INTERMACS Registry in Section \ref{subsec:intermacs} and provide a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning in Section \ref{subsec:inference_and_learning}.
-We then describe thirteen strategies to handle missing values (Section \ref{subsec:imputation}) and three strategies to develop risk prediction models after missing data have been imputed (Section \ref{subsec:modeling}).
-Next, we describe how risk prediction models and by extension the imputation strategy applied before training the model were evaluated in Section \ref{subsec:evaluation}.
-We outline an internal validation procedure that was applied to compare combinations of imputation and modeling strategies using hierarchical Bayesian models in Section \ref{subsec:internal}.
-Last, details on computation and software packages used for the current analysis are described in Section \ref{subsec:computing} -->
-
-## INTERMACS Registry
-\label{subsec:intermacs}
-
-INTERMACS is a North American observational registry for patients receiving MCS devices that began as a partnership between the National Heart, Lung, and Blood Institute, US Food and Drug Administration, the Centers for Medicaid and Medicare Services, industry, and individual hospitals with the mission of improving MCS outcomes.
-In 2018, INTERMACS became an official Society of Thoracic Surgeons database.
+<br>
 
@@
 Primary data collection is approved through University of Alabama Institutional Review Board and at individual sites.
+
+<br>
 
@@
 
+<br>
+
 ## Statistical Inference and Learning with Missing data
-\label{subsec:inference_and_learning}
 
-\paragraph{Statistical Inference} To conduct statistical inference in the context of missing data setting, analysts often create multiple imputed datasets, replicate an analysis in each of them, and then pool their results to obtain valid test statistics for hypothesis testing \cite{rubin2004multiple}.
-Imputation strategies that create a single imputed dataset (\eg imputation to the mean) have been shown to increase type I errors (\ie rejecting a true null hypothesis) for inferential statistics by artificially reducing the variance of observed data and ignoring the uncertainty attributed to missing values \cite{van2018flexible}.
-To ensure valid inference, imputation models should leverage outcome variables to predict missing predictors \cite{sterne2009multiple}.
-When very few data are missing, analysts may apply listwise deletion, \ie removing any observation with at least one missing value.
-However, listwise deletion can easily lead to biased inference \cite{van2020rebutting}.
+__Statistical Inference:__ To conduct statistical inference in the context of missing data setting, analysts often create multiple imputed datasets, replicate an analysis in each of them, and then pool their results to obtain valid test statistics for hypothesis testing [@rubin2004multiple].
+Imputation strategies that create a single imputed dataset (\eg imputation to the mean) have been shown to increase type I errors (\ie rejecting a true null hypothesis) for inferential statistics by artificially reducing the variance of observed data and ignoring the uncertainty attributed to missing values [@van2018flexible].
+To ensure valid inference, imputation models should leverage outcome variables to predict missing predictors [@sterne2009multiple].
+When very few data are missing, analysts may apply listwise deletion, *i.e.* removing any observation with at least one missing value.
+However, listwise deletion can easily lead to biased inference [@van2020rebutting].
 
-\paragraph{Statistical Learning} In the presence of missing data, the goal of supervised statistical learning is to develop a prediction function for an outcome variable that accurately generalizes to testing data, which may or may not contain missing values \cite{hastie2009elements}.
+<br>
+
+__Statistical Learning:__ In the presence of missing data, the goal of supervised statistical learning is to develop a prediction function for an outcome variable that accurately generalizes to testing data, which may or may not contain missing values [@hastie2009elements].
 Because testing data may contain missing values, listwise deletion is not a feasible strategy for statistical learning tasks.
-In contrast to statistical inference, strategies that create a single imputed dataset are often used for statistical learning \cite{kuhn2019feature}.
-Previous work has emphasized the importance of imputation strategies with greater accuracy, suggesting that more accurate imputation strategies lead to better performance of downstream models (\ie models fitted to the imputed data) \cite{jerez2010missing}.
-Others have shown that imputation to the mean, model-based imputation, and multiple imputation can provide Bayes-optimal prediction models provided certain assumptions are true, but these assumptions are difficult to validate in applied settings \cite{josse2019consistency}.
+In contrast to statistical inference, strategies that create a single imputed dataset are often used for statistical learning [@kuhn2019feature].
+Previous work has emphasized the importance of imputation strategies with greater accuracy, suggesting that more accurate imputation strategies lead to better performance of downstream models (*i.e.* models fitted to the imputed data) [@jerez2010missing].
+Others have shown that imputation to the mean, model-based imputation, and multiple imputation can provide Bayes-optimal prediction models provided certain assumptions are true, but these assumptions are difficult to validate in applied settings [@josse2019consistency].
 
-\paragraph{Using outcomes during imputation} The outcome variable should be used to impute predictor values for statistical inference, but using outcome variables for this purpose in supervised learning can have unintended consequences in model implementation.
+<br>
+
+__Using outcomes during imputation:__ The outcome variable should be used to impute predictor values for statistical inference, but using outcome variables for this purpose in supervised learning can have unintended consequences in model implementation.
 For instance, suppose we seek to predict an outcome $Y$ for a patient in the clinical setting.
@@
 On the other hand, if we already know $Y$, then we do not need to predict it.
-Due to these practical considerations and because INTERMACS is often leveraged to create prediction models for clinical practice \cite{thomas2014pre}, we did not leverage outcome variables to impute missing values of predictors in the current analysis.
+Due to these practical considerations and because INTERMACS is often leveraged to create prediction models for clinical practice [@thomas2014pre], we did not leverage outcome variables to impute missing values of predictors in the current analysis.
+
+<br>
 
 ## Missing Data Strategies
-\label{subsec:imputation}
 
-\paragraph{Single and multiple imputation} Several of the imputation methods we considered allowed for creation of single or multiple imputed datasets.
+__Single and multiple imputation:__ Several of the imputation methods we considered allowed for creation of single or multiple imputed datasets.
 For downstream models fitted to multiple sets of imputed data, we applied the modeling technique to each imputed training set, separately, and then used a pooling technique to generate a single set of predictions based on multiple imputed testing data.
@@
 
-\paragraph{Imputation to the mean} Imputing data to the mean involves three steps.
+<br>
+
+__Imputation to the mean:__ Imputing data to the mean involves three steps.
 First, numeric and nominal variables are identified.
@@
 
-\paragraph{Bayesian Regression} Imputation with Bayesian regression draws imputed values from the posterior distribution of model parameters, accounting for uncertainty in both model error and estimation \cite{rubin2004multiple}.
-Multiple imputation with chained equations (MICE), a well known technique for generating multiply imputed data \cite{azur2011multiple, van2018flexible, van2006fully}.
+<br>
+
+__Bayesian Regression:__ Imputation with Bayesian regression draws imputed values from the posterior distribution of model parameters, accounting for uncertainty in both model error and estimation [@rubin2004multiple].
+Multiple imputation with chained equations (MICE), a well known technique for generating multiply imputed data [@azur2011multiple; @van2018flexible; @van2006fully].
 In each iteration of MICE, each specified variable in the dataset is imputed using  other variables in the dataset.
@@
 
-\paragraph{Predictive Mean Matching (PMM)} PMM computes predicted values from a pre-specified model that treats one incomplete column in the data, $X$, as a dependent variable.
-For each missing value in $X$, PMM identifies a set of candidate donors based on distance in predicted values of $X$ \cite{landerman1997empirical}.
+<br>
+
+__Predictive Mean Matching (PMM):__ PMM computes predicted values from a pre-specified model that treats one incomplete column in the data, $X$, as a dependent variable.
+For each missing value in $X$, PMM identifies a set of candidate donors based on distance in predicted values of $X$ [@landerman1997empirical].
 One donor is randomly drawn from the candidates, and the observed value of the donor is taken to replace the missing value.
@@
 
-\paragraph{K-Nearest-Neighbors (KNN)} KNN imputation identifies $k$ `similar' observations (\ie neighbors) for each observation with a missing value in a given variable \cite{chen2000nearest}.
+<br>
+
+__K-Nearest-Neighbors (KNN):__ KNN imputation identifies $k$ `similar' observations (*i.e.* neighbors) for each observation with a missing value in a given variable [@chen2000nearest].
 A donor value for the current missing value is generated by sampling or aggregating the values from the $k$ nearest neighbors.
-In the current analysis, we identified 10 nearest neighbors using Gower's distance \cite{gower}.
+In the current analysis, we identified 10 nearest neighbors using Gower's distance [@gower].
 When imputing a single dataset using KNN, we aggregated values from 10 nearest neighbors using the median for numeric variables and the mode for categorical variables.
@@
 
-\paragraph{Hot Deck} Similar to KNN, hot deck imputation finds $k$ similar observations for each observation with a missing value in a given variable \cite{andridge2010hotdeck}.
+<br>
+
+__Hot Deck:__ Similar to KNN, hot deck imputation finds $k$ similar observations for each observation with a missing value in a given variable [@andridge2010hotdeck].
 However, hot deck imputation uses a less computationally intensive approach, either identifying neighbors at random or using a subset of variables to find similar observations.
@@
 
-\paragraph{Random forests} Random forests grow an ensemble of de-correlated decision trees, where each tree is grown using a bootstrapped replicate of the original training data \cite{Breiman2001, hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
-A particularly helpful feature of random forests is their ability to estimate testing error by aggregating each decision tree's prediction error on data outside of their bootstrapped sample (\ie out-of-bag error).
+<br>
+
+__Random forests:__ Random forests grow an ensemble of de-correlated decision trees, where each tree is grown using a bootstrapped replicate of the original training data [@Breiman2001; @hothorn2006survival; @strobl2007bias; @strobl2008conditional; @ishwaran2008random; @jaeger2019oblique].
+A particularly helpful feature of random forests is their ability to estimate testing error by aggregating each decision tree's prediction error on data outside of their bootstrapped sample (*i.e.* out-of-bag error).
 In the current analysis, we conduct MICE using one random forest to impute each variable, separately.
@@
 
-\paragraph{Missingness incorporated as an attribute (MIA)} MIA is a technique that uses missing status as a predictor rather than explicitly imputing missing values \cite{twala2009empirical, ding2010investigation}.
+<br>
+
+__Missingness incorporated as an attribute (MIA):__ MIA is a technique that uses missing status as a predictor rather than explicitly imputing missing values [@twala2009empirical; @ding2010investigation].
 MIA adds another category to nominal variables: "Missing".
@@
 
-<!-- \paragraph{Surrogate splitting} When decision trees search for a variable and cut-point to use for growing two new nodes in a decision tree, surrogate splitting will temporarily ignore missing values while searching for an initial split.
+<!-- __Surrogate splitting:__ When decision trees search for a variable and cut-point to use for growing two new nodes in a decision tree, surrogate splitting will temporarily ignore missing values while searching for an initial split.
 Once an initial split is chosen, surrogate splitting searches for a split on another variable with observed values for the observations with missing values on the initial splitting variable and creates a surrogate split based on the new variable that is similar to the original split.
@@
 
-\paragraph{Assumptions} Each missing data strategy poses different assumptions regarding the data and the mechanisms that lead to missing values in the data.
+<br>
+
+__Assumptions:__ Each missing data strategy poses different assumptions regarding the data and the mechanisms that lead to missing values in the data.
 Imputation using Bayesian regression makes the same distributional assumptions as the Bayesian models that are applied.
-The \texttt{miceRanger} algorithm does not make any formal distributional assumptions, as random forests are non-parametric and can thus handle skewed and multi-modal data as well as categorical data that are ordinal or non-ordinal.
+The miceRanger algorithm does not make any formal distributional assumptions, as random forests are non-parametric and can thus handle skewed and multi-modal data as well as categorical data that are ordinal or non-ordinal.
 Hot deck and KNN imputation also do not make distributional assumptions, but implicitly assume that a missing value for a given observation can be approximated by aggregating observed values from the $k$ most similar observations.
@@
 
+<br>
+
 ## Evaluating Imputation Accuracy
-\label{subsec:imputation_accuracy}
 
@@
 The observed values of data that were amputed were used to assess imputation accuracy.
-As a consequence, we only measured imputation accuracy 15\% or 30\% of additional missing values were amputed.
+As a consequence, we only measured imputation accuracy 15% or 30% of additional missing values were amputed.
 In the context of this article, the term 'amputation of data' means artificially making observed values missing.
@@
 
+<br>
+
 ## Risk Prediction Models
-\label{subsec:modeling}
 
@@
 
-A thorough description of stepwise variable selection and boosting can be found in Sections 6.1.2 and 8.2.2, respectively, of \emph{Introduction to Statistical Learning} \cite{james2013introduction}.
-Sir David Cox's PH model is one of the most frequently applied methods for the analysis of right-censored time-to-event outcomes \cite{kleinbaum2010survival}.
+A thorough description of stepwise variable selection and boosting can be found in Sections 6.1.2 and 8.2.2, respectively, of *Introduction to Statistical Learning* [@james2013introduction].
+Sir David Cox's PH model is one of the most frequently applied methods for the analysis of right-censored time-to-event outcomes [@kleinbaum2010survival].
 According to the PH assumption, the effect of a unit increase in a predictor is multiplicative with respect to a baseline hazard function.
-Boosting grows a sequence of decision trees, each using information from the previous trees in an attempt to correct their errors \cite{friedman2001greedy, chen2016xgboost}.
+Boosting grows a sequence of decision trees, each using information from the previous trees in an attempt to correct their errors [@friedman2001greedy; @chen2016xgboost].
+
+<br>
 
 ## Evaluation of Predictions
-\label{subsec:evaluation}
 
-\paragraph{The Brier score} The prognostic value of each risk prediction model was primarily assessed using the Brier score, which depends on both the discrimination and calibration of predicted risk values \cite{graf1999assessment, rufibach2010use}.
+__The Brier score:__ The prognostic value of each risk prediction model was primarily assessed using the Brier score, which depends on both the discrimination and calibration of predicted risk values [@graf1999assessment; @rufibach2010use].
 Let $\tilde{Y}_i(t)$ represent the observed status of individual $i$ at time $t > 0$ in a testing set of $M$ observations.
@@
 The Brier score is computed with \begin{equation} \label{eqn:brier_score}
-\bstat = \frac{1}{M} \sum_{i=1}^{M} \widehat{W}_i(t) \left\{ \tilde{Y}_i(t) - \widehat{S}(t \mid \bm{x}_i) \right\}^2, \end{equation} where, for the $i^{th}$ observation, $\widehat{S}(t \mid \bm{x}_i)$ is the estimated probability of survival at time $t$ according to a given risk prediction model, $\bm{x}_i$ is the set of input values for predictor variables in the model, and $\widehat{W}_i(t)$ is the inverse proportional censoring weight at time $t$ \cite{gerds2006consistent}.
+\widehat{\textrm{BS}}(t) = \frac{1}{M} \sum_{i=1}^{M} \widehat{W}_i(t) \left\{ \tilde{Y}_i(t) - \widehat{S}(t \mid \bm{x}_i) \right\}^2, \end{equation} where, for the $i^{th}$ observation, $\widehat{S}(t \mid \bm{x}_i)$ is the estimated probability of survival at time $t$ according to a given risk prediction model, $\bm{x}_i$ is the set of input values for predictor variables in the model, and $\widehat{W}_i(t)$ is the inverse proportional censoring weight at time $t$ [@gerds2006consistent].
 Thus, the Brier score is the mean squared difference between observed event status and expected event status according to a RPE at time $t$.
 Throughout the current analysis, we set $t = `r times`$ months after receiving MCS to focus on short term risk prediction.
-Models have been developed to simultaneously predict short term and long term mortality risk after receiving MCS, but these are beyond the scope of the current study \cite{blackstone1986decomposition}.
+Models have been developed to simultaneously predict short term and long term mortality risk after receiving MCS, but these are beyond the scope of the current study [@blackstone1986decomposition].
 
-\paragraph{The scaled Brier score}
+<br>
 
-The Brier score is dependent on the rate of observed events, which can make it a difficult metric to interpret.
+__The scaled Brier score:__ The Brier score is dependent on the rate of observed events, which can make it a difficult metric to interpret.
 It is often more informative to scale the Brier score based on the Brier score of a naive model.
-More specifically, for a given risk prediction model, the scaled Brier score is computed as $$\textrm{Scaled } \bstat \textrm{ of model } = 1 - \frac{\bstat \textrm{ of model}}{\bstat \textrm{ of naive model}}.$$ As the Brier score for risk prediction is analogous to mean-squared error for prediction of a continuous outcome, the scaled Brier score is analogous to the $R^2$ statistic.
-Similar to the $R^2$ statistic, a scaled $\bstat$ of 1.00 and 0.00 indicate a perfect and worthless model, respectively.
-In our analyses, a Kaplan-Meier estimate based on the training data (\ie a risk prediction model that did not use any predictor variables) provided the naive prediction.
+More specifically, for a given risk prediction model, the scaled Brier score is computed as $$\textrm{Scaled } \widehat{\textrm{BS}}(t) \textrm{ of model } = 1 - \frac{\widehat{\textrm{BS}}(t) \textrm{ of model}}{\widehat{\textrm{BS}}(t) \textrm{ of naive model}}.$$ As the Brier score for risk prediction is analogous to mean-squared error for prediction of a continuous outcome, the scaled Brier score is analogous to the $R^2$ statistic.
+Similar to the $R^2$ statistic, a scaled $\widehat{\textrm{BS}}(t)$ of 1.00 and 0.00 indicate a perfect and worthless model, respectively.
+In our analyses, a Kaplan-Meier estimate based on the training data (*i.e.* a risk prediction model that did not use any predictor variables) provided the naive prediction.
 In the current analysis, we multiply scaled Brier score values by 100 for ease of interpretation.
 
-\paragraph{Discrimination and calibration}
+<br>
 
-The discrimination of a risk prediction model measures the probability that the model will successfully identify which of two observations is at higher risk for the event of interest.
+__Discrimination and calibration:__ The discrimination of a risk prediction model measures the probability that the model will successfully identify which of two observations is at higher risk for the event of interest.
 We estimated discrimination using a time-dependent concordance (C-) index that accounted for covariate-dependent censoring.
@@
 We estimated calibration error by averaging the squared distance between expected and observed event rates according to a calibration plot.
-Full description of these evaluation metrics are available \cite{gerds2014calibration, gerds2013estimating}.
+Full description of these evaluation metrics are available [@gerds2014calibration; @gerds2013estimating].
+
+<br>
 
 ## Internal Validation via Monte-Carlo Cross-Validation (MCCV)
-\label{subsec:internal}
 
-To assess the prognostic value of each missing data strategy, we internally validated a total of 23 modeling algorithm based on combinations of imputation strategies and modeling strategies described in Sections \ref{subsec:imputation} and \ref{subsec:modeling}.
-For convenience, we use the term `modeling algorithm' to denote the combination of a missing data strategy and a modeling strategy (\eg imputation using mean/mode values followed by fitting the PH model with stepwise variable selection) \cite{kuhn2013applied}.
+To assess the prognostic value of each missing data strategy, we internally validated a total of 23 modeling algorithm based on combinations of imputation strategies and modeling strategies described above.
+For convenience, we use the term `modeling algorithm' to denote the combination of a missing data strategy and a modeling strategy (*e.g.* imputation using mean/mode values followed by fitting the PH model with stepwise variable selection) [@kuhn2013applied].
 We conducted internal validation using 200 replicates of Monte-Carlo cross validation, a resampling technique for internal validation.
 
-\paragraph{Steps taken in each replicate} In each replicate of Monte-Carlo cross validation, 50\% of the available data were used for model training and testing.
-All predictor variables with < 50\% missing values were considered for imputation and subsequent model development.
-Among these variables, artificial missingness (0\%, 15\%, or 30\% additional missing values) was induced based on patient age, with younger and older patients more likely to have missing data compared to patients who were between 40 and 65 years of age.
+<br>
+
+__Steps taken in each replicate:__ In each replicate of Monte-Carlo cross validation, 50% of the available data were used for model training and testing.
+All predictor variables with < 50% missing values were considered for imputation and subsequent model development.
+Among these variables, artificial missingness (0%, 15%, or 30% additional missing values) was induced based on patient age, with younger and older patients more likely to have missing data compared to patients who were between 40 and 65 years of age.
 Prior to imputation, 50 predictor variables were selected using a boosting model that quantified variable importance as the fractional contribution of each predictor to the model based on the total gain attributed to using the predictor while growing decision trees.
 Imputation was conducted in the training and testing sets, separately, for each imputation strategy.
-Although some imputation strategies (\eg KNN and random forests) can impute data in the testing set using models fitted to the training set, others (\eg Bayesian regression, PMM, and hot deck) cannot.
+Although some imputation strategies (*e.g.* KNN and random forests) can impute data in the testing set using models fitted to the training set, others (*e.g.* Bayesian regression, PMM, and hot deck) cannot.
 Therefore, to ensure fair comparisons in our experiment, each imputation procedure imputed data in the training set using models fitted to the training set and then imputed data in the testing set using models fitted to the testing set.
@@
 
-\paragraph{Bayesian analysis of model performance} To determine whether any of the imputation methods described above had improved upon imputation to the mean, we applied Bayesian hierarchical models to analyze differences in scaled Brier score \cite{benavoli2017time}.
+<br>
+
+__Bayesian analysis of model performance:__ To determine whether any of the imputation methods described above had improved upon imputation to the mean, we applied Bayesian hierarchical models to analyze differences in scaled Brier score [@benavoli2017time].
 This strategy provides a flexible framework to conduct hypothesis testing and also accounts for correlation occurring within each replicate of Monte-Carlo cross validation.
 Specifically, within each replicate of Monte-Carlo cross validation, the performance of different modeling algorithms are correlated because they are trained and tested using the same data.
+
+<br>
 
@@
 The number and proportion of missing values were also tabulated for the overall population and stratified by event status.
-Missing data patterns were visualized for the overall population using an UpSet plot \cite{lex2014upset}.
+Missing data patterns were visualized for the overall population using an UpSet plot [@lex2014upset].
 All of the proceeding analyses were conducted using resampling results from Monte-Carlo cross validation.
@@
 For imputation methods that created multiple datasets, scores were aggregated over each dataset to provide a single summary score.
-The distribution (\ie 25^th^, 50^th^, and 75^th^ percentile) of the scaled Brier score was estimated for each modeling algorithm.
+The distribution (*i.e.* 25^th^, 50^th^, and 75^th^ percentile) of the scaled Brier score was estimated for each modeling algorithm.
+
+<br>
 
 We split our results from Monte-carlo cross validation into four datasets based on outcome (mortality or transplant) and modeling procedure (Cox PH or boosting).
-For each dataset, we fit one hierarchical Bayesian model where the dependent variable was scaled Brier score and independent variables included the imputation strategy applied and the amount of artificially missing data (0\%, 15\%, or 30\%) induced before imputation.
+For each dataset, we fit one hierarchical Bayesian model where the dependent variable was scaled Brier score and independent variables included the imputation strategy applied and the amount of artificially missing data (0%, 15%, or 30%) induced before imputation.
 With each model, we estimated the difference in scaled Brier score of downstream models when random forests, Bayesian regression, PMM, hot deck, MIA, or KNN imputation were applied to impute missing values instead of imputation to the mean.
 
+<br>
+
 ## Computational Details
-\label{subsec:computing}
 
-SAS software (version 9.4) and Python (version 3.8.2) were used to create analytic data for the current study \cite{van1995python, sas2013}.
+SAS software (version 9.4) and Python (version 3.8.2) were used to create analytic data for the current study [@van1995python; @sas2013].
 These analyses were completed using The American Heart Association Precision Medicine Platform (https://precision.heart.org/).
-Base R (version 4.0.3) was used in combination with a number of open-source R packages (\eg \texttt{drake}, \texttt{tidyverse}, \texttt{naniar}, \texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to conduct statistical analyses and create the current manuscript \cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
-Our R code is available on GitHub (see [https://github.com/bcjaeger/INTERMACS-missing-data](https://github.com/bcjaeger/INTERMACS-missing-data)) \cite{byron_2020_4247449}.
-We used Cheaha, a high performance computing cluster at University of Alabama at Birmingham, to perform the 200 Monte-Carlo cross validation runs \cite{cheaha}.
+Base R (version 4.0.3) was used in combination with a number of open-source R packages (*e.g.* drake, tidyverse, naniar, table.glue, mice, miceRanger, and others) to conduct statistical analyses and create the current manuscript [@drake; @naniar; @mice; @miceRanger; @table.glue; @tidyverse; @rstanarm; @tidybayes; @survival; @tidymodels; @riskRegression].
+Our R code is available on GitHub (see [https://github.com/bcjaeger/INTERMACS-missing-data](https://github.com/bcjaeger/INTERMACS-missing-data)) [@byron_2020_4247449].
+We used Cheaha, a high performance computing cluster at University of Alabama at Birmingham, to perform the 200 Monte-Carlo cross validation runs [@cheaha].
 
-# Results
+<br>
+
+# RESULTS
 
@@
 
-\paragraph{Patient Characteristics} During the first `r times` months after receiving MCS, `r trunc_events$dead` patients died and `r trunc_events$txpl` had a transplant.
+<br>
+
+__Patient Characteristics:__ During the first `r times` months after receiving MCS, `r trunc_events$dead` patients died and `r trunc_events$txpl` had a transplant.
 In total, `r trunc_events$cens` patients had a follow-up time < `r times` months and were considered right-censored in the current analysis.
-The mean (standard deviation) age of patients was `r age_overall` years, `r white_overall`% of patients identified as white and `r male_overall`% were male (Table \ref{tbl_characteristics}).
+The mean (standard deviation) age of patients was `r age_overall` years, `r white_overall`% of patients identified as white and `r male_overall`% were male (Table 1).
 
-\paragraph{Missing data} Many predictor variables exhibited similar proportions of missing values in different outcome groups (Table \ref{tbl_missingness}).
+<br>
+
+__Missing data:__ Many predictor variables exhibited similar proportions of missing values in different outcome groups (Table 2).
 However, the number (percent) of missing values for surgery time was an exception, with `r inline_missingness$Overall$im_surgery_time` in the overall population, `r inline_missingness$Dead$im_surgery_time` among patients who died, and `r inline_missingness$Censored$im_surgery_time` among patients who were censored.
@@
 The collection of surgery time in the INTERMACS registry began on May 30, 2014.
-Additionally, missing values for surgery time were frequently accompanied by missing values for CV pressure (Figure \ref{fig:upset}).
+Additionally, missing values for surgery time were frequently accompanied by missing values for CV pressure (Figure 1).
 
@@
 
-  data <- strsplit(x, split = ' \\(') %>% 
+  data <- strsplit(x, split = '\\n\\(') %>% 
     unlist() %>% 
@@
 
-\paragraph{Imputation accuracy} Single imputation strategies were more accurate than their counterparts using multiple imputation in `r si_higher_accuracy$n_nominal` out of `r si_higher_accuracy$n_total` comparisons of nominal scores and `r si_higher_accuracy$n_numeric` out of `r si_higher_accuracy$n_total` comparisons of numeric scores (Table \ref{tbl_impute_accuracy}).
+<br>
+
+__Imputation accuracy:__ Single imputation strategies were more accurate than their counterparts using multiple imputation in `r si_higher_accuracy$n_nominal` out of `r si_higher_accuracy$n_total` comparisons of nominal scores and `r si_higher_accuracy$n_numeric` out of `r si_higher_accuracy$n_total` comparisons of numeric scores (Table 3).
 When an additional 15% of data were missing, single imputation KNN obtained the highest nominal accuracy (`r break_ci(top_accuracy$nominal.15$nominal_si, pre_label = 'score: ')`) and single imputation with random forests obtained the highest numeric accuracy (`r break_ci(top_accuracy$numeric.15$numeric_si, pre_label = 'score: ')`).
@@
 
+<br>
+
 ## Scaled Brier score
 
-\paragraph{Mortality risk prediction}
-
-When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$dead.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$dead.ipa.xgb.meanmode_si.0` for boosting models (Table \ref{tbl_md_strat_dead_ipa}; top panel).
+__Mortality risk prediction:__ When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$dead.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$dead.ipa.xgb.meanmode_si.0` for boosting models (Table 4; top panel).
 Multiple imputation strategies universally obtained higher scaled Brier scores versus their single imputation counterparts.
 Multiple imputation using random forests provided the highest scaled Brier score compared to other strategies, leading to a median (25^th^, 75^th^ percentile) increase in the scaled Brier score of `r inline_md_strat$dead.ipa.cph.ranger_mi.0` for Cox PH and `r inline_md_strat$dead.ipa.xgb.ranger_mi.0` for boosting models.
-These performance increments improved when an aditional 15\% and 30\% of data were amputed (Table \ref{tbl_md_strat_dead_ipa}; middle and bottom panel).
+These performance increments improved when an additional 15% and 30% of data were amputed (Table 4; middle and bottom panel).
 
-\paragraph{Transplant risk prediction}
+<br>
 
-When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$txpl.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$txpl.ipa.xgb.meanmode_si.0` for boosting models (Table \ref{tbl_md_strat_txpl_ipa}; top panel).
+__Transplant risk prediction__: When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$txpl.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$txpl.ipa.xgb.meanmode_si.0` for boosting models (Table 5; top panel).
 For Cox PH models, imputation to the mean provided the lowest scaled Brier score, and random forest imputation led to a median (25^th^, 75^th^ percentile) increase of `r inline_md_strat$txpl.ipa.cph.ranger_mi.0` versus imputation to the mean.
 For boosting models, surprisingly, imputation to the mean provided a higher scaled Brier score than all imputation methods except for multiple imputation using Bayesian regression.
-When an additional 15\% and 30\% of data were amputed, the performance increments corresponding to the use of multiple imputation increased for both Cox PH and boosting models (Table \ref{tbl_md_strat_txpl_ipa}; middle and bottom panel).
+When an additional 15% and 30% of data were amputed, the performance increments corresponding to the use of multiple imputation increased for both Cox PH and boosting models (Table 5; middle and bottom panel).
+
+<br>
 
@@
 
-\paragraph{Mortality risk prediction}
-
-Regardless of how much additional data were amputed, boosting models obtained higher median C-indices than PH models for prediction of 6-month mortality risk (Table \ref{tbl_md_strat_dead_auc}).
+__Mortality risk prediction:__ Regardless of how much additional data were amputed, boosting models obtained higher median C-indices than PH models for prediction of 6-month mortality risk (Table 6).
 In addition, all models that used multiple imputation consistently obtained higher median C-indices compared to their counterparts using single imputation.
-When 0 and 15\% additional data were amputed, median calibration error was lower for PH models compared to boosting models, but boosting models obtained lower median calibration error when 30\% additional data were amputed (Table \ref{tbl_md_strat_dead_cal_error}).
+When 0 and 15% additional data were amputed, median calibration error was lower for PH models compared to boosting models, but boosting models obtained lower median calibration error when 30\% additional data were amputed (Table 7).
 Almost all imputation strategies provided lower median calibration error compared with imputation to the mean.
 
-\paragraph{Transplant risk prediction}
+<br>
 
-For all amounts of additional data amputed, PH models obtained higher median C-indices than boosting models for prediction of 6-month transplant risk (Table \ref{tbl_md_strat_txpl_auc}).
+__Transplant risk prediction:__ For all amounts of additional data amputed, PH models obtained higher median C-indices than boosting models for prediction of 6-month transplant risk (Table 8).
 Similar to mortality risk prediction, multiple imputation strategies generally provided higher C-indices than their counterparts using single imputation strategy.
 For boosting models, MIA provided higher C-indices compared to all other single imputation strategies and had similar or superior performance compared to several multiple imputation strategies.
-Differences in calibration error were minor when no additional missing data were amputed (Table \ref{tbl_md_strat_txpl_cal_error}).
-When an additional 30\% of data were amputed, boosting models using MIA obtained lower calibration error than any other strategy.
+Differences in calibration error were minor when no additional missing data were amputed (Table 9).
+When an additional 30% of data were amputed, boosting models using MIA obtained lower calibration error than any other strategy.
+
+<br>
 
@@
 
-Adjusting for the amount of additional missing data amputed and the outcome variable, the posterior probability that an imputation strategy would improve the scaled Brier score of a downstream model relative to imputation to the mean was maximized by using multiple imputation with random forests (probability of improvement: `r inline_ipa_rf`; Figure \ref{fig:fig_md_strat_infer_ipa}).
-Similarly, multiple imputation using random forests was estimated to have the highest posterior probability of improving the C-index in comparison to using imputation to the mean (probability of improvement: `r inline_auc_rf`; Figure \ref{fig:fig_md_strat_infer_auc}).
-However, the estimated posterior probability of a reduction in calibration was >0.999 when using either single or multiple imputation with random forests compared with imputation to the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}).
+Adjusting for the amount of additional missing data amputed and the outcome variable, the posterior probability that an imputation strategy would improve the scaled Brier score of a downstream model relative to imputation to the mean was maximized by using multiple imputation with random forests (probability of improvement: `r inline_ipa_rf`; Figure 2).
+Similarly, multiple imputation using random forests was estimated to have the highest posterior probability of improving the C-index in comparison to using imputation to the mean (probability of improvement: `r inline_auc_rf`; Figure 3).
+However, the estimated posterior probability of a reduction in calibration was >0.999 when using either single or multiple imputation with random forests compared with imputation to the mean (Figure 4).
 Although imputation using random forest was estimated to be the best overall option, there was moderate to strong evidence that MIA and each multiple imputation strategy we applied would improve prognostic accuracy of downstream models compared with imputation to the mean.
 
-# Discussion
+<br>
+
+# DISCUSSION
 
@@
 
-In previous studies involving the INTERMACS data registry, imputation to the mean has been applied prior to developing a mortality risk prediction model \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
-An interesting recent study indicates that imputation to the mean can provide an asymptotically consistent prediction model, given the prediction model is flexible and non-linear \cite{josse2019consistency}.
+<br>
+
+In previous studies involving the INTERMACS data registry, imputation to the mean has been applied prior to developing a mortality risk prediction model [@hsich2012should; @cotts2014predictors; @eckman2011survival; @kirklin2017eighth; @kormos2019society; @Adamo950].
+An interesting recent study indicates that imputation to the mean can provide an asymptotically consistent prediction model, given the prediction model is flexible and non-linear [@josse2019consistency].
 However, theoretical results for finite samples have not yet been established.
@@
 
+<br>
+
 Previous research has also established evidence in favor of applying multiple imputation to improve the prognostic value of risk prediction models.
-For example, Hassan and Atiya demonstrated superior downstream prediction using an ensemble multiple imputation method on synthetic data with continuous outcomes \cite{hassan2007regression}.
+For example, Hassan and Atiya demonstrated superior downstream prediction using an ensemble multiple imputation method on synthetic data with continuous outcomes [@hassan2007regression].
 Similarly, Nanni et.
-al demonstrated superior performance in downstream prediction when missing values were imputed using their proposed ensemble multiple imputation method \cite{nanni2012classifier}.
+al demonstrated superior performance in downstream prediction when missing values were imputed using their proposed ensemble multiple imputation method [@nanni2012classifier].
 Notably, the authors artificially induced missing values in these studies and the largest real dataset that was evaluated contained less than 700 observations.
 An article by Jerez et.
-al evaluated missing data strategies based on the downstream task of fitting a neural network and predicting early breast cancer relapse \cite{jerez2010missing}.
+al evaluated missing data strategies based on the downstream task of fitting a neural network and predicting early breast cancer relapse [@jerez2010missing].
 The authors found that KNN imputation led to risk prediction models with the highest discrimination and lowest calibration error.
-Results from the current study are consistent with these previous findings but also extend their results by providing evidence from a larger source of data (\ie INTERMACS) and dealing with `real-world' missing values.
+Results from the current study are consistent with these previous findings but also extend their results by providing evidence from a larger source of data (*i.e.* INTERMACS) and dealing with 'real-world' missing values.
 
-Others have previously evaluated imputation techniques based on the accuracy with which these techniques impute missing values in the training data \cite{tutz2015improved, little2013joys, steele2018machine}.
+<br>
+
+Others have previously evaluated imputation techniques based on the accuracy with which these techniques impute missing values in the training data [@tutz2015improved; @little2013joys; @steele2018machine].
 While it is intuitive to hypothesize that more accurate imputation will provide more prognostic downstream models, our results do not support this supposition.
-For example, when an additional 30\% of missing data were amputed, none of the missing data strategies we implemented obtained higher accuracy than imputation to the mean.
-However, using \emph{any} of the multiple imputation strategies we considered instead of imputation to the mean increased prognostic accuracy of downstream models when an additional 30\% of missing data were amputed.
+For example, when an additional 30% of missing data were amputed, none of the missing data strategies we implemented obtained higher accuracy than imputation to the mean and multiple imputation strategies obtained particularly low accuracy.
+However, *all* of the multiple imputation strategies improved the prognostic accuracy of downstream models relative to imputation to the mean when an additional 30% of data were amputed prior to imputation of missing values.
 This result is likely explained by the bias variance tradeoff.
@@
 
-\paragraph{Strengths and limitations} The current analysis has a number of strengths.
+<br>
+
+__Strengths and limitations:__ The current analysis has a number of strengths.
 We leveraged the INTERMACS data registry, comprising one of the largest cohorts of patients who received MCS.
 We applied a well known resampling method to internally validate modeling algorithms for risk prediction.
-Last, we made our analysis R code available in a public repository (https://github.com/bcjaeger/INTERMACS-missing-data).
+We also made our analysis R code available in a public repository (https://github.com/bcjaeger/INTERMACS-missing-data).
 Last, the approach presented in this paper provides a general framework that can be applied in other studies where missing data are imputed prior to fitting a risk prediction model.
@@
 Also, it was not feasible to use only the training data to impute missing values in the testing data due to a lack of available software.
-Although the \texttt{miceRanger} package allows imputation of new data using existing models, few software packages for imputation allow users to implement multiple imputation with this protocol.
-Future analyses should introduce more flexible software and hands-on tutorials so that future investigators can optimize imputation of missing data.
+Although the miceRanger R package allows imputation of new data using existing models, few software packages for imputation allow users to implement multiple imputation with this protocol.
+Future analyses should introduce more flexible software and hands-on tutorials so that investigators can optimize imputation of missing data.
 
-\paragraph{Conclusion} Selecting an optimal strategy to impute missing values such as random forests can impact the prognostic accuracy of downstream risk prediction models.
+<br>
+
+__Conclusion:__ Selecting an optimal strategy to impute missing values such as random forests can impact the prognostic accuracy of downstream risk prediction models.
 In the current analysis, conducting multiple imputation using random forests emerged as an optimal strategy to impute missing values in the INTERMACS data.
 This investigation can directly inform future analyses of INTERMACS data and provide evidence quantifying the benefit of imputing missing data with sound methodology.
-\paragraph{Sources of Funding} This work is supported by the American Heart Association, grant# 18AIML34280052.
+
+<br>
+
+__Sources of Funding:__ This work is supported by the American Heart Association, grant# 18AIML34280052.
 De-identified data provided from the INTERMACS contract with from the National Heart, Lung, and Blood Institute; National Institutes of Health; and Department of Health and Human Services, under contract# HHSN268201100025C.
@@
 <!-- Table : characteristics -->
-\clearpage
-\begin{table}
-\footnotesize
-\caption{Participant characteristics overall and stratified by event status at `r times` months following surgery to receive mechanical circulatory support.
-The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
-\label{tbl_characteristics}
+\newpage
 
-```{r tbl_characteristics, results = 'asis'}
+Table 1: Participant characteristics overall and stratified by event status at `r times` months following surgery to receive mechanical circulatory support.
+The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.
+`r fp_par(line_spacing = 1)`
 
-as_kable_extra(
-  tbl_characteristics,
-  booktabs = TRUE, 
-  align = 'lcccccc'
+```{r tbl_characteristics}
+
+tbl_characteristics %>% 
+  as_flex_table() %>% 
+  width(width = 0.9) %>% 
+  width(j=1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  footnote(
+    i = 1, j = 1, 
+    ref_symbols = '*',
+    part = 'header',
+    value = as_paragraph(
+      'Table values are mean (standard deviation) or %'
+    )
   ) %>% 
-  footnote(general = 'Table values are mean (standard deviation) or %')
+  footnote(
+    i = 1, j = 1, 
+    ref_symbols = '', 
+    value = as_paragraph('Abbreviations: LVEDD = left ventricular end diastolic dimension; CV = central venous; BUN = blood urea nitrogen; CPB = cardiopulmonary bypass')
+  )
 
@@
 
-\end{table}
+\newpage
 
-<!-- Table : missingness -->
-\clearpage
+Table 2: Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.
+Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.
+This pattern is likely due to the fact that surgery time was added to the INTERMACS data after 2012.
+`r fp_par(line_spacing = 1)`
 
@@
          label != 'months_post_implant') %>%
-  gt(rowname_col = 'label') %>%
-  cols_align(align = 'center') %>%
-  gt_latex(
-    caption = 'Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.
-Many predictor variables exhibited similar proportions of missing values in different outcome groups, but surgery time was an exception.
-This pattern is likely due to the fact that surgery time was added to the INTERMACS data after 2012',
-    label = 'tbl_missingness',
-    size = 'footnotesize'
+  flextable(cwidth = 1.3) %>% 
+  set_header_labels(label = 'Variable') %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all') %>% 
+  width(width = 0.9) %>% 
+  width(j = 1, width = 2) %>% 
+  footnote(
+    i = 1, j = 1, 
+    ref_symbols = '', 
+    value = as_paragraph('Abbreviations: LVEDD = left ventricular end diastolic dimension; CV = central venous; BUN = blood urea nitrogen; CPB = cardiopulmonary bypass')
   )
@@
 
-<!-- Table: imputation accuracy -->
+\newpage
 
-\clearpage
+Table 3: Accuracy of strategies to impute artificial missing data.
+Table values are the median change in accuracy (25th, 75th percentile) relative to the accuracy of imputation to the mean.
+In general, multiple imputation strategies had lower accuracy than single imputation strategies, and few imputation strategies were more accurate than imputation to the mean.
+`r fp_par(line_spacing = 1)`
 
-```{r tbl_impute_accuracy, results = 'asis'}
+```{r}
 
@@
   mutate(
+    across(
+      ends_with("_mi"),
+      ~ if_else(is.na(.x), '---', .x)
+    ),
     md_method = factor(
@@
   ) %>%
-  arrange(md_method, additional_missing_pct) %>% 
-  gt(rowname_col = 'md_method', 
-     groupname_col = 'additional_missing_pct') %>% 
-  fmt_missing(columns = everything()) %>% 
-  cols_align('center') %>% 
-  tab_spanner(
-    label = 'Nominal variables',
-    columns = c('nominal_si', 'nominal_mi')
+  arrange(additional_missing_pct, md_method) %>% 
+  as_grouped_data(groups = 'additional_missing_pct') %>% 
+  as_flextable(hide_grouplabel = T) %>% 
+  add_header_row(
+    values = c("Imputation method", 
+               "Nominal variables", 
+               "Numeric variables"),
+    colwidths = c(1, 2, 2)
   ) %>% 
-  tab_spanner(
-    label = 'Numeric variables',
-    columns = c('numeric_si', 'numeric_mi')
+  set_header_labels(
+    nominal_si = 'Single imputation',
+    nominal_mi = 'Multiple imputation',
+    numeric_si = 'Single imputation',
+    numeric_mi = 'Multiple imputation',
+    md_method = 'Imputation method'
   ) %>% 
-  cols_label(
-    numeric_si = 'Single imputation', 
-    nominal_si = 'Single imputation', 
-    numeric_mi = 'Multiple imputation',
-    nominal_mi = 'Multiple imputation'
-  ) %>%
-  gt_latex(
-    caption = 'Accuracy of strategies to impute artificial missing data.
-Table values are the median change in accuracy (25th, 75th percentile) relative to the accuracy of imputation to the mean.
-In general, multiple imputation strategies had lower accuracy than single imputation strategies, and few imputation strategies were more accurate than imputation to the mean.',
-    label = 'tbl_impute_accuracy'
-  )
+  bg(i = ~!is.na(additional_missing_pct), bg = 'grey') %>% 
+  italic(i = ~!is.na(additional_missing_pct), italic = TRUE) %>% 
+  theme_box() %>% 
+  merge_v(part = 'header', j = 1) %>% 
+  width(width = 1.2) %>% 
+  width(j = 1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all')
 
@@
 
-\clearpage
+\newpage
 
-<!-- TODO: Add (reference) to imputation to the mean bracket -->
+Table 4: Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+Table values show the scaled Brier score when the imputation method is imputation to the mean and the change in scaled Brier score relative to imputation to the mean for other imputation strategies.
+All table values are multiplied by 100 for ease of interpretation.
+Multiple imputation with random forests leads to the highest scaled Brier score for both models and when 0%, 15%, or 30% of additional data in the training and testing sets were set to missing.
+`r fp_par(line_spacing = 1)`
 
-```{r tbl_md_strat_dead_ipa, results = 'asis'}
+```{r}
 
-make_tbl_md_strat_caption <- function(metric, outcome, msg){
-  glue('Median (25th, 75th percentile) change in {metric} when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for {outcome}.
-Table values show the {metric} for imputation to the mean.
-For other imputation strategies, table values show the change in {metric} relative to the {metric} when imputation to the mean was applied.
-All table values are multiplied by 100 for ease of interpretability.
-{msg}.')
-}
-
-gt_latex(
-  gt_tbls$md_strat$dead.ipa,
-  caption = make_tbl_md_strat_caption(
-    metric = 'scaled Brier score',
-    outcome = 'mortality',
-    msg = "Multiple imputation with random forests leads to the highest scaled Brier score for both models and when 0\\%, 15\\%, or 30\\% of additional data in the training and testing sets were set to missing"
-  ),
-  label = 'tbl_md_strat_dead_ipa'
+gt_tbls$md_strat$dead.ipa %>% 
+  arrange(additional_missing_pct, md_method) %>% 
+  mutate(
+    across(
+      where(is.character),
+      ~ if_else(is.na(.x), '---', .x)
     )
+  ) %>% 
+  as_grouped_data(groups = 'additional_missing_pct') %>% 
+  as_flextable(hide_grouplabel = TRUE) %>% 
+  add_header_row(
+    values = c("Imputation method", 
+               "Proportional hazards", 
+               "Gradient boosted decision trees"),
+    colwidths = c(1, 2, 2)
+  ) %>% 
+  set_header_labels(
+    `SI_Proportional hazards` = 'Single imputation',
+    `MI_Proportional hazards` = 'Multiple imputation',
+    `SI_Gradient boosted decision trees` = 'Single imputation',
+    `MI_Gradient boosted decision trees` = 'Multiple imputation',
+    md_method = 'Imputation method'
+  ) %>% 
+  bg(i = ~!is.na(additional_missing_pct), bg = 'grey') %>% 
+  italic(i = ~!is.na(additional_missing_pct), italic = TRUE) %>% 
+  theme_box() %>% 
+  merge_v(part = 'header', j = 1) %>% 
+  width(width = 1.2) %>% 
+  width(j = 1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all')
 
@@
 
-\clearpage
+\newpage
 
-```{r tbl_md_strat_txpl_ipa, results = 'asis'}
+Table 5: Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+Table values show the scaled Brier score when the imputation method is imputation to the mean and the change in scaled Brier score relative to imputation to the mean for other imputation strategies.
+All table values are multiplied by 100 for ease of interpretation.
+While there was very little difference in scaled Brier score values when 0% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher scaled Brier scores when 15% or 30% of additional missing data were amputed.
+`r fp_par(line_spacing = 1)`
 
-gt_latex(
-  gt_tbls$md_strat$txpl.ipa,
-  caption = make_tbl_md_strat_caption(
-    metric = 'scaled Brier score',
-    outcome = 'transplant',
-    msg = 'While there was very little difference in scaled Brier score values when 0\\% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher scaled Brier scores when 15\\% or 30\\% of additional missing data were amputed'
-  ),
-  label = 'tbl_md_strat_txpl_ipa'
+```{r}
+
+gt_tbls$md_strat$txpl.ipa %>% 
+  arrange(additional_missing_pct, md_method) %>% 
+  mutate(
+    across(
+      where(is.character),
+      ~ if_else(is.na(.x), '---', .x)
     )
+  ) %>% 
+  as_grouped_data(groups = 'additional_missing_pct') %>% 
+  as_flextable(hide_grouplabel = TRUE) %>% 
+  add_header_row(
+    values = c("Imputation method", 
+               "Proportional hazards", 
+               "Gradient boosted decision trees"),
+    colwidths = c(1, 2, 2)
+  ) %>% 
+  set_header_labels(
+    `SI_Proportional hazards` = 'Single imputation',
+    `MI_Proportional hazards` = 'Multiple imputation',
+    `SI_Gradient boosted decision trees` = 'Single imputation',
+    `MI_Gradient boosted decision trees` = 'Multiple imputation',
+    md_method = 'Imputation method'
+  ) %>% 
+  bg(i = ~!is.na(additional_missing_pct), bg = 'grey') %>% 
+  italic(i = ~!is.na(additional_missing_pct), italic = TRUE) %>% 
+  theme_box() %>% 
+  merge_v(part = 'header', j = 1) %>% 
+  width(width = 1.2) %>% 
+  width(j = 1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all')
 
@@
 
-\clearpage
+\newpage
 
-```{r tbl_md_strat_dead_auc, results = 'asis'}
+Table 6: Median (25th, 75th percentile) change in concordance index when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+Table values show the concordance index when the imputation method is imputation to the mean and the change in concordance index relative to imputation to the mean for other imputation strategies.
+All table values are multiplied by 100 for ease of interpretation.
+Multiple imputation with random forests led to the highest concordance index for boosting models when 0%, 15%, or 30% of additional data in the training and testing sets were set to missing.
+For proportional hazards models, multiple imputation with nearest neighbors or random forests was the most effective strategy.
+`r fp_par(line_spacing = 1)`
 
-gt_latex(
-  gt_tbls$md_strat$dead.auc,
-  caption = make_tbl_md_strat_caption(
-    metric = 'concordance index',
-    outcome = 'mortality',
-    msg = 'Multiple imputation with random forests led to the highest concordance index for boosting models when 0\\%, 15\\%, or 30\\% of additional data in the training and testing sets were set to missing.
-For proportional hazards models, multiple imputation with nearest neighbors or random forests was the most effective strategy'
-  ),
-  label = 'tbl_md_strat_dead_auc'
+```{r}
+
+gt_tbls$md_strat$dead.auc %>% 
+  arrange(additional_missing_pct, md_method) %>% 
+  mutate(
+    across(
+      where(is.character),
+      ~ if_else(is.na(.x), '---', .x)
     )
+  ) %>% 
+  as_grouped_data(groups = 'additional_missing_pct') %>% 
+  as_flextable(hide_grouplabel = TRUE) %>% 
+  add_header_row(
+    values = c("Imputation method", 
+               "Proportional hazards", 
+               "Gradient boosted decision trees"),
+    colwidths = c(1, 2, 2)
+  ) %>% 
+  set_header_labels(
+    `SI_Proportional hazards` = 'Single imputation',
+    `MI_Proportional hazards` = 'Multiple imputation',
+    `SI_Gradient boosted decision trees` = 'Single imputation',
+    `MI_Gradient boosted decision trees` = 'Multiple imputation',
+    md_method = 'Imputation method'
+  ) %>% 
+  bg(i = ~!is.na(additional_missing_pct), bg = 'grey') %>% 
+  italic(i = ~!is.na(additional_missing_pct), italic = TRUE) %>% 
+  theme_box() %>% 
+  merge_v(part = 'header', j = 1) %>% 
+  width(width = 1.2) %>% 
+  width(j = 1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all')
 
@@
 
-\clearpage
+\newpage
 
-```{r tbl_md_strat_dead_cal_error, results = 'asis'}
+Table 7: Median (25th, 75th percentile) change in calibration error when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality.
+Table values show the calibration error when the imputation method is imputation to the mean and the change in calibration error relative to imputation to the mean for other imputation strategies.
+All table values are multiplied by 100 for ease of interpretation.
+As more data in the training and testing sets were set to missing, single and multiple imputation with random forests emerged as the strategies with the lowest calibration error.
+`r fp_par(line_spacing = 1)`
 
-gt_latex(
-  gt_tbls$md_strat$dead.cal_error,
-  caption = make_tbl_md_strat_caption(
-    metric = 'calibration error',
-    outcome = 'mortality',
-    msg = 'As more data in the training and testing sets were set to missing, single and multiple imputation with random forests emerged as the strategies with the lowest calibration error'
-  ),
-  label = 'tbl_md_strat_dead_cal_error'
+```{r}
+
+gt_tbls$md_strat$dead.cal_error %>% 
+  arrange(additional_missing_pct, md_method) %>% 
+  mutate(
+    across(
+      where(is.character),
+      ~ if_else(is.na(.x), '---', .x)
     )
+  ) %>% 
+  as_grouped_data(groups = 'additional_missing_pct') %>% 
+  as_flextable(hide_grouplabel = TRUE) %>% 
+  add_header_row(
+    values = c("Imputation method", 
+               "Proportional hazards", 
+               "Gradient boosted decision trees"),
+    colwidths = c(1, 2, 2)
+  ) %>% 
+  set_header_labels(
+    `SI_Proportional hazards` = 'Single imputation',
+    `MI_Proportional hazards` = 'Multiple imputation',
+    `SI_Gradient boosted decision trees` = 'Single imputation',
+    `MI_Gradient boosted decision trees` = 'Multiple imputation',
+    md_method = 'Imputation method'
+  ) %>% 
+  bg(i = ~!is.na(additional_missing_pct), bg = 'grey') %>% 
+  italic(i = ~!is.na(additional_missing_pct), italic = TRUE) %>% 
+  theme_box() %>% 
+  merge_v(part = 'header', j = 1) %>% 
+  width(width = 1.2) %>% 
+  width(j = 1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all')
 
@@
 
-\clearpage
+\newpage
 
-```{r tbl_md_strat_txpl_auc, results = 'asis'}
+Table 8: Median (25th, 75th percentile) change in concordance index when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+Table values show the concordance index when the imputation method is imputation to the mean and the change in concordance index relative to imputation to the mean for other imputation strategies.
+All table values are multiplied by 100 for ease of interpretation.
+While there was little difference in concordance index when 0% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher concordance indices when 15% or 30% of additional missing data were amputed.
+`r fp_par(line_spacing = 1)`
 
-gt_latex(
-  gt_tbls$md_strat$txpl.auc,
-  caption = make_tbl_md_strat_caption(
-    metric = 'concordance index',
-    outcome = 'transplant',
-    msg = 'While there was very little difference in concordance index when 0\\% of additional data were set to missing in the training and testing sets, missingness incorporated as an attribute, predictive mean matching, random forests, and Bayesian regression provided models with higher concordance indices when 15\\% or 30\\% of additional missing data were amputed'
-  ),
-  label = 'tbl_md_strat_txpl_auc'
+```{r}
+
+gt_tbls$md_strat$txpl.auc %>% 
+  arrange(additional_missing_pct, md_method) %>% 
+  mutate(
+    across(
+      where(is.character),
+      ~ if_else(is.na(.x), '---', .x)
     )
+  ) %>% 
+  as_grouped_data(groups = 'additional_missing_pct') %>% 
+  as_flextable(hide_grouplabel = TRUE) %>% 
+  add_header_row(
+    values = c("Imputation method", 
+               "Proportional hazards", 
+               "Gradient boosted decision trees"),
+    colwidths = c(1, 2, 2)
+  ) %>% 
+  set_header_labels(
+    `SI_Proportional hazards` = 'Single imputation',
+    `MI_Proportional hazards` = 'Multiple imputation',
+    `SI_Gradient boosted decision trees` = 'Single imputation',
+    `MI_Gradient boosted decision trees` = 'Multiple imputation',
+    md_method = 'Imputation method'
+  ) %>% 
+  bg(i = ~!is.na(additional_missing_pct), bg = 'grey') %>% 
+  italic(i = ~!is.na(additional_missing_pct), italic = TRUE) %>% 
+  theme_box() %>% 
+  merge_v(part = 'header', j = 1) %>% 
+  width(width = 1.2) %>% 
+  width(j = 1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all')
 
@@
 
-\clearpage
+\newpage
 
-```{r tbl_md_strat_txpl_cal_error, results = 'asis'}
+Table 9: Median (25th, 75th percentile) change in calibration error when different imputation strategies were applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for transplant.
+Table values show the calibration error when the imputation method is imputation to the mean and the change in calibration error relative to imputation to the mean for other imputation strategies.
+All table values are multiplied by 100 for ease of interpretation.
+As more data in the training and testing sets were set to missing, single and multiple imputation with nearest neighbors emerged as the optimal strategy for proportional hazards models while missingness incorporated as an attributed emerged an the optimal strategy for boosting models.
+`r fp_par(line_spacing = 1)`
 
-gt_latex(
-  gt_tbls$md_strat$txpl.cal_error,
-  caption = make_tbl_md_strat_caption(
-    metric = 'calibration error',
-    outcome = 'transplant',
-    msg = "As more data in the training and testing sets were set to missing, single and multiple imputation with nearest neighbors emerged as the optimal strategy for proportional hazards models while missingness incorporated as an attributed emerged an the optimal strategy for boosting models"
-  ),
-  label = 'tbl_md_strat_txpl_cal_error'
+```{r}
+
+gt_tbls$md_strat$txpl.cal_error %>% 
+  arrange(additional_missing_pct, md_method) %>% 
+  mutate(
+    across(
+      where(is.character),
+      ~ if_else(is.na(.x), '---', .x)
     )
+  ) %>% 
+  as_grouped_data(groups = 'additional_missing_pct') %>% 
+  as_flextable(hide_grouplabel = TRUE) %>% 
+  add_header_row(
+    values = c("Imputation method", 
+               "Proportional hazards", 
+               "Gradient boosted decision trees"),
+    colwidths = c(1, 2, 2)
+  ) %>% 
+  set_header_labels(
+    `SI_Proportional hazards` = 'Single imputation',
+    `MI_Proportional hazards` = 'Multiple imputation',
+    `SI_Gradient boosted decision trees` = 'Single imputation',
+    `MI_Gradient boosted decision trees` = 'Multiple imputation',
+    md_method = 'Imputation method'
+  ) %>% 
+  bg(i = ~!is.na(additional_missing_pct), bg = 'grey') %>% 
+  italic(i = ~!is.na(additional_missing_pct), italic = TRUE) %>% 
+  theme_box() %>% 
+  merge_v(part = 'header', j = 1) %>% 
+  width(width = 1.2) %>% 
+  width(j = 1, width = 1.5) %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  align(align = 'center', part = 'all') %>% 
+  align(j = 1, align = 'left', part = 'all')
 
@@
 
-\clearpage
+\newpage
 
@@
 
-\clearpage
+\newpage
 
 ```{r fig_md_strat_infer_ipa, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in scaled Brier score values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
-Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0%, 15%, or 30%.
 Posterior probability that the difference in scaled Brier score exceeds 0, indicating an improvement in overall model accuracy, is printed to the right of each distribution.
-Each multiple imputation strategy and single imputation with missingness incorporated as an attribute had over 90\\% posterior predicted probability of increasing the scaled Brier score versus using imputation to the mean."}
+Each multiple imputation strategy and single imputation with missingness incorporated as an attribute had over 90% posterior predicted probability of increasing the scaled Brier score versus using imputation to the mean."}
 
@@
 
-\clearpage
+\newpage
 
 ```{r fig_md_strat_infer_auc, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in concordance index values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
-Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0%, 15%, or 30%.
 Posterior probability that the difference in concordance index exceeds 0, indicating an improvement in model discrimination, is printed to the right of each distribution.
-Multiple imputation with predictive mean matching, random forests, and Bayesian regression each had over 90\\% posterior predicted probability of increasing the concordance index versus using imputation to the mean."}
+Multiple imputation with predictive mean matching, random forests, and Bayesian regression each had over 90% posterior predicted probability of increasing the concordance index versus using imputation to the mean."}
 
@@
 
-\clearpage
+\newpage
 
 ```{r fig_md_strat_infer_cal_error, dpi = 300, fig.height = 6, fig.width = 7, fig.align='center', fig.cap = "Posterior distribution of differences in calibration error values (multiplied by 100) relative to imputation to the mean when different imputation strategies are applied before fitting a risk prediction model.
-Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0\\%, 15\\%, or 30\\%.
+Results are aggregated over scenarios where the outcome is mortality and transplant and the amount of additional missing data is 0%, 15%, or 30%.
 Posterior probability that the difference in calibration error is less than 0, indicating an improvement in model calibration, is printed to the right of each distribution.
-Every imputation strategy evaluated had over 90\\% posterior predicted probability of improving model calibration versus using imputation to the mean."}
+Every imputation strategy evaluated had over 90% posterior predicted probability of improving model calibration versus using imputation to the mean."}
 
@@
 
-\clearpage
+\newpage
+
+# REFERENCES

### 2020-12-16 R/as_gt.R

@@ -8,7 +8,8 @@ as_gt <- function(tbl_md_strat,
                   model_labels,
                   additional_missing_labels,
                   md_method_labels,
-                  rspec) {
+                  rspec,
+                  return_gt_table = TRUE) {
 
   gt_md_strat <- tbl_md_strat %>%
     map(
@@ -25,7 +26,7 @@ as_gt <- function(tbl_md_strat,
             table_val = if_else(
               md_strat == 'meanmode_si',
               true = table_glue("{est} (reference)", rspec=rspec),
-              false = table_glue("{pm(est)}{est} ({lwr}, {upr})", rspec=rspec)
+              false = table_glue("{pm(est)}{est}\n({lwr}, {upr})", rspec=rspec)
             )
           ) %>%
           select(-est, -lwr, -upr) %>%
@@ -50,28 +51,36 @@ as_gt <- function(tbl_md_strat,
             `MI_Gradient boosted decision trees`
           )
 
-        cols <- str_detect(names(gt_data), pattern = 'MI$|SI$')
-        cols <- names(gt_data)[cols]
+        if(return_gt_table){
 
-        gt(gt_data,
-           groupname_col = 'additional_missing_pct',
-           rowname_col = 'md_method') %>%
-          tab_stubhead(label = 'Imputation method') %>%
-          fmt_missing(columns = everything(),
-                      missing_text = '--') %>%
-          tab_spanner(label = 'Proportional hazards',
-                      columns = c("SI_Proportional hazards",
-                                  "MI_Proportional hazards")) %>%
-          tab_spanner(label = "Gradient boosted decision trees",
-                      columns = c("SI_Gradient boosted decision trees",
-                                  "MI_Gradient boosted decision trees")) %>%
-          cols_label(
-            "MI_Proportional hazards" = 'Multiple Imputation',
-            "SI_Proportional hazards" = 'Single Imputation',
-            "MI_Gradient boosted decision trees" = 'Multiple Imputation',
-            "SI_Gradient boosted decision trees" = 'Single Imputation',
-          ) %>%
-          cols_align('center')
+          cols <- str_detect(names(gt_data), pattern = 'MI$|SI$')
+          cols <- names(gt_data)[cols]
+
+          gt(gt_data,
+             groupname_col = 'additional_missing_pct',
+             rowname_col = 'md_method') %>%
+            tab_stubhead(label = 'Imputation method') %>%
+            fmt_missing(columns = everything(),
+                        missing_text = '--') %>%
[…]

### 2020-12-16 R/plan.R

@@ -91,14 +91,15 @@ the_plan <- drake_plan(
                   model_labels,
                   additional_missing_labels,
                   md_method_labels,
-                  rspec),
+                  rspec,
+                  return_gt_table = FALSE),
 
-  arxiv_preprint = target(
-    command = {
-      rmarkdown::render(knitr_in("doc_arxiv/doc_arxiv.Rmd"))
-      file_out("doc_arxiv/doc_arxiv.pdf")
-    }
-  )
+  # arxiv_preprint = target(
+  #   command = {
+  #     rmarkdown::render(knitr_in("doc_arxiv/doc_arxiv.Rmd"))
+  #     file_out("doc_arxiv/doc_arxiv.pdf")
+  #   }
+  # )
 
 
 )

### 2020-12-16 R/tabulate_descriptives.R

@@ -88,7 +88,7 @@ tabulate_descriptives <- function(im, times) {
     transmute(
       status,
       variable,
-      tbl_value = table_glue("{n_miss} ({pct_miss}%)", rspec = rspec),
+      tbl_value = table_glue("{n_miss}\n({pct_miss}%)", rspec = rspec),
       label = recode(
         variable,
         demo_age = "Age, years",

### 2020-12-16 R/tabulate_impute_accuracy.R

@@ -35,18 +35,18 @@ tabulate_impute_accuracy <- function(mc_cv_results) {
   data_tbl %>%
     mutate(
       nominal = table_glue(
-        "{pm(nominal_est)}{nominal_est} ({nominal_lwr}, {nominal_upr})"
+        "{pm(nominal_est)}{nominal_est}\n({nominal_lwr}, {nominal_upr})"
       ),
       numeric = table_glue(
-        "{pm(numeric_est)}{numeric_est} ({numeric_lwr}, {numeric_upr})"
+        "{pm(numeric_est)}{numeric_est}\n({numeric_lwr}, {numeric_upr})"
       )
     ) %>%
     mutate(
       across(
         .cols = c(nominal, numeric),
         .fns = ~str_replace(.x,
-                            pattern = '0\\.00 \\(0\\.00, 0\\.00\\)',
-                            replacement = '0 \\(reference\\)')
+                            pattern = fixed('0.00\n(0.00, 0.00)'),
+                            replacement = fixed('0 (reference)'))
       )
     ) %>%
     separate(md_strat, into = c('md_method', 'md_type')) %>%

### 2020-12-16 packages.R

@@ -48,6 +48,7 @@ conflict_prefer('summarize', 'dplyr')
 conflict_prefer('select',    'dplyr')
 conflict_prefer("gather",    "tidyr")
 conflict_prefer("set_names", "purrr")
+conflict_prefer("fixed", "stringr")
 conflict_prefer("all_numeric", "recipes")
 conflict_prefer("matches",     "tidyselect")
 conflict_prefer("impute",      "miceRanger")

### 2021-03-08 doc_ccqo/response_to_reviewers_r2.Rmd
R2R; revision 2; first draft

---
date: "`r Sys.Date()`"
title: "Response to reviewers"
output: 
  officedown::rdocx_document:
    reference_docx: style_manuscript_times_new_roman.docx
---

We thank the editors and reviewers for the opportunity to re-submit our manuscript. In response to the reviewer's excellent feedback, we have made minor changes to the manuscript and detailed each change in our responses to the reviewer's comment.

# Reviewer 2

__Comment__: The authors have made substantial changes in this revision. This revised analysis conducted a thorough comparison among 12 missing data imputation strategies regarding the imputation accuracy, c-statistics, calibration errors, and scaled Brier score. The authors suggest that the advanced imputation strategies, particularly random forest with multiple imputation, have a high probability of improving the downstream prediction models when comparing to the imputation to mean. However, the magnitude of improvement is very minor, the recommendation of missing imputation strategies remains unclear.

__Response__: Thank you for your review and helpful feedback. In our update, we make recommendations on which imputation strategy to use in the Discussion section and in a summary table: FILL IN

1. Can the authors comment whether the choice of imputation strategies may depend on the downstream purposes of using the missing imputation? I found it's hard to conclude a best missing imputation approach, which can vary by the % of missing data, prediction outcomes, etc. For example, for PH model of transplant, Bayesian regression with multiple imputations has a relatively high scaled Brier score, high c-statistics. For PH model of mortality, random forest with multiple imputations has a relatively high scaled Brier score, c-statistics, lower calibration errors. Furthermore, the difference across different imputing methods are very small based on the tables. Given the highest imputation accuracy with imputation to the mean, will the authors recommend using imputation to the mean when downstream modeling only involved a small amount of missingness?

__Response__: This is a good suggestion. We have commented on how the context of an analysis impacts the choice of how to impute missing values in the revised manuscript. In the Discussion Section, we write "in scenarios where few variables have missing values and the proportion of missing values is low, imputation to the mean is likely to be just as good as more computational strategies. As the degree of missingness increases, using multiple imputation and flexible imputation models appears to offer a small increment in the prognostic accuracy of downstream models."

2. Page 23. Line 1-2 Given the small magnitude of improvement and the complexity of the advanced imputation approaches, it's still reasonable to use mean imputation or treating the missingness as the attribute for developing risk prediction models. In addition, imputation to the mean has the highest imputation accuracy among all other approaches. Thus I think the avoidance of imputation to the mean is an overstatement. Can the authors provide comments that under which context should a more complex imputation method be applied?

__Response__: The reviewer points to our conclusion on page 23, lines 1-2. We believe our conclusion is consistent with the points the reviewer has made, but we did not write it clearly enough in our first revision. Josse et al. showed that you get Bayes optimal models in some scenarios when imputation to the mean is applied to the training/testing sets. This result is contingent on using a flexible model (e.g., boosting) that can create decision boundaries around the imputed mean value. Our conclusion has been modified to make this more clear: "Future INTERMACS analyses developing risk prediction models should consider imputation to the mean a valid strategy when the proportion of missing values is low or a flexible modeling strategy that can create decision boundaries around the imputed mean value is applied."

3. The scaled Brier score is closed to zero for the risk prediction models, indicating very poor prediction accuracy. The missing imputation strategies however did not improve the prediction accuracy of the prediction model regarding the scaled Brier score (Table 4 and 5). The authors have taken steps to improve c-statistics in their RSF, but the improvement is small. For prediction models with binary and continuous outcomes, will the authors have similar findings?

__Response:__ In response to the reviewer's thoughtful comments, we hypothesize that different imputation strategies would lead to small or moderate improvements in downstream classification or regression models. however, the current analysis does not examine outcomes that are continuous or binary, and therefore we do not comment on those domains in the paper.

Regarding the low improvements to prognostic accuracy in Tables 4 and 5, we note that mortality and transplant events are difficult to predict using pre-implant data. We hypothesize that downstream prognostic accuracy would have increased by a larger amount if these outcomes were more correlated to the predictors that had missing values. We have added the following text on line X page Y: "Our results are limited in scope to mortality and transplant outcomes, which are difficult to predict using data collected prior to the surgery where mechanical circulatory support is applied. Future analyses should investigate whether greater improvements are obtained based on the choice of imputation strategy for outcomes with greater correlation to the predictors that are being imputed."

4. Although the authors have provided Figure 1 showing the missing patterns among three variables, the authors haven't explored the reasons why they're missing. For example, it is possible that the variables are missing from one INTERMACS center that failed to submit the data. And the variables can also be missing for a certain time period when a center hasn't fully started their data participation or the data version did not include such variables. Under this scenario, it will make sense to perform listwise deletion based on centers or time periods.

__Response__ We appreciate the reviewer's helpful suggestion. We reviewed potential mechanisms for missingness in all of the analysis variables and found some additional notes to include in our revision. On lines X-X of page Y, we write "Data on surgery time and CV pressure was recorded less frequently in the INTERMACS registry prior to 2014, and the  combination of missing CV pressure and surgery time is the most frequently occurring missing pattern in the current analysis (Figure 1). In addition, missing values for peripheral edema occurred more frequently in 2014-2015 due to FILL IN"

<!-- table(is.na(im$pi_peripheral_edema), im$im_impl_yr) -->

5. To improve readability, I would suggest that: 
5.1 Put some details of methods into a technical appendix along with the code.

__Response__: We have developed an online appendix to the manuscript that provides code and additional methodological details. The document comprising this content is available on GitHub: FILL IN LINK.

5.2 Page 7 line 8: It will be good to give some examples of outcome variables Y.

__Response__: We have replaced the generic terms $Y$ and $X$ with "mortality risk" and "bilirubin"

5.3 If possible, summarize the features, pros, and cons for each strategy (e.g., aspects of imputation accuracy, scaled Brier score, c-index, calibration errors) and recommendation into one table.

__Response__: We have included a table that lists pros and cons for each imputation strategy we analyzed. The table also gives a recommendation for these strategies based on data from the current analysis.

### 2021-03-08 doc_ccqo/doc_ccqo.Rmd
R2R; revision 2; first draft

@@
 ```{r setup, include=FALSE}
+
+# About this document:
+
+# 1. This r-markdown file is built using officedown, an R package
+# designed to create reproducible MS word and powerpoint documents.
+
+# 2. The R project that this document lives in uses drake, an R
+# package designed to facilitate workflow by creating, storing,
+# and lazily updated analysis targets (see R/plan.R and README.Rmd)
+
+# 3. With the relevant packages downloaded and access to the INTERMACS 
+# analytic data (available on BIOLINCC), results in this publication
+# can be reproduced. However, noting that we utilized a parallel 
+# computing cluster to complete the Monte-Carlo Cross-validation
+# experiment, some results may require high computational overhead.
 

### 2021-03-08 R/plan.R

@@ -104,7 +104,3 @@ the_plan <- drake_plan(
 
 )
 
-# risk_evaluation %>%
-#   filter(additional_missing_pct == 0) %>%
-#   group_by(model, outcome, md_strat) %>%
-#   summarize(across(auc:GND_pvalue, median, na.rm = T))

### 2021-03-25 doc_ccqo/response_to_reviewers_r2.Rmd
fake data

@@
 __Response__ We appreciate the reviewer's helpful suggestion.
-We reviewed potential mechanisms for missingness in all of the analysis variables and found some additional notes to include in our revision.
+We reviewed potential mechanisms among the variables described in Figure 1.
+Due to the lack of indication for specific INTERMACS site in the de-identified data, we were not able to assess missingness based on center.
 On lines X-X of page Y, we write "Data on surgery time and CV pressure was recorded less frequently in the INTERMACS registry prior to 2014, and the  combination of missing CV pressure and surgery time is the most frequently occurring missing pattern in the current analysis (Figure 1).
-In addition, missing values for peripheral edema occurred more frequently in 2014-2015 due to FILL IN"
+In addition, missing values for peripheral edema occurred more frequently in 2014-2015 due to revisions to the data collection forms.
+CV pressure is measured using an invasive hemodynamic procedure.
+Roughly 80% of the missing values for CV pressure were attributable to this procedure not being done."
 
 <!-- table(is.na(im$pi_peripheral_edema), im$im_impl_yr) -->
+<!-- table(is.na(im$im_surgery_time), im$im_impl_yr) -->
 
@@
 
-__Response__: We have included a table that lists pros and cons for each imputation strategy we analyzed.
+__Response__: We have included a table that lists features, pros, and cons for each imputation strategy we analyzed.
 The table also gives a recommendation for these strategies based on data from the current analysis.

### 2021-03-25 doc_ccqo/doc_ccqo.Rmd
fake data

@@
 
+# inline_md_strat
+
+inline_md_strat <- tbl_md_strat %>%
+  bind_rows(.id = 'outcome') %>% 
+  mutate(tbv = table_glue("{est} ({lwr}, {upr})", rspec = rspec)) %>% 
+  split(f = list(.$outcome,
+                 .$model,
+                 .$md_strat,
+                 .$additional_missing_pct)) %>% 
+  map(pull, tbv)
+
+rspec_unscaled <- round_spec() %>% 
+  round_using_decimal(digits = 4)
+
+inline_md_strat_unscaled <- tbl_md_strat %>%
+  bind_rows(.id = 'outcome') %>% 
+  mutate(tbv = table_glue("{est/100} ({lwr/100}, {upr/100})", 
+                          rspec = rspec_unscaled)) %>% 
+  split(f = list(.$outcome,
+                 .$model,
+                 .$md_strat,
+                 .$additional_missing_pct)) %>% 
+  map(pull, tbv)
+
 ```
@@
 
+```{r}
+
+# inline results for abstract
+
+inline_auc_rf_vs_mean <- 
+  inline_md_strat_unscaled$dead.auc.cph.ranger_mi.0 %>% 
+  bracket_insert_left('25th, 75th percentile: ')
+
+prob_rf_mi_gt_mean <- 
+  fig_md_strat_infer$auc$probs$`Multiple imputation` %>% 
+  filter(m2 == 'ranger') %>% 
+  pull(prob_gt_meanmode)
+
+prob_rf_si_gt_mean <- 
+  fig_md_strat_infer$auc$probs$`Single imputation` %>% 
+  filter(m2 == 'ranger') %>% 
+  pull(prob_gt_meanmode)
+
+```
+
 __Background__ Risk prediction models play an important role in clinical decision making.
@@
 
-__Results__ Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.0008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
-The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively.
+__Results__ Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by `r inline_auc_rf_vs_mean` compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model.
+The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was `r prob_rf_si_gt_mean` and `r prob_rf_mi_gt_mean`, respectively.
 
@@
 When an additional 30% of data were missing, imputation to the mean obtained the highest numeric and nominal scores.
-
-```{r inline_md_strat}
-
-inline_md_strat <- tbl_md_strat %>%
-  bind_rows(.id = 'outcome') %>% 
-  mutate(tbv = table_glue("{est} ({lwr}, {upr})", rspec = rspec)) %>% 
-  split(f = list(.$outcome,
-                 .$model,
-                 .$md_strat,
-                 .$additional_missing_pct)) %>% 
-  map(pull, tbv)
-
-```
 

### 2021-03-25 R/colorize.R

@@ -1,0 +1,15 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param x
+##' @param color
+colorize <- function(x, color) {
+  if (knitr::is_latex_output()) {
+    sprintf("\\textcolor{%s}{%s}", color, x)
+  } else if (knitr::is_html_output()) {
+    sprintf("<span style='color: %s;'>%s</span>", color,
+            x)
+  } else x
+}

### 2021-03-25 R/make_intermacs_fake.R

@@ -1,0 +1,16 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param im
+make_intermacs_fake <- function(im) {
+
+  data_fake <- syn(data = im, method = 'cart', k = nrow(im))
+
+  write_csv(as_tibble(data_fake$syn),
+            file = 'doc_ccqo/data_fake.csv')
+
+  data_fake$syn
+
+}

### 2021-03-25 R/plan.R

@@ -49,6 +49,7 @@ the_plan <- drake_plan(
 
   # creates intermacs_clean.csv
   im = clean_intermacs(),
+  im_fake = make_intermacs_fake(im),
 
   # tabulate descriptives
   tbl_descriptives = tabulate_descriptives(im, times),

### 2021-03-25 packages.R

@@ -19,6 +19,7 @@ library(knitr)
 library(kableExtra)
 library(gtsummary)
 library(gt)
+library(synthpop)
 ## data analysis
 library(tidyverse)
 library(rstanarm)
